---
title: "Statistical Rethinking 2026 — Pipeline A01–A02"
subtitle: "Exact path counting, Bayesian updating, and predictive validation"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load-targets
library(targets)

a01_input <- tar_read(a01_input)
a01_hypotheses <- tar_read(a01_hypotheses)
a01_summary <- tar_read(a01_summary)
a02_input <- tar_read(a02_input)
a02_grid_posterior <- tar_read(a02_grid_posterior)
a02_exact_posterior <- tar_read(a02_exact_posterior)
a02_exact_predictive <- tar_read(a02_exact_predictive)
a02_grid_predictive <- tar_read(a02_grid_predictive)
a02_predictive_simulation <- tar_read(a02_predictive_simulation)
a02_summary <- tar_read(a02_summary)
```

# Pipeline contract

This first pipeline increment separates five objects:

1. validated input;
2. a forward generative model;
3. an exact or grid-based inferential calculation;
4. an independent computational check;
5. a report that reads cached targets instead of recalculating them.

# A01 — Random Allocation Game

The observed result is `r a01_input$claims` claims among `r a01_input$participants` participants.

Under the fixed-identity interpretation, all participants flip a fair coin, honest participants claim only after heads, and dishonest participants always claim. If $H$ participants are honest, the number of compatible paths is:

$$
\binom{H}{N-K}2^{N-H}.
$$

```{r}
#| label: a01-table
a01_hypotheses
```

If everyone is honest, there are `r a01_summary$all_honest_ways` compatible paths. If five fixed, labeled participants are honest, there are `r a01_summary$five_honest_fixed_ways` paths. Counting honesty assignments as additional microstates gives `r format(a01_summary$five_honest_variable_ways, big.mark = ",")` paths.

```{r}
#| label: a01-path-plot
#| fig-cap: "Compatible paths under two definitions of the hypothesis."
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(a01_hypotheses$honest, a01_hypotheses$fixed_ways,
  type = "b", xlab = "Honest participants", ylab = "Compatible paths",
  main = "Fixed identities")
plot(a01_hypotheses$honest, a01_hypotheses$variable_ways,
  type = "b", xlab = "Honest participants", ylab = "Compatible microstates",
  main = "Variable identities")
par(old_par)
```

The fixed path count is maximized at $H=$ `r paste(a01_summary$fixed_count_modes, collapse = " or ")`. Raw variable-identity counts are maximized at $H=$ `r paste(a01_summary$variable_count_modes, collapse = " or ")`.

The latter counts include $\binom{N}{H}$ possible assignments of honest identities. Because this factor changes with $H$, raw counts are not a likelihood over $H$ unless that microstate weighting is the intended prior. After conditioning on a fixed value of $H$, the claim-count likelihood is maximized at $H=$ `r paste(a01_summary$likelihood_modes, collapse = " or ")`.

> `r a01_summary$interpretation_note`

# A02 — Globe tossing

The data contain `r a02_input$water` water and `r a02_input$land` land observations. With a uniform $\operatorname{Beta}(1,1)$ prior, the exact posterior is:

$$
p\mid W,L \sim \operatorname{Beta}(4,12).
$$

Its mean is `r sprintf("%.3f", a02_summary$exact_mean)`. The grid approximation gives `r sprintf("%.3f", a02_summary$grid_mean)`.

```{r}
#| label: a02-posterior
#| fig-cap: "Grid approximation to the posterior probability of water."
plot(a02_grid_posterior$probability, a02_grid_posterior$posterior,
  type = "l", lwd = 2, xlab = "Probability of water",
  ylab = "Posterior mass")
abline(v = a02_exact_posterior$mean, lty = 2)
```

## Posterior predictive distribution

For five future tosses, integrating the binomial sampling distribution over the beta posterior produces a beta-binomial distribution.

```{r}
#| label: a02-predictive-table
transform(a02_exact_predictive,
  grid_probability = a02_grid_predictive$probability)
```

```{r}
#| label: a02-predictive-plot
#| fig-cap: "Exact and simulated posterior predictive probabilities for five future tosses."
empirical <- tabulate(a02_predictive_simulation$future_water + 1L,
  nbins = a02_input$future_tosses + 1L) / nrow(a02_predictive_simulation)
barplot(rbind(a02_exact_predictive$probability, empirical), beside = TRUE,
  names.arg = a02_exact_predictive$future_water,
  col = c("steelblue", "orange"),
  xlab = "Water outcomes in five future tosses", ylab = "Probability")
legend("topright", legend = c("Exact beta-binomial", "Posterior simulation"),
  fill = c("steelblue", "orange"), bty = "n")
```

The exact predictive mean is `r sprintf("%.3f", a02_summary$predictive_mean)` water observations out of five. The largest difference between the grid and exact predictive probabilities is `r format(a02_summary$maximum_grid_exact_difference, scientific = TRUE, digits = 3)`.

# Validation status

| Check | Status |
|---|---|
| Input contracts | Enforced before computation |
| A01 analytic versus exhaustive enumeration | Tested for every $H=0,\ldots,10$ |
| A01 likelihood normalization | Tested for every hypothesis |
| A02 posterior normalization | Tested |
| Grid posterior versus exact beta posterior | Tested |
| Grid predictive versus exact beta-binomial | Tested |
| Posterior simulation reproducibility | Tested with fixed seeds |
| MCMC diagnostics | Not applicable: A01–A02 use exact/grid calculations |
| External validation | Not applicable at this stage |
| Prior sensitivity | Deferred to a later problem |
