---
title: "A03 — Geocentric Models"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# A03 — Geocentric Models

**Course:** Statistical Rethinking 2026  
**Section:** A — Beginner  
**Lecture:** A03 — Geocentric Models  
**Duration:** approximately 1 hour 1 minute  
**Official reading:** Chapter 4 of *Statistical Rethinking*, second edition  
**Central question:** How can a model predict well while remaining descriptively incomplete or causally wrong, and how is Bayesian linear regression built, checked, and interpreted?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, Richard McElreath's A03 announcement, the uploaded second edition of *Statistical Rethinking*, and the official scripts `03_ptolemaic_model.R`, `03_gaussian_generative_sim.r`, `03_howell_plots.r`, `03_howell_new_weight_model.r`, and `03_prior_pred_OLS.r`.
>
> The official lecture description identifies the history of least-squares estimation, simple linear regression, quadratic approximation, and prior predictive distributions as core topics. These notes reconstruct the lecture's statistical structure from those materials; they are not a verbatim transcript.

---

## 1. Main thesis

The central argument is:

> **Linear regression is a powerful descriptive approximation, not a literal theory of how the world works.**

McElreath compares linear regression with Ptolemaic geocentric astronomy. The geocentric model is wrong as a physical account of the solar system, yet suitably calibrated epicycles can predict apparent planetary motion rather well.

Linear regression has the same dual character:

- it can summarize patterns;
- interpolate and predict;
- approximate complicated relationships;
- but its parameters do not automatically reveal causal mechanisms.

The workflow is:

$$
\text{generative assumptions}
\rightarrow
\text{Gaussian observation model}
\rightarrow
\text{linear model for the mean}
\rightarrow
\text{priors}
\rightarrow
\text{prior predictive simulation}
\rightarrow
\text{posterior approximation}
\rightarrow
\text{prediction and criticism}
$$

A03 moves from the one-parameter globe model to regression with several jointly uncertain parameters.

---

## 2. Relationship to the book

A03 corresponds mainly to Sections 4.1 through 4.4.

| Lecture component | Second-edition counterpart | Role |
|---|---|---|
| Ptolemaic analogy | Chapter 4 introduction | Wrong models can still predict well |
| Why Gaussian distributions arise | Section 4.1 | Aggregation and Gaussian patterns |
| Mathematical model notation | Section 4.2 | Likelihood, linear model, and priors |
| Gaussian model of a measurement | Section 4.3 | Mean and standard deviation |
| Linear prediction | Section 4.4 | Expected outcome depends on a predictor |
| Prior predictive simulation | Sections 4.3–4.4 | Checks priors before fitting |
| Quadratic approximation | Sections 4.3.2 and 4.4.2 | Approximates a multivariate posterior |
| Expected and predicted outcomes | Sections 4.4.3–4.4.5 | Separates mean uncertainty from observation uncertainty |
| Polynomial curves and splines | Section 4.5 | Secondary extension |

### 2.1 Important book-to-lecture difference

The book's main Howell regression models height from weight:

$$
h_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha
+
\beta
\left(
\text{weight}_i-\overline{\text{weight}}
\right)
$$

The official 2026 script explicitly reverses the direction and models weight from height:

$$
W_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha
+
\beta
\left(
H_i-\overline{H}
\right)
$$

These models answer different questions. In general:

$$
p(Y \mid X)
\neq
p(X \mid Y)
$$

A regression line cannot simply be inverted to obtain the reverse conditional model.

---

## 3. Why “geocentric models”?

The official Ptolemaic script animates:

1. an Earth-centered deferent-and-epicycle model;
2. a Copernican Earth–Mars model that generates apparent retrograde motion.

The comparison separates description from explanation.

A geocentric model can reproduce the apparent path of Mars. A heliocentric model gives a more useful account of why that path appears.

The statistical lesson is:

$$
\text{predictive success}
\not\Rightarrow
\text{causal truth}
$$

A regression model may be excellent for interpolation or forecasting within a narrow domain while remaining poor for intervention, extrapolation, or mechanistic explanation.

This does not make regression useless. It defines how its output should be interpreted.

---

## 4. Least squares and the Gaussian likelihood

Suppose:

$$
y_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

with:

$$
\mu_i
=
\alpha+\beta x_i
$$

The likelihood is:

$$
p(y \mid \alpha,\beta,\sigma)
=
\prod_{i=1}^{N}
\frac{1}{\sqrt{2\pi}\sigma}
\exp
\left[
-
\frac{(y_i-\alpha-\beta x_i)^2}{2\sigma^2}
\right]
$$

Its log is:

$$
\log p(y \mid \alpha,\beta,\sigma)
=
-N\log \sigma
-
\frac{1}{2\sigma^2}
\sum_{i=1}^{N}
(y_i-\alpha-\beta x_i)^2
+
\text{constant}
$$

For fixed $\sigma$, maximizing the likelihood is equivalent to minimizing:

$$
\sum_{i=1}^{N}
(y_i-\alpha-\beta x_i)^2
$$

This is the ordinary least-squares objective.

Bayesian regression adds priors and produces a joint posterior:

$$
p(\alpha,\beta,\sigma \mid y,x)
\propto
p(y \mid \alpha,\beta,\sigma,x)
\,p(\alpha,\beta,\sigma)
$$

The result is a distribution over plausible lines and residual scales, not merely one best-fitting line.

---

## 5. Why Gaussian distributions are common

Section 4.1 and `03_gaussian_generative_sim.r` motivate the Gaussian distribution through aggregation.

### 5.1 Football-field random walk

Many people begin at the center of a field. At each step, a coin toss moves each person left or right.

For person $j$ after $T$ steps:

$$
X_j
=
\sum_{t=1}^{T}S_{jt}
$$

where:

$$
S_{jt}
\in
\{-1,+1\}
$$

Central positions can arise through many sequences. Extreme positions require unusual sequences dominated by movement in one direction. The resulting distribution becomes approximately Gaussian.

### 5.2 Accumulated growth

The script also represents growth as accumulated increments:

$$
X
=
X_0
+
\sum_{t=1}^{T}\Delta_t
$$

When many small influences accumulate and no single one dominates, the aggregate often approaches a Gaussian form after centering and scaling.

This explains how adult height or weight can be approximately Gaussian without any biological mechanism literally sampling from a normal distribution.

### 5.3 Gaussian is not universal

Some processes produce skewness, heavy tails, bounds, multiple modes, or discrete outcomes. Gaussian tails are thin and can underestimate extreme events.

The Gaussian should be used because its implications are reasonable for the task, not because every continuous outcome is inherently normal.

---

## 6. The Gaussian observation model

A Gaussian variable is written:

$$
Y
\sim
\operatorname{Normal}(\mu,\sigma)
$$

with density:

$$
p(y \mid \mu,\sigma)
=
\frac{1}{\sqrt{2\pi}\sigma}
\exp
\left[
-
\frac{(y-\mu)^2}{2\sigma^2}
\right]
$$

The parameters have distinct roles:

$$
\mathbb{E}[Y]
=
\mu
$$

$$
\operatorname{Var}(Y)
=
\sigma^2
$$

When predictors enter the model, the relevant claim is conditional:

$$
Y_i
\mid
X_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

A raw outcome histogram need not be perfectly Gaussian. The model concerns residual variation around the conditional expectation.

---

## 7. Gaussian model without a predictor

Chapter 4 first models adult height with one mean and standard deviation:

$$
h_i
\sim
\operatorname{Normal}(\mu,\sigma)
$$

with priors such as:

$$
\mu
\sim
\operatorname{Normal}(178,20)
$$

$$
\sigma
\sim
\operatorname{Uniform}(0,50)
$$

The posterior is:

$$
p(\mu,\sigma \mid h)
\propto
\left[
\prod_{i=1}^{N}p(h_i \mid \mu,\sigma)
\right]
p(\mu)p(\sigma)
$$

This is already a multivariate posterior. Even if the priors factorize, the data may induce posterior dependence:

$$
p(\mu,\sigma \mid h)
\neq
p(\mu \mid h)p(\sigma \mid h)
$$

Regression adds further parameters and richer posterior geometry.

---

## 8. Linear prediction

The generic regression model is:

$$
y_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha+\beta x_i
$$

The first line is the observation model. The second is the model for the expected outcome.

In the uncentered form:

- $\alpha$ is the expected outcome when $x=0$;
- $\beta$ is the expected change in the outcome associated with one unit of $x$.

Formally:

$$
\beta
=
\frac{\partial \mu}{\partial x}
$$

This is an associational interpretation unless causal assumptions justify an interventional one.

---

## 9. Centering the predictor

The Howell models use:

$$
\mu_i
=
\alpha
+
\beta(x_i-\overline{x})
$$

When:

$$
x_i=\overline{x}
$$

then:

$$
\mu_i=\alpha
$$

Centering therefore makes the intercept the expected outcome at an average predictor value.

### 9.1 Interpretive advantage

Without centering, the intercept may refer to adult weight at zero centimeters or height at zero kilograms. Such values lie outside the meaningful domain.

### 9.2 Computational advantage

Without centering, the intercept and slope may compensate strongly for each other. Centering often reduces posterior covariance and improves numerical approximation.

### 9.3 Predictions remain equivalent

Centered and uncentered models span the same family of lines. They differ in parameter interpretation and posterior geometry, not in their possible predictions.

---

## 10. The official 2026 Howell weight model

Let:

- $W_i$ be weight in kilograms;
- $H_i$ be height in centimeters;
- $\overline{H}$ be mean adult height.

The official fitted model is:

$$
W_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha
+
\beta(H_i-\overline{H})
$$

with priors:

$$
\alpha
\sim
\operatorname{Normal}(60,10)
$$

$$
\beta
\sim
\operatorname{LogNormal}(0,1)
$$

$$
\sigma
\sim
\operatorname{Uniform}(0,10)
$$

Interpretation:

- $\alpha$ is expected weight at average adult height;
- $\beta$ is expected weight change per additional centimeter;
- $\sigma$ is residual variation among adults of the same height.

A log-normal slope prior implies:

$$
\beta>0
$$

This encodes prior knowledge that the expected adult weight–height relationship should not slope downward over the modeled range. The restriction must come from substantive knowledge, not from inspecting the sample trend.

---

## 11. Prior predictive simulation

Coefficient priors are difficult to judge in isolation. Their implications for lines and observations are more informative.

For draw $s$:

$$
\alpha^{(s)}
\sim
p(\alpha)
$$

$$
\beta^{(s)}
\sim
p(\beta)
$$

$$
\sigma^{(s)}
\sim
p(\sigma)
$$

The expected line is:

$$
\mu^{(s)}(H)
=
\alpha^{(s)}
+
\beta^{(s)}(H-\overline{H})
$$

A prior predictive observation is:

$$
W^{(s)}
\sim
\operatorname{Normal}
\left(
\mu^{(s)}(H),
\sigma^{(s)}
\right)
$$

The official scripts first generate implausible lines from broad priors and then use more defensible priors.

Prior predictive checks ask whether the model permits or favors:

- negative adult weights;
- impossible weight changes over a modest height range;
- downward relationships judged implausible beforehand;
- or unrealistic residual variation.

The key lesson is:

$$
\text{prior over parameters}
\rightarrow
\text{prior over functions and data}
$$

A modest-looking slope prior can imply extreme outcomes when multiplied by a wide predictor range.

---

## 12. Generative simulation before fitting

The official Howell script simulates data from a known regression:

$$
H_i
\sim
\operatorname{Uniform}(130,170)
$$

$$
\mu_i
=
\alpha
+
\beta(H_i-\overline{H})
$$

$$
W_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

Because the simulation chooses $\alpha$, $\beta$, and $\sigma$, their true values are known.

The fitting procedure can then be tested:

1. choose or draw true parameters;
2. generate heights;
3. generate weights;
4. fit the model with `quap`;
5. compare posterior summaries with the known values;
6. repeat across many simulations.

This tests parameter recovery in the model's small world.

A stronger robustness exercise would simulate from alternatives such as:

- heavy-tailed residuals;
- nonlinear relationships;
- heteroskedasticity;
- measurement error;
- subgroup mixtures;
- or selection bias.

---

## 13. Quadratic approximation

For the regression model:

$$
\theta
=
(\alpha,\beta,\sigma)
$$

Quadratic approximation:

1. finds the posterior mode $\widehat{\theta}$;
2. evaluates local curvature;
3. replaces the posterior near the mode with a multivariate Gaussian.

The approximation is:

$$
p(\theta \mid D)
\approx
\operatorname{MVNormal}
\left(
\widehat{\theta},
\widehat{\Sigma}
\right)
$$

with:

$$
\widehat{\Sigma}
\approx
\left[
-
\nabla^2
\log p(\theta \mid D)
\bigg|_{\theta=\widehat{\theta}}
\right]^{-1}
$$

Near a smooth maximum:

$$
\log p(\theta \mid D)
\approx
\log p(\widehat{\theta} \mid D)
-
\frac{1}{2}
(\theta-\widehat{\theta})^{\mathsf T}
\widehat{\Sigma}^{-1}
(\theta-\widehat{\theta})
$$

Exponentiating this quadratic form gives a multivariate Gaussian approximation.

### 13.1 When it works

Quadratic approximation is often adequate when the posterior is:

- unimodal;
- reasonably symmetric;
- well identified;
- and away from hard boundaries.

### 13.2 When it can fail

It can be poor for posteriors that are:

- skewed;
- multimodal;
- truncated;
- funnel-shaped;
- weakly identified;
- or strongly nonlinear.

The fitting engine is part of the practical model and must be checked.

---

## 14. Posterior geometry and covariance

Each pair $(\alpha,\beta)$ defines one line.

Before data, the prior produces a broad cloud of parameter pairs and a broad family of lines. After data, the posterior concentrates.

Under quadratic approximation, equal-density contours are ellipses. A tilted ellipse indicates posterior covariance.

If the intercept increases, the slope may need to decrease to preserve similar predictions over the observed range.

Predictions should therefore use joint posterior draws:

$$
(\alpha^{(s)},\beta^{(s)},\sigma^{(s)})
\sim
p(\alpha,\beta,\sigma \mid D)
$$

Combining unrelated marginal endpoints does not preserve posterior dependence.

---

## 15. Sequential learning of a line

The official scripts animate the movement from prior to posterior as observations are added.

At $n=0$, lines come from the prior. With each observation, the posterior becomes:

$$
p(\alpha,\beta,\sigma \mid D_n)
$$

The animation illustrates that:

- uncertainty initially spans many lines;
- data concentrate the posterior;
- intercept and slope remain dependent;
- predictor spread matters for learning the slope;
- and uncertainty grows under extrapolation.

Many observations at nearly the same $x$ can estimate the mean there while leaving $\beta$ uncertain. Information about the slope depends on variation in the predictor, not only sample size.

---

## 16. Expected outcomes versus predicted observations

For a new height $H^*$, each posterior draw gives an expected weight:

$$
\mu^{(s)}(H^*)
=
\alpha^{(s)}
+
\beta^{(s)}(H^*-\overline{H})
$$

These draws approximate:

$$
p(\mu(H^*) \mid D)
$$

For a new individual, simulate:

$$
W_{\mathrm{new}}^{(s)}
\sim
\operatorname{Normal}
\left(
\mu^{(s)}(H^*),
\sigma^{(s)}
\right)
$$

This approximates:

$$
p(W_{\mathrm{new}} \mid H^*,D)
$$

The predictive interval is wider because it includes:

1. uncertainty about the expected relationship;
2. residual variation among individuals with the same predictor value.

A precise estimate of the mean does not imply a precise individual prediction.

---

## 17. Associational versus causal slopes

For the centered weight model:

$$
\beta
=
\frac{\partial}{\partial H}
\mathbb{E}[W \mid H]
$$

This does not automatically equal:

$$
\frac{\partial}{\partial h}
\mathbb{E}
\left[
W \mid \operatorname{do}(H=h)
\right]
$$

Comparing naturally taller and shorter adults is not the same as intervening to change an adult's height.

Regression parameters acquire causal meaning only through an explicit causal model, a meaningful intervention, and identification assumptions.

This is the geocentric warning expressed in causal notation.

---

## 18. Polynomial extensions

`03_prior_pred_OLS.r` also contains quadratic and cubic models.

Quadratic:

$$
\mu_i
=
\alpha
+
\beta_1x_i
+
\beta_2x_i^2
$$

Cubic:

$$
\mu_i
=
\alpha
+
\beta_1x_i
+
\beta_2x_i^2
+
\beta_3x_i^3
$$

These models are linear in their parameters even though their predicted curves are nonlinear in $x$.

Adding terms resembles adding epicycles: the model can describe increasingly complicated patterns without identifying the generating mechanism.

Higher-order terms make prior predictive checking especially important because small coefficient changes can produce extreme curves outside the observed range.

Section 4.5 generalizes this idea with spline basis functions:

$$
\mu_i
=
\alpha
+
\sum_{k=1}^{K}w_kB_{ki}
$$

This section belongs to Chapter 4 but is secondary to the main A03 topics named in the official lecture description.

---

## 19. Common errors exposed by A03

### Error 1: Treating the fitted line as a causal mechanism

A line describes a conditional mean unless causal assumptions establish more.

### Error 2: Reversing a regression

A model for $p(Y \mid X)$ is not a model for $p(X \mid Y)$.

### Error 3: Choosing priors coefficient by coefficient

Priors should be evaluated through the functions and observations they generate.

### Error 4: Using the observed trend to justify the prior sign

Sign restrictions should come from pre-data scientific knowledge.

### Error 5: Interpreting an unsupported intercept

Centering places the intercept in a meaningful region.

### Error 6: Ignoring posterior covariance

Joint draws are required for coherent lines and predictions.

### Error 7: Confusing a mean interval with a prediction interval

Individual observations remain variable around the expected relationship.

### Error 8: Assuming Gaussian residuals from the raw histogram

The model concerns the conditional distribution.

### Error 9: Ignoring heavy tails

A Gaussian likelihood may underestimate extreme outcomes.

### Error 10: Trusting `quap` automatically

Quadratic approximation is local and can fail for non-Gaussian posterior geometry.

### Error 11: Extrapolating because the equation permits it

The mathematical line extends indefinitely; the scientific relationship usually does not.

### Error 12: Equating prediction with explanation

A geocentric model may predict appearances while misunderstanding causes.

---

## 20. CRM application: customer revenue

Let:

- $R_i$ be next-30-day revenue;
- $F_i$ be recent purchase frequency;
- $\overline{F}$ be mean recent frequency.

A baseline model is:

$$
R_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha
+
\beta(F_i-\overline{F})
$$

Prior predictive checks should ask:

- Does the model generate large negative revenues?
- Does one extra purchase imply an impossible revenue increase?
- Does the residual scale reflect customer heterogeneity?
- Is a Gaussian likelihood plausible for a nonnegative, right-skewed outcome?

The expected-revenue distribution is:

$$
p(\mu(F) \mid D)
$$

The individual predictive distribution is:

$$
p(R_{\mathrm{new}} \mid F,D)
$$

The latter is wider and more relevant for customer-level risk.

A positive slope does not prove that forcing one additional purchase would increase future revenue by $\beta$. Frequency may reflect affinity, income, tenure, product need, previous campaigns, or seasonality.

The model is descriptive unless a causal design identifies an intervention effect.

---

## 21. Implications for CausalOps

| A03 concept | CausalOps implication |
|---|---|
| Geocentric analogy | Label models as descriptive, predictive, or causal |
| Regression direction | Record the modeled conditional distribution |
| Gaussian likelihood | Check support, skewness, and tails |
| Centering | Store transformations and reference values |
| Priors | Link priors to units and domain ranges |
| Prior predictive simulation | Mandatory pre-fit plausibility report |
| Generative validation | Simulated parameter and estimand recovery |
| Quadratic approximation | Record inference engine and warnings |
| Posterior covariance | Preserve joint draws |
| Mean interval | Distinguish from predictive interval |
| Extrapolation | Flag predictions outside data support |
| Causal interpretation | Require identification before intervention claims |

A model artifact should declare its role, for example:

- `descriptive`;
- `predictive`;
- `causal_outcome_model`;
- `treatment_model`;
- `measurement_model`;
- `generative_simulator`.

A predictive model should not silently produce causal claims.

CausalOps can automate prior predictive checks by simulating:

$$
\mu^{(s)}(x)
$$

and:

$$
y^{(s)}
$$

across the intended predictor domain, flagging impossible support, explosive extrapolation, implausible slopes, and unrealistic residual variation.

---

## 22. Lecture takeaways

1. Linear regression is a useful approximation, not a literal causal theory.
2. Predictive success does not establish the true generating mechanism.
3. Least squares arises from a Gaussian observation model with a linear mean.
4. Bayesian regression adds priors and a joint posterior over lines and residual scale.
5. Gaussian distributions often arise from aggregating many small influences.
6. Gaussian models are not universal and may underestimate extreme events.
7. A regression model separates the observation distribution from the model of its mean.
8. Modeling $Y$ from $X$ differs from modeling $X$ from $Y$.
9. Centering improves interpretation and often posterior geometry.
10. Priors should be checked through their implications for functions and observations.
11. Generative simulation tests parameter recovery before real-data analysis.
12. Quadratic approximation locally replaces the posterior with a multivariate Gaussian.
13. Posterior covariance matters for coherent prediction.
14. Uncertainty about an expected outcome is narrower than uncertainty about a new observation.
15. Polynomial terms can improve fit without revealing mechanism.
16. Extrapolation is a scientific assumption.
17. A regression coefficient remains associational unless a causal design gives it interventional meaning.

---

## 23. Review questions

1. Why compare linear regression with Ptolemaic astronomy?
2. How can a model be wrong and still predict well?
3. Why does Gaussian maximum likelihood lead to least squares?
4. What does Bayesian regression add to an ordinary least-squares line?
5. Why can sums of many small influences look Gaussian?
6. What does $\sigma$ represent in Gaussian regression?
7. Why are height-on-weight and weight-on-height different models?
8. What does the intercept mean after centering the predictor?
9. Why can centering reduce intercept-slope covariance?
10. What does a log-normal slope prior imply?
11. Why should priors be checked with prior predictive simulation?
12. How can a modest coefficient prior generate implausible predictions?
13. What is the purpose of fitting simulated data with known parameters?
14. What does quadratic approximation approximate as Gaussian?
15. What does a tilted posterior ellipse indicate?
16. Why must prediction use joint posterior draws?
17. How does an interval for $\mu(H)$ differ from one for a new $W$?
18. Why can polynomial regression behave like statistical epicycles?
19. What assumptions would make a CRM slope causal rather than descriptive?
20. Why is extrapolation a scientific rather than merely mathematical decision?

---

## 24. Book reading guide

### Primary reading

- Chapter 4, *Geocentric Models*
  - Section 4.1, *Why normal distributions are normal*
  - Section 4.2, *A language for describing models*
  - Section 4.3, *Gaussian model of height*
  - Section 4.4, *Linear prediction*

### Secondary reading

- Section 4.5, *Curves from lines*

This section extends linear predictors to polynomials and splines. It is useful for understanding flexible function approximation but is not among the main topics named in the official A03 announcement.

### Read before A03

Focus on:

- the geocentric analogy;
- the football-field random walk;
- the roles of $\mu$ and $\sigma$;
- complete mathematical model notation;
- the adult Howell data;
- prior predictive simulation.

### Revisit after A03

Return to:

- the joint posterior of the Gaussian parameters;
- `quap` and quadratic approximation;
- predictor centering;
- posterior covariance among $\alpha$, $\beta$, and $\sigma$;
- `link` for expected outcomes;
- `sim` for posterior predictive outcomes;
- the distinction between mean and prediction intervals.

### Book-to-lecture translation

Book:

$$
\text{height}
\mid
\text{weight}
$$

Official 2026 script:

$$
\text{weight}
\mid
\text{height}
$$

The workflow is similar, but the conditional estimand and interpretation differ.

---

## 25. Official code guide

### `03_ptolemaic_model.R`

Animates Ptolemaic and Copernican models and motivates:

$$
\text{accurate description}
\not\Rightarrow
\text{correct mechanism}
$$

### `03_gaussian_generative_sim.r`

Simulates random walks and accumulated growth to show:

$$
\text{many small contributions}
\rightarrow
\text{approximately Gaussian aggregate}
$$

### `03_howell_plots.r`

Plots adult Howell data and samples regression lines from priors.

### `03_howell_new_weight_model.r`

Models weight from centered height, compares priors, simulates and validates known data-generating parameters, fits adult data, contrasts expected and predictive intervals, and animates posterior learning.

Core model:

$$
W_i
\sim
\operatorname{Normal}(\mu_i,\sigma)
$$

$$
\mu_i
=
\alpha
+
\beta(H_i-\overline{H})
$$

### `03_prior_pred_OLS.r`

Animates prior functions, posterior concentration, intercept-slope covariance, and quadratic and cubic extensions.

Main lesson:

$$
\text{prior over parameters}
\rightarrow
\text{prior over functions}
$$

---

## 26. Sources

- Richard McElreath, *Statistical Rethinking: A Bayesian Course with Examples in R and Stan*, second edition.
- Official `stat_rethinking_2026` repository and A03 lecture listing.
- Richard McElreath's official A03 announcement identifying the history of least squares, simple linear regression, quadratic approximation, and prior predictive distributions.
- Official A03 scripts listed above.


---

## 27. Appendix A: Solutions to Review Questions (Section 23)

### Question 1: Why compare linear regression with Ptolemaic astronomy?
**Answer:**
Ptolemaic astronomy used a system of deferents and epicycles to describe and forecast apparent planetary motions with remarkable numerical accuracy. However, its physical premise (that planets revolve in circular epicycles around a stationary Earth) was completely wrong.
Linear regression plays an analogous role in statistics: it fits parameters ($\alpha, \beta, \sigma$) that accurately interpolate and summarize patterns in observed data $p(Y \mid X)$, but it does not inherently model the true underlying physical or causal mechanism generating the data. McElreath uses this analogy to caution that predictive success within a sample or domain does not imply causal or mechanistic truth.

---

### Question 2: How can a model be wrong and still predict well?
**Answer:**
Predictive accuracy relies on statistical association and covariance between variables within a specific domain or data-generating distribution. As long as the correlation structure between predictor $X$ and outcome $Y$ remains stable, a model can yield highly accurate predictions even if it omits key confounders, misspecifies the functional form, or reverses the true direction of causation. However, because such a model does not reflect the causal architecture ($\operatorname{do}(X)$), its predictive power will collapse if the underlying data-generating environment changes or if an external intervention alters $X$.

---

### Question 3: Why does Gaussian maximum likelihood lead to least squares?
**Answer:**
Under a Gaussian likelihood assumption $y_i \sim \operatorname{Normal}(\mu_i, \sigma)$ with linear predictor $\mu_i = \alpha + \beta x_i$, the total log-likelihood is:
$$
\log p(y \mid \alpha, \beta, \sigma) = -\frac{N}{2} \log(2\pi) - N \log \sigma - \frac{1}{2\sigma^2} \sum_{i=1}^N (y_i - \alpha - \beta x_i)^2
$$
For any fixed noise scale $\sigma > 0$, maximizing the log-likelihood with respect to parameters $\alpha$ and $\beta$ is mathematically identical to minimizing the Residual Sum of Squares (RSS):
$$
(\widehat{\alpha}_{\mathrm{MLE}}, \widehat{\beta}_{\mathrm{MLE}}) = \arg\min_{\alpha, \beta} \sum_{i=1}^N (y_i - \alpha - \beta x_i)^2
$$
Thus, Ordinary Least Squares (OLS) is the Maximum Likelihood Estimator under independent Gaussian errors.

#### Multi-Language Syntax Comparison Table (Question 3)

| Framework / Language | Core Code Snippet | Computational Mechanics |
|---|---|---|
| **R (Base)** | `lm(y ~ x)` / `optim(par, function(p) -sum(dnorm(y, p[1]+p[2]*x, p[3], log=TRUE)))` | Solves normal equations directly or optimizes log-likelihood via Nelder-Mead/BFGS. |
| **Python (NumPy/SciPy)** | `scipy.optimize.minimize(lambda p: -np.sum(stats.norm.logpdf(y, p[0]+p[1]*x, p[2])))` | Solves negative log-likelihood minimization over parameter vector. |
| **PyTorch (Tensors)** | `loss = torch.sum((y - (alpha + beta * x))**2)` | Automatic differentiation minimizing MSE loss via gradient descent. |
| **Stan (C++ PPL)** | `y ~ normal(alpha + beta * x, sigma);` | Declarative log-density accumulation and gradient evaluation. |
| **PyMC (Python PPL)** | `with pm.Model(): mu = a + b*x; pm.Normal("y", mu, sigma, observed=y)` | Computes PyTensor log-probability graph and optimizes via MAP / NUTS. |

```{r}
# R Code Demonstration: MLE vs OLS Equivalency
set.seed(42)
x <- runif(50, 10, 30); y <- 5 + 2 * x + rnorm(50, 0, 3)
ols_fit <- lm(y ~ x)
nll <- function(par) -sum(dnorm(y, mean = par[1] + par[2] * x, sd = par[3], log = TRUE))
mle_fit <- optim(c(1, 1, 1), nll)
cat(sprintf("OLS Intercept: %.3f, Slope: %.3f\n", coef(ols_fit)[1], coef(ols_fit)[2]))
cat(sprintf("MLE Intercept: %.3f, Slope: %.3f\n", mle_fit$par[1], mle_fit$par[2]))
```

```{python}
# Python Code Demonstration
import numpy as np
from scipy.optimize import minimize
from scipy.stats import norm

np.random.seed(42)
x = np.random.uniform(10, 30, 50)
y = 5 + 2 * x + np.random.normal(0, 3, 50)


def nll(par):
  return -np.sum(norm.logpdf(y, loc=par[0] + par[1] * x, scale=par[2]))


res = minimize(nll, [1.0, 1.0, 1.0])
print(
    f"Python MLE -> Intercept: {res.x[0]:.3f}, Slope: {res.x[1]:.3f}, Sigma:"
    f" {res.x[2]:.3f}"
)
```

```{python}
# PyTorch Code Demonstration
import torch

torch.manual_seed(42)
x_t = torch.distributions.Uniform(10, 30).sample((50,))
y_t = 5 + 2 * x_t + torch.randn(50) * 3
a = torch.tensor(0.0, requires_grad=True)
b = torch.tensor(0.0, requires_grad=True)
optimizer = torch.optim.Adam([a, b], lr=0.1)

for _ in range(2000):
  optimizer.zero_grad()
  loss = torch.mean((y_t - (a + b * x_t)) ** 2)
  loss.backward()
  optimizer.step()

print(f"PyTorch OLS -> Intercept: {a.item():.3f}, Slope: {b.item():.3f}")
```

```{stan, output.var="stan_model"}
// Stan Code Demonstration: Gaussian Likelihood Regression
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  y ~ normal(alpha + beta * x, sigma);
}
```

```{python}
# PyMC Code Demonstration: Gaussian Likelihood Regression
import pymc as pm

with pm.Model() as model_q3:
  alpha = pm.Flat("alpha")
  beta = pm.Flat("beta")
  sigma = pm.HalfFlat("sigma")
  mu = alpha + beta * x
  y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=y)
  idata_q3 = pm.sample(draws=1000, random_seed=42)
```

---

### Question 4: What does Bayesian regression add to an ordinary least-squares line?
**Answer:**
Ordinary Least Squares produces a single point estimate for parameters $(\widehat{\alpha}, \widehat{\beta})$ and relies on asymptotic approximations for standard errors.
Bayesian regression adds:
1. **Prior Information:** Explicit prior distributions $p(\alpha, \beta, \sigma)$ that incorporate scientific domain knowledge, physical constraints, or regularization.
2. **Full Joint Posterior Distribution:** Rather than a single line, it yields a joint probability density $p(\alpha, \beta, \sigma \mid D)$ representing all plausible regression lines.
3. **Coherent Propagation of Uncertainty:** Parameter uncertainty and observation noise are combined naturally to produce exact posterior predictive distributions $p(Y_{\mathrm{new}} \mid X^*, D)$.

---

### Question 5: Why can sums of many small influences look Gaussian?
**Answer:**
By the **Central Limit Theorem (CLT)**, when a large number $T$ of independent (or weakly dependent) random fluctuations $S_1, S_2, \dots, S_T$ with finite variance are added together:
$$
X = \sum_{t=1}^T S_t
$$
the distribution of the sum $X$ converges to a Normal distribution regardless of the underlying distribution of the individual fluctuations $S_t$. In biological and physical systems, traits like height or weight are formed by accumulating thousands of micro-level genetic and environmental influences, leading naturally to Gaussian aggregate distributions.

#### Multi-Language Syntax Comparison Table (Question 5)

| Framework / Language | Core Code Snippet | Computational Mechanics |
|---|---|---|
| **R (Base)** | `rowSums(matrix(runif(N*T, -1, 1), nrow=N))` | Matrix summation over $T$ independent uniform micro-steps. |
| **Python (NumPy)** | `np.sum(np.random.uniform(-1, 1, size=(N, T)), axis=1)` | Vectorized array summation demonstrating convergence to Gaussian. |
| **PyTorch (Tensors)** | `torch.sum(torch.rand(N, T) * 2 - 1, dim=1)` | Parallel GPU/CPU tensor accumulation of random walk steps. |
| **Stan (C++ PPL)** | `generated quantities { real x_sum = sum(uniform_rng(-1, 1, T)); }` | Direct simulation of aggregate sum via Stan pseudo-random number generator. |
| **PyMC (Python PPL)** | `with pm.Model(): steps = pm.Uniform("s", -1, 1, shape=(N,T)); x = pm.Deterministic("x", steps.sum(axis=1))` | Declarative DAG node modeling sum of independent micro-distributions. |

```{r}
# R Code: Football Field Random Walk Convergence to Gaussian
set.seed(42); N <- 10000; T_steps <- 16
walk_outcomes <- rowSums(matrix(sample(c(-1, 1), N * T_steps, replace = TRUE), nrow = N))
cat(sprintf("Mean: %.3f, SD: %.3f\n", mean(walk_outcomes), sd(walk_outcomes)))
```

```{python}
# Python Code: Sum of Random Steps
import numpy as np

np.random.seed(42)
N, T_steps = 10000, 16
steps = np.random.choice([-1, 1], size=(N, T_steps))
sums = np.sum(steps, axis=1)
print(f"Python Walk Sum -> Mean: {np.mean(sums):.3f}, SD: {np.std(sums):.3f}")
```

```{python}
# PyTorch Code
import torch

torch.manual_seed(42)
steps_t = torch.randint(0, 2, (10000, 16)).float() * 2 - 1
sums_t = torch.sum(steps_t, dim=1)
print(
    f"PyTorch Walk Sum -> Mean: {sums_t.mean():.3f}, SD:"
    f" {sums_t.std():.3f}"
)
```

```{stan, output.var="stan_model"}
// Stan Code: Central Limit Theorem Simulation
data {
  int<lower=1> T_steps;
}
generated quantities {
  real x_sum = 0;
  for (t in 1:T_steps) {
    x_sum += uniform_rng(-1.0, 1.0);
  }
}
```

```{python}
# PyMC Code: CLT Sum Simulation
import pymc as pm

with pm.Model() as clt_model:
  steps = pm.Uniform("steps", lower=-1, upper=1, shape=16)
  x_sum = pm.Deterministic("x_sum", steps.sum())
  idata_clt = pm.sample_prior_predictive(samples=10000, random_seed=42)
```

---

### Question 6: What does $\sigma$ represent in Gaussian regression?
**Answer:**
In a Gaussian regression $y_i \sim \operatorname{Normal}(\mu_i, \sigma)$, the parameter $\sigma$ represents the **residual standard deviation** (observation noise scale).
- It quantifies the spread of individual observations around the conditional mean line $\mu_i = \alpha + \beta x_i$.
- It reflects unexplained variability, measurement noise, and omitted predictors.
- Unlike $\mu_i$ (which varies with $x_i$), standard Gaussian regression assumes homoskedasticity: $\sigma$ is constant across all values of predictor $x_i$.

---

### Question 7: Why are height-on-weight and weight-on-height different models?
**Answer:**
Regressing Height on Weight ($H_i \sim \operatorname{Normal}(\alpha_1 + \beta_1 W_i, \sigma_1)$) minimizes vertical squared errors in $H$, conditioning on $W$.
Regressing Weight on Height ($W_i \sim \operatorname{Normal}(\alpha_2 + \beta_2 H_i, \sigma_2)$) minimizes vertical squared errors in $W$, conditioning on $H$.
Because probability conditionals $p(H \mid W) \neq p(W \mid H)$, these models estimate fundamentally different parameters. Inverting the slope coefficient algebraically ($\beta_1 \neq 1/\beta_2$) is mathematically invalid because regression slopes depend on variance ratios: $\beta_{H \mid W} = \rho \frac{\sigma_H}{\sigma_W}$ whereas $\beta_{W \mid H} = \rho \frac{\sigma_W}{\sigma_H}$.

#### Multi-Language Syntax Comparison Table (Question 7)

| Framework / Language | Core Code Snippet | Computational Mechanics |
|---|---|---|
| **R (Base)** | `lm(height ~ weight)` vs `lm(weight ~ height)` | Demonstrates unequal slopes $\beta_1 \neq 1/\beta_2$ due to conditional error projection. |
| **Python (NumPy)** | `polyfit(w, h, 1)` vs `polyfit(h, w, 1)` | Fits separate conditional expectation lines $E[H \mid W]$ and $E[W \mid H]$. |
| **PyTorch (Tensors)** | `fit_h_on_w(w, h)` vs `fit_w_on_h(h, w)` | Minimizes $H$-loss vs $W$-loss in separate optimization passes. |
| **Stan (C++ PPL)** | `h ~ normal(a1 + b1*w, s1);` vs `w ~ normal(a2 + b2*h, s2);` | Distinct Stan likelihood blocks specifying opposite conditional distributions. |
| **PyMC (Python PPL)** | `pm.Normal("h", a1 + b1*w, s1, observed=h)` vs `pm.Normal("w", a2 + b2*h, s2, observed=w)` | Separate DAG models conditioning on different observed nodes. |

```{r}
# R Code Demonstration: Asymmetry of Regression Directions
set.seed(42)
h <- rnorm(100, 160, 10); w <- 0.7 * h + rnorm(100, -50, 5)
fit_h_on_w <- lm(h ~ w); fit_w_on_h <- lm(w ~ h)
cat(sprintf("Slope H~W: %.4f (Inverse: %.4f)\n", coef(fit_h_on_w)[2], 1/coef(fit_h_on_w)[2]))
cat(sprintf("Slope W~H: %.4f\n", coef(fit_w_on_h)[2]))
```

```{python}
# Python Code Demonstration
import numpy as np

np.random.seed(42)
h = np.random.normal(160, 10, 100)
w = 0.7 * h + np.random.normal(-50, 5, 100)
b_h_on_w = np.polyfit(w, h, 1)[0]
b_w_on_h = np.polyfit(h, w, 1)[0]
print(
    f"Slope H~W: {b_h_on_w:.4f} (1/b = {1/b_h_on_w:.4f}) | Slope W~H:"
    f" {b_w_on_h:.4f}"
)
```

```{python}
# PyTorch Code Demonstration
import torch

h_t = torch.randn(100) * 10 + 160
w_t = 0.7 * h_t + torch.randn(100) * 5 - 50
cov_hw = torch.cov(torch.stack([h_t, w_t]))
b_h_w = cov_hw[0, 1] / cov_hw[1, 1]
b_w_h = cov_hw[0, 1] / cov_hw[0, 0]
print(
    f"PyTorch Slope H~W: {b_h_w.item():.4f} (Inverse: {1/b_h_w.item():.4f}) |"
    f" Slope W~H: {b_w_h.item():.4f}"
)
```

```{stan, output.var="stan_model"}
// Stan Code: Comparing Regression Directions
data {
  int<lower=0> N;
  vector[N] H;
  vector[N] W;
}
parameters {
  real a1; real b1; real<lower=0> s1; // H ~ W
  real a2; real b2; real<lower=0> s2; // W ~ H
}
model {
  H ~ normal(a1 + b1 * W, s1);
  W ~ normal(a2 + b2 * H, s2);
}
```

```{python}
# PyMC Code Demonstration
import pymc as pm

with pm.Model() as direction_model:
  # Model 1: H ~ W
  b1 = pm.Normal("b1", 0, 10)
  h_obs = pm.Normal("h_obs", mu=100 + b1 * w, sigma=10, observed=h)
  # Model 2: W ~ H
  b2 = pm.Normal("b2", 0, 10)
  w_obs = pm.Normal("w_obs", mu=-50 + b2 * h, sigma=10, observed=w)
```

---

### Question 8: What does the intercept mean after centering the predictor?
**Answer:**
In an uncentered regression $\mu_i = \alpha + \beta x_i$, the intercept $\alpha$ represents the expected outcome when predictor $x_i = 0$ (e.g., adult weight when height is $0\text{ cm}$, which is physically impossible).
When the predictor is centered ($x_i - \overline{x}$):
$$
\mu_i = \alpha + \beta (x_i - \overline{x})
$$
Plugging in $x_i = \overline{x}$ gives $\mu_i = \alpha$. Thus, $\alpha$ becomes the **expected outcome at the sample average predictor value** (e.g., expected adult weight at mean height $\overline{H}$).

---

### Question 9: Why can centering reduce intercept-slope covariance?
**Answer:**
The posterior covariance between intercept $\alpha$ and slope $\beta$ under a linear Gaussian model is proportional to the mean of the predictor:
$$
\operatorname{Cov}(\alpha, \beta \mid D) = -\overline{x} \cdot \operatorname{Var}(\beta \mid D)
$$
If predictors $x_i$ are all positive (e.g., heights between $130$ and $180\text{ cm}$), increasing the slope $\beta$ forces the intercept $\alpha$ at $x=0$ to drop dramatically to maintain fit over the data range, creating strong negative posterior correlation.
Centering transforms the predictor so that $\overline{x - \overline{x}} = 0$. This eliminates the correlation ($\operatorname{Cov}(\alpha, \beta \mid D) = 0$), making the posterior contours axis-aligned (ellipse with no tilt) and improving numerical sampling efficiency.

#### Multi-Language Syntax Comparison Table (Question 9)

| Framework / Language | Core Code Snippet | Computational Mechanics |
|---|---|---|
| **R (Base)** | `cov(vcov(lm(y ~ x)))` vs `cov(vcov(lm(y ~ I(x - mean(x)))))` | Evaluates variance-covariance matrix of uncentered vs centered regression fits. |
| **Python (NumPy)** | `x_c = x - np.mean(x); cov_matrix = np.linalg.inv(X_c.T @ X_c)` | Direct inversion of design matrix $X^{\top}X$, demonstrating diagonal off-elements after centering. |
| **PyTorch (Tensors)** | `torch.inverse(X_centered.T @ X_centered)` | Tensor Gram matrix inversion proving zero off-diagonal covariance. |
| **Stan (C++ PPL)** | `vector[N] x_c = x - mean(x); y ~ normal(alpha + beta * x_c, sigma);` | Centered parameterization in Stan eliminating off-diagonal HMC mass matrix tilt. |
| **PyMC (Python PPL)** | `x_c = x - x.mean(); mu = a + b * x_c` | PyTensor graph construction with centered design matrix for decoupled posterior sampling. |

```{r}
# R Code: Posterior Covariance Reduction via Centering
x <- rnorm(100, 150, 10); y <- 10 + 0.5 * x + rnorm(100, 0, 2)
cov_uncentered <- vcov(lm(y ~ x))[1, 2]
cov_centered   <- vcov(lm(y ~ I(x - mean(x))))[1, 2]
cat(sprintf("Uncentered Cov(alpha, beta): %.6f\nCentered Cov(alpha, beta):   %.6f\n", cov_uncentered, cov_centered))
```

```{python}
# Python Code: Matrix Covariance Demonstration
import numpy as np

x = np.random.normal(150, 10, 100)
y = 10 + 0.5 * x + np.random.normal(0, 2, 100)
X_uncentered = np.column_stack([np.ones(100), x])
X_centered = np.column_stack([np.ones(100), x - np.mean(x)])

cov_un = np.linalg.inv(X_uncentered.T @ X_uncentered)[0, 1]
cov_c = np.linalg.inv(X_centered.T @ X_centered)[0, 1]
print(
    f"Python Uncentered Cov: {cov_un:.6f} | Python Centered Cov:"
    f" {cov_c:.6f}"
)
```

```{python}
# PyTorch Code
import torch

x_t = torch.randn(100) * 10 + 150
X_u = torch.stack([torch.ones(100), x_t], dim=1)
X_c = torch.stack([torch.ones(100), x_t - x_t.mean()], dim=1)

cov_u_t = torch.inverse(X_u.T @ X_u)[0, 1].item()
cov_c_t = torch.inverse(X_c.T @ X_c)[0, 1].item()
print(
    f"PyTorch Uncentered Cov: {cov_u_t:.6f} | PyTorch Centered Cov:"
    f" {cov_c_t:.6f}"
)
```

```{stan, output.var="stan_model"}
// Stan Code: Centered vs Uncentered Model
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
transformed data {
  vector[N] x_c = x - mean(x);
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  y ~ normal(alpha + beta * x_c, sigma);
}
```

```{python}
# PyMC Code: Centered Regression Model
import pymc as pm

x_c = x - np.mean(x)
with pm.Model() as centered_model:
  alpha = pm.Normal("alpha", mu=50, sigma=10)
  beta = pm.Normal("beta", mu=0, sigma=5)
  sigma = pm.Exponential("sigma", 1)
  mu = alpha + beta * x_c
  y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=y)
```

---

### Question 10: What does a log-normal slope prior imply?
**Answer:**
Assigning a Log-Normal prior to a regression slope parameter $\beta \sim \operatorname{LogNormal}(\mu_{\beta}, \sigma_{\beta})$ enforces $\beta > 0$ strictly across all parameter draws.
- It encodes **strictly positive monotonic relationship** between predictor $X$ and outcome $Y$.
- In the Howell height-weight example, it expresses prior scientific knowledge that expected body weight must increase with height, ruling out physically impossible downward-sloping regression lines before observing data.

---

### Question 11: Why should priors be checked with prior predictive simulation?
**Answer:**
Parameters in multi-parameter models interact nonlinearly to generate predictions. Individual prior distributions (e.g., $\alpha \sim \operatorname{Normal}(0, 100)$, $\beta \sim \operatorname{Normal}(0, 50)$) may appear "harmless" or "non-informative" individually, but when combined into a functional model $\mu_i = \alpha + \beta x_i$, they often generate wildly absurd predictions (e.g., adult heights of $-500\text{ cm}$ or $10,000\text{ kg}$).
Prior predictive simulation draws parameter tuples from joint priors *before* conditioning on data and plots the resulting functions $f(X)$ and simulated observations $Y$. This reveals whether the joint prior reflects reasonable domain science.

#### Multi-Language Syntax Comparison Table (Question 11)

| Framework / Language | Core Code Snippet | Prior Predictive Simulation Mechanics |
|---|---|---|
| **R (Base)** | `a <- rnorm(N, 60, 10); b <- rlnorm(N, 0, 1); mu <- a + b*x` | Direct Monte Carlo draw from parameter priors and function evaluation. |
| **Python (NumPy)** | `a = np.random.normal(60, 10, N); b = np.random.lognormal(0, 1, N)` | Vectorized sampling of prior parameter space and curve plotting. |
| **PyTorch (Tensors)** | `a = Normal(60, 10).sample((N,)); b = LogNormal(0, 1).sample((N,))` | Parallel tensor prior generation and line evaluation. |
| **Stan (C++ PPL)** | `generated quantities { real a_sim = normal_rng(60,10); ... }` | Stan block simulating prior predictive distribution without data likelihood. |
| **PyMC (Python PPL)** | `idata_prior = pm.sample_prior_predictive(samples=1000)` | Dedicated PyMC API drawing samples directly from unconditioned prior graph. |

```{r}
# R Code: Prior Predictive Simulation for Weight-Height Model
set.seed(42); N_sim <- 50; h_grid <- seq(130, 170, length.out = 100); h_bar <- 150
alpha_prior <- rnorm(N_sim, 60, 10); beta_prior <- rlnorm(N_sim, 0, 0.5)
plot(NULL, xlim = c(130, 170), ylim = c(0, 120), xlab = "Height (cm)", ylab = "Prior Weight (kg)")
for (i in 1:N_sim) lines(h_grid, alpha_prior[i] + beta_prior[i] * (h_grid - h_bar), col = rgb(0.2,0.4,0.8,0.3))
```

```{python}
# Python Code: Prior Predictive Simulation
import matplotlib.pyplot as plt
import numpy as np

np.random.seed(42)
N_sim = 50
h_grid = np.linspace(130, 170, 100)
h_bar = 150
alpha_p = np.random.normal(60, 10, N_sim)
beta_p = np.random.lognormal(0, 0.5, N_sim)
lines = alpha_p[:, None] + beta_p[:, None] * (h_grid - h_bar)
print(f"Python Prior Predictive Lines shape: {lines.shape}")
```

```{python}
# PyTorch Code
import torch

alpha_t = torch.distributions.Normal(60, 10).sample((50, 1))
beta_t = torch.distributions.LogNormal(0, 0.5).sample((50, 1))
h_grid_t = torch.linspace(130, 170, 100).unsqueeze(0)
lines_t = alpha_t + beta_t * (h_grid_t - 150)
print(f"PyTorch Prior Lines shape: {lines_t.shape}")
```

```{stan, output.var="stan_model"}
// Stan Code: Prior Predictive Simulation
data {
  int<lower=1> N_grid;
  vector[N_grid] h_grid;
  real h_bar;
}
generated quantities {
  real alpha_sim = normal_rng(60, 10);
  real beta_sim = lognormal_rng(0, 0.5);
  vector[N_grid] mu_sim = alpha_sim + beta_sim * (h_grid - h_bar);
}
```

```{python}
# PyMC Code: Prior Predictive Check
import pymc as pm

with pm.Model() as prior_check_model:
  alpha = pm.Normal("alpha", mu=60, sigma=10)
  beta = pm.LogNormal("beta", mu=0, sigma=0.5)
  sigma = pm.Uniform("sigma", lower=0, upper=10)
  prior_idata = pm.sample_prior_predictive(samples=1000, random_seed=42)
```

---

### Question 12: How can a modest coefficient prior generate implausible predictions?
**Answer:**
A slope prior like $\beta \sim \operatorname{Normal}(0, 10)$ might seem small when considered in isolation. However, when multiplied by a wide predictor range (e.g., height difference $\Delta x = 100\text{ cm}$), a 2-standard-deviation slope draw of $\beta = 20$ predicts a weight difference of $\Delta Y = 20 \times 100 = 2000\text{ kg}$ across the sample!
Because predictions compound slope coefficients over the full span of $X$, seemingly modest priors on slope parameters can imply physically impossible changes in $Y$.

---

### Question 13: What is the purpose of fitting simulated data with known parameters?
**Answer:**
Fitting simulated data generated from known parameters ($\alpha_{\mathrm{true}}, \beta_{\mathrm{true}}, \sigma_{\mathrm{true}}$) is called **generative validation** (or parameter recovery testing).
Its purpose is to verify that:
1. The statistical fitting algorithm (`quap`, MCMC, PyMC, Stan) is implemented correctly without bugs.
2. The model design is mathematically identifiable (parameters can be recovered from data).
3. The sample size $N$ provides sufficient information to estimate the target parameters accurately.
Testing parameter recovery in the model's "small world" is a mandatory diagnostic step before applying the pipeline to real empirical data.

#### Multi-Language Syntax Comparison Table (Question 13)

| Framework / Language | Core Code Snippet | Parameter Recovery Mechanics |
|---|---|---|
| **R (Base)** | `sim_y <- a_true + b_true*(x-150) + rnorm(N,0,s_true); lm(sim_y ~ ...)` | Generates synthetic response data and checks recovery of true parameters. |
| **Python (NumPy)** | `y_sim = a_true + b_true*x_c + np.random.normal(0, s_true, N)` | Vectorized synthetic data generation and recovery check via OLS/optimization. |
| **PyTorch (Tensors)** | `y_sim = a_true + b_true*x_c + torch.randn(N)*s_true` | Tensor-based synthetic data simulation and parameter recovery optimization. |
| **Stan (C++ PPL)** | `y ~ normal(alpha + beta * x_c, sigma);` fitted on simulated inputs | Validates HMC sampler by checking if 95% posterior intervals cover true data-generating values. |
| **PyMC (Python PPL)** | `pm.sample()` on synthetic dataset `y_sim` | MCMC recovery verification using ArviZ summary vs known ground-truth values. |

```{r}
# R Code: Parameter Recovery Simulation
set.seed(42); N <- 200; a_true <- 55.0; b_true <- 0.6; s_true <- 4.0
x <- runif(N, 130, 170); x_c <- x - mean(x)
y_sim <- a_true + b_true * x_c + rnorm(N, 0, s_true)
fit_sim <- lm(y_sim ~ x_c)
cat(sprintf("True Alpha: %.1f -> Est: %.3f\nTrue Beta:  %.1f -> Est: %.3f\nTrue SD:    %.1f -> Est: %.3f\n",
            a_true, coef(fit_sim)[1], b_true, coef(fit_sim)[2], s_true, summary(fit_sim)$sigma))
```

```{python}
# Python Code: Parameter Recovery
import numpy as np

np.random.seed(42)
N = 200
a_true, b_true, s_true = 55.0, 0.6, 4.0
x = np.random.uniform(130, 170, N)
x_c = x - np.mean(x)
y_sim = a_true + b_true * x_c + np.random.normal(0, s_true, N)
b_est, a_est = np.polyfit(x_c, y_sim, 1)
print(
    f"True: ({a_true}, {b_true}) | Estimated: ({a_est:.3f},"
    f" {b_est:.3f})"
)
```

```{python}
# PyTorch Code
import torch

torch.manual_seed(42)
N = 200
a_true, b_true, s_true = 55.0, 0.6, 4.0
x_t = torch.distributions.Uniform(130, 170).sample((N,))
x_c_t = x_t - x_t.mean()
y_sim_t = a_true + b_true * x_c_t + torch.randn(N) * s_true
cov_m = torch.cov(torch.stack([x_c_t, y_sim_t]))
b_rec = cov_m[0, 1] / cov_m[0, 0]
a_rec = y_sim_t.mean() - b_rec * x_c_t.mean()
print(f"PyTorch Recovered -> Alpha: {a_rec.item():.3f}, Beta: {b_rec.item():.3f}")
```

```{stan, output.var="stan_model"}
// Stan Code: Parameter Recovery Check
data {
  int<lower=0> N;
  vector[N] x_c;
  vector[N] y_sim;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  y_sim ~ normal(alpha + beta * x_c, sigma);
}
```

```{python}
# PyMC Code: Parameter Recovery Check
import arviz as az
import pymc as pm

with pm.Model() as recovery_model:
  alpha = pm.Normal("alpha", 50, 20)
  beta = pm.Normal("beta", 0, 5)
  sigma = pm.HalfNormal("sigma", 10)
  y_obs = pm.Normal("y_obs", mu=alpha + beta * x_c, sigma=sigma, observed=y_sim)
  idata_rec = pm.sample(draws=1000, random_seed=42)

print(az.summary(idata_rec, var_names=["alpha", "beta", "sigma"]))
```

---

### Question 14: What does quadratic approximation approximate as Gaussian?
**Answer:**
Quadratic approximation (`quap` or Laplace approximation) replaces the **entire joint posterior distribution** $p(\theta \mid D)$ with a multivariate Gaussian distribution centered at the posterior mode (MAP estimate $\widehat{\theta}$):
$$
p(\theta \mid D) \approx \operatorname{MVNormal}\left(\widehat{\theta}, \Sigma\right)
$$
where $\Sigma = \left[-\mathbf{H}\right]^{-1}$ is the inverse Hessian matrix of negative second-derivatives of the log-posterior evaluated at the mode $\widehat{\theta}$.
It approximates the log-posterior shape as a quadratic bowl near its peak.

---

### Question 15: What does a tilted posterior ellipse indicate?
**Answer:**
In a 2D contour plot of joint posterior density for two parameters (such as intercept $\alpha$ and slope $\beta$), a **tilted ellipse** indicates **posterior correlation / covariance** between those parameters ($\operatorname{Cov}(\alpha, \beta \mid D) \neq 0$).
- An ellipse tilted top-left to bottom-right indicates negative correlation: increasing $\alpha$ requires decreasing $\beta$ to maintain compatibility with data.
- An axis-aligned ellipse (no tilt) indicates posterior independence between $\alpha$ and $\beta$.

---

### Question 16: Why must prediction use joint posterior draws?
**Answer:**
Parameters in a fitted posterior distribution are interdependent ($\operatorname{Cov}(\alpha, \beta \mid D) \neq 0$).
If predictions are made by sampling $\alpha^{(s)}$ and $\beta^{(s)}$ independently from their separate marginal distributions, parameter correlations are destroyed. This pair mismatch produces regression lines with unrealistic combinations (e.g., high intercept combined with high slope) that lie outside the true posterior density, generating artificially wide, distorted prediction intervals. Joint posterior draws $(\alpha^{(s)}, \beta^{(s)}, \sigma^{(s)})$ preserve the true correlation structure.

#### Multi-Language Syntax Comparison Table (Question 16)

| Framework / Language | Core Code Snippet | Joint Sampling Mechanics |
|---|---|---|
| **R (Base / rethinking)** | `post <- extract.samples(fit); lines <- post$a + post$b * (x - x_bar)` | Draws joint rows of $(\alpha^{(s)}, \beta^{(s)})$ preserving covariance. |
| **Python (NumPy / SciPy)** | `samples = np.random.multivariate_normal(mean, cov, size=N)` | Joint sampling from multivariate normal posterior covariance matrix. |
| **PyTorch (Tensors)** | `MultivariateNormal(mean_vec, cov_mat).sample((N,))` | Parallel tensor joint sampling from full multivariate Gaussian posterior. |
| **Stan (C++ PPL)** | `as.data.frame(extract(stan_fit))` | Extracts correlated MCMC posterior draw vectors across parameter dimensions. |
| **PyMC (Python PPL)** | `draws = idata.posterior[["alpha", "beta"]].to_dataframe()` | ArviZ InferenceData joint trace extraction preserving chain dependencies. |

```{r}
# R Code: Joint Posterior Sampling vs Independent Sampling
set.seed(42); library(MASS)
mu_vec <- c(55, 0.6); cov_mat <- matrix(c(1.0, -0.4, -0.4, 0.2), nrow = 2)
joint_draws <- mvrnorm(1000, mu_vec, cov_mat) # Correct joint draws
indep_draws <- cbind(rnorm(1000, 55, 1.0), rnorm(1000, 0.6, sqrt(0.2))) # Wrong independent draws
cat(sprintf("Joint Draws Covariance:       %.4f\nIndependent Draws Covariance: %.4f\n", cov(joint_draws)[1,2], cov(indep_draws)[1,2]))
```

```{python}
# Python Code: Joint Draws
import numpy as np

np.random.seed(42)
mu_vec = np.array([55.0, 0.6])
cov_mat = np.array([[1.0, -0.4], [-0.4, 0.2]])
joint_samples = np.random.multivariate_normal(mu_vec, cov_mat, size=1000)
print(f"Python Joint Samples Covariance: {np.cov(joint_samples.T)[0, 1]:.4f}")
```

```{python}
# PyTorch Code Demonstration
import torch

mu_t = torch.tensor([55.0, 0.6])
cov_t = torch.tensor([[1.0, -0.4], [-0.4, 0.2]])
mvn = torch.distributions.MultivariateNormal(mu_t, cov_t)
joint_t = mvn.sample((1000,))
print(
    "PyTorch Joint Samples Covariance:"
    f" {torch.cov(joint_t.T)[0, 1].item():.4f}"
)
```

```{stan, output.var="stan_model"}
// Stan Code: Joint Posterior Draws Sampling
data { int<lower=0> N; vector[N] x; vector[N] y; }
parameters { real alpha; real beta; real<lower=0> sigma; }
model { y ~ normal(alpha + beta * x, sigma); }
generated quantities {
  // Joint predictive line for a new point x_new = 160
  real mu_new = alpha + beta * 160;
}
```

```{python}
# PyMC Code: Joint Posterior Predictive Evaluation
import pymc as pm

with pm.Model() as joint_pred_model:
  alpha = pm.Normal("alpha", 50, 10)
  beta = pm.Normal("beta", 0, 2)
  sigma = pm.HalfNormal("sigma", 5)
  mu = alpha + beta * x_c
  y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=y)
  idata_joint = pm.sample(draws=2000, random_seed=42)

# Extract joint parameter arrays
a_draws = idata_joint.posterior["alpha"].values.flatten()
b_draws = idata_joint.posterior["beta"].values.flatten()
```

---

### Question 17: How does an interval for $\mu(H)$ differ from one for a new $W$?
**Answer:**
1. **Interval for expected mean $\mu(H^*)$ (Epistemic Uncertainty):** Quantifies uncertainty in the location of the true average regression line at height $H^*$. It is formed by sampling $\mu^{(s)}(H^*) = \alpha^{(s)} + \beta^{(s)}(H^* - \overline{H})$ across joint posterior parameter draws. As sample size $N \to \infty$, this interval shrinks to zero width.
2. **Interval for a new observation $W_{\mathrm{new}}$ (Epistemic + Aleatoric Uncertainty):** Quantifies uncertainty for a specific, newly sampled individual of height $H^*$. It is formed by sampling $W_{\mathrm{new}}^{(s)} \sim \operatorname{Normal}(\mu^{(s)}(H^*), \sigma^{(s)})$. Even as $N \to \infty$, this interval remains bounded by residual observation variability $\sigma$.

#### Multi-Language Syntax Comparison Table (Question 17)

| Framework / Language | Core Code Snippet | Predictive Interval Mechanics |
|---|---|---|
| **R (Base)** | `mu_draws <- a + b*x_new` vs `W_sim <- rnorm(N, mu_draws, sigma)` | Contrasts expected mean interval (`mu_draws`) with posterior predictive interval (`W_sim`). |
| **Python (NumPy)** | `mu_sim = a + b * x_new; w_sim = np.random.normal(mu_sim, sigma)` | Simulates mean expectations vs individual outcome predictions. |
| **PyTorch (Tensors)** | `mu_t = a_t + b_t * x_new; w_t = Normal(mu_t, sigma_t).sample()` | Tensor sampling of expected mean vs individual predictive distributions. |
| **Stan (C++ PPL)** | `generated quantities { real mu_new = a + b*x_new; real W_new = normal_rng(mu_new, sigma); }` | Inline Stan generation of mean expectation vs individual observation predictions. |
| **PyMC (Python PPL)** | `pm.sample_posterior_predictive()` | Generates posterior predictive data nodes incorporating $\sigma$ noise. |

```{r}
# R Code: Mean Interval vs Prediction Interval
set.seed(42); N_draws <- 10000
a_draws <- rnorm(N_draws, 55, 0.5); b_draws <- rnorm(N_draws, 0.6, 0.05); s_draws <- rnorm(N_draws, 4, 0.2)
mu_draws <- a_draws + b_draws * (160 - 150) # Expected mean height
w_pred_draws <- rnorm(N_draws, mu_draws, s_draws) # Individual prediction
cat(sprintf("89%% Interval for Expected Mean mu(160): [%.2f, %.2f]\n", quantile(mu_draws, 0.055), quantile(mu_draws, 0.945)))
cat(sprintf("89%% Interval for New Individual W_new:  [%.2f, %.2f]\n", quantile(w_pred_draws, 0.055), quantile(w_pred_draws, 0.945)))
```

```{python}
# Python Code: Mean vs Prediction Interval
import numpy as np

np.random.seed(42)
N_draws = 10000
a_draws = np.random.normal(55, 0.5, N_draws)
b_draws = np.random.normal(0.6, 0.05, N_draws)
s_draws = np.random.normal(4.0, 0.2, N_draws)

mu_draws = a_draws + b_draws * (160 - 150)
w_pred_draws = np.random.normal(mu_draws, s_draws)

print(
    f"Mean mu(160) 89% HPDI: {np.percentile(mu_draws, [5.5, 94.5]).round(2)}"
)
print(
    "Individual W_new 89% HPDI:"
    f" {np.percentile(w_pred_draws, [5.5, 94.5]).round(2)}"
)
```

```{python}
# PyTorch Code
import torch

a_t = torch.randn(10000) * 0.5 + 55.0
b_t = torch.randn(10000) * 0.05 + 0.6
s_t = torch.randn(10000) * 0.2 + 4.0

mu_t = a_t + b_t * (160 - 150)
w_pred_t = torch.normal(mu_t, s_t)

print(
    "PyTorch Mean 89% CI:  "
    f" {torch.quantile(mu_t, torch.tensor([0.055, 0.945])).numpy().round(2)}"
)
print(
    "PyTorch Pred 89% CI:  "
    f" {torch.quantile(w_pred_t, torch.tensor([0.055, 0.945])).numpy().round(2)}"
)
```

```{stan, output.var="stan_model"}
// Stan Code: Mean vs Predictive Quantity Generation
data { real x_target; real x_bar; }
parameters { real alpha; real beta; real<lower=0> sigma; }
model { /* Likelihood */ }
generated quantities {
  real mu_target = alpha + beta * (x_target - x_bar);
  real y_target_sim = normal_rng(mu_target, sigma);
}
```

```{python}
# PyMC Code: Posterior Predictive Generation
import pymc as pm

with pm.Model() as pred_diff_model:
  alpha = pm.Normal("alpha", 55, 1)
  beta = pm.Normal("beta", 0.6, 0.1)
  sigma = pm.HalfNormal("sigma", 5)
  mu = alpha + beta * (160 - 150)
  y_new = pm.Normal("y_new", mu=mu, sigma=sigma)
  idata_pred_diff = pm.sample_posterior_predictive(
      pm.sample(1000, random_seed=42), random_seed=42
  )
```

---

### Question 18: Why can polynomial regression behave like statistical epicycles?
**Answer:**
Polynomial regression adds terms ($\beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \dots$) to curve the line and improve in-sample fit.
Like Ptolemaic epicycles, polynomial terms can fit arbitrary wiggles and complex curves without representing any physical or biological mechanism. Furthermore, polynomial curves exhibit dangerous wild oscillations (Runge's phenomenon) at data boundaries and produce catastrophic extrapolation errors outside the observed predictor range.

---

### Question 19: What assumptions would make a CRM slope causal rather than descriptive?
**Answer:**
A regression slope $\beta$ of next-30-day revenue $R$ on purchase frequency $F$ is strictly descriptive ($\mathbb{E}[R \mid F]$). To interpret $\beta$ causally ($\mathbb{E}[R \mid \operatorname{do}(F)]$):
1. **No Unobserved Confounding:** There must be no unmeasured variable $Z$ (e.g., customer income, brand affinity, marketing exposure) that influences both purchase frequency $F$ and revenue $R$.
2. **Causal Direction:** Purchase frequency $F$ must temporally and causally precede revenue $R$ (no reverse causation or simultaneous feedback).
3. **Well-Defined Intervention:** The operational action that alters $F$ (e.g., a discount voucher) must not alter revenue through secondary off-DAG pathways.

---

### Question 20: Why is extrapolation a scientific rather than merely mathematical decision?
**Answer:**
Mathematically, a fitted function $y = f(x)$ can be evaluated at any numeric value $x \in (-\infty, \infty)$. However, statistical models only capture associations within the empirical support of observed data.
Extrapolating outside observed data assumes that the physical, biological, or behavioral data-generating process remains unchanged in the unobserved regime. Because statistics cannot verify unobserved regimes, deciding whether a model's predictions remain valid outside the data range requires substantive scientific domain knowledge, not mathematical calculation.

---

## 28. Appendix B: Solution to Homework A03

### Problem Statement
> **Homework A03 Question:**  
> People get heavier with age. Using the Howell1 data, I want you to repeat the prior predictive simulation from Lecture A03 (and Chapter 4 of the book), but using only the height and age variables. The estimand should be the influence of age on height. Construct one or more plots showing the prior predictive distribution for height as a function of age, and try to justify your prior choices with some science or other.

---

### 1. Mathematical Construction of Prior Trajectories

#### Scientific Anchor Points
To construct scientifically defensible priors for human height $H$ (in cm) as a function of age $A$ (in years), we establish key biological anchor points:
1. **Conception ($A = -0.75$ years / $-9$ months):** Height (length) $H(-0.75) \approx 0\text{ cm}$.
2. **Birth ($A = 0$ years):** Average newborn length $\alpha \approx 50\text{ cm}$ (SD $\approx 5\text{ cm}$).
3. **Terminal Adult Height ($A \ge 20$ years):** Average adult height $H_{\max} \approx 150-160\text{ cm}$ (SD $\approx 10\text{ cm}$).

#### Model Formulations

##### Formulation 1: Piecewise Linear Model (Fixed Growth Cutoff $A_{\mathrm{cut}} = 20$)
Growth occurs linearly from birth ($A=0$) to adulthood ($A=20$), after which height plateaus:
$$
\mathbb{E}[H_i \mid A_i] = \begin{cases} \alpha + \beta A_i & \text{if } A_i < 20 \\ \alpha + 20 \beta & \text{if } A_i \ge 20 \end{cases}
$$
- **Prior for $\alpha$ (Birth Height):** $\alpha \sim \operatorname{Normal}(50, 5)$
- **Prior for $\beta$ (Growth Rate):** To reach $150\text{ cm}$ at age 20 from $50\text{ cm}$ at birth, total growth is $100\text{ cm}$ over 20 years $\implies \beta \approx 5\text{ cm/year}$.
  Prior: $\beta \sim \operatorname{Normal}(5, 1)$ or $\beta \sim \operatorname{LogNormal}(\log(5), 0.2)$ to enforce positive growth.

##### Formulation 2: Piecewise Model with Random Terminal Growth Age ($x$)
Allows individual variability in the age of terminal growth completion:
$$
x \sim \operatorname{Normal}(20, 2)
$$
$$
\mathbb{E}[H_i \mid A_i] = \begin{cases} \alpha + \beta A_i & \text{if } A_i < x \\ \alpha + \beta x & \text{if } A_i \ge x \end{cases}
$$

##### Formulation 3: Asymptotic Biological Growth Model (von Bertalanffy)
Differential equation for growth velocity declining proportionally to remaining growth potential:
$$
\frac{dH}{dA} = \beta (H_{\max} - H) \implies H(A) = H_{\max} \left(1 - \exp\left(-\beta (A - A_0)\right)\right)
$$
- $H_{\max} \sim \operatorname{Normal}(150, 10)$ (Terminal adult height)
- $A_0 = -0.75$ (Age at zero height / conception)
- $\beta \sim \operatorname{Normal}(0.12, 0.02)$ (Growth rate constant, derived from $x = e^{-\beta(20 - A_0)} \approx 0.10 \implies \beta \approx 0.12$).

---

### 2. Multi-Language Implementation Comparison Table (Homework A03)

| Framework / Language | Core Code Implementation | Prior Predictive Mechanics |
|---|---|---|
| **R (Base)** | `pmin(age, 20)`, `pmin(age, x_cut)`, `Hmax*(1 - exp(-b*(age+0.75)))` | Base R graphical curve overlays for fixed cutoff, random cutoff, and von Bertalanffy curves. |
| **Python (NumPy / Matplotlib)** | `np.minimum(age, 20)`, `np.minimum(age, x_cut)`, `Hmax * (1 - np.exp(-b * (age+0.75)))` | Vectorized 2D matrix calculation comparing piecewise (fixed/random cutoff) and biological growth prior curves. |
| **PyTorch (Tensors)** | `torch.clamp(age, max=20)`, `torch.minimum(age, x_cut)`, `Hmax * (1 - torch.exp(...))` | Tensor broadcast calculation across fixed piecewise, random cutoff, and asymptotic growth formulations. |
| **Stan (C++ PPL)** | `fmin(age_grid[i], 20.0)`, `fmin(age_grid[i], x_cutoff_sim)`, `Hmax * (1 - exp(...))` | Direct Stan generated quantities block sampling prior growth trajectories for all 3 model formulations. |
| **PyMC (Python PPL)** | `pm.math.minimum(age, 20)`, `pm.math.minimum(age, x_cutoff)`, `Hmax * (1 - pm.math.exp(...))` | PyTensor symbolic graphs for fixed piecewise, random cutoff, and von Bertalanffy prior predictive sampling. |

---

### 3. Complete Executable Code Solutions

#### R Code Solution

```{r}
# ==============================================================================
# Homework A03 - R Solution: Prior Predictive Simulation of Height vs Age
# ==============================================================================

set.seed(42)
n_sim <- 50
age_grid <- seq(-0.75, 80, length.out = 200)

# Common Priors for Piecewise Models (Model 1 & Model 2)
alpha_draws <- rnorm(n_sim, mean = 50, sd = 5)
beta_draws  <- rnorm(n_sim, mean = 5, sd = 1)

# Model 1: Piecewise Linear Model (Fixed Cutoff at Age 20)
plot(NULL, xlim = c(0, 80), ylim = c(0, 200),
     xlab = "Age (years)", ylab = "Height (cm)",
     main = "Homework A03: Model 1 - Piecewise Linear (Fixed Cutoff 20)")
abline(h = c(50, 150), lty = 2, col = "gray60")

for (i in 1:n_sim) {
  h_pred1 <- alpha_draws[i] + beta_draws[i] * pmin(age_grid, 20)
  lines(age_grid, h_pred1, col = rgb(0.8, 0.2, 0.2, 0.4), lwd = 1.5)
}

# Model 2: Piecewise Model with Random Growth Cutoff Age (x ~ Normal(20, 2))
x_cutoff_draws <- rnorm(n_sim, mean = 20, sd = 2)

plot(NULL, xlim = c(0, 80), ylim = c(0, 200),
     xlab = "Age (years)", ylab = "Height (cm)",
     main = "Homework A03: Model 2 - Piecewise Linear (Random Cutoff x)")
abline(h = c(50, 150), lty = 2, col = "gray60")

for (i in 1:n_sim) {
  h_pred2 <- alpha_draws[i] + beta_draws[i] * pmin(age_grid, x_cutoff_draws[i])
  lines(age_grid, h_pred2, col = rgb(0.2, 0.7, 0.3, 0.4), lwd = 1.5)
}

# Model 3: Asymptotic von Bertalanffy Growth Model
Hmax_draws <- rnorm(n_sim, mean = 150, sd = 10)
b_draws    <- rnorm(n_sim, mean = 0.12, sd = 0.02)

plot(NULL, xlim = c(-1, 80), ylim = c(0, 200),
     xlab = "Age (years)", ylab = "Height (cm)",
     main = "Homework A03: Model 3 - von Bertalanffy Growth Prior Predictive")
abline(h = 150, lty = 2, col = "gray60")

for (i in 1:n_sim) {
  h_curve <- Hmax_draws[i] * (1 - exp(-b_draws[i] * (age_grid + 0.75)))
  lines(age_grid, h_curve, col = rgb(0.2, 0.5, 0.8, 0.4), lwd = 1.5)
}
```

---

#### Python Code Solution (NumPy & SciPy)

```{python}
# ==============================================================================
# Homework A03 - Python Solution (NumPy & SciPy)
# ==============================================================================
import numpy as np

np.random.seed(42)
n_sim = 50
age_grid = np.linspace(-0.75, 80, 200)

# Model 1: Piecewise Linear Model (Fixed Cutoff at Age 20)
alpha_draws = np.random.normal(50, 5, size=n_sim)
beta_draws = np.random.normal(5, 1, size=n_sim)
H_m1 = alpha_draws[:, None] + beta_draws[:, None] * np.minimum(
    age_grid[None, :], 20
)

# Model 2: Piecewise Model with Random Growth Cutoff Age x ~ Normal(20, 2)
x_cutoff_draws = np.random.normal(20, 2, size=n_sim)
H_m2 = alpha_draws[:, None] + beta_draws[:, None] * np.minimum(
    age_grid[None, :], x_cutoff_draws[:, None]
)

# Model 3: von Bertalanffy Asymptotic Growth Model
Hmax_draws = np.random.normal(150, 10, size=n_sim)
b_draws = np.random.normal(0.12, 0.02, size=n_sim)
H_m3 = Hmax_draws[:, None] * (
    1 - np.exp(-b_draws[:, None] * (age_grid[None, :] + 0.75))
)

print("--- Homework A03 Python Simulation (3 Formulations) ---")
print(f"Age grid range: [{age_grid.min():.2f}, {age_grid.max():.2f}] years")
print(f"Model 1 (Fixed Cutoff 20) Mean Adult Height (Age 20): {np.mean(H_m1[:, age_grid >= 20]):.2f} cm")
print(f"Model 2 (Random Cutoff x) Mean Adult Height (Age 25): {np.mean(H_m2[:, age_grid >= 25]):.2f} cm")
print(f"Model 3 (Biological)      Mean Adult Height (Age 20): {np.mean(H_m3[:, age_grid >= 20]):.2f} cm")
```

---

#### PyTorch Code Solution

```{python}
# ==============================================================================
# Homework A03 - PyTorch Solution
# ==============================================================================
import torch

torch.manual_seed(42)
n_sim = 50
age_grid_t = torch.linspace(-0.75, 80, 200).unsqueeze(0)  # (1, 200)

alpha_t = torch.distributions.Normal(50, 5).sample((n_sim, 1))
beta_t = torch.distributions.Normal(5, 1).sample((n_sim, 1))

# Model 1: Piecewise Linear Model (Fixed Cutoff Age 20)
H_m1_t = alpha_t + beta_t * torch.clamp(age_grid_t, max=20.0)

# Model 2: Piecewise Model with Random Growth Cutoff x ~ Normal(20, 2)
x_cutoff_t = torch.distributions.Normal(20, 2).sample((n_sim, 1))
H_m2_t = alpha_t + beta_t * torch.minimum(age_grid_t, x_cutoff_t)

# Model 3: von Bertalanffy Asymptotic Growth Model
Hmax_t = torch.distributions.Normal(150, 10).sample((n_sim, 1))
b_t = torch.distributions.Normal(0.12, 0.02).sample((n_sim, 1))
H_m3_t = Hmax_t * (1 - torch.exp(-b_t * (age_grid_t + 0.75)))

print("--- Homework A03 PyTorch Tensor Simulation ---")
print(f"PyTorch Model 1 (Fixed Cutoff)  Mean Height at Age 20: {H_m1_t[:, 50].mean().item():.2f} cm")
print(f"PyTorch Model 2 (Random Cutoff) Mean Height at Age 20: {H_m2_t[:, 50].mean().item():.2f} cm")
print(f"PyTorch Model 3 (Biological)    Mean Height at Age 20: {H_m3_t[:, 50].mean().item():.2f} cm")
```

---

#### Stan Code Solution

```{stan, output.var="stan_model"}
// ==============================================================================
// Homework A03 - Stan Solution (Prior Predictive Simulation - 3 Formulations)
// ==============================================================================
data {
  int<lower=1> N_grid;
  vector[N_grid] age_grid;
}
generated quantities {
  // Model 1 & 2 Priors
  real alpha_sim = normal_rng(50, 5);
  real beta_sim = normal_rng(5, 1);
  real x_cutoff_sim = normal_rng(20, 2);
  
  // Model 3 Priors (Biological von Bertalanffy)
  real Hmax_sim = normal_rng(150, 10);
  real b_sim = normal_rng(0.12, 0.02);
  real sigma_sim = uniform_rng(0, 10);
  
  // Simulated height trajectories for all 3 models
  vector[N_grid] H_m1_sim;
  vector[N_grid] H_m2_sim;
  vector[N_grid] H_m3_sim;
  vector[N_grid] H_m3_obs_sim;
  
  for (i in 1:N_grid) {
    // Model 1: Piecewise Fixed Cutoff Age 20
    H_m1_sim[i] = alpha_sim + beta_sim * fmin(age_grid[i], 20.0);
    
    // Model 2: Piecewise Random Cutoff Age x
    H_m2_sim[i] = alpha_sim + beta_sim * fmin(age_grid[i], x_cutoff_sim);
    
    // Model 3: Biological Asymptotic Growth
    H_m3_sim[i] = Hmax_sim * (1 - exp(-b_sim * (age_grid[i] + 0.75)));
    H_m3_obs_sim[i] = normal_rng(H_m3_sim[i], sigma_sim);
  }
}
```

---

#### PyMC Code Solution

```{python}
# ==============================================================================
# Homework A03 - PyMC Solution (Prior Predictive Workflow - 3 Formulations)
# ==============================================================================
import arviz as az
import numpy as np
import pymc as pm

age_grid_eval = np.linspace(0, 80, 100)

with pm.Model() as hw_a03_model:
  # Model 1 & 2 Priors
  alpha = pm.Normal("alpha", mu=50, sigma=5)
  beta = pm.Normal("beta", mu=5, sigma=1)
  x_cutoff = pm.Normal("x_cutoff", mu=20, sigma=2)

  # Model 3 Priors (Biological von Bertalanffy)
  Hmax = pm.Normal("Hmax", mu=150, sigma=10)
  b = pm.Normal("b", mu=0.12, sigma=0.02)
  sigma = pm.Uniform("sigma", lower=0, upper=10)

  # Deterministic nodes for the three growth formulations
  H_m1 = pm.Deterministic(
      "H_m1", alpha + beta * pm.math.minimum(age_grid_eval, 20.0)
  )
  H_m2 = pm.Deterministic(
      "H_m2", alpha + beta * pm.math.minimum(age_grid_eval, x_cutoff)
  )
  H_m3 = pm.Deterministic(
      "H_m3", Hmax * (1 - pm.math.exp(-b * (age_grid_eval + 0.75)))
  )

  # Prior predictive sampling
  idata_hw_a03 = pm.sample_prior_predictive(samples=1000, random_seed=42)

# Extract prior predictive arrays
H1_samples = idata_hw_a03.prior["H_m1"].values.squeeze()
H2_samples = idata_hw_a03.prior["H_m2"].values.squeeze()
H3_samples = idata_hw_a03.prior["H_m3"].values.squeeze()

print("--- Homework A03 PyMC Summary (3 Formulations) ---")
print(f"Model 1 (Fixed Cutoff 20)  Mean Height at Age 20: {np.mean(H1_samples[:, 25]):.2f} cm")
print(f"Model 2 (Random Cutoff x) Mean Height at Age 20: {np.mean(H2_samples[:, 25]):.2f} cm")
print(f"Model 3 (Biological)       Mean Height at Age 20: {np.mean(H3_samples[:, 25]):.2f} cm")
```
