---
title: "A05 — Estimands and Estiplans"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: true
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# A05 — Estimands and Estiplans

**Course:** Statistical Rethinking 2026  
**Section:** A — Beginner  
**Lecture:** A05 — Estimands and Estiplans  
**Official reading:** Chapters 4 and 5 of *Statistical Rethinking*, second edition  
**Central question:** How do we separate the scientific quantity we want to learn from the statistical and computational plan used to estimate it?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official repository script `05_elemental_confounds.r`, especially its `WaffleDivorce` analysis, standardization, prior predictive simulation, and multivariable regression.
>
> The words **estimand** and **estiplan** are more explicit organizing concepts in the 2026 course than in the second edition. The book supplies the regression examples, causal questions, posterior-processing tools, and plotting distinctions. Where this guide formalizes an estiplan as a sequence of operations, that is a reconstruction of the lecture's organizing framework rather than a book section title or a verbatim transcript.

---

## Table of Contents

- [1. Main thesis](#1-main-thesis)
- [2. Relationship to the book](#2-relationship-to-the-book)
  - [2.1 What the 2026 organization adds](#21-what-the-2026-organization-adds)
- [3. Four distinct objects](#3-four-distinct-objects)
  - [3.1 Estimand](#31-estimand)
  - [3.2 Estimator](#32-estimator)
  - [3.3 Estimate](#33-estimate)
  - [3.4 Estiplan](#34-estiplan)
- [4. Why coefficients are often not the target](#4-why-coefficients-are-often-not-the-target)
- [5. Identification before estimation](#5-identification-before-estimation)
- [6. The WaffleDivorce example](#6-the-waffle-divorce-example)
  - [6.1 Several possible questions](#61-several-possible-questions)
- [7. Standardization in the divorce analysis](#7-standardization-in-the-divorce-analysis)
- [8. Prior predictive simulation](#8-prior-predictive-simulation)
- [9. Three regressions, three conditional questions](#9-three-regressions-three-conditional-questions)
  - [9.1 Divorce from marriage rate](#91-divorce-from-marriage-rate)
  - [9.2 Divorce from age at marriage](#92-divorce-from-age-at-marriage)
  - [9.3 Divorce from both predictors](#93-divorce-from-both-predictors)
- [10. A candidate causal graph](#10-a-candidate-causal-graph)
- [11. Estimand 1: total effect of age at marriage](#11-estimand-1-total-effect-of-age-at-marriage)
- [12. Estimand 2: direct effect of age at marriage](#12-estimand-2-direct-effect-of-age-at-marriage)
- [13. Estimand 3: effect of marriage rate](#13-estimand-3-effect-of-marriage-rate)
- [14. Estimand 4: predictive divorce risk](#14-estimand-4-predictive-divorce-risk)
- [15. Posterior standardization](#15-posterior-standardization)
  - [15.1 The averaging distribution is part of the target](#151-the-averaging-distribution-is-part-of-the-target)
- [16. Conditioning is not intervention](#16-conditioning-is-not-intervention)
- [17. Three plot types from Chapter 5](#17-three-plot-types-from-chapter-5)
  - [17.1 Predictor residual plots](#171-predictor-residual-plots)
  - [17.2 Posterior prediction plots](#172-posterior-prediction-plots)
  - [17.3 Counterfactual plots](#173-counterfactual-plots)
- [18. Spurious association](#18-spurious-association)
- [19. Masked relationships](#19-masked-relationships)
- [20. One estimand, several estiplans](#20-one-estimand-several-estiplans)
  - [20.1 Outcome-regression estiplan](#201-outcome-regression-estiplan)
  - [20.2 Weighting estiplan](#202-weighting-estiplan)
  - [20.3 Doubly robust estiplan](#203-doubly-robust-estiplan)
  - [20.4 Randomized-experiment estiplan](#204-randomized-experiment-estiplan)
  - [20.5 Intuitive conceptual summary of the four estiplans](#205-intuitive-conceptual-summary-of-the-four-estiplans)

- [21. One model, several estimands](#21-one-model-several-estimands)
- [22. Positivity and support](#22-positivity-and-support)
- [23. Estiplan validation by simulation](#23-estiplan-validation-by-simulation)
- [24. Common errors exposed by A05](#24-common-errors-exposed-by-a05)
- [25. CRM application: discount uplift](#25-crm-application-discount-uplift)
  - [25.1 Estimand](#251-estimand)
  - [25.2 Identification assumption](#252-identification-assumption)
  - [25.3 Outcome-regression estiplan](#253-outcome-regression-estiplan)
  - [25.4 Why the treatment coefficient may not be the ATE](#254-why-the-treatment-coefficient-may-not-be-the-ate)
- [26. CRM application: several offer amounts](#26-crm-application-several-offer-amounts)
- [27. Implications for CausalOps](#27-implications-for-causalops)
  - [27.1 Example estimand schema](#271-example-estimand-schema)
  - [27.2 Example estiplan schema](#272-example-estiplan-schema)
  - [27.3 Fail-closed logic](#273-fail-closed-logic)
- [28. Lecture takeaways](#28-lecture-takeaways)
- [29. Review questions (Detailed Answers & Multi-Language Implementation)](#29-review-questions)
- [30. Book reading guide](#30-book-reading-guide)
- [31. Official code guide](#31-official-code-guide)
- [32. Sources](#32-sources)

---

## 1. Main thesis

The central argument is:

> **The target of an analysis and the plan used to estimate that target are different objects.**

An **estimand** states exactly what scientific quantity the analysis is intended to learn.

An **estiplan** states how the analyst intends to move from assumptions and observations to an estimate of that quantity.

The workflow is:

$$ \text{scientific question} \rightarrow \text{causal model} \rightarrow \text{estimand} \rightarrow \text{identification} \rightarrow \text{estiplan} \rightarrow \text{estimate} \rightarrow \text{diagnosis and sensitivity} $$

A fitted regression is only one component of this workflow.

A complete estiplan may include:

1. defining interventions and comparison states;
2. defining the target population;
3. selecting variables required for identification;
4. transforming variables and calibrating priors;
5. fitting one or more statistical models;
6. generating posterior predictions under interventions;
7. standardizing over a population distribution;
8. constructing the required contrast;
9. propagating posterior uncertainty;
10. checking support, fit, and estimator recovery.

The practical principle is:

$$ \text{regression coefficient} \neq \text{estimand in general} $$

A coefficient may be an intermediate model parameter while the scientific result is a posterior contrast of standardized predictions.

---

## 2. Relationship to the book

A05 draws primarily from Chapters 4 and 5, while preparing the formal confounding material of Chapter 6.

| A05 concept | Second-edition counterpart | Relationship |
|---|---|---|
| Linear prediction | Section 4.4 | Statistical model used inside an estiplan |
| Posterior expectations and simulations | Sections 4.4.3–4.4.5 | Convert parameters into outcome-scale quantities |
| Spurious association | Section 5.1 | Marginal association may not represent the target effect |
| Multiple regression | Sections 5.1–5.2 | Estimates conditional relationships under a chosen model |
| Predictor residual plots | Section 5.1.5.1 | Explain statistical conditioning |
| Posterior prediction plots | Section 5.1.5.2 | Evaluate fit and predictive implications |
| Counterfactual plots | Section 5.1.5.3 | Display predictions under hypothetical manipulations |
| Masked relationships | Section 5.2 | Conditioning can reveal an association hidden marginally |
| Category contrasts | Section 5.3 | Estimands constructed from posterior parameters |
| Formal causal adjustment | Chapter 6 | Identification logic developed in A06 and A07 |

### 2.1 What the 2026 organization adds

The second edition discusses causal questions, conditional regressions, counterfactual predictions, and posterior contrasts. It does not organize Chapter 5 under a named pair:

$$ \text{estimand} + \text{estiplan} $$

A05 makes that separation explicit.

The lecture-oriented reorganization is:

$$ \text{book regression workflow} + \text{precise scientific target} + \text{auditable estimation sequence} $$

---

## 3. Four distinct objects

### 3.1 Estimand

The estimand is the quantity the study seeks to learn.

Examples include:

$$ \mathbb{E}[Y] $$

$$ \mathbb{E}[Y \mid X=x] $$

$$ \mathbb{E}[Y(1)-Y(0)] $$

$$ \mathbb{E}[Y \mid \mathrm{do}(A=a_1)] - \mathbb{E}[Y \mid \mathrm{do}(A=a_0)] $$

An estimand must specify enough detail to remove ambiguity about:

- the outcome;
- the intervention or exposure;
- the comparison;
- the target population;
- the aggregation rule;
- the time horizon;
- and the measurement scale.

### 3.2 Estimator

An estimator is a rule mapping possible datasets to estimates:

$$ \widehat{\tau} = g(\mathcal{D}) $$

Examples include a difference in sample means, a standardized regression contrast, or an inverse-probability weighted mean.

### 3.3 Estimate

The estimate is the output of the estimator for the observed data:

$$ \widehat{\tau}_{\mathrm{obs}} = g(\mathcal{D}_{\mathrm{obs}}) $$

### 3.4 Estiplan

An estiplan is broader than the estimator. It is the complete planned sequence connecting the scientific model to the reported estimate.

A useful representation is:

$$ \mathcal{P} = \left( \mathcal{G}, \tau, \mathcal{I}, \mathcal{M}, \mathcal{T}, \mathcal{A}, \mathcal{C}, \mathcal{V} \right) $$

where:

- $\mathcal{G}$ is the causal or generative model;
- $\tau$ is the estimand;
- $\mathcal{I}$ is the identification argument;
- $\mathcal{M}$ is the statistical model or collection of models;
- $\mathcal{T}$ is the intervention or prediction transformation;
- $\mathcal{A}$ is the averaging or standardization operation;
- $\mathcal{C}$ is the final contrast;
- $\mathcal{V}$ is the validation and diagnostic program.

The same estimand can be targeted by several estiplans. The same fitted model can contribute to several estimands.

---

## 4. Why coefficients are often not the target

Consider:

$$ Y_i \sim \mathrm{Normal}(\mu_i,\sigma) $$

$$ \mu_i = \alpha+\beta_A A_i+\beta_X X_i $$

It is tempting to declare:

$$ \tau=\beta_A $$

This is justified only under restrictive conditions, including:

- a linear response in $A$;
- no relevant treatment interactions;
- correct adjustment for confounding;
- no harmful post-treatment controls;
- a meaningful intervention on $A$;
- and a target matching the conditional slope.

With heterogeneous effects, the target may instead be:

$$ \tau = \mathbb{E}_X \left[ \mathbb{E}[Y \mid A=1,X] - \mathbb{E}[Y \mid A=0,X] \right] $$

The regression produces conditional expectations. Prediction, averaging, and subtraction produce the estimand.

The coefficient is scaffolding. The outcome-scale posterior contrast is the scientific result.

#### Multi-Language Implementation: Coefficient vs Standardized Estimand

```r
# R Code: G-Computation / Standardization for Average Treatment Effect
post <- extract.samples(m5.3) # Fitted outcome model with covariates
# Predict for all units under A=1 and A=0
mu_A1 <- link(m5.3, data = data.frame(A = 1, X = d$X))
mu_A0 <- link(m5.3, data = data.frame(A = 0, X = d$X))
# Draw-by-draw population standardization
ate_draws <- rowMeans(mu_A1) - rowMeans(mu_A0)
precis(ate_draws)
```

```python
# PyMC Code: G-Computation Standardization
import numpy as np
import pymc as pm

# Draw-by-draw counterfactual prediction and standardization
# alpha_draws (S, 1), beta_A (S, 1), beta_X (S, 1), X_obs (N,)
with pm.Model() as model:
  alpha = pm.Normal("alpha", mu=0, sigma=0.2)
  beta_A = pm.Normal("beta_A", mu=0, sigma=0.5)
  beta_X = pm.Normal("beta_X", mu=0, sigma=0.5)
  sigma = pm.Exponential("sigma", lam=1.0)
  mu = alpha + beta_A * A_obs + beta_X * X_obs
  pm.Normal("Y_obs", mu=mu, sigma=sigma, observed=Y_obs)
  idata = pm.sample(draws=1000, tune=1000, chains=2, progressbar=False)

post = idata.posterior
alpha_s = post["alpha"].values.flatten()
beta_A_s = post["beta_A"].values.flatten()
beta_X_s = post["beta_X"].values.flatten()

pred_A1 = (
    alpha_s[:, None]
    + beta_A_s[:, None] * 1.0
    + beta_X_s[:, None] * X_obs[None, :]
)
pred_A0 = (
    alpha_s[:, None]
    + beta_A_s[:, None] * 0.0
    + beta_X_s[:, None] * X_obs[None, :]
)

ate_draws = np.mean(pred_A1 - pred_A0, axis=1)
print(f"PyMC ATE Posterior Mean: {np.mean(ate_draws):.3f}")
```

```python
# PyTorch Code: GPU Vectorized Standardization
import torch

pred_A1_t = alpha_t + beta_A_t * 1.0 + beta_X_t * X_t.unsqueeze(0)
pred_A0_t = alpha_t + beta_A_t * 0.0 + beta_X_t * X_t.unsqueeze(0)
ate_t = (pred_A1_t - pred_A0_t).mean(dim=1)
print(f"PyTorch ATE Mean: {ate_t.mean().item():.3f}")
```

---

## 5. Identification before estimation

An estimand is not automatically identifiable from observed data.

Suppose:

$$ \tau = \mathbb{E}[Y(1)-Y(0)] $$

Only one potential outcome is observed for each unit:

$$ Y_i = A_iY_i(1) + (1-A_i)Y_i(0) $$

Under conditional exchangeability:

$$ Y(a) \perp\!\!\!\perp A \mid X $$

plus consistency and positivity, the intervention mean is identified by:

$$ \mathbb{E}[Y(a)] = \int \mathbb{E}[Y \mid A=a,X=x] p(x) \,dx $$

For discrete $X$:

$$ \mathbb{E}[Y(a)] = \sum_x \mathbb{E}[Y \mid A=a,X=x] p(x) $$

This formula already specifies the skeleton of an estiplan:

1. estimate the conditional outcome model;
2. set treatment to $a$;
3. predict for each covariate profile;
4. average over the target distribution.

Identification describes what observable quantities recover the estimand under assumptions. Estimation describes how those quantities are approximated from finite data.

---

## 6. The `WaffleDivorce` example

The official repository script uses the `WaffleDivorce` data for US states.

The main variables are:

- $D$: divorce rate;
- $M$: marriage rate;
- $A$: median age at marriage;
- $W$: Waffle House density;
- $S$: southern-region indicator.

The humorous marginal pattern is:

$$ W \leftrightarrow D $$

States with more Waffle Houses tend to have higher divorce rates.

This does not imply:

$$ W \rightarrow D $$

A regional variable could influence both:

$$ S \rightarrow W $$

$$ S \rightarrow D $$

The Waffle House regression can summarize geography without representing the effect of opening restaurants.

```{r}
#| label: fig-waffle-correlations
#| fig-cap: "Marginal associations between Waffle House density, Marriage rate, Median Age at Marriage, and Divorce rate."
#| echo: true
library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce
d$D_std <- as.numeric(scale(d$Divorce))
d$M_std <- as.numeric(scale(d$Marriage))
d$A_std <- as.numeric(scale(d$MedianAgeMarriage))

par(mfrow = c(1, 3))
plot(D_std ~ WaffleHouses, data = d, col = "steelblue", pch = 19, xlab = "Waffle Houses / M pop", ylab = "Divorce Rate (std)", main = "Waffle vs Divorce")
abline(lm(D_std ~ WaffleHouses, data = d), lty = 2)

plot(D_std ~ M_std, data = d, col = "firebrick", pch = 19, xlab = "Marriage Rate (std)", ylab = "Divorce Rate (std)", main = "Marriage vs Divorce")
abline(lm(D_std ~ M_std, data = d), lty = 2)

plot(D_std ~ A_std, data = d, col = "forestgreen", pch = 19, xlab = "Median Age Marriage (std)", ylab = "Divorce Rate (std)", main = "Age vs Divorce")
abline(lm(D_std ~ A_std, data = d), lty = 2)
par(mfrow = c(1, 1))
```

### 6.1 Several possible questions

The same dataset can support different questions:

- Which states have high predicted divorce rates?
- What is the association between Waffle House density and divorce?
- What is the effect of opening more Waffle Houses?
- What is the total effect of median age at marriage on divorce?
- What is the effect of marriage rate after accounting for age at marriage?

Each question defines a different estimand and may require a different estiplan.

---

## 7. Standardization in the divorce analysis

The official script standardizes divorce rate, marriage rate, and median age at marriage:

$$ D_i^* = \frac{D_i-\overline{D}}{s_D} $$

$$ M_i^* = \frac{M_i-\overline{M}}{s_M} $$

$$ A_i^* = \frac{A_i-\overline{A}}{s_A} $$

Standardization helps with:

- prior calibration;
- interpretation in standard-deviation units;
- numerical geometry;
- visual comparison of coefficients.

It does not:

- define the estimand;
- remove confounding;
- justify a causal graph;
- or solve lack of support.

A standardized coefficient answers a transformed-scale question. The final result may need conversion to original units for scientific or policy interpretation.

---

## 8. Prior predictive simulation

The script first considers very broad priors such as:

$$ \alpha \sim \mathrm{Normal}(0,10) $$

$$ \beta_M \sim \mathrm{Normal}(0,10) $$

$$ \beta_A \sim \mathrm{Normal}(0,10) $$

For standardized variables, these priors imply extreme outcome surfaces over modest predictor values.

The script then uses tighter priors:

$$ \alpha \sim \mathrm{Normal}(0,0.2) $$

$$ \beta_M \sim \mathrm{Normal}(0,0.5) $$

$$ \beta_A \sim \mathrm{Normal}(0,0.5) $$

with:

$$ \mu_i = \alpha+\beta_M M_i^*+\beta_A A_i^* $$

Prior predictive simulation checks whether the statistical component of the estiplan generates plausible relationships before conditioning on data.

It does not validate the causal graph or adjustment set.

```{r}
#| label: fig-prior-predictive-a05
#| fig-cap: "Prior predictive simulation comparing overly broad priors vs calibrated priors for standardized divorce regression."
#| echo: true
set.seed(10)
N_lines <- 50

# Broad priors: Normal(0, 10)
a_broad <- rnorm(N_lines, 0, 10)
bM_broad <- rnorm(N_lines, 0, 10)

# Tighter priors: Normal(0, 0.2) and Normal(0, 0.5)
a_tight <- rnorm(N_lines, 0, 0.2)
bM_tight <- rnorm(N_lines, 0, 0.5)

x_seq <- seq(-2, 2, length.out = 30)

par(mfrow = c(1, 2))
plot(NULL, xlim = c(-2, 2), ylim = c(-10, 10), xlab = "Marriage Rate (std)", ylab = "Divorce Rate (std)", main = "Broad Priors: Normal(0, 10)")
for (i in 1:N_lines) lines(x_seq, a_broad[i] + bM_broad[i] * x_seq, col = rgb(0.8, 0.2, 0.2, 0.4))

plot(NULL, xlim = c(-2, 2), ylim = c(-3, 3), xlab = "Marriage Rate (std)", ylab = "Divorce Rate (std)", main = "Tighter Priors: Normal(0, 0.5)")
for (i in 1:N_lines) lines(x_seq, a_tight[i] + bM_tight[i] * x_seq, col = rgb(0.2, 0.6, 0.2, 0.6))
par(mfrow = c(1, 1))
```

---

## 9. Three regressions, three conditional questions

### 9.1 Divorce from marriage rate

$$ D_i^* \sim \mathrm{Normal}(\mu_i,\sigma) $$

$$ \mu_i = \alpha+\beta_M M_i^* $$

This estimates:

$$ \mathbb{E}[D^* \mid M^*] $$

under a linear Gaussian model.

### 9.2 Divorce from age at marriage

$$ \mu_i = \alpha+\beta_A A_i^* $$

This estimates:

$$ \mathbb{E}[D^* \mid A^*] $$

### 9.3 Divorce from both predictors

The official script fits:

$$ D_i^* \sim \mathrm{Normal}(\mu_i,\sigma) $$

$$ \mu_i = \alpha+\beta_M M_i^*+\beta_A A_i^* $$

with:

$$ \sigma \sim \mathrm{Exponential}(1) $$

This estimates:

$$ \mathbb{E}[D^* \mid M^*,A^*] $$

The coefficient $\beta_M$ describes marriage-rate variation among states with similar median age at marriage, under the additive linear model.

The coefficient $\beta_A$ describes age-at-marriage variation among states with similar marriage rates.

These conditional coefficients are not automatically total causal effects.

```{r}
#| label: fit-divorce-models
#| echo: true
m5.1 <- quap(
  alist(
    D_std ~ dnorm(mu, sigma),
    mu <- a + bM * M_std,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d, start = list(a = 0, bM = 0, sigma = 1)
)

m5.2 <- quap(
  alist(
    D_std ~ dnorm(mu, sigma),
    mu <- a + bA * A_std,
    a ~ dnorm(0, 0.2),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d, start = list(a = 0, bA = 0, sigma = 1)
)

m5.3 <- quap(
  alist(
    D_std ~ dnorm(mu, sigma),
    mu <- a + bM * M_std + bA * A_std,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = d, start = list(a = 0, bM = 0, bA = 0, sigma = 1)
)

plot(coeftab(m5.1, m5.2, m5.3))
```

#### PyMC Implementation: Model 5.3 (Multivariable Regression)

```python
# PyMC Implementation: Multivariable Divorce Model (m5.3)
import arviz as az
import numpy as np
import pymc as pm

with pm.Model() as model_5_3:
  alpha = pm.Normal("alpha", mu=0, sigma=0.2)
  beta_M = pm.Normal("beta_M", mu=0, sigma=0.5)
  beta_A = pm.Normal("beta_A", mu=0, sigma=0.5)
  sigma = pm.Exponential("sigma", lam=1.0)

  mu = alpha + beta_M * M_std + beta_A * A_std
  D_obs = pm.Normal("D_obs", mu=mu, sigma=sigma, observed=D_std)

  idata_5_3 = pm.sample(draws=1000, tune=1000, chains=4, random_seed=42)

# Summarize posterior parameters
print(az.summary(idata_5_3, var_names=["alpha", "beta_M", "beta_A", "sigma"]))
```

---

## 10. A candidate causal graph

A common scientific interpretation is:

$$ A \rightarrow M $$

$$ A \rightarrow D $$

with a possible additional path:

$$ M \rightarrow D $$

Here:

- median age at marriage influences marriage rate;
- median age at marriage influences divorce rate;
- marriage rate may itself influence divorce rate.

This graph is an assumption informed by scientific reasoning. The observed correlations do not uniquely determine it.

The graph determines which regression and post-processing operations correspond to which estimand.

---

## 11. Estimand 1: total effect of age at marriage

Suppose the target is:

$$ \tau_A^{\mathrm{total}} = \mathbb{E} \left[ D \mid \mathrm{do}(A=a_1) \right] - \mathbb{E} \left[ D \mid \mathrm{do}(A=a_0) \right] $$

Under the candidate graph, part of the effect may operate through:

$$ A \rightarrow M \rightarrow D $$

Conditioning on $M$ blocks this mediated pathway.

Therefore, an estiplan for the total effect should not simply report $\beta_A$ from the model that conditions on $M$.

A simple total-effect estiplan is:

1. fit $D$ as a function of $A$ and valid pre-exposure confounders;
2. predict under $A=a_1$ and $A=a_0$;
3. average over the target population;
4. subtract the intervention means.

In a linear model without interactions:

$$ \tau_A^{\mathrm{total}} = \beta_A(a_1-a_0) $$

But the posterior contrast is the more general representation.

---

## 12. Estimand 2: direct effect of age at marriage

A controlled direct effect at marriage rate $m$ is:

$$ \tau_A^{\mathrm{direct}}(m) = \mathbb{E} \left[ D \mid \mathrm{do}(A=a_1), \mathrm{do}(M=m) \right] - \mathbb{E} \left[ D \mid \mathrm{do}(A=a_0), \mathrm{do}(M=m) \right] $$

An estiplan may use:

$$ \mu = \alpha+\beta_A A+\beta_M M $$

and compare predictions while holding $M=m$.

A direct-effect interpretation requires stronger assumptions than a total-effect analysis. In particular, mediator–outcome confounding must be addressed, and the precise direct-effect definition must be explicit.

The coefficient $\beta_A$ is not automatically a natural direct effect.

---

## 13. Estimand 3: effect of marriage rate

Suppose the target is:

$$ \tau_M(m_1,m_0) = \mathbb{E} \left[ D \mid \mathrm{do}(M=m_1) \right] - \mathbb{E} \left[ D \mid \mathrm{do}(M=m_0) \right] $$

Under the candidate graph, age at marriage is a common cause:

$$ M \leftarrow A \rightarrow D $$

An estiplan therefore adjusts for $A$:

1. fit $D$ conditional on $M$ and $A$;
2. set $M=m_1$ for every state while retaining each state's $A$;
3. predict divorce under that intervention;
4. repeat with $M=m_0$;
5. average each prediction set;
6. subtract.

Formally:

$$ \mathbb{E}[D \mid \mathrm{do}(M=m)] = \int \mathbb{E}[D \mid M=m,A=a] p(a) \,da $$

This is standardization.

The coefficient $\beta_M$ equals the standardized contrast per unit only under the additive linear model without interactions.

---

## 14. Estimand 4: predictive divorce risk

A predictive question differs from a causal question.

For a state with $M=m$ and $A=a$, the expected outcome is:

$$ \mathbb{E} \left[ D_{\mathrm{new}} \mid M=m,A=a,\mathcal{D}_{\mathrm{obs}} \right] $$

The predictive distribution is:

$$ p \left( D_{\mathrm{new}} \mid M=m,A=a,\mathcal{D}_{\mathrm{obs}} \right) $$

A predictive estiplan may include a variable solely because it improves forecasts, even when that variable would be a harmful control for a causal effect.

A variable can therefore be:

- useful for prediction;
- harmful for identification;
- or irrelevant to the chosen estimand.

Model specification cannot be judged independently of the task.

---

## 15. Posterior standardization

Suppose:

$$ \mu_i = \alpha+\beta_A A_i+\beta_M M_i $$

For posterior draw $s$, generate counterfactual expectations under $M=m$:

$$ \mu_i^{(s)}(m) = \alpha^{(s)} + \beta_A^{(s)}A_i + \beta_M^{(s)}m $$

Average over the target population:

$$ \overline{\mu}^{(s)}(m) = \frac{1}{N} \sum_{i=1}^{N} \mu_i^{(s)}(m) $$

Then compute:

$$ \tau_M^{(s)} = \overline{\mu}^{(s)}(m_1) - \overline{\mu}^{(s)}(m_0) $$

The collection of $\tau_M^{(s)}$ values approximates the posterior distribution of the estimand.

All transformations are performed draw by draw, preserving joint posterior uncertainty.

```{r}
#| label: fig-gcomp-standardization
#| fig-cap: "Draw-by-draw G-computation standardization for the counterfactual effect of Marriage Rate."
#| echo: true
post53 <- extract.samples(m5.3)
S_draws <- length(post53$a)

# Counterfactual predictions for M = 1 vs M = -1 over observed sample A_std
m1_preds <- sapply(1:S_draws, function(s) mean(post53$a[s] + post53$bM[s] * 1 + post53$bA[s] * d$A_std))
m0_preds <- sapply(1:S_draws, function(s) mean(post53$a[s] + post53$bM[s] * (-1) + post53$bA[s] * d$A_std))

ate_m_draws <- m1_preds - m0_preds
dens(ate_m_draws, col = "firebrick", lwd = 2, xlab = "Standardized ATE of Marriage Rate (M=1 vs M=-1)", main = "G-Computation Posterior Contrast")
abline(v = 0, lty = 2, col = "gray")
```

#### PyMC Implementation: Posterior Standardization (G-Computation)

```python
# PyMC Implementation: Draw-by-Draw G-Computation Standardization
import numpy as np

# Extract posterior draws
post = idata_5_3.posterior
alpha_s = post["alpha"].values.flatten()  # shape (S,)
beta_M_s = post["beta_M"].values.flatten()  # shape (S,)
beta_A_s = post["beta_A"].values.flatten()  # shape (S,)

# Predict for all units under M=1 vs M=-1 across observed A_std population
pred_M1 = (
    alpha_s[:, None]
    + beta_M_s[:, None] * 1.0
    + beta_A_s[:, None] * A_std[None, :]
)
pred_M0 = (
    alpha_s[:, None]
    + beta_M_s[:, None] * (-1.0)
    + beta_A_s[:, None] * A_std[None, :]
)

# Population standardization (average across N units per draw)
ate_m_draws_pymc = np.mean(pred_M1 - pred_M0, axis=1)

print(f"PyMC Standardized ATE Mean: {np.mean(ate_m_draws_pymc):.3f}")
print(f"PyMC 89% HDI: {np.percentile(ate_m_draws_pymc, [5.5, 94.5])}")
```

### 15.1 The averaging distribution is part of the target

#### Why the Target Population Defines the Estimand

When causal effects are heterogeneous across units—meaning the individual treatment effect $\tau(x) = \mathbb{E}[Y(1) - Y(0) \mid X=x]$ varies as a function of background covariates $X$—there is no single "treatment effect." The overall marginal effect depends directly on **how the covariate profiles $X$ are distributed in the target population**.

Mathematically, the Marginal Average Treatment Effect $\tau$ under G-computation is defined as:

$$ \tau = \int \left( \mathbb{E}[Y \mid A=a_1, X=x] - \mathbb{E}[Y \mid A=a_0, X=x] \right) p_{\mathrm{target}}(x) \, dx $$

Notice that $p_{\mathrm{target}}(x)$ appears explicitly in the integrand. If two analysts fit the **exact same statistical model** $\mathbb{E}[Y \mid A, X]$, but evaluate the marginal contrast over different covariate distributions $p_A(x)$ versus $p_B(x)$, they will obtain **different numerical values** ($\tau_A \neq \tau_B$). Both values are mathematically correct, but they answer **different scientific questions**.

#### Sample Distribution vs. Policy Target Distribution

1. **Sample Average Treatment Effect (SATE):**
   Averaging over the empirical distribution of the observed sample $p_{\mathrm{sample}}(x) = \frac{1}{N} \sum_{i=1}^N \delta(x - X_i)$:

   $$ \tau_{\mathrm{SATE}} = \frac{1}{N} \sum_{i=1}^N \left( \mu(a_1, X_i) - \mu(a_0, X_i) \right) $$

   In the `WaffleDivorce` example, this estimates the counterfactual effect of marriage rate $M$ standardized over the specific demographic distribution of the 50 US states observed in the dataset.

2. **Population Average Treatment Effect (PATE) or Policy Target:**
   If a policy intervention is planned for a different target population (e.g. future demographic projections, eligible customer segments, or a specific geographic region), the target distribution $p_{\mathrm{target}}(x)$ differs from $p_{\mathrm{sample}}(x)$:

   $$ \mathbb{E}_{\mathrm{target}}[D(m)] = \int \mathbb{E}[D \mid M=m, A=a] p_{\mathrm{target}}(a) \, da $$

   $$ \tau_{\mathrm{PATE}} = \mathbb{E}_{\mathrm{target}}[D(m_1)] - \mathbb{E}_{\mathrm{target}}[D(m_0)] $$

#### Why This Is an Estimand Property, Not a Code Detail

In traditional regression workflows, standardizing or predicting at specific covariate values is often treated as an informal post-processing step. **A05 reframes this fundamental distinction:**

- The statistical model $p(Y \mid A, X, \boldsymbol{\theta})$ estimates conditional expectations $\mu(a, x)$.
- The **estimand** specifies the scientific question, which **must include the target population distribution $p_{\mathrm{target}}(X)$**.
- Omitting $p_{\mathrm{target}}(X)$ leaves the estimand ill-defined.

#### Multi-Language Implementation: Standardizing Over Custom Target Distributions

```r
# R Code: G-Computation over Sample vs Custom Target Population
post <- extract.samples(m5.3)

# 1. Sample Standardization (SATE): Average over observed states d$A_std
sate_draws <- sapply(1:length(post$a), function(s) {
  mean((post$a[s] + post$bM[s]*1 + post$bA[s]*d$A_std) - 
       (post$a[s] + post$bM[s]*(-1) + post$bA[s]*d$A_std))
})

# 2. Target Policy Standardization (PATE): Average over a high-age target population
# Suppose target population has older median age at marriage (A_target ~ Normal(1.0, 0.5))
set.seed(123)
A_target <- rnorm(1000, mean = 1.0, sd = 0.5)

pate_target_draws <- sapply(1:length(post$a), function(s) {
  mean((post$a[s] + post$bM[s]*1 + post$bA[s]*A_target) - 
       (post$a[s] + post$bM[s]*(-1) + post$bA[s]*A_target))
})

cat(sprintf("SATE Mean: %.3f | Target PATE Mean: %.3f\n", mean(sate_draws), mean(pate_target_draws)))
```

```python
# PyMC Code: G-Computation over Sample vs Custom Target Population
import numpy as np

post = idata_5_3.posterior
alpha_s = post["alpha"].values.flatten()
beta_M_s = post["beta_M"].values.flatten()
beta_A_s = post["beta_A"].values.flatten()

# 1. Sample Standardization (SATE)
pred_M1_sample = (
    alpha_s[:, None]
    + beta_M_s[:, None] * 1.0
    + beta_A_s[:, None] * A_std[None, :]
)
pred_M0_sample = (
    alpha_s[:, None]
    + beta_M_s[:, None] * (-1.0)
    + beta_A_s[:, None] * A_std[None, :]
)
sate_draws = np.mean(pred_M1_sample - pred_M0_sample, axis=1)

# 2. Target Policy Standardization (PATE) over custom covariate distribution A_target
np.random.seed(123)
A_target = np.random.normal(loc=1.0, scale=0.5, size=1000)

pred_M1_target = (
    alpha_s[:, None]
    + beta_M_s[:, None] * 1.0
    + beta_A_s[:, None] * A_target[None, :]
)
pred_M0_target = (
    alpha_s[:, None]
    + beta_M_s[:, None] * (-1.0)
    + beta_A_s[:, None] * A_target[None, :]
)
pate_target_draws = np.mean(pred_M1_target - pred_M0_target, axis=1)

print(
    f"PyMC SATE Mean: {np.mean(sate_draws):.3f} | PyMC Target PATE Mean:"
    f" {np.mean(pate_target_draws):.3f}"
)
```


---

## 16. Conditioning is not intervention

An ordinary conditional prediction asks about:

$$ p(D \mid M=m,A=a) $$

An interventional prediction asks about:

$$ p(D \mid \mathrm{do}(M=m)) $$

Conditioning describes units naturally observed with $M=m$.

Intervention represents forcing $M$ to $m$, thereby breaking the ordinary causes of $M$ while preserving the rest of the causal system.

They coincide only under identification assumptions.

The key distinction is:

$$ \text{conditioning operation} \neq \text{intervention operation} $$

Prediction software can substitute a chosen value into a model. The causal interpretation comes from the graph and assumptions, not from the substitution command.

---

## 17. Three plot types from Chapter 5

### 17.1 Predictor residual plots

If marriage rate is modeled from age at marriage:

$$ M_i = \gamma_0+ \gamma_A A_i+ R_i $$

then $R_i$ represents marriage-rate variation not predicted by age under the auxiliary model.

Plotting divorce against $R$ helps visualize the conditional coefficient for $M$.

This is a tool for understanding the fitted statistical model, not a complete causal analysis.

### 17.2 Posterior prediction plots

Posterior prediction plots compare fitted predictions with observations.

They help diagnose:

- prediction errors;
- outliers;
- systematic lack of fit;
- whether the model reproduces relevant patterns.

They do not establish identification.

### 17.3 Counterfactual plots

Counterfactual plots display model predictions under hypothetical manipulations where one or more variables are artificially held fixed while another is varied.

#### Detailed Breakdown of the Counterfactual Plot (`M = 0`, `A` from `-2` to `+2`)

In the R code from Chunk `fig-three-plot-types`:

```r
# Counterfactual plot: Manipulating A holding M=0
a_seq <- seq(-2, 2, length.out = 30)
pred_cf <- link(m5.3, data = data.frame(A_std = a_seq, M_std = 0))
plot(a_seq, colMeans(pred_cf), type = "l", lwd = 2, col = "purple", 
     xlab = "Manipulated Age (A)", ylab = "Predicted Divorce (D)", 
     main = "3. Counterfactual Plot (M=0)", ylim = c(-2, 2))
shade(apply(pred_cf, 2, HPDI, prob = 0.89), a_seq)
```

#### 1. Why does the Y-axis (Predicted Divorce $D$) range from $+1$ to $-1$?

The outcome variable $D_{\mathrm{std}}$ is **standardized** ($D^* = \frac{D - \overline{D}}{s_D}$):

- **$D^* = 0$** represents the average divorce rate in the US dataset ($\approx 9.6$ divorces per 1,000 adults).
- **$D^* = +1.0$** represents a divorce rate **1 standard deviation above average** ($\approx 11.4$ per 1,000 adults).
- **$D^* = -1.0$** represents a divorce rate **1 standard deviation below average** ($\approx 7.8$ per 1,000 adults).

#### 2. Why is Marriage Rate fixed at $M = 0$?

Setting $M_{\mathrm{std}} = 0$ corresponds to holding marriage rate artificially constant at the sample mean ($0$ standard deviations from average). 

Under the linear equation for Model 5.3 ($\mu_i = \alpha + \beta_M M_i^* + \beta_A A_i^*$), setting $M^* = 0$ reduces the model prediction to:

$$ \mu_i = \alpha + \beta_A A_i^* $$

This isolates the **controlled direct effect** of median age at marriage $A$, holding marriage frequency $M$ constant.

> [!IMPORTANT]
> **Crucial Pitfall Clarification: Why $M^* = 0$ does NOT mean 0 marriages!**  
> 
> It is a very common point of confusion to think that $M = 0$ means a population where nobody gets married ($0$ marriages), which would intuitively imply zero divorces ($D = 0$).
> 
> In this model:
> 1. **$M_{\mathrm{std}} = 0$ is a Standardized Z-Score (Mean = 0):** $M^*$ is defined as $M^* = \frac{M - \overline{M}}{s_M}$. Therefore, $M^* = 0$ represents the **sample average (mean) marriage rate** ($\approx 20.1$ marriages per 1,000 adults per year), NOT zero marriages!
> 2. **What zero marriages would actually be on the Z-scale:** A state with literally zero marriages ($M = 0$) would have a standardized Z-score of $M^* = \frac{0 - 20.1}{3.8} \approx -5.3$ standard deviations.
> 3. **Annual Rates vs Existing Stock:** Furthermore, $M$ measures *new marriages in the current year*, whereas divorces $D$ occur among people who got married in *previous years*.
> 
> Therefore, setting `M_std = 0` in `link(m5.3, data = data.frame(A_std = a_seq, M_std = 0))` means: **"Holding Marriage Rate fixed at the average US state marriage rate (mean frequency of marriage)."**


#### 3. Why does the curve slope downwards from $+1.2$ at $A = -2$ down to $-1.2$ at $A = +2$?

The estimated posterior coefficient for age at marriage is strongly negative ($\beta_A \approx -0.6$). 

- **At $A^* = -2$ (young median age at marriage, e.g. ~22 years old):** The model predicts $\mu \approx 0 + (-0.6)(-2) = +1.2$. States where people marry very young are predicted to have high divorce rates (+1.2 SDs above average).
- **At $A^* = 0$ (average age at marriage, e.g. ~24.5 years old):** The model predicts $\mu \approx 0$.
- **At $A^* = +2$ (older median age at marriage, e.g. ~27 years old):** The model predicts $\mu \approx 0 + (-0.6)(+2) = -1.2$. States where people marry older are predicted to have very low divorce rates (-1.2 SDs below average).

The total decrease across the $4$ standard-deviation swing in age is $2.4$ standard deviations ($\approx 4.3$ fewer divorces per 1,000 adults).

#### 4. The Shaded Region (89% HPDI)

The purple shaded band represents the **89% Highest Posterior Density Interval (HPDI)** calculated draw by draw across joint posterior samples of $(\alpha, \beta_A, \beta_M)$, capturing joint parameter uncertainty.

#### 5. Crucial Estiplan Caveats

A counterfactual plot can be computed perfectly in software by substituting values into `link()`, but **it remains scientifically wrong if the underlying DAG or identification strategy is invalid**:

- The plot reflects a true causal intervention $\mathbb{E}[D \mid \mathrm{do}(A=a), \mathrm{do}(M=0)]$ **only if** the candidate DAG $A \rightarrow M \rightarrow D$ and $A \rightarrow D$ is correctly specified and no unobserved confounders (e.g. regional economy or religious demographics $S$) exist.
- Software substitution is a mathematical operation; causal validity is an identification property of the estiplan.

```{r}
#| label: fig-three-plot-types
#| fig-cap: "Predictor residual plot (left), Posterior prediction plot (center), and Counterfactual plot (right)."
#| echo: true
par(mfrow = c(1, 3))

# 1. Predictor residual plot
m5.4 <- quap(alist(M_std ~ dnorm(mu, sigma), mu <- a + bA * A_std, a ~ dnorm(0, 0.2), bA ~ dnorm(0, 0.5), sigma ~ dexp(1)), data = d, start = list(a=0, bA=0, sigma=1))
mu_m <- link(m5.4)
m_resid <- d$M_std - colMeans(mu_m)
plot(d$D_std ~ m_resid, xlab = "Marriage Rate Residuals (given Age)", ylab = "Divorce Rate (std)", main = "1. Predictor Residual Plot", pch = 19, col = "steelblue")
abline(lm(d$D_std ~ m_resid), lty = 2)

# 2. Posterior prediction plot
mu_post <- link(m5.3)
mu_post_mean <- colMeans(mu_post)
plot(d$D_std ~ mu_post_mean, xlab = "Posterior Predicted Divorce", ylab = "Observed Divorce", main = "2. Posterior Prediction Plot", pch = 19, col = "forestgreen")
abline(a = 0, b = 1, lty = 2)

# 3. Counterfactual plot: Manipulating A holding M=0
a_seq <- seq(-2, 2, length.out = 30)
pred_cf <- link(m5.3, data = data.frame(A_std = a_seq, M_std = 0))
plot(a_seq, colMeans(pred_cf), type = "l", lwd = 2, col = "purple", xlab = "Manipulated Age (A)", ylab = "Predicted Divorce (D)", main = "3. Counterfactual Plot (M=0)", ylim = c(-2, 2))
shade(apply(pred_cf, 2, HPDI, prob = 0.89), a_seq)
par(mfrow = c(1, 1))
```

---

## 18. Spurious association

Under:

$$ M \leftarrow A \rightarrow D $$

$M$ and $D$ can be associated even when:

$$ M \not\rightarrow D $$

The marginal regression:

$$ D \sim M $$

may therefore produce a nonzero slope when the causal effect of $M$ is zero.

The multivariable regression:

$$ D \sim M+A $$

computes a conditional association under the chosen model.

Multiple regression does not discover the correct graph automatically. It performs the comparison requested by the analyst.

---

## 19. Masked relationships

Section 5.2 presents the converse pattern: an association can be weak marginally but stronger after conditioning on another predictor.

The book's milk-energy example uses:

- $K$: milk energy;
- $N$: neocortex percentage;
- $B$: body mass.

Body mass and neocortex are associated with one another and have relationships with milk energy in opposite directions. Marginal patterns can partially cancel.

A multiple regression asks:

- among species with similar body mass, how does milk energy vary with neocortex percentage?
- among species with similar neocortex percentage, how does milk energy vary with body mass?

This is primarily a **book counterpart**. The official A05-related script is more clearly centered on the divorce example.

The lesson is that conditioning can remove spurious association or reveal masked association. Causal interpretation still depends on the graph.

---

## 20. One estimand, several estiplans

A fundamental architectural insight of Lecture A05 is that **the estimand is an invariant scientific target, while an estiplan is a specific, auditable estimation procedure**.

Suppose our scientific target is the Marginal Average Treatment Effect (ATE):

$$ \tau = \mathbb{E}[Y(1)-Y(0)] $$

Because $\tau$ defines *what* we want to learn rather than *how* we compute it, multiple completely different estiplans can target the exact same estimand. Each estiplan relies on different model choices, mathematical mechanics, identification assumptions, and diagnostic failure modes.

---

### 20.1 Outcome-Regression Estiplan (G-Computation / Standardization)

#### Mechanism
1. **Model the Outcome Surface:** Fit a Bayesian or parametric regression model $\mathbb{E}[Y \mid A, X, \boldsymbol{\theta}]$ for outcome $Y$ conditional on treatment $A$ and confounders $X$.
2. **Generate Counterfactual Predictions:** For every unit $i = 1, \ldots, N$ in the target population, generate predicted outcomes under intervention $A_i = 1$ and under $A_i = 0$, holding their observed $X_i$ fixed.
3. **Standardize and Contrast:** Average individual counterfactual predictions over the population distribution $p(X)$, and subtract the intervention means draw by draw:

$$ \tau_{\mathrm{G-comp}}^{(s)} = \frac{1}{N} \sum_{i=1}^N \left( \mu^{(s)}(1, X_i) - \mu^{(s)}(0, X_i) \right) $$

#### Assumptions & Vulnerabilities
- **Primary Assumption:** Correct functional specification of the outcome surface $\mathbb{E}[Y \mid A, X]$.
- **Failure Mode:** Highly sensitive to outcome model mis-specification (e.g., omitting non-linearities or interactions). In regions with poor covariate overlap, predictions rely heavily on unverified parametric extrapolation.

---

### 20.2 Weighting Estiplan (Inverse Probability Weighting - IPW)

#### Mechanism
1. **Model Treatment Assignment:** Fit a propensity score model $e(X) = \Pr(A=1 \mid X)$ predicting treatment assignment from confounders $X$.
2. **Construct Inverse-Probability Weights:** Assign weights $w_i = \frac{A_i}{e(X_i)} + \frac{1-A_i}{1-e(X_i)}$ to create a pseudo-population where treatment $A$ is independent of $X$.
3. **Compute Weighted Contrast:** Calculate the difference between weighted treatment group means:

$$ \widehat{\tau}_{\mathrm{IPW}} = \frac{\sum_{i=1}^N A_i Y_i w_i}{\sum_{i=1}^N A_i w_i} - \frac{\sum_{i=1}^N (1-A_i) Y_i w_i}{\sum_{i=1}^N (1-A_i) w_i} $$

#### Assumptions & Vulnerabilities
- **Primary Assumption:** Correct specification of the propensity model $e(X)$ and strong positivity ($0 < \epsilon \le e(X) \le 1 - \epsilon$).
- **Failure Mode:** Extreme propensity scores (near 0 or 1) produce massive weights, destroying estimator precision and causing high variance.

---

### 20.3 Doubly Robust Estiplan (Augmented IPW - AIPW)

#### Mechanism
Combines the outcome regression model $\mu(a, x)$ and the propensity model $e(x)$ into an augmented efficient score:

$$ \widehat{\tau}_{\mathrm{AIPW}} = \frac{1}{N} \sum_{i=1}^N \left[ \mu(1, X_i) - \mu(0, X_i) + \frac{A_i \left(Y_i - \mu(1, X_i)\right)}{e(X_i)} - \frac{(1-A_i)\left(Y_i - \mu(0, X_i)\right)}{1-e(X_i)} \right] $$

#### Assumptions & Vulnerabilities
- **Double Robustness Property:** Unbiased and consistent if **either** the outcome model $\mu(a, x)$ OR the propensity model $e(x)$ is correctly specified.
- **Failure Mode:** Can still exhibit severe bias if *both* models are mis-specified, or in extreme overlap violation regimes where $e(X) \approx 0$ or $1$.

---

### 20.4 Randomized-Experiment Estiplan

#### Mechanism
1. **Physical Intervention Assignment:** Randomize treatment $A$ so that $Y(1), Y(0) \perp\!\!\!\perp A$ unconditionally.
2. **Simple Arm Contrast:** Calculate the raw unadjusted difference in sample arm means:

$$ \widehat{\tau}_{\mathrm{RCT}} = \overline{Y}_{A=1} - \overline{Y}_{A=0} $$

#### Assumptions & Vulnerabilities
- **Primary Assumption:** Physical randomization integrity, zero differential attrition, and full treatment adherence.
- **Failure Mode:** Subject to non-compliance, measurement error, or sample selection bias if the RCT sample does not match the intended deployment population.

---

### Comparison of Estiplans Targeting the Same ATE ($\tau$)

| Estiplan | Key Component Model | Primary Identification Requirement | Primary Failure Mode |
|---|---|---|---|
| **1. Outcome Regression (G-Comp)** | Outcome model $\mu(A, X)$ | Correct outcome response specification | Model mis-specification & Extrapolation |
| **2. Propensity Weighting (IPW)** | Treatment assignment $e(X)$ | Correct propensity specification & Overlap | Extreme weights & High variance |
| **3. Doubly Robust (AIPW)** | Outcome $\mu(A, X)$ + Propensity $e(X)$ | Correctness of *at least one* model | Dual mis-specification & Severe positivity violation |
| **4. Randomized Experiment** | Physical randomization mechanism | Randomization integrity & Adherence | Non-compliance & Sample attrition |

---

#### Multi-Language Implementation: Comparing G-Computation vs IPW vs AIPW Estiplans

```r
# R Code: Executing 3 Different Estiplans for the Same ATE Estimand
# Data setup: A (treatment), X (confounder), Y (outcome)
set.seed(42)
N <- 500
X <- rnorm(N)
A <- rbinom(N, 1, prob = plogis(0.5 * X))
Y <- 1.5 * A + 0.8 * X + rnorm(N)

# --- Estiplan 1: Outcome Regression (G-Computation) ---
fit_out <- lm(Y ~ A + X)
pred_1 <- predict(fit_out, newdata = data.frame(A = 1, X = X))
pred_0 <- predict(fit_out, newdata = data.frame(A = 0, X = X))
ate_gcomp <- mean(pred_1 - pred_0)

# --- Estiplan 2: Propensity Score Weighting (IPW) ---
fit_ps <- glm(A ~ X, family = binomial)
ps <- predict(fit_ps, type = "response")
w1 <- A / ps
w0 <- (1 - A) / (1 - ps)
ate_ipw <- sum(Y * w1) / sum(w1) - sum(Y * w0) / sum(w0)

# --- Estiplan 3: Doubly Robust (AIPW) ---
mu1 <- predict(fit_out, newdata = data.frame(A = 1, X = X))
mu0 <- predict(fit_out, newdata = data.frame(A = 0, X = X))
ate_aipw <- mean((mu1 - mu0) + A * (Y - mu1) / ps - (1 - A) * (Y - mu0) / (1 - ps))

cat(sprintf("ATE Estimand Target = 1.50\nEstiplan 1 (G-Comp): %.3f\nEstiplan 2 (IPW):    %.3f\nEstiplan 3 (AIPW):   %.3f\n", 
            ate_gcomp, ate_ipw, ate_aipw))
```

```python
# PyMC / Python Code: Comparing Estiplans for the Same ATE Estimand
import numpy as np
import statsmodels.api as sm

# 1. G-Computation Estiplan
# Using fitted posterior outcome parameters
mu1_py = alpha_s[:, None] + beta_A_s[:, None] * 1.0 + beta_X_s[:, None] * X[None, :]
mu0_py = alpha_s[:, None] + beta_A_s[:, None] * 0.0 + beta_X_s[:, None] * X[None, :]
ate_gcomp_py = np.mean(np.mean(mu1_py - mu0_py, axis=1))

# 2. IPW Estiplan
ps_model = sm.Logit(A_obs, sm.add_constant(X_obs)).fit(disp=False)
ps_py = ps_model.predict()
w1_py = A_obs / ps_py
w0_py = (1 - A_obs) / (1 - ps_py)
ate_ipw_py = np.sum(Y_obs * w1_py) / np.sum(w1_py) - np.sum(Y_obs * w0_py) / np.sum(
    w0_py
)

# 3. AIPW Estiplan
mu1_hat = np.mean(mu1_py, axis=0)
mu0_hat = np.mean(mu0_py, axis=0)
ate_aipw_py = np.mean(
    (mu1_hat - mu0_hat)
    + A_obs * (Y_obs - mu1_hat) / ps_py
    - (1 - A_obs) * (Y_obs - mu0_hat) / (1 - ps_py)
)

print(
    f"PyMC Estiplan 1 (G-Comp): {ate_gcomp_py:.3f} | Estiplan 2 (IPW):"
    f" {ate_ipw_py:.3f} | Estiplan 3 (AIPW): {ate_aipw_py:.3f}"
)
```

### 20.5 Intuitive Conceptual Summary of the Four Estiplans

To synthesize how these four estiplans differ in intuition, mechanism, and modeling strategy:

1. **Section 20.1 — Outcome Regression (G-Computation): *"Generative / Simulation Mode"***  
   You act like a generative engine. Inside your statistical model $\mu(A, X)$, you simulate potential outcomes for *every single observation* under both $A_i=1$ and $A_i=0$, creating a complete counterfactual table across the target population before averaging the contrast.

2. **Section 20.2 — Propensity Weighting (IPW): *"Weighted Observed Reality"***  
   You do not model the outcome process $Y$ or simulate counterfactuals. You accept the observed outcomes $Y_i$ as naturally recorded, but re-weight each unit by the inverse propensity score $1/e(X_i)$ to build a balanced, unconfounded pseudo-population.

3. **Section 20.3 — Doubly Robust (AIPW): *"Dual Protection & Variance Reduction"***  
   Combines G-computation and IPW. It provides **Double Robustness** (the ATE estimator is unbiased if *either* the outcome model OR the propensity model is correctly specified) and **Variance Reduction** (weighting is applied only to the residual prediction errors $Y_i - \mu(A_i, X_i)$, eliminating the high variance caused by extreme propensity weights).

4. **Section 20.4 — Randomized Experiment (RCT / A/B Test): *"Physical Random Assignment"***  
   Physical randomization forces $A \perp\!\!\!\perp Y(1), Y(0)$ unconditionally by design. No backdoor adjustment or propensity modeling is required; the raw difference in sample arm means directly identifies the estimand.



---

## 21. One model, several estimands

Given a fitted outcome model:

$$ p(Y \mid A,X,\mathcal{D}_{\mathrm{obs}}) $$

one can derive several targets.

### Conditional treatment effect

$$ \tau(x) = \mathbb{E}[Y \mid A=1,X=x] - \mathbb{E}[Y \mid A=0,X=x] $$

### Sample average treatment effect

$$ \tau_{\mathrm{SATE}} = \frac{1}{N} \sum_{i=1}^{N} \tau(X_i) $$

### Population average treatment effect

$$ \tau_{\mathrm{PATE}} = \int \tau(x)p_{\mathrm{population}}(x) \,dx $$

### Average treatment effect on the treated

$$ \tau_{\mathrm{ATT}} = \mathbb{E}[\tau(X) \mid A=1] $$

### Policy value

For policy $\pi(X)$:

$$ V(\pi) = \mathbb{E} \left[ Y\left(\pi(X)\right) \right] $$

The fitted model does not decide which output is scientifically relevant. The estimand does.

---

## 22. Positivity and support

A standardization estiplan requires relevant support:

$$ 0 < \Pr(A=a \mid X=x) < 1 $$

for covariate profiles in the target population.

If no states with high age at marriage also have high marriage rates, some intervention predictions may depend mainly on extrapolation.

A regression equation will still produce numbers outside the observed support.

That does not mean the estimand is reliably learned.

An estiplan should therefore include:

- treatment-support plots;
- covariate-overlap diagnostics;
- extrapolation flags;
- sensitivity to trimming;
- possible restriction of the target population.

---

## 23. Estiplan validation by simulation

A credible estiplan should be tested on synthetic data where the final estimand is known.

For simulation $r$:

1. specify a causal generative model;
2. generate intervention outcomes;
3. generate observed assignment and data;
4. execute the complete estiplan;
5. compare the estimate with the known estimand.

Let the true and estimated targets be:

$$ \tau^{(r)} $$

and:

$$ \widehat{\tau}^{(r)} $$

Then examine bias:

$$ \mathrm{Bias} = \frac{1}{R} \sum_{r=1}^{R} \left( \widehat{\tau}^{(r)}-\tau^{(r)} \right) $$

and interval coverage:

$$ \widehat{C} = \frac{1}{R} \sum_{r=1}^{R} \mathbb{I} \left[ L_r \leq \tau^{(r)} \leq U_r \right] $$

Validation should test the full pipeline, including prediction, standardization, and contrasts, not only recovery of regression parameters.

```{r}
#| label: fig-sbc-estiplan-recovery
#| fig-cap: "Estiplan recovery experiment across 100 simulation repetitions for G-computation ATE estimation."
#| echo: true
set.seed(42)
R <- 100
N <- 100
true_ate <- -0.5

sim_results <- replicate(R, {
  A <- rnorm(N, 0, 1)
  M <- 0.4 * A + rnorm(N, 0, 1)
  D <- true_ate * A + 0.0 * M + rnorm(N, 0, 1)
  
  dat_s <- list(D = D, A = A, M = M)
  fit_s <- quap(alist(D ~ dnorm(mu, sigma), mu <- a + bA * A + bM * M, a ~ dnorm(0, 0.2), bA ~ dnorm(0, 0.5), bM ~ dnorm(0, 0.5), sigma ~ dexp(1)), data = dat_s, start = list(a=0, bA=0, bM=0, sigma=1))
  post_s <- extract.samples(fit_s)
  hpdi_val <- as.numeric(HPDI(post_s$bA, prob = 0.89))
  covered <- as.numeric((true_ate >= hpdi_val[1]) && (true_ate <= hpdi_val[2]))
  c(est = mean(post_s$bA), lower = hpdi_val[1], upper = hpdi_val[2], covered = covered)
})

cov_rate <- mean(sim_results["covered", ])
cat(sprintf("SBC Estiplan Recovery - Mean ATE: %.4f | 89%% Coverage: %.1f%%\n", mean(sim_results["est", ]), cov_rate * 100))

plot(1:R, sim_results["est", ], ylim = range(sim_results[c("lower", "upper"), ]), pch = 19, cex = 0.6, xlab = "Simulation (r)", ylab = "Estimated ATE", main = sprintf("Estiplan Recovery (Coverage = %.1f%%)", cov_rate * 100), col = ifelse(sim_results["covered", ] == 1, "steelblue", "firebrick"))
for (r in 1:R) lines(c(r, r), c(sim_results["lower", r], sim_results["upper", r]), col = ifelse(sim_results["covered", r] == 1, rgb(0.2, 0.4, 0.8, 0.5), rgb(0.8, 0.2, 0.2, 0.8)))
abline(h = true_ate, lty = 2, lwd = 2)
```

#### PyMC Implementation: Estiplan Simulation Recovery Validation

```python
# PyMC Implementation: Simulation-Based Estiplan Recovery Validation
import numpy as np
import pymc as pm


def run_pymc_estiplan_validation(R=50, N=100, true_ate=-0.5):
  covered_count = 0
  estimates = []

  for r in range(R):
    # Generative model simulation under DAG A -> M -> D and A -> D
    A_sim = np.random.normal(0, 1, size=N)
    M_sim = 0.4 * A_sim + np.random.normal(0, 1, size=N)
    D_sim = true_ate * A_sim + 0.0 * M_sim + np.random.normal(0, 1, size=N)

    with pm.Model() as sim_model:
      a = pm.Normal("a", mu=0, sigma=0.2)
      bA = pm.Normal("bA", mu=0, sigma=0.5)
      bM = pm.Normal("bM", mu=0, sigma=0.5)
      sig = pm.Exponential("sig", lam=1.0)

      mu = a + bA * A_sim + bM * M_sim
      pm.Normal("D_obs", mu=mu, sigma=sig, observed=D_sim)

      idata_sim = pm.sample(draws=500, tune=500, chains=2, progressbar=False)

    bA_draws = idata_sim.posterior["bA"].values.flatten()
    hdi = np.percentile(bA_draws, [5.5, 94.5])

    est_mean = np.mean(bA_draws)
    estimates.append(est_mean)
    if hdi[0] <= true_ate <= hdi[1]:
      covered_count += 1

  print(
      f"PyMC Estiplan Recovery - Mean ATE: {np.mean(estimates):.3f} | 89%"
      f" Coverage: {covered_count / R * 100:.1f}%"
  )
```

---

## 24. Common errors exposed by A05

### Error 1: Starting with the model

“Fit a multiple regression” is not a scientific question.

### Error 2: Naming a coefficient as the estimand

The target may require prediction, averaging, and contrast operations.

### Error 3: Defining an exposure without a comparison

An effect requires intervention states such as:

$$ a_1 \quad\text{versus}\quad a_0 $$

### Error 4: Omitting the target population

Sample, eligible, treated, and deployment populations can have different average effects.

### Error 5: Conditioning on a mediator for a total-effect target

The adjustment changes the estimand.

### Error 6: Omitting a common cause

A marginal association can be spurious.

### Error 7: Assuming multiple regression discovers causality

Regression computes the conditional comparison chosen by the analyst.

### Error 8: Confusing conditional prediction with intervention prediction

Setting a predictor value in software does not create causal identification.

### Error 9: Ignoring overlap

The model can extrapolate into unsupported intervention regions.

### Error 10: Reporting only parameter summaries

The scientific result often lives on the outcome scale as a posterior contrast.

### Error 11: Validating only predictive fit

Good retrodiction does not establish estimand recovery.

### Error 12: Losing uncertainty during post-processing

Predictions, averages, and contrasts should be computed draw by draw.

---

## 25. CRM application: discount uplift

Suppose the question is:

> What is the average 30-day incremental margin from assigning a €10 discount rather than no discount to eligible customers?

Let:

- $A=1$ denote the €10 offer;
- $A=0$ denote no offer;
- $Y$ denote 30-day net margin;
- $E=1$ denote eligibility.

### 25.1 Estimand

$$ \tau = \mathbb{E} \left[ Y(1)-Y(0) \mid E=1 \right] $$

The eligibility restriction is part of the target.

### 25.2 Identification assumption

Let $X$ contain observed common causes of offer assignment and margin, such as tenure, prior orders, product affinity, and previous campaign response.

The assumption is:

$$ Y(a) \perp\!\!\!\perp A \mid X,E=1 $$

### 25.3 Outcome-regression estiplan

1. define the eligible population;
2. fit $p(Y \mid A,X,E=1)$;
3. draw posterior parameters;
4. predict every eligible customer under $A=1$;
5. predict every eligible customer under $A=0$;
6. subtract customer-level predictions;
7. average the contrasts;
8. summarize posterior uncertainty;
9. diagnose overlap and extrapolation;
10. validate on simulated CRM data.

For posterior draw $s$:

$$ \tau^{(s)} = \frac{1}{N_E} \sum_{i:E_i=1} \left[ \mu_i^{(s)}(1) - \mu_i^{(s)}(0) \right] $$

### 25.4 Why the treatment coefficient may not be the ATE

With effect heterogeneity:

$$ \mu_i = \alpha + \beta_A A_i + \beta_X X_i + \beta_{AX}A_iX_i $$

the conditional effect is:

$$ \tau_i = \beta_A+\beta_{AX}X_i $$

The population estimand requires averaging over eligible customers.

```python
# PyMC Code: CRM Uplift G-Computation Estiplan
import numpy as np
import pymc as pm

# 1. Fit outcome regression model with covariates (tenure, prior orders)
with pm.Model() as crm_model:
  alpha = pm.Normal("alpha", mu=0, sigma=10)
  beta_A = pm.Normal("beta_A", mu=0, sigma=5)
  beta_X = pm.Normal("beta_X", mu=0, sigma=2, shape=2)
  sigma = pm.Exponential("sigma", lam=1)

  mu = (
      alpha
      + beta_A * offer_obs
      + pm.math.dot(X_obs, beta_X)
  )
  pm.Normal("Y_obs", mu=mu, sigma=sigma, observed=Y_obs)
  idata = pm.sample(draws=500, tune=500, chains=2, progressbar=False)

# 2. G-Computation Counterfactual Predictions & Standardization
# Predict for all eligible customers under offer_10 (A=1) and no_offer (A=0)
posterior_samples = idata.posterior
# Draw-by-draw customer-level contrast averaging produces ATE posterior distribution
```

---

## 26. CRM application: several offer amounts

Suppose:

$$ A \in \{0,5,10,20\} $$

A decision-relevant estimand is policy value:

$$ V(a) = \mathbb{E} \left[ Y(a)-C(a) \right] $$

where $C(a)$ includes discount and contact costs.

The incremental value is:

$$ \Delta_V(a_1,a_0) = V(a_1)-V(a_0) $$

An estiplan can:

1. model outcomes flexibly by offer and customer features;
2. predict every customer's outcome under each offer;
3. subtract offer costs;
4. average value over the target population;
5. compute posterior regret for each action;
6. choose the offer or policy with lowest expected regret.

The estimand is not “the offer coefficient.” It is the decision-relevant contrast.

```{r}
#| label: fig-crm-uplift
#| fig-cap: "Expected Revenue, Offer Cost, and Net Profit across 4 discount levels (€0, €5, €10, €20)."
#| echo: true
alpha_means <- c(30.0, 42.0, 48.0, 50.0)
costs <- c(0, 5, 10, 20)
margin <- 1.0

exp_revenue <- alpha_means * margin
exp_profit <- exp_revenue - costs
opt_action <- which.max(exp_profit)

barplot(rbind(exp_revenue, costs, exp_profit), beside = TRUE,
        names.arg = c("€0 Offer", "€5 Offer", "€10 Offer", "€20 Offer"),
        col = c("steelblue", "firebrick", "forestgreen"),
        ylab = "Euros (€)", main = "CRM Offer Net Profit Optimization")
legend("topleft", legend = c("Revenue", "Offer Cost", "Net Profit"), fill = c("steelblue", "firebrick", "forestgreen"), bty = "n")
```

#### PyMC Implementation: Multi-Offer Net Profit Policy Optimization

```python
# PyMC Code: Multi-Offer Policy Value & Net Profit Optimization
import numpy as np
import pymc as pm

offers = np.array([0, 5, 10, 20])
costs = np.array([0, 5, 10, 20])

with pm.Model() as multi_offer_model:
  alpha = pm.Normal("alpha", mu=30, sigma=10)
  beta_offer = pm.Normal("beta_offer", mu=1.5, sigma=0.5)
  beta_sat = pm.Normal("beta_sat", mu=-0.03, sigma=0.01)
  sigma = pm.Exponential("sigma", lam=0.1)

  mu_offers = alpha + beta_offer * offers + beta_sat * (offers**2)
  pm.Normal("Y_obs", mu=mu_offers[offer_idx], sigma=sigma, observed=Y_crm)

  idata_crm = pm.sample(draws=1000, tune=1000, chains=2, progressbar=False)

# Posterior expected net profit calculation across candidate offers
post_alpha = idata_crm.posterior["alpha"].values.flatten()
post_bo = idata_crm.posterior["beta_offer"].values.flatten()
post_bs = idata_crm.posterior["beta_sat"].values.flatten()

expected_profits = []
for a_idx, offer_val in enumerate(offers):
  rev_draws = post_alpha + post_bo * offer_val + post_bs * (offer_val**2)
  net_profit_draws = rev_draws - costs[a_idx]
  expected_profits.append(np.mean(net_profit_draws))

best_offer = offers[np.argmax(expected_profits)]
print(
    f"PyMC Optimal Offer: €{best_offer} with Expected Net Profit:"
    f" €{max(expected_profits):.2f}"
)
```

---

## 27. Implications for CausalOps

A05 strongly motivates separate workflow artifacts.

| A05 concept | CausalOps artifact |
|---|---|
| Scientific question | `StudyArtifact` |
| Target quantity | `EstimandArtifact` |
| Intervention definition | `InterventionArtifact` |
| Target population | `PopulationArtifact` |
| Causal graph | `DAGArtifact` |
| Identification assumptions | `IdentificationArtifact` |
| Complete operation sequence | `EstiplanArtifact` |
| Statistical models | `ModelArtifact` |
| Averaging distribution | `TargetDistributionArtifact` |
| Posterior result | `EstimateArtifact` |
| Recovery and coverage | `ValidationArtifact` |
| Overlap and extrapolation | `SupportDiagnosticArtifact` |
| Unverified assumptions | `SensitivityArtifact` |

### 27.1 Example estimand schema

```yaml
estimand_id: ate_margin_offer10_vs_none
outcome: net_margin_30d
intervention:
  left: offer_10
  right: no_offer
contrast: difference
population:
  filter: eligible == true
aggregation: customer_average
scale: euros_per_eligible_customer
```

### 27.2 Example estiplan schema

```yaml
estiplan_id: gcomp_margin_v1
estimand_id: ate_margin_offer10_vs_none
identification:
  strategy: backdoor_adjustment
  adjustment_set:
    - tenure
    - prior_orders_90d
    - prior_campaign_response
    - product_affinity
models:
  outcome_model: bayesian_outcome_regression_v3
operations:
  - fit_outcome_model
  - predict_all_under_offer_10
  - predict_all_under_no_offer
  - contrast_customer_predictions
  - average_over_eligible_population
  - summarize_posterior
validation:
  - prior_predictive_check
  - posterior_predictive_check
  - overlap_check
  - simulation_recovery
```

### 27.3 Fail-closed logic

CausalOps should refuse or downgrade a causal estimate when:

- the intervention is undefined;
- the comparison state is missing;
- the target population is unspecified;
- no identification argument is recorded;
- required adjustment variables are unavailable;
- positivity is severely violated;
- post-treatment variables are included for a total-effect target;
- the declared estiplan fails simulation recovery.

This is the operational meaning of making the workflow auditable.

---

## 28. Lecture takeaways

1. An estimand is the scientific quantity to be learned.
2. An estiplan is the complete plan connecting assumptions and data to an estimate.
3. A statistical model is only one component of an estiplan.
4. Identification must be established before estimation.
5. Regression coefficients are not automatically causal effects or final estimands.
6. Prediction, averaging, and contrast operations often construct the actual target.
7. The target population is part of the estimand.
8. The Waffle House pattern illustrates how association can be detached from causal meaning.
9. Multiple regression computes conditional relationships; it does not discover the causal graph.
10. A total-effect estiplan should not condition on a mediator.
11. A confounded exposure may require adjustment and standardization.
12. Conditional prediction and intervention prediction are different operations.
13. Predictor residual plots explain statistical conditioning, not causality.
14. Posterior prediction plots assess fit and prediction, not identification.
15. Counterfactual plots express causal implications only under the assumed graph.
16. One estimand can be targeted by several estiplans.
17. One fitted model can support several estimands through different averaging distributions and contrasts.
18. Support and positivity are properties of the estiplan, not optional diagnostics.
19. Simulation should test recovery of the final estimand, not only model parameters.
20. An auditable causal system must store estimand and estiplan separately.

---

## 29. Review questions (Detailed Answers & Multi-Language Implementation) {#29-review-questions}

### Summary List of Review Questions

1. [**Question 1:** What is the difference between an estimand and an estimate?](#q1)
2. [**Question 2:** How does an estiplan differ from an estimator?](#q2)
3. [**Question 3:** Why is a regression model only one component of an estiplan?](#q3)
4. [**Question 4:** Under what conditions can a slope coefficient equal an average causal effect?](#q4)
5. [**Question 5:** Which assumptions identify the g-formula?](#q5)
6. [**Question 6:** Why does the Waffle House association not identify an intervention effect?](#q6)
7. [**Question 7:** What does standardization accomplish, and what does it not accomplish?](#q7)
8. [**Question 8:** What do the three divorce regressions estimate statistically?](#q8)
9. [**Question 9:** Under $A \rightarrow M \rightarrow D$ and $A \rightarrow D$, why should a total-effect model for $A$ avoid conditioning on $M$?](#q9)
10. [**Question 10:** Why does estimating the effect of $M$ require adjustment for $A$ under that graph?](#q10)
11. [**Question 11:** How is posterior standardization performed draw by draw?](#q11)
12. [**Question 12:** Why is the target covariate distribution part of the estimand?](#q12)
13. [**Question 13:** How does a counterfactual plot differ from an ordinary prediction plot?](#q13)
14. [**Question 14:** What is a predictor residual plot useful for?](#q14)
15. [**Question 15:** Why does multiple regression not determine the causal graph?](#q15)
16. [**Question 16:** How can conditioning reveal a masked relationship?](#q16)
17. [**Question 17:** Name three estiplans that can target the same ATE.](#q17)
18. [**Question 18:** How can one outcome model support SATE, PATE, ATT, and policy value?](#q18)
19. [**Question 19:** Why is overlap required for standardization?](#q19)
20. [**Question 20:** What should simulation validation evaluate beyond parameter recovery?](#q20)
21. [**Question 21:** In a CRM offer study, why is the treatment coefficient not necessarily incremental profit?](#q21)
22. [**Question 22:** Which fields belong in an `EstimandArtifact`?](#q22)
23. [**Question 23:** Which operations belong in an `EstiplanArtifact`?](#q23)
24. [**Question 24:** When should CausalOps fail closed rather than report a causal estimate?](#q24)

---

### Question 1: What is the difference between an estimand and an estimate? {#q1}
**Answer:**  
An **estimand** is the precise, theoretical scientific quantity that an investigation seeks to learn (e.g. $\tau = \mathbb{E}[Y(1) - Y(0) \mid E=1]$). It is defined independently of any sample data or statistical algorithm. An **estimate** is the concrete numerical value produced by applying an estimator/estiplan to observed data ($\widehat{\tau}_{\mathrm{obs}} = -0.42$).

---

### Question 2: How does an estiplan differ from an estimator? {#q2}
**Answer:**  
An **estimator** is a specific mathematical function mapping data to a value ($\widehat{\tau} = g(\mathcal{D})$). An **estiplan** is the broader, auditable sequence of operations connecting causal assumptions to the final estimate: defining interventions, identifying backdoor sets, specifying prior/likelihood models, generating draw-by-draw counterfactual predictions, standardizing over target populations, computing contrasts, and conducting simulation-based calibration.

---

### Question 3: Why is a regression model only one component of an estiplan? {#q3}
**Answer:**  
A regression model merely estimates conditional distributions $p(Y \mid A, X)$. Converting those conditional parameters into a scientific answer requires identification assumptions (DAG validation), variable selection, counterfactual intervention setting ($A=a$), population standardization over $p(X)$, and contrast calculation. The regression parameters are intermediate building blocks.

---

### Question 4: Under what conditions can a slope coefficient equal an average causal effect? {#q4}
**Answer:**  
A slope coefficient $\beta_A$ in $\mu = \alpha + \beta_A A + \beta_X X$ equals the ATE if and only if:
1. The true outcome relationship is strictly linear in $A$.
2. There are no treatment-covariate interaction terms ($\beta_{AX} = 0$).
3. The adjustment set $X$ fully blocks all backdoor confounding paths between $A$ and $Y$.
4. No post-treatment mediators or colliders are included in $X$.
5. Positivity holds across the covariate space.

---

### Question 5: Which assumptions identify the g-formula? {#q5}
**Answer:**  
The g-formula $\mathbb{E}[Y(a)] = \int \mathbb{E}[Y \mid A=a, X=x] p(x) dx$ is identified under three core assumptions:
1. **Conditional Exchangeability (Unconfoundedness)**: $Y(a) \perp\!\!\!\perp A \mid X$.
2. **Positivity (Overlap)**: $0 < \Pr(A=a \mid X=x) < 1$ for all $x$ where $p(x) > 0$.
3. **Consistency**: If $A_i = a$, then $Y_i(a) = Y_i$.

#### Multi-Language Implementation (Question 5: G-Formula / Standardization)

```r
# R Code: G-Formula / Standardization Implementation
post <- extract.samples(m5.3)
pred_a1 <- sapply(1:nrow(post$a), function(s) mean(post$a[s] + post$bA[s] * 1 + post$bM[s] * d$M_std))
pred_a0 <- sapply(1:nrow(post$a), function(s) mean(post$a[s] + post$bA[s] * 0 + post$bM[s] * d$M_std))
ate_draws <- pred_a1 - pred_a0
precis(ate_draws)
```

```python
# PyMC Code: G-Formula Standardization
import numpy as np
import pymc as pm

post = idata_5_3.posterior
alpha_s = post["alpha"].values.flatten()
bA_s = post["beta_A"].values.flatten()
bM_s = post["beta_M"].values.flatten()

# Draw-by-draw standardization
pred_a1 = alpha_s[:, None] + bA_s[:, None] * 1.0 + bM_s[:, None] * M_std[None, :]
pred_a0 = alpha_s[:, None] + bA_s[:, None] * 0.0 + bM_s[:, None] * M_std[None, :]
ate_draws = np.mean(pred_a1 - pred_a0, axis=1)
print(f"PyMC G-Formula ATE: {np.mean(ate_draws):.3f}")
```

```python
# PyTorch Code
import torch

pred_a1_t = alpha_t + bA_t * 1.0 + bM_t * M_t.unsqueeze(0)
pred_a0_t = alpha_t + bA_t * 0.0 + bM_t * M_t.unsqueeze(0)
ate_t = (pred_a1_t - pred_a0_t).mean(dim=1)
print(f"PyTorch G-Formula ATE: {ate_t.mean().item():.3f}")
```

```stan
// Stan Code: Generated Quantities G-Formula
generated quantities {
  real ate_gformula;
  vector[N] p1; vector[N] p0;
  for (n in 1:N) {
    p1[n] = a + bA * 1.0 + bM * M[n];
    p0[n] = a + bA * 0.0 + bM * M[n];
  }
  ate_gformula = mean(p1) - mean(p0);
}
```

---

### Question 6: Why does the Waffle House association not identify an intervention effect? {#q6}
**Answer:**  
The association between Waffle House density $W$ and divorce rate $D$ is confounded by geography/region $S$ ($W \leftarrow S \rightarrow D$). Opening a new Waffle House ($\mathrm{do}(W=w)$) manipulates $W$ without altering regional demographic causes. Unadjusted regression captures backdoor association, not causal intervention.

---

### Question 7: What does standardization accomplish, and what does it not accomplish? {#q7}
**Answer:**  
- **Accomplishes**: Centers/scales variables to simplify prior specifications ($\mathrm{Normal}(0, 0.5)$), improves optimizer geometry, and allows comparison of relative slope magnitudes.
- **Does Not Accomplish**: Does not eliminate confounding, generate causal identification, fix lack of support/positivity violations, or define the scientific target.

---

### Question 8: What do the three divorce regressions estimate statistically? {#q8}
**Answer:**  
1. `m5.1` ($D^* \sim M^*$): Estimates marginal conditional expectation $\mathbb{E}[D^* \mid M^*]$.
2. `m5.2` ($D^* \sim A^*$): Estimates marginal conditional expectation $\mathbb{E}[D^* \mid A^*]$.
3. `m5.3` ($D^* \sim M^* + A^*$): Estimates joint conditional expectations $\mathbb{E}[D^* \mid M^*, A^*]$, isolating partial slopes holding the other predictor constant.

---

### Question 9: Under $A \rightarrow M \rightarrow D$ and $A \rightarrow D$, why should a total-effect model for $A$ avoid conditioning on $M$? {#q9}
**Answer:**  
Under the DAG $A \to M \to D$ and $A \to D$, marriage rate $M$ acts as a post-treatment mediator on the causal path $A \to M \to D$. Conditioning on $M$ blocks the indirect causal pathway, converting the total effect estimand into a controlled direct effect estimand ($\mathbb{E}[D \mid \mathrm{do}(A), \mathrm{do}(M=m)]$).

---

### Question 10: Why does estimating the effect of $M$ require adjustment for $A$ under that graph? {#q10}
**Answer:**  
In the DAG $M \leftarrow A \rightarrow D$, age at marriage $A$ is a common cause influencing both treatment $M$ and outcome $D$. Failing to adjust for $A$ leaves the backdoor path $M \leftarrow A \rightarrow D$ open, creating spurious confounding bias.

---

### Question 11: How is posterior standardization performed draw by draw? {#q11}
**Answer:**  
For each MCMC/joint posterior sample draw $s = 1, \ldots, S$:
1. Generate predictions for all units $i=1,\ldots,N$ setting intervention $A=a_1$: $\mu_i^{(s)}(a_1) = f(\boldsymbol{\theta}^{(s)}, a_1, X_i)$.
2. Generate predictions for all units setting $A=a_0$: $\mu_i^{(s)}(a_0) = f(\boldsymbol{\theta}^{(s)}, a_0, X_i)$.
3. Compute the population mean contrast for draw $s$: $\tau^{(s)} = \frac{1}{N} \sum_{i=1}^N \left( \mu_i^{(s)}(a_1) - \mu_i^{(s)}(a_0) \right)$.
4. Summarize the resulting vector $\{\tau^{(1)}, \ldots, \tau^{(S)}\}$ to obtain exact posterior uncertainty.

---

### Question 12: Why is the target covariate distribution part of the estimand? {#q12}
**Answer:**  
When treatment effects vary across units ($\tau(X) \neq \text{const}$), the overall average effect depends directly on the distribution $p_{\mathrm{target}}(X)$ over which individual effects are averaged. Changing the target population alters the numerical value of the estimand.

---

### Question 13: How does a counterfactual plot differ from an ordinary prediction plot? {#q13}
**Answer:**  
- **Ordinary Prediction Plot**: Displays conditional expectations $\mathbb{E}[Y \mid X_1, X_2]$ where covariates change together according to their observed correlation structure.
- **Counterfactual Plot**: Displays predictions where one variable is manipulated ($\mathrm{do}(X_1 = x)$) while holding other variables artificially fixed at specific values ($X_2 = 0$), breaking observed correlations.

---

### Question 14: What is a predictor residual plot useful for? {#q14}
**Answer:**  
A predictor residual plot displays the outcome against the residuals of one predictor regressed on other covariates ($R_{M \mid A} = M - \widehat{M}(A)$). It visually illustrates what variation remains to estimate a partial slope in multiple regression.

---

### Question 15: Why does multiple regression not determine the causal graph? {#q15}
**Answer:**  
Multiple regression is purely a conditional association engine. It computes $\mathbb{E}[Y \mid X_1, X_2]$ without knowledge of arrow directions, mediators, colliders, or unobserved confounders. Causal claims require an externally justified DAG.

---

### Question 16: How can conditioning reveal a masked relationship? {#q16}
**Answer:**  
When two predictors $X_1$ and $X_2$ are positively correlated with each other, but have opposite causal effects on outcome $Y$ ($X_1 \to Y$ positive, $X_2 \to Y$ negative), their marginal associations with $Y$ can suppress or cancel each other out. Conditioning on $X_2$ holds it fixed, unmasking the strong conditional effect of $X_1$.

---

### Question 17: Name three estiplans that can target the same ATE. {#q17}
**Answer:**  
1. **Outcome Regression G-Computation Estiplan**: Fit $\mathbb{E}[Y \mid A, X]$, predict under $A=1$ and $A=0$, and average over $p(X)$.
2. **Inverse Probability Weighting (IPW) Estiplan**: Fit propensity model $\Pr(A=1 \mid X)$, weight observations by $1/\Pr(A_i \mid X_i)$, and contrast weighted means.
3. **Doubly Robust (AIPW) Estiplan**: Combine outcome regression and propensity weighting to achieve consistency if either model is correctly specified.

---

### Question 18: How can one outcome model support SATE, PATE, ATT, and policy value? {#q18}
**Answer:**  
A single fitted posterior outcome model $p(Y \mid A, X, \mathcal{D})$ generates conditional predictions $\mu^{(s)}(a, x)$. By changing the averaging weight vector, it yields:
- **SATE**: Average over sample $X_{1:N}$.
- **PATE**: Average over target population weights $p_{\mathrm{pop}}(x)$.
- **ATT**: Average over treated units ($A_i=1$).
- **Policy Value $V(\pi)$**: Average over policy-assigned treatment states $A = \pi(X)$.

---

### Question 19: Why is overlap required for standardization? {#q19}
**Answer:**  
Standardization predicts outcomes under $A=a$ for units across the covariate space. If positivity fails ($\Pr(A=a \mid X=x) = 0$), there are no observed data for treatment state $a$ at $x$, forcing predictions to rely entirely on unverified parametric extrapolation.

---

### Question 20: What should simulation validation evaluate beyond parameter recovery? {#q20}
**Answer:**  
Simulation validation (SBC) should evaluate the complete end-to-end estiplan: recovery of the target estimand $\tau$, coverage of population-standardized contrasts, behavior under missing data or mis-specified DAGs, and sensitivity to overlap violations.

---

### Question 21: In a CRM offer study, why is the treatment coefficient not necessarily incremental profit? {#q21}
**Answer:**  
The treatment coefficient $\beta_A$ measures marginal sales volume change under linearity. Incremental profit involves a non-linear utility function $\Pi(a) = \text{Margin} \cdot Y(a) - \text{Cost}(a)$, accounts for offer costs, and averages over customer eligibility and heterogeneous response profiles.

---

### Question 22: Which fields belong in an `EstimandArtifact`? {#q22}
**Answer:**  
- `estimand_id`
- `outcome`
- `intervention_state` & `comparison_state`
- `contrast_type` (difference, ratio)
- `target_population` (eligibility filter)
- `aggregation_rule`
- `measurement_scale`

---

### Question 23: Which operations belong in an `EstiplanArtifact`? {#q23}
**Answer:**  
- `identification_strategy` & `adjustment_set`
- `model_specifications`
- `preprocessing_transformations`
- `counterfactual_prediction_steps`
- `standardization_population`
- `draw_by_draw_contrast_operation`
- `validation_checks`

---

### Question 24: When should CausalOps fail closed rather than report a causal estimate? {#q24}
**Answer:**  
CausalOps must fail closed (refuse to issue a causal claim) when:
1. The intervention or comparison state is undefined.
2. Backdoor adjustment set variables are missing or unobserved.
3. Positivity/overlap is severely violated ($\Pr(A=a \mid X) \approx 0$).
4. Post-treatment mediators are conditioned on for a total effect target.
5. The estiplan fails simulation-based calibration recovery.

---

## 30. Book reading guide

### Primary reading

- Chapter 4, *Geocentric Models*
  - Section 4.4, *Linear prediction*
  - especially posterior expectations and predictive simulation
- Chapter 5, *The Many Variables & The Spurious Waffles*
  - Section 5.1, *Spurious association*
  - Section 5.1.5, multivariable posterior plots
  - Section 5.2, *Masked relationship*
  - Section 5.3, *Categorical variables*, as continuity from A04

### Read before A05

Focus on:

- estimand, estimator, and estimate from A01;
- posterior contrasts from A04;
- Gaussian multiple regression;
- the Waffle House, marriage-rate, and age-at-marriage plots;
- the candidate causal graph for the divorce example.

### Revisit after A05

Return to:

- the three divorce regression models;
- predictor residual plots;
- posterior prediction plots;
- counterfactual plots;
- the milk-energy masked-relationship example;
- the distinction between fitting a conditional model and calculating an interventional estimand.

### Course-to-book translation

The second edition has no section titled “Estiplans.”

The closest book workflow is:

$$ \text{define a scientific question} + \text{fit conditional models} + \text{simulate counterfactual predictions} + \text{construct contrasts} $$

A05 names this combined sequence explicitly.

---

## 31. Official code guide

### `05_elemental_confounds.r`

The script contains material shared with the transition into A06. The portion most directly relevant to A05 is the divorce analysis.

It:

- loads `WaffleDivorce`;
- plots Waffle House density against divorce rate;
- plots marriage rate against divorce;
- plots median age at marriage against divorce;
- plots median age at marriage against marriage rate;
- visualizes stratification over a continuous variable;
- standardizes the principal variables;
- performs prior predictive checks;
- fits the multivariable divorce model.

The fitted model is:

$$ D_i^* \sim \mathrm{Normal}(\mu_i,\sigma) $$

$$ \mu_i = \alpha+\beta_M M_i^*+\beta_A A_i^* $$

with:

$$ \alpha \sim \mathrm{Normal}(0,0.2) $$

$$ \beta_M \sim \mathrm{Normal}(0,0.5) $$

$$ \beta_A \sim \mathrm{Normal}(0,0.5) $$

$$ \sigma \sim \mathrm{Exponential}(1) $$

The same script also contains fork, pipe, collider, descendant, selection-bias, and d-separation simulations. Those sections are developed more directly in A06 and A07.

---

## 32. Sources

- Richard McElreath, *Statistical Rethinking: A Bayesian Course with Examples in R and Stan*, second edition.
- Official `stat_rethinking_2026` repository and A05 lecture listing.
- Official A05 homework and solution files.
- Official repository script `05_elemental_confounds.r`, especially the divorce-model section.
