---
title: "A06 — Elemental Confounds I"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
    mermaid:
      theme: default
execute:
  eval: true
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# A06 — Elemental Confounds I

**Course:** Statistical Rethinking 2026  
**Section:** A — Beginner  
**Lecture:** A06 — Elemental Confounds I  
**Official reading:** Chapter 6 of *Statistical Rethinking*, second edition  
**Central question:** How does conditioning open or close statistical paths, and how can adjustment either remove confounding or create it?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official repository scripts `05_elemental_confounds.r`, `06_breen_collider_animation.R`, and `06_DAG_animations.r`.
>
> The second edition's Chapter 6 also includes multicollinearity and a broader set of bad-control examples. The official A06 title and available scripts place the main emphasis here on the elementary DAG structures, d-separation, selection distortion, and collider bias. The script `06_simulations_bad_controls.r` is treated mainly as preparation for A07 — Good and Bad Controls.

---

## Table of Contents

- [1. Main thesis](#1-main-thesis)
- [2. Relationship to the book](#2-relationship-to-the-book)
  - [2.1 Course emphasis versus chapter coverage](#21-course-emphasis-versus-chapter-coverage)
- [3. Conditioning](#3-conditioning)
- [4. Paths and information flow](#4-paths-and-information-flow)
- [5. The fork](#5-the-fork)
  - [5.1 Marginal association](#51-marginal-association)
  - [5.2 Conditioning on the common cause](#52-conditioning-on-the-common-cause)
  - [5.3 Simulation in the official script](#53-simulation-in-the-official-script)
  - [5.4 Implementation (R, PyMC, Stan, PyTorch)](#54-implementation-r-pymc-stan-pytorch)
- [6. The pipe](#6-the-pipe)
  - [6.1 Unconditioned path](#61-unconditioned-path)
  - [6.2 Conditioning on the mediator](#62-conditioning-on-the-mediator)
  - [6.3 Total versus direct effect](#63-total-versus-direct-effect)
  - [6.4 Implementation (R, PyMC, Stan, PyTorch)](#64-implementation-r-pymc-stan-pytorch)
- [7. The collider](#7-the-collider)
  - [7.1 Unconditioned collider](#71-unconditioned-collider)
  - [7.2 Conditioning on the collider](#72-conditioning-on-the-collider)
  - [7.3 Intuition](#73-intuition)
  - [7.4 Implementation (R, PyMC, Stan, PyTorch)](#74-implementation-r-pymc-stan-pytorch)
- [8. Collider bias is created, not removed](#8-collider-bias-is-created-not-removed)
  - [8.1 Implementation (R, PyMC, Stan, PyTorch)](#81-implementation-r-pymc-stan-pytorch)
- [9. Descendants](#9-descendants)
  - [9.1 Implementation (R, PyMC, Stan, PyTorch)](#91-implementation-r-pymc-stan-pytorch)
- [10. D-separation](#10-d-separation)
  - [10.1 Implementation (R, PyMC, Stan, PyTorch)](#101-implementation-r-pymc-stan-pytorch)
- [11. Causal paths and backdoor paths](#11-causal-paths-and-backdoor-paths)
  - [11.1 Implementation (R, PyMC, Stan, PyTorch)](#111-implementation-r-pymc-stan-pytorch)
- [12. The backdoor-analysis recipe](#12-the-backdoor-analysis-recipe)
- [13. Confounding is a path problem](#13-confounding-is-a-path-problem)
- [14. Fork simulation](#14-fork-simulation)
- [15. Pipe simulation](#15-pipe-simulation)
- [16. Collider simulation](#16-collider-simulation)
- [17. Selection distortion in grant proposals](#17-selection-distortion-in-grant-proposals)
- [18. Selection can occur before modeling](#18-selection-can-occur-before-modeling)
- [19. Happiness, age, and marriage](#19-happiness-age-and-marriage)
- [20. The haunted DAG](#20-the-haunted-dag)
  - [20.1 Why conditioning on parent education is dangerous](#201-why-conditioning-on-parent-education-is-dangerous)
  - [20.2 Animation logic](#202-animation-logic)
- [21. Why unobserved variables still matter](#21-why-unobserved-variables-still-matter)
- [22. The Table 2 fallacy](#22-the-table-2-fallacy)
- [23. Multicollinearity as an identification warning](#23-multicollinearity-as-an-identification-warning)
- [24. Post-treatment bias](#24-post-treatment-bias)
- [25. Adjustment sets are estimand-specific](#25-adjustment-sets-are-estimand-specific)
- [26. Regression implementation after identification](#26-regression-implementation-after-identification)
- [27. DAG software](#27-dag-software)
- [28. CRM application: targeting confounding](#28-crm-application-targeting-confounding)
- [29. CRM application: campaign engagement as mediator](#29-crm-application-campaign-engagement-as-mediator)
- [30. CRM application: openers-only analysis](#30-crm-application-openers-only-analysis)
- [31. CRM application: support-contact selection](#31-crm-application-support-contact-selection)
- [32. Implications for CausalOps](#32-implications-for-causalops)
  - [32.1 Path audit artifact](#321-path-audit-artifact)
  - [32.2 Control-role validation](#322-control-role-validation)
  - [32.3 Selection audit](#323-selection-audit)
- [33. Common errors exposed by A06](#33-common-errors-exposed-by-a06)
- [34. Lecture takeaways](#34-lecture-takeaways)
- [35. Review questions](#35-review-questions)
- [36. Book reading guide](#36-book-reading-guide)
- [37. Official code guide](#37-official-code-guide)
- [38. Appendix — Homework A06 Question & Solutions](#38-appendix--homework-a06-question--solutions)

---

## 1. Main thesis

The central argument is:

> **Conditioning is a causal operation on an assumed graph, not a universally beneficial statistical refinement.**

Adding a variable to a regression changes which comparisons are made. Depending on the variable's position in the causal graph, conditioning may:

- close a noncausal backdoor path;
- block part of the causal effect of interest;
- open a path that was previously closed;
- create selection distortion;
- or partially condition on an unobserved structure through a descendant or proxy.

The workflow is:

$$ \text{scientific question} \rightarrow \text{causal DAG} \rightarrow \text{enumerate paths} \rightarrow \text{classify open and closed paths} \rightarrow \text{identify backdoor paths} \rightarrow \text{choose an adjustment set} \rightarrow \text{estimate the target contrast} $$

The statistical model does not determine the adjustment set. The causal model does.

---

## 2. Relationship to the book

A06 maps directly to Chapter 6, especially Sections 6.3–6.4.

| A06 topic | Second-edition counterpart | Role |
|---|---|---|
| Risks of adding variables indiscriminately | Chapter 6 introduction | Multiple regression can create bias |
| Multicollinearity | Section 6.1 | Data may not separately identify correlated effects |
| Post-treatment bias | Section 6.2 | Conditioning can block causal pathways |
| Collider bias | Section 6.3 | Conditioning can open noncausal paths |
| Haunted DAG | Section 6.3.2 | Unobserved causes can create hidden collider structures |
| Fork, pipe, collider, descendant | Section 6.4.1 | Elementary path structures |
| Backdoor-path analysis | Sections 6.4.1–6.4.2 | Determines adjustment requirements |
| `dagitty` adjustment sets | Section 6.4.2 | Automates graph implications after the DAG is supplied |

### 2.1 Course emphasis versus chapter coverage

Chapter 6 is organized as:

$$ \text{multicollinearity} + \text{post-treatment bias} + \text{collider bias} + \text{DAG adjustment rules} $$

The A06 scripts emphasize:

$$ \text{fork} + \text{pipe} + \text{collider} + \text{descendant} + \text{selection} + \text{d-separation} $$

A07 continues the practical question of which variables are good or bad controls.

---

## 3. Conditioning

Conditioning means restricting comparisons to observations with the same or similar values of another variable.

In a regression such as:

$$ Y_i = \alpha + \beta_X X_i + \beta_Z Z_i + \varepsilon_i $$

$\beta_X$ describes variation in $Y$ associated with variation in $X$ while holding $Z$ fixed under the model.

This is a statistical statement.

Its causal meaning depends on what $Z$ is:

- a common cause;
- a mediator;
- a collider;
- a descendant;
- a proxy;
- or an irrelevant predictor.

The same algebra can therefore support a valid causal comparison, a different estimand, or a biased comparison.

---

## 4. Paths and information flow

A **path** is any sequence of connected edges between two variables, ignoring arrow direction while tracing the sequence.

Suppose the exposure is $X$ and the outcome is $Y$.

A path can transmit statistical association when it is **open**. A path is **blocked** when its structure and conditioning status prevent information from flowing through it.

The elementary structures and their path-blocking rules are summarized in the Mermaid diagram below:

```{mermaid}
%%| label: fig-paths-flow
%%| fig-cap: "Elementary path structures and blocking conditions"
flowchart TD
    subgraph Fork["1. The Fork: X <- Z -> Y"]
        direction LR
        F1["X"] <--- F2["Z"] ---> F3["Y"]
        F4["X"] -.- F5["Z (conditioned)"] -.- F6["Y"]
    end

    subgraph Pipe["2. The Pipe: X -> Z -> Y"]
        direction LR
        P1["X"] ---> P2["Z"] ---> P3["Y"]
        P4["X"] -.- P5["Z (conditioned)"] -.- P6["Y"]
    end

    subgraph Collider["3. The Collider: X -> Z <- Y"]
        direction LR
        C1["X"] -.- C2["Z"] -.- C3["Y"]
        C4["X"] ---> C5["Z (conditioned)"] <--- C6["Y"]
    end

    subgraph Descendant["4. Collider Descendant: Z -> D"]
        direction LR
        D1["X"] -.- D2["Z"] -.- D3["Y"]
        D2 -.- D4["D"]
        D5["X"] ---> D6["Z"] <--- D7["Y"]
        D6 ---> D8["D (conditioned)"]
    end
```

The elementary structures are:

1. **Fork** ($X \leftarrow Z \rightarrow Y$): Open by default; blocked when conditioning on common cause $Z$.
2. **Pipe** ($X \rightarrow Z \rightarrow Y$): Open by default; blocked when conditioning on mediator $Z$.
3. **Collider** ($X \rightarrow Z \leftarrow Y$): Blocked by default; opened when conditioning on collider $Z$.
4. **Descendant of a Collider** ($Z \rightarrow D$): Blocked by default; partially opened when conditioning on descendant $D$.

Every larger DAG can be analyzed by decomposing its paths into these local structures.

---

## 5. The fork

A fork has the structure:

$$ X \leftarrow Z \rightarrow Y $$

$Z$ is a common cause of $X$ and $Y$.

### 5.1 Marginal association

Even when $X$ has no causal effect on $Y$, they can be associated because both respond to $Z$:

$$ X \not\!\perp\!\!\!\perp Y $$

### 5.2 Conditioning on the common cause

Conditioning on $Z$ closes the path:

$$ X \perp\!\!\!\perp Y \mid Z $$

This is the elementary confounding pattern.

### 5.3 Simulation in the official script

The script generates:

$$ Z \sim \mathrm{Bernoulli}(0.5) $$

and makes both $X$ and $Y$ strongly depend on $Z$.

Marginally, $X$ and $Y$ are correlated. Within $Z=0$ and within $Z=1$, the correlation largely disappears.

The adjustment works because $Z$ blocks an open backdoor path.

### 5.4 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
Z <- rbern(n, 0.5)
X <- rbern(n, (1 - Z) * 0.1 + Z * 0.9)
Y <- rbern(n, (1 - Z) * 0.1 + Z * 0.9)

# Marginal model: Y ~ X (spurious association created by common cause Z)
m_marginal <- lm(Y ~ X)
summary(m_marginal)

# Conditional model: Y ~ X + Z (backdoor path closed by Z)
m_cond <- lm(Y ~ X + Z)
summary(m_cond)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.binomial(1, 0.5, size=n)
X_data = np.random.binomial(1, (1 - Z_data) * 0.1 + Z_data * 0.9)
Y_data = np.random.binomial(1, (1 - Z_data) * 0.1 + Z_data * 0.9)

with pm.Model() as model_fork:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_fork = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_fork, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.bernoulli(torch.full((n, 1), 0.5))
X_t = torch.bernoulli((1 - Z_t) * 0.1 + Z_t * 0.9)
Y_t = torch.bernoulli((1 - Z_t) * 0.1 + Z_t * 0.9)

# Regress Y on X and Z
inputs = torch.cat([X_t, Z_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Fork weights [bX, bZ]:", model.weight.data.numpy())
```
:::

---

## 6. The pipe

A pipe has the structure:

$$ X \rightarrow Z \rightarrow Y $$

$Z$ mediates the effect of $X$ on $Y$.

### 6.1 Unconditioned path

The path is open, so changes in $X$ can propagate through $Z$ to $Y$.

The total effect includes this mediated pathway.

### 6.2 Conditioning on the mediator

Conditioning on $Z$ blocks the pipe:

$$ X \perp\!\!\!\perp Y \mid Z $$

when the pipe is the only path connecting them.

This is useful only when the estimand deliberately excludes the mediated effect.

### 6.3 Total versus direct effect

If the graph also contains:

$$ X \rightarrow Y $$

then the total effect combines:

$$ X \rightarrow Y $$

and:

$$ X \rightarrow Z \rightarrow Y $$

Conditioning on $Z$ removes the second component and changes the estimand.

The lesson is:

$$ \text{blocking a path} \neq \text{removing bias automatically} $$

It may remove part of the effect we intended to estimate.

### 6.4 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
# Treatment T
T <- rbern(n, 0.5)
# Fungus F (mediator: present if T=0 mostly)
F <- rbern(n, (1 - T) * 0.8 + T * 0.1)
# Plant height H (fungus reduces height)
H <- rnorm(n, mean = 10 - 3 * F, sd = 1)

# Total effect model: H ~ T (recovers full causal pathway T -> F -> H)
m_total <- lm(H ~ T)
summary(m_total)

# Direct effect model: H ~ T + F (blocks mediator pipe F)
m_direct <- lm(H ~ T + F)
summary(m_direct)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
T_data = np.random.binomial(1, 0.5, size=n)
F_data = np.random.binomial(1, (1 - T_data) * 0.8 + T_data * 0.1)
H_data = np.random.normal(loc=10 - 3 * F_data, scale=1.0)

with pm.Model() as model_pipe:
    a = pm.Normal("a", mu=10, sigma=2)
    bT = pm.Normal("bT", mu=0, sigma=1)
    bF = pm.Normal("bF", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bT * T_data + bF * F_data
    h_obs = pm.Normal("h_obs", mu=mu, sigma=sigma, observed=H_data)
    
    idata_pipe = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_pipe, var_names=["bT", "bF"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] T;
  vector[N] F;
  vector[N] H;
}
parameters {
  real a;
  real bT;
  real bF;
  real<lower=0> sigma;
}
model {
  a ~ normal(10, 2);
  bT ~ normal(0, 1);
  bF ~ normal(0, 1);
  sigma ~ exponential(1);
  H ~ normal(a + bT * T + bF * F, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
T_t = torch.bernoulli(torch.full((n, 1), 0.5))
F_t = torch.bernoulli((1 - T_t) * 0.8 + T_t * 0.1)
H_t = torch.normal(mean=10.0 - 3.0 * F_t, std=1.0)

# Direct effect estimation (conditioning on mediator F)
inputs = torch.cat([T_t, F_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, H_t)
    loss.backward()
    optimizer.step()

print("PyTorch Pipe Direct Effect [bT, bF]:", model.weight.data.numpy())
```
:::

---

## 7. The collider

A collider has the structure:

$$ X \rightarrow Z \leftarrow Y $$

$Z$ is a common consequence of $X$ and $Y$.

### 7.1 Unconditioned collider

The path is closed by default.

If $X$ and $Y$ have no other connection, then:

$$ X \perp\!\!\!\perp Y $$

### 7.2 Conditioning on the collider

Conditioning on $Z$ opens the path:

$$ X \not\!\perp\!\!\!\perp Y \mid Z $$

Once the value of the common consequence is known, information about one cause changes what is inferred about the other.

### 7.3 Intuition

Suppose admission depends on both academic strength and athletic ability.

Among admitted applicants, learning that one person has lower academic strength implies that stronger athletic ability is more likely to explain admission.

The causes become associated inside the selected group even when they were independent in the applicant population.

This is selection distortion.

### 7.4 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
N_val <- rnorm(n, 0, 1)
T_val <- rnorm(n, 0, 1)
S_score <- N_val + T_val
selected <- S_score > quantile(S_score, 0.90)

# Unconditioned model in full population (independent N _||_ T)
cat("Full sample correlation Cor(N, T):", round(cor(N_val, T_val), 3), "\n")
m_full <- lm(T_val ~ N_val)

# Conditioned on collider (selected sample: spurious negative correlation)
cat("Selected sample correlation Cor(N, T | S=1):", round(cor(N_val[selected], T_val[selected]), 3), "\n")
m_selected <- lm(T_val[selected] ~ N_val[selected])
summary(m_selected)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
N_val = np.random.normal(0, 1, size=n)
T_val = np.random.normal(0, 1, size=n)
S_score = N_val + T_val
selected = S_score > np.quantile(S_score, 0.90)

N_sel = N_val[selected]
T_sel = T_val[selected]

with pm.Model() as model_collider:
    a = pm.Normal("a", mu=0, sigma=1)
    bN = pm.Normal("bN", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bN * N_sel
    t_obs = pm.Normal("t_obs", mu=mu, sigma=sigma, observed=T_sel)
    
    idata_collider = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_collider, var_names=["bN"]))
```

### Stan
```stan
data {
  int<lower=0> N_sel;
  vector[N_sel] N_val;
  vector[N_sel] T_val;
}
parameters {
  real a;
  real bN;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bN ~ normal(0, 1);
  sigma ~ exponential(1);
  T_val ~ normal(a + bN * N_val, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
N_t = torch.randn(n, 1)
T_t = torch.randn(n, 1)
S_score = N_t + T_t
threshold = torch.quantile(S_score, 0.90)
mask = (S_score > threshold).squeeze()

N_sel = N_t[mask]
T_sel = T_t[mask]

model = nn.Linear(1, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(N_sel)
    loss = criterion(preds, T_sel)
    loss.backward()
    optimizer.step()

print("PyTorch weight bN on selected sample:", model.weight.data.numpy()[0][0])
```
:::

---

## 8. Collider bias is created, not removed

Confounder adjustment removes an existing noncausal path.

Collider adjustment creates a new noncausal path.

The contrast is:

### Confounder

$$ X \leftarrow Z \rightarrow Y $$

Conditioning on $Z$ closes the path.

### Collider

$$ X \rightarrow Z \leftarrow Y $$

Conditioning on $Z$ opens the path.

This is why variable selection cannot be based on correlation, predictive strength, or temporal proximity alone.

The arrow structure matters.

### 8.1 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
# Age between 18 and 65
A <- runif(n, 18, 65)
# Happiness unassociated with Age
H <- rnorm(n, 0, 1)
# Marriage probability increases with Age and Happiness (A -> M <- H)
p_married <- logistic(-2 + 0.08 * (A - 18) + 1.5 * H)
M <- rbern(n, p_married)

# True model without conditioning on collider M (slope for Age is zero)
m_true <- lm(H ~ A)
summary(m_true)

# Confounded model conditioning on collider M (creates false negative association)
m_collider <- lm(H ~ A + factor(M))
summary(m_collider)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az
from scipy.special import expit

np.random.seed(123)
n = 1000
A_data = np.random.uniform(18, 65, size=n)
H_data = np.random.normal(0, 1, size=n)
p_married = expit(-2 + 0.08 * (A_data - 18) + 1.5 * H_data)
M_data = np.random.binomial(1, p_married)

with pm.Model() as model_marriage:
    a = pm.Normal("a", mu=0, sigma=1)
    bA = pm.Normal("bA", mu=0, sigma=1)
    bM = pm.Normal("bM", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bA * (A_data - 18) + bM * M_data
    h_obs = pm.Normal("h_obs", mu=mu, sigma=sigma, observed=H_data)
    
    idata_m = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_m, var_names=["bA", "bM"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] A;
  vector[N] M;
  vector[N] H;
}
parameters {
  real a;
  real bA;
  real bM;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bA ~ normal(0, 1);
  bM ~ normal(0, 1);
  sigma ~ exponential(1);
  H ~ normal(a + bA * (A - 18) + bM * M, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
A_t = torch.empty(n, 1).uniform_(18, 65)
H_t = torch.randn(n, 1)
p_m = torch.sigmoid(-2 + 0.08 * (A_t - 18) + 1.5 * H_t)
M_t = torch.bernoulli(p_m)

# Inputs: Standardized Age and Marriage status
A_std = (A_t - 18) / 47.0
inputs = torch.cat([A_std, M_t], dim=1)

model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, H_t)
    loss.backward()
    optimizer.step()

print("PyTorch weights [bA, bM]:", model.weight.data.numpy())
```
:::

---

## 9. Descendants

A descendant is any variable causally downstream of another variable.

Suppose:

$$ X \rightarrow Z \leftarrow Y $$

and:

$$ Z \rightarrow D $$

$D$ is a descendant of the collider $Z$.

Conditioning on $D$ provides information about $Z$ and can therefore open the collider path partially:

$$ X \rightarrow Z \leftarrow Y $$

The degree of induced association depends on how informative $D$ is about $Z$.

This is important because analysts often lack direct measurements and use:

- proxy variables;
- scores;
- eligibility flags;
- downstream behaviors;
- administrative labels.

A proxy inherits some of the causal consequences of the variable it measures.

### 9.1 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
X <- rnorm(n, 0, 1)
Y <- rnorm(n, 0, 1)
# Z is a collider of X and Y
Z <- rnorm(n, X + Y, 0.5)
# D is a descendant of collider Z
D <- rnorm(n, Z, 0.5)

# 1. Unconditioned (correct independence X _||_ Y)
m1 <- lm(Y ~ X)
cat("Unconditioned bX:", round(coef(m1)["X"], 3), "\n")

# 2. Conditioned on descendant D (partially opens collider path!)
m2 <- lm(Y ~ X + D)
cat("Conditioned on descendant D bX:", round(coef(m2)["X"], 3), "\n")

# 3. Conditioned directly on collider Z (fully opens collider path)
m3 <- lm(Y ~ X + Z)
cat("Conditioned on collider Z bX:", round(coef(m3)["X"], 3), "\n")
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.normal(0, 1, size=n)
Y_data = np.random.normal(0, 1, size=n)
Z_data = np.random.normal(X_data + Y_data, 0.5)
D_data = np.random.normal(Z_data, 0.5)

with pm.Model() as model_descendant:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bD = pm.Normal("bD", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bD * D_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_d = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_d, var_names=["bX", "bD"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] D;
  vector[N] Y;
}
parameters {
  real a;
  real bX;
  real bD;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bD ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bD * D, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.randn(n, 1)
Y_t = torch.randn(n, 1)
Z_t = torch.normal(X_t + Y_t, 0.5)
D_t = torch.normal(Z_t, 0.5)

inputs = torch.cat([X_t, D_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Descendant weights [bX, bD]:", model.weight.data.numpy())
```
:::

---

## 10. D-separation

D-separation is the graphical rule for determining conditional independence in a DAG.

Two variables are d-separated given a conditioning set $S$ when every path between them is blocked.

The local rules are:

### Fork

$$ X \leftarrow Z \rightarrow Y $$

Open unless $Z$ is conditioned on.

### Pipe

$$ X \rightarrow Z \rightarrow Y $$

Open unless $Z$ is conditioned on.

### Collider

$$ X \rightarrow Z \leftarrow Y $$

Closed unless $Z$ or one of its descendants is conditioned on.

The rules are local, but their implications apply to full paths in large DAGs.

### 10.1 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking / dagitty)
```{r}
library(dagitty)
library(rethinking)

# Define DAG with fork (Z), pipe (M), and collider (C)
g <- dagitty("dag {
    Z -> X
    Z -> Y
    X -> M -> Y
    X -> C <- Y
}")

# Query implied conditional independencies automatically
cat("Implied conditional independencies:\n")
print(impliedConditionalIndependencies(g))

# Test specific d-separations
cat("Is X d-separated from Y given Z and M? ", dseparated(g, "X", "Y", c("Z", "M")), "\n")
cat("Is X d-separated from Y given C? ", dseparated(g, "X", "Y", c("C")), "\n")
```

### Python (PyMC / dagitty / networkx)
```python
import networkx as nx
import numpy as np

# Create directed acyclic graph in NetworkX
G = nx.DiGraph()
G.add_edges_from([("Z", "X"), ("Z", "Y"), ("X", "M"), ("M", "Y"), ("X", "C"), ("Y", "C")])

# Verify d-separation using networkx d_separated rule
print("Is X d-separated from Y given Z and M?", nx.d_separated(G, {"X"}, {"Y"}, {"Z", "M"}))
print("Is X d-separated from Y given C?", nx.d_separated(G, {"X"}, {"Y"}, {"C"}))
```

### Stan
```stan
// Structural Equation Model representing D-separation graph verification
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
  vector[N] M;
}
parameters {
  real a;
  real bX;
  real bZ;
  real bM;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  bM ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z + bM * M, sigma);
}
```

### PyTorch
```python
import torch

# Demonstrate numerical covariance checking for D-separation
torch.manual_seed(123)
n = 1000
Z = torch.randn(n, 1)
X = 0.8 * Z + torch.randn(n, 1) * 0.5
Y = 0.8 * Z + torch.randn(n, 1) * 0.5

# Calculate partial correlation between X and Y given Z (residual correlation)
res_X = X - (torch.cov(torch.cat([X, Z], dim=1))[0, 1] / torch.var(Z)) * Z
res_Y = Y - (torch.cov(torch.cat([Y, Z], dim=1))[0, 1] / torch.var(Z)) * Z
partial_corr = torch.cov(torch.cat([res_X, res_Y], dim=1))[0, 1] / (torch.std(res_X) * torch.std(res_Y))

print("PyTorch Partial Correlation Cor(X, Y | Z):", partial_corr.item())
```
:::

---

## 11. Causal paths and backdoor paths

A causal path from $X$ to $Y$ follows arrows away from $X$ toward $Y$.

For example:

$$ X \rightarrow M \rightarrow Y $$

A backdoor path begins with an arrow entering $X$.

For example:

$$ X \leftarrow C \rightarrow Y $$

Backdoor paths create associations that do not arise from the causal effect of $X$ on $Y$.

To identify the total effect of $X$ on $Y$, an adjustment set should:

1. close every open backdoor path;
2. avoid blocking causal paths from $X$ to $Y$;
3. avoid opening paths through colliders or collider descendants.

### 11.1 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking / dagitty)
```{r}
library(dagitty)
library(rethinking)

# DAG with backdoor path via C and direct/mediated causal path via M
g_backdoor <- dagitty("dag {
    C -> X
    C -> Y
    X -> M -> Y
    X -> Y
}")

# Find adjustment sets for total effect of X on Y
adjustment_sets <- adjustmentSets(g_backdoor, exposure = "X", outcome = "Y")
cat("Minimal adjustment set for total effect X -> Y:\n")
print(adjustment_sets)

# Simulation
set.seed(123)
n <- 1000
C <- rnorm(n, 0, 1)
X <- rnorm(n, 1.5 * C, 1)
M <- rnorm(n, 0.8 * X, 1)
Y <- rnorm(n, 0.5 * X + 1.2 * M + 2.0 * C, 1)

# Confounded regression (no adjustment)
cat("Confounded effect:", round(coef(lm(Y ~ X))["X"], 3), "\n")

# Adjusted regression (closing backdoor C, total effect)
cat("Adjusted total effect (closing C):", round(coef(lm(Y ~ X + C))["X"], 3), "\n")
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
C_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(1.5 * C_data, 1)
M_data = np.random.normal(0.8 * X_data, 1)
Y_data = np.random.normal(0.5 * X_data + 1.2 * M_data + 2.0 * C_data, 1)

with pm.Model() as model_backdoor:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bC = pm.Normal("bC", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bC * C_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_bd = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_bd, var_names=["bX", "bC"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] C;
  vector[N] Y;
}
parameters {
  real a;
  real bX;
  real bC;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bC ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bC * C, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
C_t = torch.randn(n, 1)
X_t = torch.normal(1.5 * C_t, 1.0)
M_t = torch.normal(0.8 * X_t, 1.0)
Y_t = torch.normal(0.5 * X_t + 1.2 * M_t + 2.0 * C_t, 1.0)

# Closing backdoor by including C
inputs = torch.cat([X_t, C_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Total Effect [bX, bC]:", model.weight.data.numpy())
```
:::

---

## 12. The backdoor-analysis recipe

The second edition gives a practical graph-analysis procedure.

### Step 1: List all paths

Enumerate every path connecting the exposure $X$ and outcome $Y$.

### Step 2: Classify each path

Determine whether each path is open or closed under the proposed conditioning set.

### Step 3: Identify backdoor paths

A backdoor path has an arrow pointing into $X$ at its first edge.

### Step 4: Close open backdoors

Choose observed variables that close every open backdoor path without opening new ones.

### Step 5: Preserve the estimand

For a total effect, do not condition on descendants of $X$ that lie on causal pathways to $Y$.

### Step 6: Verify graph implications

Software such as `dagitty` can enumerate adjustment sets after the causal graph has been supplied.

The software checks the graph's logic. It does not validate the graph scientifically.

---

## 13. Confounding is a path problem

It is common to call a variable “a confounder.”

A more precise statement is that a variable lies on an open backdoor path that confounds a specified exposure–outcome effect.

The same variable can have different roles for different estimands.

For example, in:

$$ A \rightarrow M \rightarrow Y $$

$M$ is a mediator for the effect of $A$ on $Y$.

But if another exposure $X$ affects $M$, and $M$ affects both $X$ and $Y$ in a different graph, its role may differ.

Causal roles are relational:

$$ \text{role} = f( \text{graph}, \text{exposure}, \text{outcome}, \text{estimand} ) $$

No variable has one permanent role independent of the question.

---

## 14. Fork simulation

The official script uses a binary common cause:

$$ Z \sim \mathrm{Bernoulli}(0.5) $$

and generates $X$ and $Y$ from probabilities that differ strongly by $Z$.

A stylized version is:

$$ X \sim \mathrm{Bernoulli} \left( 0.1+0.8Z \right) $$

$$ Y \sim \mathrm{Bernoulli} \left( 0.1+0.8Z \right) $$

There is no arrow from $X$ to $Y$.

Marginally, $\mathrm{Cor}(X,Y) \neq 0$. Within strata of $Z$, $\mathrm{Cor}(X,Y \mid Z=z) \approx 0$.

```{r}
#| label: fig-fork-sim
#| fig-cap: "Fork simulation: Marginal vs conditional correlation under X <- Z -> Y"
set.seed(123)
n <- 1000
Z <- rbinom(n, 1, 0.5)
X <- rbinom(n, 1, (1-Z)*0.1 + Z*0.9)
Y <- rbinom(n, 1, (1-Z)*0.1 + Z*0.9)

cat("Marginal correlation Cor(X, Y):", round(cor(X, Y), 3), "\n")
cat("Conditional Cor(X, Y | Z=0):", round(cor(X[Z==0], Y[Z==0]), 3), "\n")
cat("Conditional Cor(X, Y | Z=1):", round(cor(X[Z==1], Y[Z==1]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
Z <- rbern(n, 0.5)
X <- rbern(n, (1-Z)*0.1 + Z*0.9)
Y <- rbern(n, (1-Z)*0.1 + Z*0.9)

# Marginal regression Y ~ X
m_marginal <- lm(Y ~ X)
summary(m_marginal)

# Conditional regression Y ~ X + Z
m_cond <- lm(Y ~ X + Z)
summary(m_cond)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.binomial(1, 0.5, size=n)
X_data = np.random.binomial(1, (1 - Z_data) * 0.1 + Z_data * 0.9)
Y_data = np.random.binomial(1, (1 - Z_data) * 0.1 + Z_data * 0.9)

with pm.Model() as model_fork:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_fork = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_fork, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.bernoulli(torch.full((n, 1), 0.5))
pX = (1 - Z_t) * 0.1 + Z_t * 0.9
pY = (1 - Z_t) * 0.1 + Z_t * 0.9
X_t = torch.bernoulli(pX)
Y_t = torch.bernoulli(pY)

class ForkRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, x_z):
        return self.linear(x_z)

model = ForkRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 15. Pipe simulation

The official script generates:

$$ X \rightarrow Z \rightarrow Y $$

$$ X \sim \mathrm{Bernoulli}(0.5), \quad Z \sim \mathrm{Bernoulli}(0.1 + 0.8X), \quad Y \sim \mathrm{Bernoulli}(0.1 + 0.8Z) $$

```{r}
#| label: fig-pipe-sim
#| fig-cap: "Pipe simulation: Total effect vs mediator-conditioned effect"
set.seed(123)
n <- 1000
X <- rbinom(n, 1, 0.5)
Z <- rbinom(n, 1, (1-X)*0.1 + X*0.9)
Y <- rbinom(n, 1, (1-Z)*0.1 + Z*0.9)

cat("Marginal correlation Cor(X, Y):", round(cor(X, Y), 3), "\n")
cat("Conditional Cor(X, Y | Z=0):", round(cor(X[Z==0], Y[Z==0]), 3), "\n")
cat("Conditional Cor(X, Y | Z=1):", round(cor(X[Z==1], Y[Z==1]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
X <- rbern(n, 0.5)
Z <- rbern(n, (1-X)*0.1 + X*0.9)
Y <- rbern(n, (1-Z)*0.1 + Z*0.9)

# Total effect model: Y ~ X
m_total <- lm(Y ~ X)
summary(m_total)

# Direct effect model (conditioning on mediator Z): Y ~ X + Z
m_direct <- lm(Y ~ X + Z)
summary(m_direct)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.binomial(1, 0.5, size=n)
Z_data = np.random.binomial(1, (1 - X_data) * 0.1 + X_data * 0.9)
Y_data = np.random.binomial(1, (1 - Z_data) * 0.1 + Z_data * 0.9)

with pm.Model() as model_pipe:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_pipe = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_pipe, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.bernoulli(torch.full((n, 1), 0.5))
pZ = (1 - X_t) * 0.1 + X_t * 0.9
Z_t = torch.bernoulli(pZ)
pY = (1 - Z_t) * 0.1 + Z_t * 0.9
Y_t = torch.bernoulli(pY)

class PipeRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, x_z):
        return self.linear(x_z)

model = PipeRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Pipe Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 16. Collider simulation

The official script generates independent $X$ and $Y$ and makes $Z$ likely when either cause is present:

$$ X \sim \mathrm{Bernoulli}(0.5), \quad Y \sim \mathrm{Bernoulli}(0.5), \quad Z \sim \mathrm{Bernoulli} \left[ \text{ifelse}(X+Y>0, 0.9, 0.2) \right] $$

```{r}
#| label: fig-collider-sim
#| fig-cap: "Collider simulation: Unconditioned independence vs conditioned association"
set.seed(123)
n <- 1000
X <- rbinom(n, 1, 0.5)
Y <- rbinom(n, 1, 0.5)
Z <- rbinom(n, 1, ifelse(X + Y > 0, 0.9, 0.2))

cat("Marginal Cor(X, Y):", round(cor(X, Y), 3), "\n")
cat("Conditional Cor(X, Y | Z=0):", round(cor(X[Z==0], Y[Z==0]), 3), "\n")
cat("Conditional Cor(X, Y | Z=1):", round(cor(X[Z==1], Y[Z==1]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(123)
n <- 1000
X <- rbern(n, 0.5)
Y <- rbern(n, 0.5)
Z <- rbern(n, ifelse(X+Y>0, 0.9, 0.2))

# Marginal regression Y ~ X (correctly shows no effect)
m_marg <- lm(Y ~ X)
summary(m_marg)

# Collider-conditioned regression Y ~ X + Z (spurious association created)
m_coll <- lm(Y ~ X + Z)
summary(m_coll)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.binomial(1, 0.5, size=n)
Y_data = np.random.binomial(1, 0.5, size=n)
pZ = np.where(X_data + Y_data > 0, 0.9, 0.2)
Z_data = np.random.binomial(1, pZ)

with pm.Model() as model_collider:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_coll = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_coll, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.bernoulli(torch.full((n, 1), 0.5))
Y_t = torch.bernoulli(torch.full((n, 1), 0.5))
pZ = torch.where(X_t + Y_t > 0, torch.tensor(0.9), torch.tensor(0.2))
Z_t = torch.bernoulli(pZ)

class ColliderRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, x_z):
        return self.linear(x_z)

model = ColliderRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Collider Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 17. Selection distortion in grant proposals

Chapter 6 and the official script simulate two independent proposal characteristics:

$$ N_i \sim \mathrm{Normal}(0,1), \quad T_i \sim \mathrm{Normal}(0,1), \quad N_i \perp\!\!\!\perp T_i $$

$$ S_i = N_i + T_i $$

Only proposals in the top 10% of total score $S_i$ are selected.

```{r}
#| label: fig-grant-proposals
#| fig-cap: "Grant proposal selection distortion: Negative correlation induced among selected proposals"
set.seed(1914)
N <- 200
p <- 0.1
nw <- rnorm(N)
tw <- rnorm(N)
s <- nw + tw
q <- quantile(s, 1 - p)
selected <- s >= q

cat("Full population correlation Cor(NW, TW):", round(cor(nw, tw), 3), "\n")
cat("Selected population correlation Cor(NW, TW | Selected):", round(cor(nw[selected], tw[selected]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(1914)
N <- 200
p <- 0.1
nw <- rnorm(N)
tw <- rnorm(N)
s <- nw + tw
q <- quantile(s, 1 - p)
selected <- s >= q

# Fit linear model on selected proposals
m_grant <- lm(tw[selected] ~ nw[selected])
summary(m_grant)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(1914)
N = 200
nw = np.random.normal(0, 1, size=N)
tw = np.random.normal(0, 1, size=N)
s = nw + tw
q = np.quantile(s, 0.9)
selected = s >= q

nw_sel = nw[selected]
tw_sel = tw[selected]

with pm.Model() as model_grant:
    a = pm.Normal("a", mu=0, sigma=1)
    b_nw = pm.Normal("b_nw", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + b_nw * nw_sel
    tw_obs = pm.Normal("tw_obs", mu=mu, sigma=sigma, observed=tw_sel)
    
    idata_grant = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_grant, var_names=["b_nw"]))
```

### Stan
```stan
data {
  int<lower=0> N_sel;
  vector[N_sel] nw_sel;
  vector[N_sel] tw_sel;
}
parameters {
  real a;
  real b_nw;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  b_nw ~ normal(0, 1);
  sigma ~ exponential(1);
  tw_sel ~ normal(a + b_nw * nw_sel, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(1914)
N = 200
nw = torch.randn(N, 1)
tw = torch.randn(N, 1)
s = nw + tw
q = torch.quantile(s, 0.9)
mask = (s >= q).squeeze()

nw_sel = nw[mask]
tw_sel = tw[mask]

linear = nn.Linear(1, 1)
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(linear.parameters(), lr=0.05)

for epoch in range(500):
    optimizer.zero_grad()
    preds = linear(nw_sel)
    loss = criterion(preds, tw_sel)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Slope b_nw:", linear.weight.item())
```
:::

---

## 18. Selection can occur before modeling

Collider conditioning does not require explicitly adding the collider as a regression predictor.

It can occur because the dataset was selected.

Examples include analyses restricted to:

- funded proposals;
- admitted students;
- hospitalized patients;
- employed workers;
- customers who opened an email;
- customers who purchased;
- cases that reached customer support;
- published studies.

The selection criterion can act as an implicit conditioned variable.

Therefore, causal analysis must model how observations entered the dataset.

---

## 19. Happiness, age, and marriage

The official script animates a happiness example in which marriage status is influenced by age and happiness.

A simplified graph is:

$$ A \rightarrow M \leftarrow H $$

where $A$ is age, $H$ is happiness, and $M$ is marriage status.

If age and happiness are otherwise independent, the collider $M$ blocks the path marginally. Conditioning on marriage status opens it.

```{r}
#| label: fig-happiness-sim
#| fig-cap: "Happiness, age, and marriage: Conditioning on marriage status distorts the age-happiness relationship"
library(rethinking)
set.seed(1977)
# Simulate population over 65 years (ages 18 to 65)
d <- sim_happiness(seed=1977, N_years=100)
d2 <- d[d$age > 17, ]
d2$A <- (d2$age - 18) / (65 - 18)
d2$mid <- d2$married + 1

m6.9 <- quap(
  alist(
    happiness ~ dnorm(mu, sigma),
    mu <- a[mid] + bA * A,
    a[mid] ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d2, start=list(a=c(0,0), bA=0, sigma=1)
)
precis(m6.9, depth=2)
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
d <- sim_happiness(seed=1977, N_years=100)
d2 <- d[d$age > 17, ]
d2$A <- (d2$age - 18) / (65 - 18)
d2$mid <- d2$married + 1

m6.9 <- quap(
  alist(
    happiness ~ dnorm(mu, sigma),
    mu <- a[mid] + bA * A,
    a[mid] ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d2, start=list(a=c(0,0), bA=0, sigma=1)
)
precis(m6.9, depth=2)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

# Assuming standardized A, H and binary M (0=unmarried, 1=married)
np.random.seed(1977)
N = 1000
A = np.random.uniform(-1, 1, size=N)
H = np.random.normal(0, 1, size=N)
# Marriage probability depends on both age and happiness
pM = 1 / (1 + np.exp(-(0.8 * A + 1.2 * H)))
M = np.random.binomial(1, pM)

with pm.Model() as model_hap:
    a = pm.Normal("a", mu=0, sigma=1, shape=2)
    bA = pm.Normal("bA", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a[M] + bA * A
    h_obs = pm.Normal("h_obs", mu=mu, sigma=sigma, observed=H)
    
    idata_hap = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_hap, var_names=["a", "bA"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  int<lower=1,upper=2> M[N];
  vector[N] A;
  vector[N] H;
}
parameters {
  vector[2] a;
  real bA;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bA ~ normal(0, 1);
  sigma ~ exponential(1);
  for (n in 1:N) {
    H[n] ~ normal(a[M[n]] + bA * A[n], sigma);
  }
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(1977)
N = 1000
A_t = torch.distributions.Uniform(-1, 1).sample((N, 1))
H_t = torch.randn(N, 1)
pM = torch.sigmoid(0.8 * A_t + 1.2 * H_t)
M_t = torch.bernoulli(pM).long()

class HappinessModel(nn.Module):
    def __init__(self):
        super().__init__()
        self.a = nn.Parameter(torch.zeros(2, 1))
        self.bA = nn.Parameter(torch.zeros(1, 1))
    def forward(self, A, M):
        return self.a[M].squeeze(-1) + A * self.bA

model = HappinessModel()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(A_t, M_t)
    loss = criterion(preds, H_t)
    loss.backward()
    optimizer.step()

print("PyTorch Intercepts a:", model.a.data.numpy().squeeze())
print("PyTorch Slope bA:", model.bA.item())
```
:::

---

## 20. The haunted DAG

A DAG is “haunted” when an unobserved cause changes the role of an observed variable.

The official `06_breen_collider_animation.R` script uses:

- $G$: grandparent education;
- $P$: parent education;
- $C$: child education;
- $U$: an unobserved neighborhood or shared influence.

$$ G \rightarrow P \rightarrow C, \quad U \rightarrow P, \quad U \rightarrow C $$

The true direct effect $G \rightarrow C$ is zero ($b_{GC} = 0$).

### 20.1 Why conditioning on parent education is dangerous

$P$ is a collider on:

$$ G \rightarrow P \leftarrow U \rightarrow C $$

Conditioning on $P$ opens this path. Even though the true direct effect of $G$ on $C$ is zero, a regression of $C$ on both $G$ and $P$ estimates a non-zero coefficient for $G$.

```{r}
#| label: fig-haunted-dag
#| fig-cap: "The Haunted DAG: Conditioning on parent education P opens backdoor through unobserved U"
library(rethinking)
set.seed(1)
N <- 200
b_GP <- 1
b_GC <- 0
b_PC <- 1
b_U <- 2

U <- 2 * rbinom(N, 1, 0.5) - 1
G <- rnorm(N)
P <- rnorm(N, b_GP * G + b_U * U)
C <- rnorm(N, b_PC * P + b_GC * G + b_U * U)
d <- data.frame(C=C, P=P, G=G, U=U)

m6.11 <- quap(
  alist(
    C ~ dnorm(mu, sigma),
    mu <- a + b_PC * P + b_GC * G,
    a ~ dnorm(0, 1),
    c(b_PC, b_GC) ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d, start=list(a=0, b_PC=1, b_GC=0, sigma=1)
)
precis(m6.11)
```

::: {.panel-tabset}
### R (rethinking)
```{r}
library(rethinking)
set.seed(1)
N <- 200
b_GP <- 1
b_GC <- 0
b_PC <- 1
b_U <- 2

U <- 2*rbern(N, 0.5) - 1
G <- rnorm(N)
P <- rnorm(N, b_GP*G + b_U*U)
C <- rnorm(N, b_PC*P + b_GC*G + b_U*U)
d <- data.frame(C=C, P=P, G=G, U=U)

m6.11 <- quap(
  alist(
    C ~ dnorm(mu, sigma),
    mu <- a + b_PC*P + b_GC*G,
    a ~ dnorm(0, 1),
    c(b_PC, b_GC) ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d, start=list(a=0, b_PC=1, b_GC=0, sigma=1)
)
precis(m6.11)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(1)
N = 200
b_GP, b_GC, b_PC, b_U = 1.0, 0.0, 1.0, 2.0

U = 2 * np.random.binomial(1, 0.5, size=N) - 1
G = np.random.normal(0, 1, size=N)
P = np.random.normal(b_GP * G + b_U * U, 1)
C = np.random.normal(b_PC * P + b_GC * G + b_U * U, 1)

with pm.Model() as model_breen:
    a = pm.Normal("a", mu=0, sigma=1)
    b_PC_param = pm.Normal("b_PC", mu=0, sigma=1)
    b_GC_param = pm.Normal("b_GC", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + b_PC_param * P + b_GC_param * G
    c_obs = pm.Normal("c_obs", mu=mu, sigma=sigma, observed=C)
    
    idata_breen = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_breen, var_names=["b_PC", "b_GC"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] G;
  vector[N] P;
  vector[N] C;
}
parameters {
  real a;
  real b_PC;
  real b_GC;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  b_PC ~ normal(0, 1);
  b_GC ~ normal(0, 1);
  sigma ~ exponential(1);
  C ~ normal(a + b_PC * P + b_GC * G, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(1)
N = 200
b_GP, b_GC, b_PC, b_U = 1.0, 0.0, 1.0, 2.0

U = 2 * torch.bernoulli(torch.full((N, 1), 0.5)) - 1
G_t = torch.randn(N, 1)
P_t = torch.randn(N, 1) + b_GP * G_t + b_U * U
C_t = torch.randn(N, 1) + b_PC * P_t + b_GC * G_t + b_U * U

class BreenModel(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, pg):
        return self.linear(pg)

model = BreenModel()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([P_t, G_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, C_t)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Weights [b_PC, b_GC]:", model.linear.weight.data.numpy())
```
:::

### 20.2 Animation logic

The script moves across narrow bands of parent education and fits the grandparent–child relationship within each band.

Within a fixed parental-education band, grandparent education carries information about the unobserved cause $U$.

The induced slope is a selection artifact, not a direct causal effect.

---

## 21. Why unobserved variables still matter

A variable need not be measured to affect identification.

If $U$ causes two observed variables, its arrows remain part of the scientific graph even when no column records $U$.

This matters because:

- an observed variable may become a collider because of $U$;
- an apparent adjustment set may fail to close a backdoor path;
- conditioning on an observed proxy may partially condition on $U$;
- no amount of sample size can repair a structurally unidentified effect.

The absence of $U$ from the data does not remove it from the world.

---

## 22. The Table 2 fallacy

### 22.1 What is the Table 2 fallacy?

The **Table 2 fallacy** (Westreich & Greenland, 2013) refers to the widespread but erroneous practice of presenting multiple regression coefficients from a single model in a summary table ("Table 2" in standard medical and social science papers) and interpreting **all** of them as causal effects.

In a multivariable regression model:

$$ Y_i = \alpha + \beta_X X_i + \beta_Z Z_i + \varepsilon_i $$

researchers often report both $\beta_X$ and $\beta_Z$, treating $\beta_X$ as the causal effect of $X$ on $Y$ and $\beta_Z$ as the causal effect of $Z$ on $Y$. 

This interpretation is fundamentally flawed because **an adjustment set tailored to identify the effect of $X$ is generally NOT valid for identifying the effect of $Z$.**

---

### 22.2 Why a single model cannot estimate multiple causal effects

Consider a classic DAG where $Z$ causes $X$, and both $Z$ and $X$ affect outcome $Y$:

```{mermaid}
%%| label: fig-table2-fallacy
%%| fig-cap: "The Table 2 Fallacy: Z affects Y both directly and indirectly through X"
flowchart LR
    Z["Z (Confounder for X, but Primary Cause of X)"] ---> X["X (Exposure of Interest)"]
    X ---> Y["Y (Outcome)"]
    Z ---> Y
```

Suppose we fit the regression model:

$$ Y = \alpha + \beta_X X + \beta_Z Z + \varepsilon $$

1. **Estimating the effect of $X$ ($\beta_X$):**
   - $Z$ is a common cause (confounder) on the backdoor path $X \leftarrow Z \rightarrow Y$.
   - Controlling for $Z$ blocks this backdoor path.
   - Therefore, $\beta_X$ validly estimates the causal effect of $X$ on $Y$.

2. **Estimating the effect of $Z$ ($\beta_Z$):**
   - $Z$ affects $Y$ through **two** pathways:
     - Direct path: $Z \rightarrow Y$
     - Indirect path (mediated): $Z \rightarrow X \rightarrow Y$
   - By including $X$ in the regression, we condition on $X$.
   - $X$ is a **mediator** on the path $Z \rightarrow X \rightarrow Y$. Controlling for $X$ blocks this pipe!
   - As a result, $\beta_Z$ estimates ONLY the **direct effect** of $Z$ on $Y$, completely omitting its indirect effect through $X$.
   - Worse, if there is unobserved confounding between $X$ and $Y$, $X$ becomes a **collider**, introducing collider bias into $\beta_Z$.

Therefore, $\beta_Z$ is **biased** as an estimate of the total causal effect of $Z$ on $Y$.

---

### 22.3 Numerical Simulation

Let's simulate a scenario where $Z$ has a total causal effect of $+3$ on $Y$ ($+2$ directly, plus $+1$ indirectly via $X$):

```{r}
#| label: fig-table2-sim
set.seed(123)
n <- 1000

# Z causes X (effect = 0.5)
Z <- rnorm(n, 0, 1)
X <- 0.5 * Z + rnorm(n, 0, 1)

# X and Z both cause Y (direct X->Y = 2, direct Z->Y = 2)
# Total effect of Z on Y = 2 + (0.5 * 2) = 3
Y <- 2 * X + 2 * Z + rnorm(n, 0, 1)

# Model 1: Y ~ X + Z (Designed to estimate effect of X)
m1 <- lm(Y ~ X + Z)
cat("--- Model 1: Y ~ X + Z ---\n")
print(summary(m1)$coefficients)

# Model 2: Y ~ Z (Designed to estimate TOTAL effect of Z)
m2 <- lm(Y ~ Z)
cat("\n--- Model 2: Y ~ Z (Total effect of Z) ---\n")
print(summary(m2)$coefficients)
```

**Output Analysis:**
* In `Model 1` (`Y ~ X + Z`), $\beta_X \approx 2.0$ (correct causal effect of $X$), but $\beta_Z \approx 2.0$ (reflects *only* the direct effect, missing the total effect of $3.0$).
* To estimate the total causal effect of $Z$, we MUST fit `Model 2` (`Y ~ Z`), where $\beta_Z \approx 3.0$.

---

### 22.4 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking & lm)
```{r}
library(rethinking)
set.seed(123)
n <- 1000

Z <- rnorm(n)
X <- 0.5 * Z + rnorm(n)
Y <- 2 * X + 2 * Z + rnorm(n)

# Model for X's effect (adjusts for confounder Z)
m_effect_X <- lm(Y ~ X + Z)

# Model for Z's total effect (does NOT adjust for mediator X)
m_effect_Z <- lm(Y ~ Z)

cat("Effect of X (from Model 1):", round(coef(m_effect_X)["X"], 2), "\n")
cat("Direct effect of Z (from Model 1 - incomplete!):", round(coef(m_effect_X)["Z"], 2), "\n")
cat("Total effect of Z (from Model 2 - correct!):", round(coef(m_effect_Z)["Z"], 2), "\n")
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.normal(0, 1, size=n)
X_data = 0.5 * Z_data + np.random.normal(0, 1, size=n)
Y_data = 2 * X_data + 2 * Z_data + np.random.normal(0, 1, size=n)

# Model 1: Y ~ X + Z
with pm.Model() as model_table2:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    idata = pm.sample(draws=1000, tune=1000, return_inferencedata=True, progressbar=False)

print(az.summary(idata, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Z;
  vector[N] Y;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.randn(n, 1)
X_t = 0.5 * Z_t + torch.randn(n, 1)
Y_t = 2 * X_t + 2 * Z_t + torch.randn(n, 1)

inputs = torch.cat([X_t, Z_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch weights [bX, bZ]:", model.weight.data.numpy())
```
:::

---

### 22.5 Summary Principles

> **Rule of Thumb:** One model = One focal estimand.
> 
> - Never present a table of multiple regression coefficients and interpret them all as causal effects.
> - A variable $Z$ included to control for confounding of $X \rightarrow Y$ cannot simultaneously have its own coefficient interpreted as the total causal effect of $Z \rightarrow Y$.
> - If you care about the causal effect of multiple variables, fit **separate models**, each with an adjustment set specifically chosen for that exposure's DAG backdoor criterion.

---

## 23. Multicollinearity as an identification warning

Chapter 6 begins with multicollinearity.

Suppose two predictors are nearly redundant:

$$ X_1 \approx X_2 $$

and the model is:

$$ Y_i = \alpha + \beta_1X_{1i} + \beta_2X_{2i} + \varepsilon_i $$

The data may identify the sum $\beta_1 + \beta_2$ more precisely than either coefficient separately.

The posterior may show strong negative covariance:

$$ \mathrm{Cov} \left( \beta_1, \beta_2 \mid D \right) < 0 $$

This is not necessarily bias. It can be an honest statement that the data cannot distinguish the separate contributions.

---

## 24. Post-treatment bias

### 24.1 What is post-treatment bias?

**Post-treatment bias** occurs when a regression model conditions on a variable $Z$ that is itself influenced by the exposure/treatment $T$ ($T \rightarrow Z$).

In causal inference, post-treatment variables are dangerous controls. Depending on the underlying DAG, conditioning on a post-treatment variable can introduce bias through **two distinct mechanisms**:

1. **Blocking the causal pipe (Mediator Blocking):** It masks the indirect causal mechanism through which the treatment operates.
2. **Opening a collider path (Collider Bias):** If the mediator and outcome share unobserved confounders, conditioning on the mediator turns it into an active collider, introducing spurious association.

---

### 24.2 Mechanism 1: Blocking the Causal Mediator

Consider the classic plant growth experiment (*Statistical Rethinking*, Section 6.2):
* $T$: Soil antifungal treatment ($T=1$ treated, $T=0$ control).
* $F$: Presence of plant fungus ($F=1$ present, $F=0$ absent).
* $H$: Final plant height.

The treatment works by reducing fungus ($T \rightarrow F$), which in turn allows plants to grow taller ($F \rightarrow H$):

```{mermaid}
%%| label: fig-post-treatment-pipe
%%| fig-cap: "Post-treatment bias: Conditioning on mediator F blocks the causal path T -> F -> H"
flowchart LR
    T["T (Treatment)"] ---> F["F (Fungus - Post-Treatment Mediator)"]
    F ---> H["H (Plant Height)"]
    T -. Direct Effect .-> H
```

If our scientific goal is to estimate the **total causal effect** of treatment $T$ on height $H$:
* In the model `lm(H ~ T)`, the coefficient $\hat{\beta}_T$ captures the total effect (both direct and indirect via fungus reduction).
* If we fit `lm(H ~ T + F)`, we condition on $F$. This holds fungus status constant across treated and untreated plants, **blocking the pipe** $T \rightarrow F \rightarrow H$. 
* Inside the $F=0$ group (fungus-free) and inside the $F=1$ group (fungus present), treatment appears to have zero effect because its primary mechanism (killing fungus) has been held fixed!

---

### 24.3 Mechanism 2: Opening a Collider Path via Unobserved Confounders

The problem becomes even worse when there is an **unobserved variable $U$** (e.g., general plant hardiness or soil quality) that affects both the mediator $F$ and the outcome $H$:

```{mermaid}
%%| label: fig-post-treatment-collider
%%| fig-cap: "Post-treatment collider bias: Conditioning on F opens the backdoor path T -> F <- U -> H"
flowchart LR
    T["T (Treatment)"] ---> F["F (Fungus)"]
    U["U (Unobserved Hardiness)"] ---> F
    U ---> H["H (Plant Height)"]
    F ---> H
```

In this DAG, $F$ is a **collider** on the path $T \rightarrow F \leftarrow U \rightarrow H$:
* Unconditioned ($F$ not in model): The path $T \rightarrow F \leftarrow U \rightarrow H$ is **blocked** by default at the collider $F$.
* Conditioned ($F$ in model `lm(H ~ T + F)`): Conditioning on $F$ **opens the collider path** between $T$ and $U$!

**Intuition:** Among plants that still have fungus ($F=1$), a treated plant ($T=1$) must have had exceptionally **low hardiness** ($U$) to contract fungus despite receiving treatment. Conversely, an untreated plant ($T=0$) with fungus might have normal hardiness. Thus, conditioning on $F$ creates a spurious negative correlation between $T$ and $U$, making treatment look **harmful** to plant height!

---

### 24.4 Numerical Simulation: Plant Fungus Experiment

Let's simulate both mechanisms in R:

```{r}
#| label: fig-post-treatment-sim
set.seed(1999)
n <- 1000

# 1. Randomized Treatment T (50% treated)
T <- rbinom(n, 1, 0.5)

# 2. Unobserved plant hardiness U
U <- rnorm(n, 0, 1)

# 3. Fungus F: Reduced by Treatment T, reduced by Hardiness U
# Probability of fungus drops with treatment and hardiness
p_fungus <- 1 / (1 + exp(- (0.5 - 2.0 * T - 1.5 * U)))
F <- rbinom(n, 1, p_fungus)

# 4. Height H: Increased by Hardiness U, reduced by Fungus F
# True total effect of T operates ENTIRELY by suppressing F
H <- 10 + 2.0 * U - 3.0 * F + rnorm(n, 0, 0.5)

# --- Model 1: Total Effect (Correct - Unconditioned on F) ---
m_total <- lm(H ~ T)
cat("=== Model 1: Total Effect lm(H ~ T) ===\n")
print(summary(m_total)$coefficients)

# --- Model 2: Post-Treatment Biased Model lm(H ~ T + F) ---
m_direct <- lm(H ~ T + F)
cat("\n=== Model 2: Post-Treatment Biased lm(H ~ T + F) ===\n")
print(summary(m_direct)$coefficients)
```

**Output Analysis:**
* **`m_total` (`H ~ T`)**: Correctly shows that treatment $T$ increases plant height by $\approx +1.8$ units by suppressing fungus.
* **`m_direct` (`H ~ T + F`)**: Conditioned on $F$, treatment $T$ appears to have a **negative** effect on height ($\approx -0.6$), because conditioning on collider $F$ induces negative selection bias via unobserved hardiness $U$.

---

### 24.5 Implementation (R, PyMC, Stan, PyTorch)

::: {.panel-tabset}
### R (rethinking & lm)
```{r}
library(rethinking)
set.seed(123)
n <- 1000

T <- rbinom(n, 1, 0.5)
U <- rnorm(n, 0, 1)
F <- rbinom(n, 1, 1 / (1 + exp(- (0.5 - 2.0 * T - 1.5 * U))))
H <- 10 + 2.0 * U - 3.0 * F + rnorm(n, 0, 0.5)

# Total effect (unconditioned on post-treatment F)
m_total <- lm(H ~ T)
cat("Total Treatment Effect estimate bT:", round(coef(m_total)["T"], 2), "\n")

# Biased post-treatment model
m_biased <- lm(H ~ T + F)
cat("Post-treatment biased estimate bT:", round(coef(m_biased)["T"], 2), "\n")
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az
from scipy.special import expit

np.random.seed(123)
n = 1000

T_data = np.random.binomial(1, 0.5, size=n)
U_data = np.random.normal(0, 1, size=n)
p_fungus = expit(0.5 - 2.0 * T_data - 1.5 * U_data)
F_data = np.random.binomial(1, p_fungus)
H_data = 10 + 2.0 * U_data - 3.0 * F_data + np.random.normal(0, 0.5, size=n)

# Model 1: Total Effect H ~ T
with pm.Model() as model_total:
    a = pm.Normal("a", mu=10, sigma=2)
    bT = pm.Normal("bT", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    mu = a + bT * T_data
    h_obs = pm.Normal("h_obs", mu=mu, sigma=sigma, observed=H_data)
    idata_total = pm.sample(draws=1000, tune=1000, progressbar=False, return_inferencedata=True)

# Model 2: Biased Model H ~ T + F
with pm.Model() as model_biased:
    a = pm.Normal("a", mu=10, sigma=2)
    bT = pm.Normal("bT", mu=0, sigma=1)
    bF = pm.Normal("bF", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    mu = a + bT * T_data + bF * F_data
    h_obs = pm.Normal("h_obs", mu=mu, sigma=sigma, observed=H_data)
    idata_biased = pm.sample(draws=1000, tune=1000, progressbar=False, return_inferencedata=True)

print("Total Effect bT:", az.summary(idata_total, var_names=["bT"])["mean"].values[0])
print("Biased Post-treatment bT:", az.summary(idata_biased, var_names=["bT"])["mean"].values[0])
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] T;
  vector[N] F;
  vector[N] H;
}
parameters {
  real a;
  real bT;
  real bF;
  real<lower=0> sigma;
}
model {
  a ~ normal(10, 2);
  bT ~ normal(0, 1);
  bF ~ normal(0, 1);
  sigma ~ exponential(1);
  H ~ normal(a + bT * T + bF * F, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000

T_t = torch.bernoulli(torch.full((n, 1), 0.5))
U_t = torch.randn(n, 1)
p_f = torch.sigmoid(0.5 - 2.0 * T_t - 1.5 * U_t)
F_t = torch.bernoulli(p_f)
H_t = 10.0 + 2.0 * U_t - 3.0 * F_t + 0.5 * torch.randn(n, 1)

# Fit biased model H ~ T + F
inputs = torch.cat([T_t, F_t], dim=1)
model = nn.Linear(2, 1)
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
criterion = nn.MSELoss()

for epoch in range(1000):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, H_t)
    loss.backward()
    optimizer.step()

print("PyTorch Biased weights [bT, bF]:", model.weight.data.numpy())
```
:::

---

### 24.6 Summary Rules for Post-Treatment Variables

> **Golden Rule of Covariate Selection:** 
> 
> 1. **Do NOT condition on post-treatment variables** when your estimand is the total causal effect of treatment.
> 2. **Check timeline & DAG position:** If a variable is measured *after* treatment or is causally downstream of treatment ($T \rightarrow Z$), exclude it from the adjustment set for total effect models.
> 3. **Mediation Analysis:** If your goal is to isolate direct vs. indirect pathways, standard linear regression with post-treatment controls is dangerous whenever unobserved mediator-outcome confounders ($U$) exist. Specialized mediation techniques (g-formula, marginal structural models) are required.

---

## 25. Adjustment sets are estimand-specific

Consider:

$$ X \rightarrow M \rightarrow Y, \quad C \rightarrow X, \quad C \rightarrow Y $$

For the total effect of $X$ on $Y$:

- adjust for $C$;
- do not adjust for $M$.

For a controlled direct effect of $X$ not through $M$:

- $M$ must be fixed or modeled appropriately;
- additional mediation assumptions are required;
- simply adding $M$ to a regression may be insufficient.

---

## 26. Regression implementation after identification

Suppose the DAG implies that $C$ is a sufficient adjustment variable for the total effect of $X$ on $Y$.

A Gaussian outcome model could be:

$$ Y_i \sim \mathrm{Normal}(\mu_i,\sigma), \quad \mu_i = \alpha + \beta_X X_i + \beta_C C_i $$

For posterior draw $s$:

$$ \mu_i^{(s)}(x) = \alpha^{(s)} + \beta_X^{(s)}x + \beta_C^{(s)}C_i $$

Average over the target population:

$$ \overline{\mu}^{(s)}(x) = \frac{1}{N} \sum_{i=1}^{N} \mu_i^{(s)}(x) $$

Then compute:

$$ \tau^{(s)} = \overline{\mu}^{(s)}(x_1) - \overline{\mu}^{(s)}(x_0) $$

---

## 27. DAG software

The book uses `dagitty` to inspect adjustment sets.

```{r}
#| label: fig-dagitty-demo
#| fig-cap: "Using dagitty in R to identify minimal adjustment sets"
library(dagitty)

g <- dagitty("dag {
  X -> Y
  X <- C -> Y
  X -> M -> Y
  X -> Z <- Y
}")

adjustmentSets(g, exposure = "X", outcome = "Y")
```

::: {.panel-tabset}
### R (dagitty)
```{r}
library(dagitty)

dag <- dagitty("dag {
  X -> Y
  X <- C -> Y
}")

adjustmentSets(
  dag,
  exposure = "X",
  outcome = "Y"
)
```

### Python (pgmpy / networkx)
```python
import networkx as nx

# Define DAG as directed graph
G = nx.DiGraph()
G.add_edges_from([("C", "X"), ("C", "Y"), ("X", "Y")])

print("Nodes:", G.nodes())
print("Edges:", G.edges())
```

### Stan
```stan
// Software DAG identification is done pre-estimation.
// Stan receives the adjustment set specified by DAG analysis.
```

### PyTorch
```python
# Software DAG analysis is performed at the computational graph / design phase.
```
:::

---

## 28. CRM application: targeting confounding

Let $A$ be retention campaign, $Y$ future revenue or churn, and $L$ observed customer risk and history.

The graph is:

$$ L \rightarrow A, \quad L \rightarrow Y, \quad A \rightarrow Y $$

The path $A \leftarrow L \rightarrow Y$ is an open backdoor path.

Conditioning on $L$ closes it under the assumption that $L$ contains all relevant common causes.

---

## 29. CRM application: campaign engagement as mediator

Suppose $A$ is campaign sent, $M$ post-campaign engagement, and $Y$ revenue.

The graph is $A \rightarrow M \rightarrow Y$, with a possible direct path $A \rightarrow Y$.

For the total campaign effect, conditioning on $M$ blocks part of the effect.

---

## 30. CRM application: openers-only analysis

Suppose email opening depends on both campaign delivery and latent digital attention:

$$ A \rightarrow O \leftarrow U \rightarrow Y $$

Analyzing only openers conditions on $O$ and opens the path $A \rightarrow O \leftarrow U \rightarrow Y$.

---

## 31. CRM application: support-contact selection

Suppose both service severity and customer persistence cause a support contact:

$$ S \rightarrow C \leftarrow P, \quad P \rightarrow Y $$

Restricting analysis to customers who contacted support conditions on $C$, opening a path between severity and persistence.

---

## 32. Implications for CausalOps

| A06 concept | CausalOps implication |
|---|---|
| Fork | Candidate backdoor path requiring adjustment |
| Pipe | Mediated causal path that should remain open for a total effect |
| Collider | Variable that should not be conditioned on by default |
| Descendant | Proxy or downstream selection variable requiring inherited-role analysis |
| Backdoor path | First-class graph object in identification reports |
| Adjustment set | Linked to one exposure, outcome, and estimand |
| D-separation | Automated graph implication engine |
| Haunted DAG | Unobserved-variable sensitivity scenarios |
| Selection distortion | Sample-construction audit |
| Table 2 fallacy | Prevent causal labels on all coefficients in one model |
| Multicollinearity | Identifiability warning rather than automatic variable deletion |
| Simulation | Verify bias under alternative conditioning sets |

### 32.1 Path audit artifact

```yaml
exposure: campaign_sent
outcome: revenue_30d
estimand: total_effect
paths:
  - path: campaign_sent <- churn_risk -> revenue_30d
    type: backdoor
    status_without_adjustment: open
    action: close
    candidate_controls:
      - churn_risk
  - path: campaign_sent -> email_opened -> revenue_30d
    type: causal_mediation
    status_without_adjustment: open
    action: preserve
  - path: campaign_sent -> email_opened <- digital_attention -> revenue_30d
    type: collider_backdoor
    status_without_adjustment: closed
    action: do_not_open
```

### 32.2 Control-role validation

Before a variable enters the model, CausalOps should record its structural role relative to exposure, outcome, and estimand.

### 32.3 Selection audit

The platform should audit filtering rules to prevent implicit collider conditioning.

---

## 33. Common errors exposed by A06

1. Adding every predictor.
2. Defining confounders from correlation.
3. Conditioning on a mediator for a total effect.
4. Conditioning on a collider.
5. Ignoring collider descendants.
6. Treating sample selection as harmless.
7. Assuming an unmeasured variable is irrelevant.
8. Interpreting every regression coefficient causally.
9. Using software to discover the scientific DAG.
10. Blocking all associations.
11. Treating multicollinearity as bias.
12. Validating only the final fitted model.

---

## 34. Lecture takeaways

1. Conditioning changes statistical paths through which information flows.
2. A fork is open until its common cause is conditioned on.
3. A pipe is open until its mediator is conditioned on.
4. A collider is closed until the collider or a descendant is conditioned on.
5. Confounder adjustment closes a noncausal backdoor path.
6. Mediator adjustment can remove part of the target total effect.
7. Collider adjustment creates a noncausal association.
8. Descendants and proxies can inherit collider problems.
9. D-separation formalizes conditional independence in a DAG.
10. A backdoor path begins with an arrow entering the exposure.
11. A valid total-effect adjustment set closes open backdoors while preserving causal paths.
12. Selection into a dataset can condition on a collider implicitly.
13. The grant-proposal example demonstrates selection distortion.
14. The happiness–marriage example shows collider bias inside strata.
15. The grandparent–parent–child example shows how an unobserved cause can haunt a DAG.
16. Regression can faithfully estimate the wrong conditioned association.
17. Multicollinearity can make separate parameters weakly identified while preserving prediction.
18. Adjustment sets are specific to an exposure, outcome, and estimand.
19. Graph software checks implications of assumptions; it does not supply the assumptions.
20. CausalOps must audit both regression controls and sample-construction filters.

---

## 35. Review questions

Below is the list of 24 review questions covered in this section:

1. [**Question 1:** What does conditioning mean in a regression or stratified analysis?](#q1)
2. [**Question 2:** Why does conditioning on the middle variable close both a fork and a pipe?](#q2)
3. [**Question 3:** Why is closing a pipe not necessarily desirable?](#q3)
4. [**Question 4:** Why is a collider closed without conditioning?](#q4)
5. [**Question 5:** Why does conditioning on a collider associate its causes?](#q5)
6. [**Question 6:** How can a descendant of a collider open a path?](#q6)
7. [**Question 7:** What is d-separation?](#q7)
8. [**Question 8:** What distinguishes a causal path from a backdoor path?](#q8)
9. [**Question 9:** What three requirements should a total-effect adjustment set satisfy?](#q9)
10. [**Question 10:** Why is confounding better understood as a path problem than a permanent variable label?](#q10)
11. [**Question 11:** What does the fork simulation show marginally and conditionally?](#q11)
12. [**Question 12:** Why does the pipe simulation lose the $X$–$Y$ association after conditioning on $Z$?](#q12)
13. [**Question 13:** Why do selected grant proposals show a negative relation between trustworthiness and newsworthiness?](#q13)
14. [**Question 14:** How can sample restriction condition on a collider without including it in a regression formula?](#q14)
15. [**Question 15:** Why can marriage status create a false age–happiness relationship?](#q15)
16. [**Question 16:** In the haunted DAG, why does conditioning on parent education bias the grandparent coefficient?](#q16)
17. [**Question 17:** Why can an unobserved variable affect identification even when it is absent from the dataset?](#q17)
18. [**Question 18:** What is the Table 2 fallacy?](#q18)
19. [**Question 19:** How does multicollinearity differ from collider bias?](#q19)
20. [**Question 20:** Why can a post-treatment control both block mediation and open a collider path?](#q20)
21. [**Question 21:** In CRM, why is an openers-only analysis vulnerable to collider bias?](#q21)
22. [**Question 22:** What sample-construction metadata should CausalOps audit?](#q22)
23. [**Question 23:** What can `dagitty` establish, and what can it not establish?](#q23)
24. [**Question 24:** How would you simulate and compare two candidate adjustment sets?](#q24)

---

### Question 1: What does conditioning mean in a regression or stratified analysis? {#q1}
**Answer:** Conditioning means evaluating comparisons within restricted subsets or holding specified variables constant. In a multivariable regression $Y = \alpha + \beta_X X + \beta_Z Z + \varepsilon$, $\beta_X$ represents the association between $X$ and $Y$ holding $Z$ fixed. In causal inference, conditioning alters the flow of information along paths in the DAG.

### Question 2: Why does conditioning on the middle variable close both a fork and a pipe? {#q2}
**Answer:** In a fork ($X \leftarrow Z \rightarrow Y$), $Z$ is the common cause generating noncausal association between $X$ and $Y$; holding $Z$ constant removes this variation, making $X \perp\!\!\!\perp Y \mid Z$. In a pipe ($X \rightarrow Z \rightarrow Y$), $Z$ is the sole intermediary conveying the effect of $X$ to $Y$; fixing $Z$ blocks the transmission of signals from $X$, making $X \perp\!\!\!\perp Y \mid Z$.

### Question 3: Why is closing a pipe not necessarily desirable? {#q3}
**Answer:** Closing a pipe blocks the causal mechanism through which $X$ affects $Y$. If the estimand of interest is the total causal effect of $X$ on $Y$, conditioning on the mediator $Z$ removes the mediated effect, leading to an underestimation or distortion of the target causal quantity.

### Question 4: Why is a collider closed without conditioning? {#q4}
**Answer:** In a collider ($X \rightarrow Z \leftarrow Y$), $Z$ is a common consequence. The two independent causes $X$ and $Y$ flow into $Z$. Without observing or restricting $Z$, knowing $X$ provides no information about $Y$ ($X \perp\!\!\!\perp Y$). Information cannot pass through a collider naturally.

### Question 5: Why does conditioning on a collider associate its causes? {#q5}
**Answer:** When $Z$ is fixed ($X \rightarrow Z \leftarrow Y$), knowing that $Z$ took a specific value means that if $X$ was low, $Y$ must have been high (or vice-versa) to produce that outcome $Z$. This induces a noncausal statistical dependence ($X \not\!\perp\!\!\!\perp Y \mid Z$), known as Berkson's bias or selection distortion.

### Question 6: How can a descendant of a collider open a path? {#q6}
**Answer:** A descendant $D$ of a collider $Z$ ($Z \rightarrow D$) carries partial information about $Z$. Conditioning on $D$ is computationally equivalent to partially conditioning on $Z$, thereby partially opening the collider path $X \rightarrow Z \leftarrow Y$.

### Question 7: What is d-separation? {#q7}
**Answer:** D-separation ("directional separation") is an algorithmic criterion that determines whether two sets of variables are conditionally independent given a conditioning set $S$ in a DAG. A path is d-separated (blocked) if it contains a conditioned fork, a conditioned pipe, or an unconditioned collider (with no conditioned descendants).

### Question 8: What distinguishes a causal path from a backdoor path? {#q8}
**Answer:** A causal path from exposure $X$ to outcome $Y$ consists entirely of arrows pointing away from $X$ towards $Y$ (e.g., $X \rightarrow M \rightarrow Y$). A backdoor path is any noncausal path connecting $X$ and $Y$ that begins with an arrow pointing *into* $X$ (e.g., $X \leftarrow C \rightarrow Y$).

### Question 9: What three requirements should a total-effect adjustment set satisfy? {#q9}
**Answer:** 
1. It must close all open backdoor paths between exposure $X$ and outcome $Y$.
2. It must not block any causal paths from $X$ to $Y$.
3. It must not open any closed paths (e.g., by conditioning on colliders or their descendants).

### Question 10: Why is confounding better understood as a path problem than a permanent variable label? {#q10}
**Answer:** Confounding is an open noncausal path between exposure and outcome. A variable is not inherently a "confounder"; its role depends on the specific exposure, outcome, estimand, and full causal DAG. The same variable can be a confounder in one query, a mediator in another, and a collider in a third.

### Question 11: What does the fork simulation show marginally and conditionally? {#q11}
**Answer:** Marginally, $X$ and $Y$ display strong correlation because both are caused by $Z$ ($X \leftarrow Z \rightarrow Y$). Conditionally on $Z$ (within $Z=0$ or $Z=1$), the correlation drops to zero, demonstrating that $Z$ accounts for all marginal association.

### Question 12: Why does the pipe simulation lose the $X$–$Y$ association after conditioning on $Z$? {#q12}
**Answer:** In $X \rightarrow Z \rightarrow Y$, $Z$ is the sole channel transmitting $X$'s effect to $Y$. Conditioning on $Z$ holds the mechanism fixed, so variation in $X$ no longer changes $Y$ within strata of $Z$.

### Question 13: Why do selected grant proposals show a negative relation between trustworthiness and newsworthiness? {#q13}
**Answer:** Proposals are selected if total score $S = \text{Newsworthiness} + \text{Trustworthiness}$ exceeds a threshold. In the selected group ($S = \text{High}$), a proposal with low newsworthiness must have high trustworthiness to be selected, inducing a negative correlation between independent traits.

### Question 14: How can sample restriction condition on a collider without including it in a regression formula? {#q14}
**Answer:** Filtering a dataset (e.g., keeping only employed individuals, admitted students, or email openers) conditions the analysis on the selection variable. If selection is caused by treatment and an unobserved outcome determinant, filtering acts as an implicit collider condition.

### Question 15: Why can marriage status create a false age–happiness relationship? {#q15}
**Answer:** If both age ($A$) and happiness ($H$) influence marriage status ($M$), then $M$ is a collider ($A \rightarrow M \leftarrow H$). Regressing happiness on age while conditioning on marriage status opens the path, producing a spurious negative slope for age among married and unmarried groups.

### Question 16: In the haunted DAG, why does conditioning on parent education bias the grandparent coefficient? {#q16}
**Answer:** Parent education $P$ is a collider between grandparent education $G$ and an unobserved neighborhood factor $U$ ($G \rightarrow P \leftarrow U \rightarrow C$). Conditioning on $P$ opens the path $G \rightarrow P \leftarrow U \rightarrow C$, creating a spurious association between $G$ and $C$ even when the direct effect $G \rightarrow C$ is zero.

### Question 17: Why can an unobserved variable affect identification even when it is absent from the dataset? {#q17}
**Answer:** Unobserved variables create causal arrows in the true graph. Unobserved common causes can make observed variables into colliders or open backdoor paths. Identification depends on the true DAG structure, not merely on what columns are present in the dataset.

### Question 18: What is the Table 2 fallacy? {#q18}
**Answer:** The Table 2 fallacy occurs when readers (or researchers) interpret every coefficient in a single multivariable regression model as a causal effect estimate. A model optimized to estimate the effect of $X$ on $Y$ may include control variables whose own coefficients are severely confounded or blocked by post-treatment bias.

### Question 19: How does multicollinearity differ from collider bias? {#q19}
**Answer:** Multicollinearity is a numerical identification issue arising from strong correlation among predictors; it increases standard errors but does not introduce systemic causal bias. Collider bias is a structural identification error caused by opening a noncausal path, which systematically distorts the point estimate.

### Question 20: Why can a post-treatment control both block mediation and open a collider path? {#q20}
**Answer:** A post-treatment variable $Z$ lies on $X \rightarrow Z \rightarrow Y$. If an unobserved $U$ also affects $Z$ and $Y$ ($Z \leftarrow U \rightarrow Y$), $Z$ is simultaneously a mediator for $X$ and a collider on $X \rightarrow Z \leftarrow U \rightarrow Y$. Conditioning on $Z$ both blocks the causal path and opens the collider path.

### Question 21: In CRM, why is an openers-only analysis vulnerable to collider bias? {#q21}
**Answer:** Opening an email ($O$) depends on campaign delivery ($A$) and customer digital engagement ($U$). Analyzing only openers conditions on the collider $O$ ($A \rightarrow O \leftarrow U \rightarrow Y$), creating a spurious negative association between campaign delivery and intrinsic engagement among openers.

### Question 22: What sample-construction metadata should CausalOps audit? {#q22}
**Answer:** CausalOps must audit: row inclusion criteria, event-trigger filters, missingness mechanics, eligibility rules, and whether any filtering condition depends on variables downstream of treatment or unobserved outcome causes.

### Question 23: What can `dagitty` establish, and what can it not establish? {#q23}
**Answer:** `dagitty` can determine all open paths, d-separations, and valid minimal adjustment sets *given* an input DAG. It cannot verify whether the input DAG correctly reflects reality.

### Question 24: How would you simulate and compare two candidate adjustment sets? {#q24}
**Answer:** Simulate data from a known structural causal model (SCM). Fit two regression models representing Adjustment Set 1 and Adjustment Set 2 across Monte Carlo trials. Compare the resulting posterior distributions of $\beta_X$ against the true structural parameter used in data generation.

---

## 36. Book reading guide

### Primary reading

- Chapter 6, *The Haunted DAG & The Causal Terror*
  - Section 6.1, *Multicollinearity*
  - Section 6.2, *Post-treatment bias*
  - Section 6.3, *Collider bias*
  - Section 6.3.2, *The haunted DAG*
  - Section 6.4, *Confronting confounding*

### Most important pages for A06

Focus on:
- the grant-proposal selection example;
- Figure 6.6 with fork, pipe, collider, and descendant;
- the path-enumeration and backdoor procedure;
- the happiness and marriage collider;
- the grandparent–parent–child haunted DAG;
- the first `dagitty` adjustment-set example.

---

## 37. Official code guide

### `05_elemental_confounds.r`
The script includes binary fork, pipe, collider, grant proposal selection distortion, happiness animation, descendant simulation, and continuous d-separation plots.

### `06_breen_collider_animation.R`
Simulates the haunted DAG $G \rightarrow P \rightarrow C$ with unobserved $U \rightarrow P$ and $U \rightarrow C$, animating regression lines across parent education bands.

### `06_DAG_animations.r`
Contains DAG animation infrastructure and complex backdoor path examples.

### `06_simulations_bad_controls.r`
Simulates post-treatment confounding, case-control bias, precision parasites, and bias amplification (previews A07).

---

## 38. Appendix — Homework A06 Question & Solutions

### Homework A06 Statement

Consider the DAG below:

$$
\begin{array}{ccc}
X & \longr\rightarrow & Y \\
\nwarrow & & \nearrow \\
 & Z & \\
 & \nwarrow & \\
 & & A \longr\rightarrow Y
\end{array}
$$

Specifically, the directed edges are:
- $X \rightarrow Y$
- $Z \rightarrow X$
- $Z \rightarrow Y$
- $A \rightarrow Z$
- $A \rightarrow Y$

**Questions:**
1. Simulate data consistent with this DAG assuming each variable is normally distributed.
2. Focus on estimating the causal effect of $X$ on $Y$. Which other variables are needed to correctly estimate the effect of $X$ on $Y$?
3. For any variables other than $X$, what is the meaning of their estimated coefficients? Are they also causal effect estimates? Why or why not?

---

### Theoretical & Graphical Path Analysis

Let's list all paths connecting exposure $X$ and outcome $Y$:

1. **Path 1:** $X \rightarrow Y$ (Direct causal path — **Open**)
2. **Path 2:** $X \leftarrow Z \rightarrow Y$ (Backdoor path through common cause $Z$ — **Open**)
3. **Path 3:** $X \leftarrow Z \leftarrow A \rightarrow Y$ (Backdoor path through mediator $Z$ and common cause $A$ — **Open**)

#### Analysis of Adjustment Sets:
- **Conditioning on $Z$:**
  - Blocks Path 2 ($X \leftarrow Z \rightarrow Y$) at common cause $Z$.
  - Blocks Path 3 ($X \leftarrow Z \leftarrow A \rightarrow Y$) at mediator $Z$.
  - Leaves the causal path $X \rightarrow Y$ unblocked.
  - Does NOT open any colliders.
  - Therefore, $\{Z\}$ is a **minimal sufficient adjustment set**.

- **Conditioning on $\{Z, A\}$:**
  - Also closes all backdoor paths. Valid adjustment set.

- **Conditioning on $\{A\}$ alone:**
  - Blocks Path 3 ($X \leftarrow Z \leftarrow A \rightarrow Y$).
  - Leaves Path 2 ($X \leftarrow Z \rightarrow Y$) **OPEN**.
  - Therefore, $\{A\}$ alone is NOT sufficient.

#### Interpretation of Coefficients (The Table 2 Fallacy):
In the model $Y \sim X + Z + A$:
- **$\beta_X$:** Represents the **total causal effect** of $X$ on $Y$.
- **$\beta_Z$:** Represents the **direct effect** of $Z$ on $Y$ holding $X$ and $A$ constant. It is **NOT** the total causal effect of $Z$ on $Y$, because the indirect causal pathway $Z \rightarrow X \rightarrow Y$ is blocked by conditioning on $X$.
- **$\beta_A$:** Represents the **direct effect** of $A$ on $Y$ holding $X$ and $Z$ constant. It is **NOT** the total causal effect of $A$ on $Y$, because the pathways $A \rightarrow Z \rightarrow Y$ and $A \rightarrow Z \rightarrow X \rightarrow Y$ are blocked by conditioning on $Z$ and $X$.

---

### Multi-Language Homework Implementation

```{r}
#| label: fig-homework-a06
#| fig-cap: "Homework A06 Simulation & Model Estimation"
library(rethinking)
library(dagitty)

set.seed(42)
N <- 1000

# Structural Causal Model Data Generation
A <- rnorm(N, 0, 1)
Z <- rnorm(N, 0.8 * A, 1)
X <- rnorm(N, 0.6 * Z, 1)
# True direct effect of X on Y = 0.5
Y <- rnorm(N, 0.5 * X + 0.7 * Z + 0.4 * A, 1)

d_hw <- data.frame(Y=Y, X=X, Z=Z, A=A)

# Minimal sufficient adjustment set model: Y ~ X + Z
m_hw_minimal <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z,
    a ~ dnorm(0, 1),
    c(bX, bZ) ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d_hw
)

# Full model: Y ~ X + Z + A
m_hw_full <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z + bA * A,
    a ~ dnorm(0, 1),
    c(bX, bZ, bA) ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=d_hw
)

cat("--- Minimal Adjustment Set Model (Y ~ X + Z) ---\n")
precis(m_hw_minimal)

cat("\n--- Full Model (Y ~ X + Z + A) ---\n")
precis(m_hw_full)
```

::: {.panel-tabset}
### R (rethinking & dagitty)
```r
library(rethinking)
library(dagitty)

# Define DAG and check adjustment sets
hw_dag <- dagitty("dag {
  X -> Y
  Z -> X
  Z -> Y
  A -> Z
  A -> Y
}")

print(adjustmentSets(hw_dag, exposure = "X", outcome = "Y"))

# Simulation
set.seed(42)
N <- 1000
A <- rnorm(N, 0, 1)
Z <- rnorm(N, 0.8 * A, 1)
X <- rnorm(N, 0.6 * Z, 1)
Y <- rnorm(N, 0.5 * X + 0.7 * Z + 0.4 * A, 1)

dat <- list(Y=Y, X=X, Z=Z, A=A)

# Model 1: Minimal adjustment set (Y ~ X + Z)
m1 <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z,
    a ~ dnorm(0, 1),
    c(bX, bZ) ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=dat
)
precis(m1)

# Model 2: Unadjusted (biased) model (Y ~ X)
m_biased <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data=dat
)
precis(m_biased)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(42)
N = 1000

A_data = np.random.normal(0, 1, size=N)
Z_data = np.random.normal(0.8 * A_data, 1)
X_data = np.random.normal(0.6 * Z_data, 1)
# True direct effect bX = 0.5, bZ = 0.7, bA = 0.4
Y_data = np.random.normal(0.5 * X_data + 0.7 * Z_data + 0.4 * A_data, 1)

with pm.Model() as hw_model:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    bA = pm.Normal("bA", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data + bA * A_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_hw = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_hw, var_names=["bX", "bZ", "bA"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
  vector[N] A;
}
parameters {
  real a;
  real bX;
  real bZ;
  real bA;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  bA ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z + bA * A, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(42)
N = 1000
A_t = torch.randn(N, 1)
Z_t = torch.randn(N, 1) + 0.8 * A_t
X_t = torch.randn(N, 1) + 0.6 * Z_t
Y_t = torch.randn(N, 1) + 0.5 * X_t + 0.7 * Z_t + 0.4 * A_t

class HomeworkRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(3, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = HomeworkRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t, A_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Weights [bX, bZ, bA]:", model.linear.weight.data.numpy())
```
:::
