---
title: "A07 — Good and Bad Controls"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: true
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# A07 — Good and Bad Controls

**Course:** Statistical Rethinking 2026  
**Section:** A — Beginner  
**Lecture:** A07 — Good and Bad Controls  
**Duration:** approximately 1 hour 1 minute  
**Official reading:** Chapter 6 of *Statistical Rethinking*, second edition  
**Secondary transition:** Chapter 7 material on overfitting, regularization, and out-of-sample evaluation  
**Central question:** Which variables should be included in a statistical model for a specific causal estimand, and how can an apparently useful control create bias or destroy precision?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official scripts `06_simulations_bad_controls.r`, `07_copernican_model.R`, and `07_overfitting_animations.r`.
>
> The official A07 title and public description emphasize beneficial and harmful controls, including confounders, mediators, and colliders. The bad-control taxonomy below follows the simulations explicitly present in `06_simulations_bad_controls.r`: post-treatment confounding, case-control or outcome-descendant bias, precision parasites, and bias amplification.
>
> The repository also contains chapter-7 Copernican and overfitting scripts immediately after these materials. They are included as a clearly marked transition because they illuminate a separate question: whether additional variables or flexibility improve out-of-sample prediction. They should not be confused with causal control selection, and this guide does not claim that every chapter-7 animation appeared in the A07 recording.

---

## Table of Contents

- [1. Main thesis](#1-main-thesis)
- [2. Relationship to the book](#2-relationship-to-the-book)
  - [2.1 Two questions that must remain separate](#21-two-questions-that-must-remain-separate)
- [3. What statistical control means](#3-what-statistical-control-means)
- [4. Control quality has two dimensions](#4-control-quality-has-two-dimensions)
  - [4.1 Bias](#41-bias)
  - [4.2 Precision](#42-precision)
- [5. A genuinely good control: an observed common cause](#5-a-genuinely-good-control-an-observed-common-cause)
  - [5.1 Good does not mean perfect](#51-good-does-not-mean-perfect)
- [6. A useful precision variable](#6-a-useful-precision-variable)
- [7. Bad control 1: post-treatment confounding](#7-bad-control-1-post-treatment-confounding)
  - [7.1 Unadjusted model](#71-unadjusted-model)
  - [7.2 Adjusted model](#72-adjusted-model)
  - [7.3 Why “post-treatment” is not the complete explanation](#73-why-post-treatment-is-not-the-complete-explanation)
  - [7.4 Practical takeaway: how to avoid this problem](#74-practical-takeaway-how-to-avoid-this-problem)
- [8. Blocking mediation and creating bias simultaneously](#8-blocking-mediation-and-creating-bias-simultaneously)
- [9. Bad control 2: case-control or outcome-descendant bias](#9-bad-control-2-case-control-or-outcome-descendant-bias)
  - [9.1 The DAG and structural mechanism ($X \rightarrow Y \rightarrow Z$)](#91-the-dag-and-structural-mechanism)
  - [9.2 Mathematical intuition: Why $\beta_X$ shrinks](#92-mathematical-intuition-why-b_x-shrinks)
  - [9.3 Connection to case-control sampling](#93-connection-to-case-control-sampling)
  - [9.4 Simulation and multi-language implementation](#94-simulation-and-multi-language-implementation)
  - [9.5 Practical takeaway: detecting and avoiding leakage](#95-practical-takeaway-detecting-and-avoiding-leakage)
- [10. Outcome descendants and leakage](#10-outcome-descendants-and-leakage)
- [11. Bad or neutral control 3: the precision parasite](#11-bad-or-neutral-control-3-the-precision-parasite)
  - [11.1 The DAG and structural mechanism ($Z \rightarrow X \rightarrow Y$)](#111-the-dag-and-structural-mechanism)
  - [11.2 Variance inflation & the math of precision loss](#112-variance-inflation--the-math-of-precision-loss)
  - [11.3 Simulation and multi-language implementation](#113-simulation-and-multi-language-implementation)
  - [11.4 Practical takeaway: when to exclude exposure-only predictors](#114-practical-takeaway-when-to-exclude-exposure-only-predictors)
- [12. Why an exposure predictor can harm precision](#12-why-an-exposure-predictor-can-harm-precision)
- [13. Bad control 4: the bias amplifier](#13-bad-control-4-the-bias-amplifier)
  - [13.1 The DAG and structural mechanism ($Z \rightarrow X \leftarrow U \rightarrow Y$)](#131-the-dag-and-structural-mechanism)
  - [13.2 Mathematical intuition: Why bias gets amplified](#132-mathematical-intuition-why-bias-gets-amplified)
  - [13.3 Instrument-like variables: IV vs OLS control](#133-instrument-like-variables-iv-vs-ols-control)
  - [13.4 Simulation and multi-language implementation](#134-simulation-and-multi-language-implementation)
  - [13.5 Practical takeaway: never control for instruments in OLS](#135-practical-takeaway-never-control-for-instruments-in-ols)
- [14. Residualization view of bias amplification](#14-residualization-view-of-bias-amplification)
- [15. Pre-treatment is not sufficient for safety](#15-pre-treatment-is-not-sufficient-for-safety)
- [16. Post-treatment is a warning, not a full taxonomy](#16-post-treatment-is-a-warning-not-a-full-taxonomy)
- [17. Randomization does not make every control safe](#17-randomization-does-not-make-every-control-safe)
- [18. The Table 2 fallacy revisited](#18-the-table-2-fallacy-revisited)
- [19. A practical control-classification table](#19-a-practical-control-classification-table)
- [20. Simulation is the decisive teaching tool](#20-simulation-is-the-decisive-teaching-tool)
- [21. General simulation template for controls](#21-general-simulation-template-for-controls)
- [22. Control selection is not feature selection](#22-control-selection-is-not-feature-selection)
- [23. Transition: the Copernican principle for model evaluation](#23-transition-the-copernican-principle-for-model-evaluation)
  - [23.1 Causal and predictive humility](#231-causal-and-predictive-humility)
- [24. Overfitting is different from bad-control bias](#24-overfitting-is-different-from-bad-control-bias)
- [25. The official polynomial overfitting animation](#25-the-official-polynomial-overfitting-animation)
- [26. Leave-one-out cross-validation](#26-leave-one-out-cross-validation)
  - [26.1 What LOO can establish](#261-what-loo-can-establish)
- [27. In-sample and out-of-sample error](#27-in-sample-and-out-of-sample-error)
- [28. Regularization](#28-regularization)
  - [28.1 Regularization is not causal repair](#281-regularization-is-not-causal-repair)
- [29. WAIC and PSIS](#29-waic-and-psis)
- [30. Why predictive model comparison cannot choose the DAG](#30-why-predictive-model-comparison-cannot-choose-the-dag)
- [31. CRM application: risk-score adjustment](#31-crm-application-risk-score-adjustment)
- [32. CRM application: email-open control](#32-crm-application-email-open-control)
- [33. CRM application: eligibility as a collider](#33-crm-application-eligibility-as-a-collider)
- [34. CRM application: channel preference as precision parasite](#34-crm-application-channel-preference-as-precision-parasite)
- [35. CRM application: instrument-like assignment rule as bias amplifier](#35-crm-application-instrument-like-assignment-rule-as-bias-amplifier)
- [36. Implications for CausalOps](#36-implications-for-causalops)
  - [36.1 Proposed control-role schema](#361-proposed-control-role-schema)
  - [36.2 Candidate-set comparison](#362-candidate-set-comparison)
  - [36.3 Separate scoreboards](#363-separate-scoreboards)
- [37. Common errors exposed by A07](#37-common-errors-exposed-by-a07)
- [38. Lecture takeaways](#38-lecture-takeaways)
- [39. Review questions](#39-review-questions)
- [40. Book reading guide](#40-book-reading-guide)
- [41. Official code guide](#41-official-code-guide)
- [42. Appendix — Homework A07 Question & Solutions](#42-appendix--homework-a07-question--solutions)
- [43. Annex — Instrumental Variables (IV) and 2SLS Structural Logic](#43-annex--instrumental-variables-iv-and-2sls-structural-logic)
  - [43.1 The Structural Causal Model (DAG)](#431-the-structural-causal-model-dag)
  - [43.2 The Wald Estimator (The Ratio)](#432-the-wald-estimator-the-ratio)
  - [43.3 Two-Stage Least Squares (2SLS)](#433-two-stage-least-squares-2sls)
  - [43.4 Critical Failure Modes in Implementation](#434-critical-failure-modes-in-implementation)

---

## 1. Main thesis

The central argument is:

> **A control variable is good or bad only relative to a causal graph, an exposure, an outcome, and an estimand.**

There is no universal rule that more controls are better.

A control can:

- remove confounding;
- improve precision;
- leave the estimand unchanged but reduce precision;
- block part of the causal effect;
- open a collider path;
- amplify bias from an unmeasured common cause;
- or redefine the target population through selection.

The workflow is:

$$ \text{causal question} \rightarrow \text{estimand} \rightarrow \text{causal DAG} \rightarrow \text{candidate control} \rightarrow \text{path analysis} \rightarrow \text{bias and precision analysis} \rightarrow \text{simulation validation} \rightarrow \text{estimation} $$

The practical lesson is:

$$ \text{predictive association with }Y \not\Rightarrow \text{valid causal control} $$

and:

$$ \text{measured before treatment} \not\Rightarrow \text{safe to control} $$

---

## 2. Relationship to the book

A07 continues Chapter 6 and transitions toward Chapter 7.

| A07 topic | Second-edition counterpart | Role |
|---|---|---|
| Post-treatment controls | Section 6.2 | Controls can block causal paths or open colliders |
| Collider and selection bias | Section 6.3 | Conditioning can create association |
| Haunted DAG | Section 6.3.2 | Unobserved causes alter observed-variable roles |
| Backdoor adjustment | Section 6.4 | Identifies controls needed for a target effect |
| Variable selection by causal role | Chapter 6 overall | Replaces automatic adjustment rules |
| Overfitting | Sections 7.1–7.2 | Additional flexibility lowers training error but may worsen prediction |
| Regularization | Sections 7.2–7.3 | Priors constrain unstable complexity |
| Cross-validation, WAIC, PSIS | Sections 7.2–7.5 | Approximate out-of-sample predictive performance |

### 2.1 Two questions that must remain separate

A control-selection question asks:

> Does including $Z$ help identify the causal effect of $X$ on $Y$?

A predictive-model question asks:

> Does including $Z$ improve predictions for new observations?

The answers can differ. A collider may improve prediction while damaging causal identification. A valid confounder may be essential for causal identification even if it barely improves predictive accuracy. A variable can be causally neutral but increase the variance of the treatment-effect estimate.

---

## 3. What statistical control means

Consider:

$$ Y_i \sim \mathrm{Normal}(\mu_i,\sigma), \quad \mu_i = \alpha + \beta_X X_i + \beta_Z Z_i $$

Including $Z$ means that $\beta_X$ is identified from variation in $X$ among units with the same or similar values of $Z$, under the model.

This is statistical control. It is not equivalent to experimental control. An experiment manipulates or holds constant aspects of the data-generating process. A regression conditions on recorded values after the data have been generated.

---

## 4. Control quality has two dimensions

A candidate control should be evaluated along at least two dimensions:

### 4.1 Bias

Does conditioning on $Z$ move the estimate toward or away from the target causal effect?

$$ \mathrm{Bias} \left( \widehat{\tau} \right) = \mathbb{E} \left[ \widehat{\tau} \right] - \tau $$

### 4.2 Precision

How variable is the estimate across repeated samples? Mean squared error combines both:

$$ \mathrm{MSE} \left( \widehat{\tau} \right) = \mathrm{Bias} \left( \widehat{\tau} \right)^2 + \mathrm{Var} \left( \widehat{\tau} \right) $$

Causal identification places bias first. Precision cannot rescue a systematically wrong estimand.

---

## 5. A genuinely good control: an observed common cause

Suppose $Z \rightarrow X$, $Z \rightarrow Y$, and $X \rightarrow Y$. The backdoor path $X \leftarrow Z \rightarrow Y$ is open without adjustment. Conditioning on $Z$ closes it.

```{r}
#| label: fig-good-control
#| fig-cap: "Good Control: Observed common cause Z removes backdoor bias"
set.seed(123)
f_good <- function(n=100, bZX=0.8, bZY=0.8, bXY=0.5) {
  Z <- rnorm(n)
  X <- rnorm(n, bZX * Z)
  Y <- rnorm(n, bXY * X + bZY * Z)
  bX_unadj <- coef(lm(Y ~ X))['X']
  bX_adj <- coef(lm(Y ~ X + Z))['X']
  return(c(bX_unadj, bX_adj))
}

sim_good <- replicate(1000, f_good())
cat("Unadjusted mean bX (Biased, True=0.5):", round(mean(sim_good[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Unbiased, True=0.5):", round(mean(sim_good[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
Z <- rnorm(n)
X <- rnorm(n, 0.8 * Z)
Y <- rnorm(n, 0.5 * X + 0.8 * Z)

# Unadjusted model (biased due to open backdoor path X <- Z -> Y)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (unbiased, backdoor path closed)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(0.8 * Z_data, 1)
Y_data = np.random.normal(0.5 * X_data + 0.8 * Z_data, 1)

with pm.Model() as model_good:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_good = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_good, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.randn(n, 1)
X_t = torch.randn(n, 1) + 0.8 * Z_t
Y_t = torch.randn(n, 1) + 0.5 * X_t + 0.8 * Z_t

class GoodControlRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = GoodControlRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Good Control Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

### 5.1 Good does not mean perfect

Adjustment can still fail when $Z$ is measured with error, important common causes remain unobserved, positivity is weak, or the conditional outcome model is misspecified.

---

## 6. A useful precision variable

Suppose $Z$ predicts $Y$ but is independent of $X$ ($Z \rightarrow Y, X \rightarrow Y$). Conditioning on $Z$ is not required to remove confounding, but it explains outcome variation unrelated to $X$, reducing residual variance and increasing precision for $\beta_X$.

```{r}
#| label: fig-precision-variable
#| fig-cap: "Useful Precision Variable: Z -> Y improves precision of bX"
set.seed(123)
f_prec <- function(n=100, bXY=0.5, bZY=1.2) {
  Z <- rnorm(n)
  X <- rnorm(n)
  Y <- rnorm(n, bXY * X + bZY * Z, sd=0.5)
  bX_unadj <- coef(lm(Y ~ X))['X']
  bX_adj <- coef(lm(Y ~ X + Z))['X']
  se_unadj <- summary(lm(Y ~ X))$coefficients['X', 'Std. Error']
  se_adj <- summary(lm(Y ~ X + Z))$coefficients['X', 'Std. Error']
  return(c(bX_unadj, se_unadj, bX_adj, se_adj))
}

sim_prec <- replicate(1000, f_prec())
cat("Unadjusted mean bX:", round(mean(sim_prec[1,]), 3), "Mean SE:", round(mean(sim_prec[2,]), 3), "\n")
cat("Adjusted for Z mean bX:", round(mean(sim_prec[3,]), 3), "Mean SE (Reduced SE!):", round(mean(sim_prec[4,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
Z <- rnorm(n)
X <- rnorm(n)
Y <- rnorm(n, 0.5 * X + 1.2 * Z, sd = 0.5)

# Unadjusted model (unbiased, but higher variance)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (unbiased, lower standard error due to precision variable Z)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(0, 1, size=n)
Y_data = np.random.normal(0.5 * X_data + 1.2 * Z_data, 0.5)

with pm.Model() as model_precision:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_precision = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_precision, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.randn(n, 1)
X_t = torch.randn(n, 1)
Y_t = 0.5 * torch.randn(n, 1) + 0.5 * X_t + 1.2 * Z_t

class PrecisionVariableRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = PrecisionVariableRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Precision Variable Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 7. Bad control 1: post-treatment confounding

The first official bad-control simulation is:

$$ X \rightarrow Z, \quad U \rightarrow Z, \quad U \rightarrow Y $$

with a possible path $Z \rightarrow Y$.

```{r}
#| label: fig-bad-control-1
#| fig-cap: "Bad Control 1: Post-treatment confounding simulation"
set.seed(123)
f_post <- function(n=100, bXZ=1, bZY=0) {
  X <- rnorm(n)
  u <- rnorm(n)
  Z <- rnorm(n, bXZ * X + u)
  Y <- rnorm(n, bZY * Z + u)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim1 <- replicate(1000, f_post(n=100, bZY=0))
cat("Unadjusted mean bX (True=0):", round(mean(sim1[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Biased):", round(mean(sim1[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
X <- rnorm(n)
U <- rnorm(n)
Z <- rnorm(n, X + U)
Y <- rnorm(n, 0 * Z + U) # True total effect of X on Y is 0

# Unadjusted model
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (opens collider X -> Z <- U -> Y)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.normal(0, 1, size=n)
U_data = np.random.normal(0, 1, size=n)
Z_data = np.random.normal(X_data + U_data, 1)
Y_data = np.random.normal(U_data, 1)

with pm.Model() as model_post:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_post = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_post, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.randn(n, 1)
U_t = torch.randn(n, 1)
Z_t = torch.randn(n, 1) + X_t + U_t
Y_t = torch.randn(n, 1) + U_t

class PostTreatmentRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = PostTreatmentRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Post-Treatment Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

### 7.1 Unadjusted model

The regression $Y \sim X$ recovers an effect near zero because $X$ is independent of $U$.

### 7.2 Adjusted model

The regression $Y \sim X + Z$ conditions on the collider $X \rightarrow Z \leftarrow U \rightarrow Y$, opening a noncausal path.

### 7.3 Why “post-treatment” is not the complete explanation

$Z$ is post-treatment, but it is harmful here because conditioning (1) may block mediation when $\beta_{ZY}\neq 0$, and (2) opens the collider path through $U$.

### 7.4 Practical takeaway: how to avoid this problem

To avoid post-treatment confounding in practice:

1. **Avoid controlling for post-treatment variables when estimating a total effect:** Never include variables measured after treatment $X$ in a total-effect regression model $Y \sim X$.
2. **Audit the DAG for post-treatment colliders:** Check whether post-treatment variable $Z$ shares unobserved causes $U$ with outcome $Y$. If $U \rightarrow Z$ and $U \rightarrow Y$, conditioning on $Z$ creates collider bias ($X \rightarrow Z \leftarrow U \rightarrow Y$).
3. **Use g-methods for direct effects under unobserved confounding:** If isolating a direct effect past mediator $Z$ is required under mediator-outcome confounding $U$, standard OLS regression fails. Use g-computation, marginal structural models, or targeted learning instead.

---

## 8. Blocking mediation and creating bias simultaneously

When $\beta_{ZY}\neq 0$, $Z$ is also a mediator ($X \rightarrow Z \rightarrow Y$). For the total effect of $X$ on $Y$, conditioning on $Z$ blocks a causal path and opens a noncausal collider path ($X \rightarrow Z \leftarrow U \rightarrow Y$).

---

## 9. Bad control 2: case-control or outcome-descendant bias

The second major bad control pattern occurs when a candidate variable $Z$ is a downstream consequence (descendant) of the outcome $Y$:

$$ X \longrightarrow Y \longrightarrow Z $$

### 9.1 The DAG and structural mechanism ($X \rightarrow Y \rightarrow Z$) {#91-the-dag-and-structural-mechanism}

In this structure:
- $X$ is the exposure or treatment of interest;
- $Y$ is the primary target outcome;
- $Z$ is a post-outcome variable caused by $Y$.

Common real-world examples include:
- **Medicine:** $X =$ New medication, $Y =$ Disease recovery, $Z =$ Post-discharge follow-up status or lab report generated after recovery.
- **E-Commerce / CRM:** $X =$ Promotional discount email, $Y =$ Customer purchase, $Z =$ Order confirmation email generated or post-purchase customer satisfaction survey.
- **Epidemiology:** $X =$ Environmental toxin, $Y =$ Cancer development, $Z =$ Hospital admission record.

Because $Z$ lies downstream of $Y$, any variation in $Z$ is driven by $Y$. Controlling for $Z$ in a regression $Y \sim X + Z$ conditions on a variable that has absorbed the variance of the outcome itself.

---

### 9.2 Mathematical intuition: Why $\beta_X$ shrinks {#92-mathematical-intuition-why-b_x-shrinks}

Consider the linear structural equations:

$$ Y = \beta_{XY} X + \varepsilon_Y $$
$$ Z = \beta_{YZ} Y + \varepsilon_Z $$

When we regress $Y$ on both $X$ and $Z$:

$$ Y = \alpha + \beta_X X + \beta_Z Z + e $$

Because $Z = \beta_{YZ} Y + \varepsilon_Z$, conditioning on $Z$ partials out the variation in $Y$ using $Z$. Since $Z$ contains $Y$ (plus noise $\varepsilon_Z$), holding $Z$ constant artificially restricts the variance of $Y$. 

Specifically:
- As the association between $Y$ and $Z$ increases ($\beta_{YZ} \to \infty$), $Z$ becomes a near-perfect proxy for $Y$.
- Controlling for $Z$ forces $Z$ to absorb the outcome variance, driving the estimated treatment effect $\widehat{\beta}_X$ toward zero (attenuation bias).
- If there is also an unobserved confounder $U$ affecting both $X$ and $Z$ ($X \leftarrow U \rightarrow Z$), conditioning on $Z$ additionally opens a collider path $X \leftarrow U \rightarrow Z \leftarrow Y$, introducing severe non-causal selection bias.

---

### 9.3 Connection to case-control sampling {#93-connection-to-case-control-sampling}

In epidemiological **case-control designs**, subjects are sampled based on their outcome status $Y$ (e.g., selecting 100 patients with disease $Y=1$ and 100 healthy controls $Y=0$).

Conditioning on an outcome descendant $Z$ in a regression is structurally equivalent to conditioning sample selection on a proxy of $Y$:

$$ X \longrightarrow Y \longrightarrow S_{\text{sample}} $$

When sample inclusion or variable adjustment depends on a proxy of $Y$, the empirical distribution $p(Y \mid X)$ in the conditioned sample no longer mirrors the true population data-generating process. This distorts odds ratios, slopes, and causal contrasts.

---

### 9.4 Simulation and multi-language implementation {#94-simulation-and-multi-language-implementation}

Below, we simulate $X \rightarrow Y \rightarrow Z$ with a true causal effect $\beta_{XY} = 1.0$ and $Y \rightarrow Z$ with $\beta_{YZ} = 1.0$.

- **Unadjusted Model ($Y \sim X$):** Correctly recovers $\widehat{\beta}_X \approx 1.0$.
- **Adjusted Model ($Y \sim X + Z$):** Heavily biased! $Z$ absorbs outcome variance, shrinking $\widehat{\beta}_X$ toward $0.5$ (or $0.0$ if $Z$ is a deterministic proxy).

```{r}
#| label: fig-bad-control-2
#| fig-cap: "Bad Control 2: Case-control / outcome-descendant bias simulation"
library(rethinking)
set.seed(123)
f_case <- function(n=100, bXY=1, bYZ=1) {
  X <- rnorm(n)
  Y <- rnorm(n, bXY * X)
  Z <- rnorm(n, bYZ * Y)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim2 <- replicate(1000, f_case())
cat("Unadjusted mean bX (True=1.0):", round(mean(sim2[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Biased, Attenuated):", round(mean(sim2[2,]), 3), "\n")

dens(sim2[1,], lwd=3, col="steelblue", xlab="Estimated bX", main="Outcome-Descendant Bias: Unadjusted vs Adjusted", xlim=c(0.2, 1.5))
dens(sim2[2,], lwd=3, col="firebrick", add=TRUE)
abline(v=1.0, lty=2, col="gray")
legend("topright", legend=c("Unadjusted (Y ~ X)", "Adjusted for Z (Y ~ X + Z)"), col=c("steelblue", "firebrick"), lwd=3)
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
X <- rnorm(n)
Y <- rnorm(n, 1 * X)
Z <- rnorm(n, 1 * Y) # Z is a descendant of outcome Y

# Unadjusted model
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (conditions on outcome descendant Z)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.normal(0, 1, size=n)
Y_data = np.random.normal(1.0 * X_data, 1)
Z_data = np.random.normal(1.0 * Y_data, 1)

with pm.Model() as model_case:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_case = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_case, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.randn(n, 1)
Y_t = torch.randn(n, 1) + 1.0 * X_t
Z_t = torch.randn(n, 1) + 1.0 * Y_t

class OutcomeDescendantRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = OutcomeDescendantRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Outcome-Descendant Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

### 9.5 Practical takeaway: detecting and avoiding leakage {#95-practical-takeaway-detecting-and-avoiding-leakage}

To avoid outcome-descendant bias in practice:

1. **Temporal auditing:** Ensure every candidate control $Z$ is measured strictly **before** the outcome $Y$ occurs. Any variable measured simultaneously with or after $Y$ is a potential outcome descendant.
2. **Prevent target leakage:** In automated ML / feature engineering pipelines, audit features to ensure no post-outcome artifacts (e.g., invoice numbers, post-treatment support logs) are fed into causal effect estimators.
3. **Keep predictive and causal feature sets separate:** A feature $Z$ that boosts $R^2$ in prediction by acting as a proxy for $Y$ must be excluded when estimating the causal effect of $X$ on $Y$.

---

## 10. Outcome descendants and leakage

In predictive modeling, an outcome descendant may dramatically improve apparent accuracy. For causal inference, it distorts the exposure coefficient. For prospective prediction, it is unavailable at scoring time.

---

## 11. Bad or neutral control 3: the precision parasite

The third major bad or neutral control pattern occurs when a variable $Z$ is a cause of exposure $X$ but has no independent direct connection to outcome $Y$:

$$ Z \longrightarrow X \longrightarrow Y $$

### 11.1 The DAG and structural mechanism ($Z \rightarrow X \rightarrow Y$) {#111-the-dag-and-structural-mechanism}

In this structure:
- $X$ is the exposure or treatment of interest;
- $Y$ is the target outcome;
- $Z$ is an exogenous predictor of $X$ (an "instrument-like" cause of exposure) that affects $Y$ *only* through $X$.

Real-world examples include:
- **Marketing / Business:** $X =$ Customer subscription upgrade, $Y =$ 12-month retention, $Z =$ Regional promotional campaign budget used to incentivize upgrades.
- **Medicine / Pharmacology:** $X =$ Statin usage, $Y =$ Cardiovascular event risk, $Z =$ Regional physician prescribing habit or drug coupon eligibility.
- **Economics / Policy:** $X =$ Years of education completed, $Y =$ Hourly wage, $Z =$ Distance to nearest college during high school.

Because $Z$ has no open backdoor path to $Y$ and does not confound $X$ and $Y$, adjusting for $Z$ is unnecessary for causal identification.

---

### 11.2 Variance inflation & the math of precision loss {#112-variance-inflation--the-math-of-precision-loss}

Controlling for $Z$ in the model $Y = \alpha + \beta_X X + \beta_Z Z$ has two distinct consequences:

1. **Bias:** $\widehat{\beta}_X$ remains **unbiased** ($\mathbb{E}[\widehat{\beta}_X] = \beta_{XY}$).
2. **Precision:** The variance of $\widehat{\beta}_X$ **increases** (standard error inflates!).

In ordinary least squares (OLS), the sampling variance of $\widehat{\beta}_X$ is given by:

$$ \mathrm{Var} \left( \widehat{\beta}_X \right) = \frac{\sigma^2} {\sum_{i=1}^N \left( X_i - \widehat{X}_i(Z) \right)^2} = \frac{\sigma^2} {\sum_{i=1}^N (X_i - \overline{X})^2 \left( 1 - R_{X \cdot Z}^2 \right)} $$

Where:
- $\sigma^2 = \mathrm{Var}(Y \mid X, Z)$ is the outcome residual variance;
- $R_{X \cdot Z}^2$ is the proportion of variance in $X$ explained by $Z$.

Because $Z$ affects $Y$ *only through $X$*, including $Z$ in the regression does **not** reduce the outcome residual variance $\sigma^2$. However, controlling for $Z$ partials out $Z$'s variation from $X$, shrinking the denominator by $(1 - R_{X \cdot Z}^2)$.

As a result, $\mathrm{Var}(\widehat{\beta}_X)$ grows wider! $Z$ acts as a **precision parasite**: it drains identifying variation from exposure $X$ without offering any outcome variance reduction in return.

---

### 11.3 Simulation and multi-language implementation {#113-simulation-and-multi-language-implementation}

Below, we simulate $Z \rightarrow X \rightarrow Y$ with $N=50$ observations across 1,000 Monte Carlo trials:
- **Unadjusted Model ($Y \sim X$):** Unbiased ($\text{mean} \approx 1.0$), narrower standard error ($\text{SD} \approx 0.10$).
- **Adjusted Model ($Y \sim X + Z$):** Unbiased ($\text{mean} \approx 1.0$), but **wider standard error** ($\text{SD} \approx 0.15$, a $+45\%$ variance expansion!).

```{r}
#| label: fig-bad-control-3
#| fig-cap: "Bad Control 3: Precision parasite simulation showing variance inflation"
library(rethinking)
set.seed(123)
f_parasite <- function(n=50, bZX=1, bXY=1) {
  Z <- rnorm(n)
  X <- rnorm(n, bZX * Z)
  Y <- rnorm(n, bXY * X)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim3 <- replicate(1000, f_parasite())
cat("Unadjusted mean bX:", round(mean(sim3[1,]), 3), "SD (Narrower):", round(sd(sim3[1,]), 3), "\n")
cat("Adjusted for Z mean bX:", round(mean(sim3[2,]), 3), "SD (Wider - Parasite!):", round(sd(sim3[2,]), 3), "\n")

# Density plot showing variance expansion
dens(sim3[1,], lwd=3, col="steelblue", xlab="Estimated bX", main="Precision Parasite: Variance Inflation", xlim=c(0.4, 1.6), ylim=c(0, 4.5))
dens(sim3[2,], lwd=3, col="firebrick", add=TRUE)
abline(v=1.0, lty=2, col="gray")
legend("topright", legend=c("Unadjusted (More precise)", "Adjusted for Z (Precision Parasite)"), col=c("steelblue", "firebrick"), lwd=3)
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 50
Z <- rnorm(n)
X <- rnorm(n, 1 * Z)
Y <- rnorm(n, 1 * X)

# Unadjusted model (more precise, lower SD)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (less precise, precision parasite Z inflates SE)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 50
Z_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(1.0 * Z_data, 1)
Y_data = np.random.normal(1.0 * X_data, 1)

with pm.Model() as model_parasite:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_parasite = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_parasite, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 50
Z_t = torch.randn(n, 1)
X_t = torch.randn(n, 1) + 1.0 * Z_t
Y_t = torch.randn(n, 1) + 1.0 * X_t

class PrecisionParasiteRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = PrecisionParasiteRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Precision Parasite Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

### 11.4 Practical takeaway: when to exclude exposure-only predictors {#114-practical-takeaway-when-to-exclude-exposure-only-predictors}

To optimize statistical precision in causal models:

1. **Exclude exposure-only causes from OLS controls:** If a pre-treatment variable $Z$ affects $X$ but has no direct link to $Y$ and does not block any backdoor paths, exclude $Z$ from ordinary regression adjustment to preserve maximum degrees of freedom and statistical power.
2. **Distinguish outcome precision predictors from precision parasites:** 
   - $Z \rightarrow Y$ (independent of $X$): **Good precision control!** Shrinks residual $\sigma^2$ and narrows $\mathrm{SE}(\widehat{\beta}_X)$.
   - $Z \rightarrow X$ (independent of $Y$): **Bad precision parasite!** Shrinks denominator $(1 - R_{X \cdot Z}^2)$ and widens $\mathrm{SE}(\widehat{\beta}_X)$.

---

## 12. Why an exposure predictor can harm precision

For a linear regression:

$$ \mathrm{Var} \left( \widehat{\beta}_X \right) \propto \frac{\sigma^2} {\sum_i \left( X_i-\widehat{X}_i(Z) \right)^2} $$

A variable $Z$ that strongly predicts $X$ without predicting $Y$ directly shrinks the denominator without shrinking outcome residual variance, increasing uncertainty.

---

## 13. Bad control 4: the bias amplifier

The fourth major bad control pattern occurs when an unobserved confounder $U$ creates baseline bias between $X$ and $Y$, and an exogenous exposure cause $Z$ is added as a control:

$$ Z \longrightarrow X \longleftarrow U \longrightarrow Y, \quad X \longrightarrow Y $$

### 13.1 The DAG and structural mechanism ($Z \rightarrow X \leftarrow U \rightarrow Y$) {#131-the-dag-and-structural-mechanism}

In this structure:
- $X$ is the exposure or treatment of interest;
- $Y$ is the target outcome;
- $U$ is an **unobserved common cause** (unmeasured confounder) of $X$ and $Y$;
- $Z$ is an exogenous variable that influences treatment assignment $X$ ($Z \rightarrow X$) with no direct link to $Y$ or $U$.

Real-world examples include:
- **Medicine / Surgery:** $X =$ Surgical procedure vs medication, $Y =$ 1-year mortality, $U =$ Unmeasured patient frailty or health status, $Z =$ Distance to regional surgical hospital.
- **CRM / Direct Marketing:** $X =$ Direct mail catalog sent, $Y =$ Customer spend, $U =$ Unmeasured intrinsic brand loyalty, $Z =$ Randomized mailing wave / batch number.
- **Policy Evaluation:** $X =$ Participation in job training program, $Y =$ Post-program earnings, $U =$ Unmeasured intrinsic motivation, $Z =$ Regional lottery ticket number or administrative assignment rule.

Because $U$ is unobserved, the unadjusted baseline regression $Y \sim X$ is already confounded. Adding $Z$ as a regression control does *not* fix $U$—instead, it dramatically **amplifies** the bias from $U$!

---

### 13.2 Mathematical intuition: Why bias gets amplified {#132-mathematical-intuition-why-bias-gets-amplified}

Consider the linear structural model:

$$ X = \beta_{ZX} Z + \gamma_U U + \varepsilon_X $$
$$ Y = \beta_{XY} X + \delta_U U + \varepsilon_Y $$

Where $\beta_{XY}$ is the true causal effect of $X$ on $Y$, and $\delta_U U$ is the unobserved confounding term.

#### 1. Unadjusted Regression ($Y \sim X$):
The unadjusted OLS slope is:

$$ \widehat{\beta}_X^{\text{unadj}} = \frac{\mathrm{Cov}(X, Y)}{\mathrm{Var}(X)} = \beta_{XY} + \delta_U \frac{\mathrm{Cov}(X, U)}{\mathrm{Var}(X)} $$

Since $\mathrm{Var}(X) = \beta_{ZX}^2 \mathrm{Var}(Z) + \gamma_U^2 \mathrm{Var}(U) + \sigma_{\varepsilon_X}^2$, the baseline bias is:

$$ \mathrm{Bias}_{\text{unadj}} = \frac{\delta_U \gamma_U \mathrm{Var}(U)}{\beta_{ZX}^2 \mathrm{Var}(Z) + \gamma_U^2 \mathrm{Var}(U) + \sigma_{\varepsilon_X}^2} $$

#### 2. Adjusted Regression ($Y \sim X + Z$):
By Frisch-Waugh-Lovell residualization, controlling for $Z$ strips out $Z$'s exogenous variation ($\beta_{ZX}^2 \mathrm{Var}(Z)$) from $X$. The residual variation in $X$ after controlling for $Z$ is:

$$ X_{\perp Z} = \gamma_U U + \varepsilon_X $$

Now, the adjusted slope is:

$$ \widehat{\beta}_X^{\text{adj}} = \beta_{XY} + \delta_U \frac{\mathrm{Cov}(X_{\perp Z}, U)}{\mathrm{Var}(X_{\perp Z})} = \beta_{XY} + \frac{\delta_U \gamma_U \mathrm{Var}(U)}{\gamma_U^2 \mathrm{Var}(U) + \sigma_{\varepsilon_X}^2} $$

#### The Bias Amplification Mechanism:
Notice that the numerator of the bias term remains unchanged ($\delta_U \gamma_U \mathrm{Var}(U)$), but the denominator has **shrunk** because $\beta_{ZX}^2 \mathrm{Var}(Z)$ was removed!

$$ | \mathrm{Bias}_{\text{adj}} | > | \mathrm{Bias}_{\text{unadj}} | $$

> **Intuitive Poison Analogy:**  
> Think of treatment $X$ as a fluid composed of three parts: "Clean Signal" ($Z$), "Noise" ($\varepsilon_X$), and "Poison" ($U$). The baseline regression is confounded because of the poison $U$.  
> When you control for $Z$, you remove the clean signal $Z$ from $X$. The remaining fluid ($X_{\perp Z}$) now contains a **much higher concentration of poison ($U$)** relative to total variance. The resulting regression estimate of $X$ becomes *more severely biased* than if $Z$ had been left completely unadjusted!

---

### 13.3 Instrument-like variables: IV vs OLS control {#133-instrument-like-variables-iv-vs-ols-control}

$Z$ in this DAG satisfies the formal topological definition of an **Instrumental Variable (IV)**:
1. $Z \rightarrow X$ (Relevance)
2. $Z \perp\!\!\!\perp U$ (Exogenous assignment)
3. $Z \perp\!\!\!\perp Y \mid (X, U)$ (Exclusion restriction)

The crucial lesson is: **How you use $Z$ matters completely!**

- **Correct Use (Instrumental Variables / 2SLS):**  
  Use $Z$ to isolate exogenous variation in $X$ via ratio $\widehat{\beta}_{\text{IV}} = \frac{\mathrm{Cov}(Z, Y)}{\mathrm{Cov}(Z, X)}$, which **eliminates** confounding bias from $U$!
- **Incorrect Use (Ordinary Least Squares Control):**  
  Add $Z$ as an ordinary covariate control in $Y \sim X + Z$, which **amplifies** confounding bias from $U$!

---

### 13.4 Simulation and multi-language implementation {#134-simulation-and-multi-language-implementation}

Below, we simulate $Z \rightarrow X \leftarrow U \rightarrow Y$ with true effect $\beta_{XY} = 0.0$ and unobserved confounding from $U$:
- **Unadjusted Model ($Y \sim X$):** Confounded by $U$ ($\text{mean } \widehat{\beta}_X \approx 0.33$).
- **Adjusted Model ($Y \sim X + Z$):** **Bias Amplified!** ($\text{mean } \widehat{\beta}_X \approx 0.50$, a $+50\%$ increase in bias!).

```{r}
#| label: fig-bad-control-4
#| fig-cap: "Bad Control 4: Bias amplifier simulation showing amplified unobserved confounding"
library(rethinking)
set.seed(123)
f_amp <- function(n=100, bZX=1, bXY=0) {
  Z <- rnorm(n)
  u <- rnorm(n)
  X <- rnorm(n, bZX * Z + u)
  Y <- rnorm(n, bXY * X + u) # True effect bXY = 0
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim4 <- replicate(1000, f_amp(bXY=0))
cat("Unadjusted mean bX (Confounded by U, True=0.0):", round(mean(sim4[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Bias Amplified!):", round(mean(sim4[2,]), 3), "\n")

# Density plot showing bias amplification
dens(sim4[1,], lwd=3, col="steelblue", xlab="Estimated bX", main="Bias Amplifier: Unadjusted vs Adjusted Bias", xlim=c(-0.2, 0.9), ylim=c(0, 5.5))
dens(sim4[2,], lwd=3, col="firebrick", add=TRUE)
abline(v=0.0, lty=2, col="gray")
legend("topright", legend=c("Unadjusted (Baseline Bias from U)", "Adjusted for Z (Amplified Bias)"), col=c("steelblue", "firebrick"), lwd=3)
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
Z <- rnorm(n)
U <- rnorm(n)
X <- rnorm(n, 1 * Z + U)
Y <- rnorm(n, 0 * X + U) # True total effect is 0

# Unadjusted model (shows baseline bias from U)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model for Z (amplifies bias from U!)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.normal(0, 1, size=n)
U_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(1.0 * Z_data + U_data, 1)
Y_data = np.random.normal(U_data, 1)

with pm.Model() as model_amp:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_amp = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_amp, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.randn(n, 1)
U_t = torch.randn(n, 1)
X_t = torch.randn(n, 1) + 1.0 * Z_t + U_t
Y_t = torch.randn(n, 1) + U_t

class BiasAmplifierRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = BiasAmplifierRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Bias Amplifier Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

### 13.5 Practical takeaway: never control for instruments in OLS {#135-practical-takeaway-never-control-for-instruments-in-ols}

To avoid bias amplification in practice:

1. **Never include instrument-like variables as ordinary OLS regression controls:** If $Z$ affects $X$ but not $Y$ directly, including $Z$ in $Y \sim X + Z$ will worsen unobserved confounding bias from $U$.
2. **Use IV / 2SLS estimation instead of regression adjustment:** If $Z$ is a valid instrument, estimate the effect via 2SLS or structural Bayesian modeling, rather than adding $Z$ as a linear covariate control.
3. **Audit assignment mechanisms:** Administrative lottery numbers, distance to facilities, or random promotional batch IDs should be recognized as instruments, not standard regression control variables.

---

## 14. Residualization view of bias amplification

The bias-amplifying nature of an instrument-like control $Z$ ($Z \rightarrow X \leftarrow U \rightarrow Y$) can be understood deeply through the geometric lens of the **Frisch-Waugh-Lovell (FWL) theorem**.

By the FWL theorem, fitting the multivariable OLS model $Y = \alpha + \beta_X X + \beta_Z Z + \varepsilon$ is algebraically identical to estimating a simple univariate regression of the residual outcome $Y_{\perp Z}$ on the residual exposure $X_{\perp Z}$:

$$\widehat{\beta}_X = \frac{\mathrm{Cov}(Y_{\perp Z}, X_{\perp Z})}{\mathrm{Var}(X_{\perp Z})}$$

Let the structural equations for exposure $X$ and outcome $Y$ be:

$$X = \gamma Z + \delta U + \varepsilon_X$$
$$Y = \tau X + \lambda U + \varepsilon_Y$$

When we residualize $X$ on $Z$, we purge the portion of $X$'s variation explained by $Z$:

$$X_{\perp Z} = X - \widehat{\mathbb{E}}[X \mid Z] = \delta U + \varepsilon_X$$

Notice what has happened geometrically:
1. **Exogenous variation removed**: The clean, variation-inducing term $\gamma Z$ has been completely stripped away from $X$.
2. **Confounding fraction concentrated**: The residual variation $X_{\perp Z}$ is now entirely composed of unobserved confounding noise $\delta U$ and idiosyncratic error $\varepsilon_X$.

Substituting $X_{\perp Z}$ into the estimator yields the expected value of $\widehat{\beta}_X$:

$$\mathbb{E}[\widehat{\beta}_X \mid Z] = \tau + \frac{\mathrm{Cov}(\lambda U + \varepsilon_Y, \delta U + \varepsilon_X)}{\mathrm{Var}(X_{\perp Z})} = \tau + \frac{\lambda \delta \, \mathrm{Var}(U)}{\mathrm{Var}(X)(1 - R^2_{X \cdot Z})}$$

Because $0 < 1 - R^2_{X \cdot Z} < 1$, dividing the original unadjusted confounding bias by $(1 - R^2_{X \cdot Z})$ strictly increases the magnitude of the bias. Residualization acts as a **variance compressor on clean variation**, leaving the treatment variable dominated by latent bias.

---

## 15. Pre-treatment is not sufficient for safety

A widespread heuristic in observational research states: *"If a variable $Z$ is measured before treatment $X$, it is safe to include as a control variable."* This temporal heuristic is dangerously incomplete.

While temporal ordering rules out post-treatment mediation ($X \rightarrow Z$), pre-treatment timing alone provides **zero guarantee** against causal bias or precision loss.

### Three Pre-Treatment Hazards

1. **Pre-treatment Colliders (M-Bias)**:
   A variable $Z$ measured before $X$ can be a collider between two unobserved pre-treatment variables $U_1$ and $U_2$:
   $$U_1 \rightarrow Z \leftarrow U_2, \quad U_1 \rightarrow Y, \quad U_2 \rightarrow X$$
   Conditioning on pre-treatment variable $Z$ opens the noncausal backdoor path $X \leftarrow U_2 \leftrightarrow Z \leftrightarrow U_1 \rightarrow Y$, inducing M-bias where none existed in the unadjusted model.

2. **Pre-treatment Instruments (Bias Amplifiers)**:
   A pre-treatment variable $Z$ (such as random lottery eligibility or policy change timing) determined before $X$ satisfies $Z \rightarrow X \leftarrow U \rightarrow Y$. Adjusting for $Z$ in OLS amplifies pre-existing confounding bias from $U$.

3. **Pre-treatment Precision Parasites**:
   A pre-treatment variable $Z$ causing $X$ ($Z \rightarrow X \rightarrow Y$) without affecting $Y$ directly shrinks $\mathrm{Var}(X \mid Z)$, inflating parameter variance without closing any backdoor path.

> [!WARNING]
> **Takeaway**: Causal safety depends entirely on **graphical topology (DAG structural paths)**, never on temporal measurement order alone.

---

## 16. Post-treatment is a warning, not a full taxonomy

The standard rule *"Never adjust for post-treatment variables"* is a valuable heuristic for beginners because post-treatment variables are frequently mediators ($X \rightarrow Z \rightarrow Y$), colliders ($X \rightarrow Z \leftarrow U \rightarrow Y$), or outcome descendants ($X \rightarrow Y \rightarrow Z$).

However, viewing "post-treatment" as a complete taxonomy creates confusion in longitudinal and multi-stage settings.

### Longitudinal Causal Modeling & Time-Varying Confounding

In long-term studies, variables measured at time $t_2$ are post-treatment relative to $X_1$, but pre-treatment relative to $X_2$:

$$X_1 \longrightarrow L_2 \longrightarrow X_2 \longrightarrow Y$$
$$X_1 \longrightarrow Y, \quad L_2 \longrightarrow Y, \quad U \longrightarrow L_2 \text{ and } U \longrightarrow Y$$

Here, $L_2$ is a time-varying confounder affected by prior treatment $X_1$. Standard linear regression faces a fundamental catch-22:
- **If you adjust for $L_2$**: You block the causal effect of $X_1$ mediated through $L_2$, and you open a collider path if $L_2$ shares unobserved confounders $U$ with $Y$.
- **If you omit $L_2$**: You leave the causal effect of $X_2$ on $Y$ confounded by $L_2$.

### Beyond Ordinary Regression: G-Methods

Resolving post-treatment complexity requires specialized causal inference frameworks designed by James Robins and colleagues, known as **g-methods**:
- **G-computation formula** (standardization across time-varying counterfactuals)
- **Marginal Structural Models (MSMs)** with Inverse Probability Weighting (IPW)
- **G-estimation** of Structural Nested Models

Ordinary multivariable regression cannot resolve time-varying confounding affected by prior treatment.

---

## 17. Randomization does not make every control safe

Randomized Controlled Trials (RCTs) are the gold standard of causal inference because physical randomization breaks all incoming arrows to treatment $X$, guaranteeing baseline unconfoundedness ($X \perp\!\!\!\perp U$).

In an unadjusted RCT analysis, the difference in sample means is an unbiased estimator of the Average Treatment Effect (ATE):

$$\widehat{\text{ATE}}_{\text{raw}} = \overline{Y}_{X=1} - \overline{Y}_{X=0}$$

However, researchers frequently fit adjusted regressions on RCT data to improve precision or account for compliance. **Post-randomization adjustments can destroy experimental unconfoundedness.**

### The Post-Randomization Collider Trap

Suppose $Z$ is a variable measured after randomization (e.g., treatment adherence, patient side-effects, or post-baseline biomarker level). Unobserved patient frailty $U$ affects both $Z$ and final health outcome $Y$:

$$X \longrightarrow Z \longleftarrow U \longrightarrow Y$$

```
    X (Randomized Treatment)
    │
    ▼
    Z (Post-treatment Adherence / Side-effect) ◄─── U (Unobserved Frailty)
    │                                              │
    └──────────────────────────────────────────────┴──► Y (Health Outcome)
```

In the unadjusted model $Y \sim X$, $X$ is independent of $U$ because $X$ was randomized. The estimator $\widehat{\beta}_X$ is completely unbiased.

If the researcher fits the adjusted model $Y \sim X + Z$:
1. $Z$ is a collider between randomized treatment $X$ and unobserved frailty $U$.
2. Conditioning on $Z$ opens the path $X \rightarrow Z \leftarrow U \rightarrow Y$.
3. Within strata of $Z$, treatment $X$ and unobserved frailty $U$ become conditionally correlated ($\mathrm{Cov}(X, U \mid Z) \neq 0$).

> [!CAUTION]
> **Critical Lesson**: Adjusting for post-randomization variables re-introduces confounding into a perfectly randomized experiment!

---

## 18. The Table 2 fallacy revisited

In epidemiological and social science literature, multivariable regression results are often presented in a standard format known as "Table 2", listing estimated coefficients $\widehat{\beta}_1, \widehat{\beta}_2, \dots, \widehat{\beta}_k$ for the primary exposure along with all control variables.

Readers (and authors) frequently fall into the **Table 2 Fallacy** (Westreich & Greenland, 2013): assuming that *every* coefficient reported in a single multivariable regression table can be interpreted as a causal effect.

### Mathematical & Causal Proof of the Fallacy

Consider the DAG where $Z$ is a common cause of exposure $X$ and outcome $Y$, but $Z$ itself is confounded by an unobserved variable $W$:

$$W \longrightarrow Z \longrightarrow X \longrightarrow Y$$
$$W \longrightarrow Y$$

```
       W (Unobserved Social Status)
      ╱ ╲
     ▼   ▼
     Z ─► X ─► Y
```

To estimate the causal effect of $X$ on $Y$, we fit:

$$Y = \alpha + \beta_X X + \beta_Z Z + \varepsilon$$

- **For exposure $X$**: Conditioning on $Z$ blocks the backdoor path $X \leftarrow Z \leftarrow W \rightarrow Y$. Therefore, $\widehat{\beta}_X$ is an **unbiased estimate** of the causal effect of $X$ on $Y$.
- **For control $Z$**: The coefficient $\widehat{\beta}_Z$ does **NOT** estimate the causal effect of $Z$ on $Y$. The backdoor path from $Z$ to $Y$ via $W$ ($Z \leftarrow W \rightarrow Y$) remains open because $W$ is unobserved! Furthermore, $X$ lies on the causal pathway from $Z$ to $Y$ ($Z \rightarrow X \rightarrow Y$), so controlling for $X$ blocks part of $Z$'s total causal effect.

### Reporting Guidelines

| Variable in Model | Causal Interpretation of Coefficient $\widehat{\beta}$ | Valid? |
|---|---|---|
| Primary Exposure $X$ | Causal effect of $X$ on $Y$ (given $Z$ blocks backdoors for $X$) | **Valid** |
| Control Variable $Z$ | Causal effect of $Z$ on $Y$ | **INVALID (Table 2 Fallacy)** |

> [!IMPORTANT]
> **Takeaway**: A single statistical model can only be optimized for **ONE** causal estimand at a time. Evaluating the causal effect of $Z$ requires a separate model with its own DAG-derived adjustment set.

---

## 19. A practical control-classification table

The table below provides a comprehensive taxonomy of candidate control roles, summarizing their structural impact on causal bias, statistical precision, and default operational recommendations.

| Candidate Variable Role | Structural DAG Pattern | Impact on Bias (Total Effect $X \rightarrow Y$) | Impact on Precision ($\mathrm{Var}(\widehat{\beta}_X)$) | Recommended Action |
|---|---|---|---|---|
| **Common Cause (Confounder)** | $X \leftarrow Z \rightarrow Y$ | **Removes backdoor bias** | May increase or decrease variance | **MUST ADJUST** |
| **Outcome-Only Cause** | $Z \rightarrow Y \quad (Z \perp\!\!\!\perp X)$ | Neutral (no bias) | **Improves precision** (reduces $\sigma^2$) | **ADJUST** (Precision Booster) |
| **Exposure-Only Cause (No Hidden Confounding)** | $Z \rightarrow X \rightarrow Y$ | Neutral (no bias) | **Reduces precision** (VIF inflation) | **EXCLUDE** (Precision Parasite) |
| **Exposure-Only Cause (Hidden $X$-$Y$ Confounding)** | $Z \rightarrow X \leftarrow U \rightarrow Y$ | **AMPLIFIES BIASED ERROR** | **Reduces precision** | **NEVER ADJUST** in OLS (Use IV/2SLS) |
| **Mediator** | $X \rightarrow Z \rightarrow Y$ | **Blocks total causal effect** | Varies | **EXCLUDE** for Total Effect |
| **Collider** | $X \rightarrow Z \leftarrow Y$ | **Opens noncausal path** (Induces bias) | May artificially shrink SE | **NEVER ADJUST** |
| **Mediator-Confounder Collider** | $X \rightarrow Z \leftarrow U \rightarrow Y$ | **Opens backdoor path via $U$** | Varies | **NEVER ADJUST** in OLS |
| **Outcome Descendant** | $X \rightarrow Y \rightarrow Z$ | **Distorts target estimand** (Selection bias) | Appears highly predictive | **NEVER ADJUST** |
| **Pre-Treatment M-Collider** | $U_1 \rightarrow Z \leftarrow U_2$ | **Opens M-bias path** | Varies | **EXCLUDE** unless $U_1$ or $U_2$ observed |

---

## 20. Simulation is the decisive teaching tool

Observational datasets in the real world never reveal their true underlying causal parameters $\tau$, nor do they allow us to observe latent confounders $U$. Consequently, we cannot verify whether an observational model is unbiased simply by looking at real-world regression output.

**Monte Carlo simulation from known Structural Causal Models (SCMs)** is the decisive tool for learning, teaching, and validating causal control selection:

1. **Ground Truth Benchmark**: In a simulation, we explicitly define the true causal parameter (e.g., $\tau = 0.5$) and the structural equations governing $U, Z, X, Y$.
2. **Empirical Evaluation**: By generating $R = 1,000$ independent datasets from the DAG, we fit competing regression specifications and empirically compute:
   - **Empirical Bias**: $\mathrm{Bias}(\widehat{\beta}_X) = \frac{1}{R}\sum_{r=1}^R \widehat{\beta}_{X,r} - \tau$
   - **Empirical Variance**: $\mathrm{Var}(\widehat{\beta}_X) = \frac{1}{R-1}\sum_{r=1}^R (\widehat{\beta}_{X,r} - \overline{\widehat{\beta}}_X)^2$
   - **Root Mean Squared Error (RMSE)**: $\mathrm{RMSE} = \sqrt{\mathrm{Bias}^2 + \mathrm{Var}(\widehat{\beta}_X)}$
   - **95% Interval Coverage**: Proportion of replicates where the 95% interval contains true $\tau$.

Simulation converts abstract theoretical claims about backdoor paths, colliders, and precision parasites into concrete, empirical numbers.

---

## 21. General simulation template for controls

Below is a standardized, production-ready R simulation template designed to compare candidate control sets against any proposed DAG ground truth.

```r
# ==============================================================================
# General Monte Carlo Simulation Template for Evaluating Control Sets
# ==============================================================================
suppressPackageStartupMessages({
  library(rethinking)
})

simulate_control_benchmark <- function(n_sim = 1000, N = 200, true_tau = 0.5) {
  
  # Storage for estimates
  results <- data.frame(
    unadjusted = numeric(n_sim),
    good_control = numeric(n_sim),
    bad_control = numeric(n_sim)
  )
  
  set.seed(2026)
  for (i in 1:n_sim) {
    # 1. Structural Data Generating Process (SCM)
    U <- rnorm(N, 0, 1)                  # Unobserved Confounder
    Z <- rnorm(N, 0, 1)                  # Candidate Control (e.g., Collider or Instrument)
    X <- rnorm(N, 0.6 * Z + 0.5 * U, 1)  # Exposure
    Y <- rnorm(N, true_tau * X + 0.7 * U, 1) # Outcome
    
    # 2. Fit Candidate Adjustment Sets
    m_unadjusted   <- lm(Y ~ X)
    m_good_control <- lm(Y ~ X + U)     # Ideal adjustment if U observed
    m_bad_control  <- lm(Y ~ X + Z)     # Candidate model
    
    # 3. Store Point Estimates
    results$unadjusted[i]   <- coef(m_unadjusted)["X"]
    results$good_control[i] <- coef(m_good_control)["X"]
    results$bad_control[i]  <- coef(m_bad_control)["X"]
  }
  
  # 4. Compute Benchmark Metrics
  eval_metrics <- function(vec, target) {
    bias <- mean(vec) - target
    sd_val <- sd(vec)
    rmse <- sqrt(bias^2 + sd_val^2)
    c(Mean = mean(vec), Bias = bias, SD = sd_val, RMSE = rmse)
  }
  
  summary_table <- rbind(
    Unadjusted   = eval_metrics(results$unadjusted, true_tau),
    Good_Control = eval_metrics(results$good_control, true_tau),
    Bad_Control  = eval_metrics(results$bad_control, true_tau)
  )
  
  return(round(summary_table, 3))
}

# Run Benchmark
simulate_control_benchmark(n_sim = 1000, N = 200, true_tau = 0.5)
```

---

## 22. Control selection is not feature selection

A major source of confusion in modern data science is conflating **causal control selection** with **predictive feature selection** (e.g., Stepwise regression, LASSO, Random Forest feature importance, or cross-validated RMSE optimization).

The two tasks have fundamentally opposing objectives:

```
┌─────────────────────────────────────────────────────────────────────────────┐
│                            PREDICTIVE GOAL                                  │
│  "Which covariates Z_k minimize out-of-sample error E[(Y - Y_hat)^2]?"     │
└─────────────────────────────────────────────────────────────────────────────┘
                                   VS
┌─────────────────────────────────────────────────────────────────────────────┐
│                              CAUSAL GOAL                                    │
│  "Which adjustment set Z_k identifies the structural parameter dY/dX?"     │
└─────────────────────────────────────────────────────────────────────────────┘
```

### Why Predictive Selection Fails Causally

1. **Predictive selection LOVES Colliders**: In $X \rightarrow Z \leftarrow Y$, collider $Z$ contains strong joint information about $X$ and $Y$. Including $Z$ drastically improves $R^2$ and lowers out-of-sample prediction error for $Y$, but it **destroys causal identification** by opening a spurious noncausal path.
2. **Predictive selection LOVES Mediators**: In $X \rightarrow Z \rightarrow Y$, mediator $Z$ is highly predictive of $Y$. Including $Z$ improves prediction but **blocks the total causal effect** of $X$.
3. **Predictive selection DISLIKES Precision Boosters**: An independent cause of $X$ ($Z \rightarrow X$) might be selected by LASSO because it predicts $X$, even though including it in a model for $Y$ acts as a **precision parasite**.

> [!IMPORTANT]
> **Takeaway**: Cross-validation and predictive loss criteria **cannot select a causal DAG**. Variable inclusion must be justified by structural causal theory, not predictive accuracy metrics.

---

## 23. Transition: the Copernican principle for model evaluation

Having established the principles of causal control selection (Chapter 6 of *Statistical Rethinking*), we transition to model evaluation, out-of-sample prediction, and overfitting (Chapter 7 material).

The **Copernican Principle** in statistics states:

> **The observed sample is not the privileged center of the statistical universe; it is merely one random realization drawn from an unobserved, population-level Data Generating Process.**

When we fit a model to a dataset of size $N$, the model parameter estimates $\widehat{\boldsymbol{\theta}}$ adapt to two distinct signals:
1. **Structural Signal**: The true underlying population relationship $f(X)$.
2. **Sample Noise**: The accidental, unrepeatable statistical fluctuations specific to those $N$ sample rows.

A model that fits sample noise too aggressively achieves stellar in-sample metrics ($R^2 \rightarrow 1$, $\mathrm{MSE}_{\text{in}} \rightarrow 0$), but fails catastrophically when deployed to predict unseen out-of-sample data.

---

## 24. Overfitting is different from bad-control bias

To build clear modeling intuition, we must distinguish between two fundamentally different types of estimation error:

| Dimension | Structural Causal Bias (Bad-Control Bias) | Overfitting (Sampling Variance Error) |
|---|---|---|
| **Root Cause** | DAG misspecification (conditioning on colliders, mediators, instruments) | Fitting sample noise due to excess model capacity relative to sample size $N$ |
| **Asymptotic Behavior** | **Persists as $N \rightarrow \infty$** (Asymptotic identification error) | **Disappears as $N \rightarrow \infty$** (Variance shrinks to 0) |
| **Remedy** | Correct DAG analysis, backdoor criterion, IV/g-methods | Regularizing priors, cross-validation (LOO), information criteria (WAIC/PSIS) |
| **Predictive Signal** | Can sometimes *improve* predictive fit (e.g., colliders) | Degrades out-of-sample predictive generalization |

> [!WARNING]
> Regularization and cross-validation can cure **overfitting**, but they can **NEVER cure bad-control bias**.

---

## 25. The official polynomial overfitting animation

The code chunk below demonstrates polynomial overfitting on a small dataset ($N=15$). As model degree increases from linear ($\text{degree}=1$) to cubic ($\text{degree}=3$), in-sample Mean Squared Error (MSE) strictly decreases. However, high-degree polynomials oscillate wildly between data points, degrading out-of-sample prediction.

```{r}
#| label: fig-overfitting-demo
#| fig-cap: "Polynomial overfitting demo: High-degree polynomials overfit in-sample"
set.seed(123)
N <- 15
x <- seq(-2, 2, length.out=N)
y <- rnorm(N, 0.5 * x, 0.5)

m_poly1 <- lm(y ~ x)
m_poly3 <- lm(y ~ poly(x, 3))
cat("Linear In-Sample MSE:", round(mean(residuals(m_poly1)^2), 3), "\n")
cat("Polynomial Deg 3 In-Sample MSE:", round(mean(residuals(m_poly3)^2), 3), "\n")
```

As shown in the output, the degree-3 polynomial achieves a lower in-sample MSE than the linear model. Yet, the linear model represents the true underlying data-generating process ($y = 0.5x + \varepsilon$), making it vastly superior for out-of-sample generalization.

---

## 26. Leave-one-out cross-validation

**Leave-One-Out Cross-Validation (LOO-CV)** evaluates a probabilistic model's out-of-sample predictive generalization without requiring a separate test dataset.

### Mathematical Estimand

LOO-CV estimates the **expected log pointwise predictive density ($\mathrm{elpd}_{\mathrm{LOO}}$)** for out-of-sample data:

$$\mathrm{elpd}_{\mathrm{LOO}} = \sum_{i=1}^N \log p(y_i \mid D_{-i}) = \sum_{i=1}^N \log \int p(y_i \mid \boldsymbol{\theta}) p(\boldsymbol{\theta} \mid D_{-i}) d\boldsymbol{\theta}$$

where $D_{-i}$ represents the dataset with observation $i$ removed, and $p(\boldsymbol{\theta} \mid D_{-i})$ is the posterior distribution trained on the remaining $N-1$ observations.

### Computational Efficiency via PSIS-LOO

Refitting a Bayesian MCMC model $N$ times is computationally prohibitive for large datasets. Modern Bayesian software uses **Pareto-Smoothed Importance Sampling (PSIS-LOO)** to approximate $\mathrm{elpd}_{\mathrm{LOO}}$ from a single posterior fit on the full dataset:

1. **Raw Importance Weights**: $w_i^s = \frac{1}{p(y_i \mid \boldsymbol{\theta}^s)}$ for posterior draw $s \in \{1, \dots, S\}$.
2. **Pareto Smoothing**: Extreme importance weights in the upper tail are fitted to a Generalized Pareto distribution to stabilize variance.
3. **Pareto $k$ Diagnostic**:
   - $k \le 0.5$: **Good** — Importance sampling estimate is reliable.
   - $0.5 < k \le 0.7$: **Ok** — Acceptable estimation reliability.
   - $k > 0.7$: **High Risk** — Observation $i$ is an extreme outlier / highly influential point. The importance sampling variance is infinite; model $D_{-i}$ must be explicitly refitted.

---

## 27. In-sample and out-of-sample error

As model flexibility (number of parameters or polynomial degree) increases, in-sample error and out-of-sample error diverge sharply.

```
  Deviance / Error
     ▲
     │       /  Out-of-Sample Error (U-Shape)
     │      /
     │     ╱      Minimum Out-of-Sample Error
     │    ╱            ▼
     │   │          .  .  .
     │   │        .         .
     │   │       .           \
     │───┼───────┼────────────┼──────────────────────────►
     │   │      │              \  In-Sample Error (Monotonic Drop)
     │   └──────┴───────────────┴────────
     └────────────────────────────────────────► Model Flexibility (Parameters)
```

### Deviance Dynamics

- **In-Sample Deviance** ($D_{\text{in}} = -2 \sum_{i=1}^N \log p(y_i \mid \widehat{\boldsymbol{\theta}})$): Decreases monotonically as parameters are added. A model with $N$ parameters can achieve $D_{\text{in}} = 0$ by interpolating every data point.
- **Out-of-Sample Deviance** ($D_{\text{out}}$): Follows a classic **U-shaped curve**:
  1. *Underfitting*: High $D_{\text{in}}$ and high $D_{\text{out}}$ (model lacks capacity to capture structural signal).
  2. *Optimal Balance*: Minimum $D_{\text{out}}$ (model captures structural signal while ignoring noise).
  3. *Overfitting*: Low $D_{\text{in}}$ but surging $D_{\text{out}}$ (model fits sample noise).

---

## 28. Regularization

**Regularization** prevents overfitting by placing weakly informative or skepticism-encoding priors on parameters, constraining model capacity without discarding variables.

### Prior Shrinkage Mechanics

Consider a linear regression with flat priors vs regularizing Gaussian priors:

$$\text{Flat Prior: } \beta_k \sim \mathrm{Uniform}(-\infty, \infty) \quad \Longrightarrow \quad \text{Fits sample noise freely}$$
$$\text{Regularizing Prior: } \beta_k \sim \mathrm{Normal}(0, 0.5) \quad \Longrightarrow \quad \text{Shrinks } \beta_k \text{ toward zero}$$

In Bayesian posterior estimation, a Gaussian prior centered at zero adds a quadratic penalty $\frac{\beta_k^2}{2\sigma^2}$ to the log posterior (equivalent to L2 Ridge Regularization in frequentist optimization):

$$\log p(\boldsymbol{\theta} \mid D) = \sum_{i=1}^N \log p(y_i \mid \boldsymbol{\theta}) - \sum_{k} \frac{\beta_k^2}{2\sigma_\beta^2} + \text{const}$$

### What Regularization Can and Cannot Do

- **CAN DO**: Shrinks noisy estimates toward zero, reduces parameter variance, and drastically improves out-of-sample prediction ($D_{\text{out}}$).
- **CANNOT DO**: **Regularization does not repair a bad DAG.** If a model includes a collider or mediator, regularizing priors shrink the magnitude of the biased parameter, but they **cannot remove structural bias**.

---

## 29. WAIC and PSIS

Information criteria provide analytical approximations of out-of-sample predictive performance without explicit data splitting.

### Widely Applicable Information Criterion (WAIC)

WAIC is a fully Bayesian information criterion that computes out-of-sample performance on a log-pointwise scale:

$$\mathrm{WAIC} = -2 \left( \mathrm{lppd} - p_{\mathrm{WAIC}} \right)$$

where:
- $\mathrm{lppd} = \sum_{i=1}^N \log \left( \frac{1}{S} \sum_{s=1}^S p(y_i \mid \boldsymbol{\theta}^s) \right)$ is the **log pointwise predictive density**.
- $p_{\mathrm{WAIC}} = \sum_{i=1}^N \mathrm{Var}_{\boldsymbol{\theta}} \left( \log p(y_i \mid \boldsymbol{\theta}) \right)$ is the **effective number of parameters** (measuring model flexibility/overfitting risk).

### PSIS vs WAIC Comparison

| Feature | WAIC | PSIS-LOO |
|---|---|---|
| **Theoretical Foundation** | Asymptotic approximation of LOO | Importance-sampling evaluation of LOO |
| **Outlier Diagnostic** | Sensitive to sample variance | **Pareto $k$ diagnostic** for each data row |
| **Robustness** | Fails with strong outliers ($p_{\mathrm{WAIC}} > 8$) | Smooths extreme weights using Pareto distributions |
| **Recommended Usage** | Secondary historical check | **Primary modern standard** (R `loo`, Python `ArviZ`) |

---

## 30. Why predictive model comparison cannot choose the DAG

A common fatal misconception in applied data science is attempting to use model selection metrics (LOO, WAIC, $R^2$, or cross-validation) to "discover" or select the correct causal DAG.

### The Predictive Selection Paradox

Suppose we are evaluating two candidate models for outcome $Y$:
- **Model A (Causally Identified)**: Adjusts only for true confounder $C$. ($Y \sim X + C$)
- **Model B (Collider-Biased)**: Adjusts for $C$ AND a post-treatment collider $Z_{\text{collider}}$. ($Y \sim X + C + Z_{\text{collider}}$)

Because collider $Z_{\text{collider}}$ ($X \rightarrow Z_{\text{collider}} \leftarrow Y$) shares statistical information with outcome $Y$, including $Z_{\text{collider}}$ in Model B drastically increases the log pointwise predictive density $\mathrm{lppd}$.

```
  Model Specification          WAIC / LOO Score       Causal Effect Estimate (beta_X)
  ─────────────────────────────────────────────────────────────────────────────────────
  Model A (Correct DAG)        WAIC = 450.2           beta_X = 0.50 (UNBIASED)
  Model B (Collider Included)  WAIC = 390.1 (WINS!)   beta_X = 0.88 (BIASED)
```

WAIC and LOO will decisively select **Model B** as superior because Model B predicts $Y$ more accurately!

> [!CAUTION]
> **Fundamental Law of Causal Inference**: Predictive criteria (LOO, WAIC, AIC, $R^2$) reward information leakage and collider bias. **They measure predictive forecasting accuracy, not causal identification.** The validity of a control set can ONLY be established through DAG structural theory.

---

## 31. CRM application: risk-score adjustment

In Customer Relationship Management (CRM) and subscription analytics, marketing teams frequently use customer risk or churn scores to target retention campaigns $A$.

### Pre-Campaign vs Post-Campaign Risk Scores

Understanding temporal measurement relative to treatment assignment is critical:

```
  Baseline Data            Treatment                Post-Treatment Outcome
  ─────────────            ─────────                ──────────────────────
  Pre-Campaign Risk R_0 ──► Campaign Send A ───────► Post-Campaign Risk R_1
  │                         │                        │
  └─────────────────────────┼────────────────────────┴──► Churn / Revenue Y
                            ▼
```

1. **Pre-Campaign Baseline Risk Score ($R_0$)**:
   Calculated *prior* to campaign launch. High risk score $R_0$ increases probability of receiving treatment $A$ ($R_0 \rightarrow A$) and independently increases baseline churn $Y$ ($R_0 \rightarrow Y$).
   - **Role**: $R_0$ is a classic **common cause (confounder)** ($A \leftarrow R_0 \rightarrow Y$).
   - **Action**: **MUST ADJUST for $R_0$** to block the baseline backdoor path.

2. **Post-Campaign Risk Score ($R_1$)**:
   Calculated *after* campaign deployment. Treatment $A$ reduces post-campaign risk ($A \rightarrow R_1$), and $R_1$ shares unobserved customer sentiment $U$ with churn ($R_1 \leftarrow U \rightarrow Y$).
   - **Role**: $R_1$ is both a **mediator** and a **collider** ($A \rightarrow R_1 \leftarrow U \rightarrow Y$).
   - **Action**: **NEVER ADJUST for $R_1$** in ordinary regression. Adjusting for $R_1$ blocks campaign mediation and opens a collider backdoor path.

---

## 32. CRM application: email-open control

A ubiquitous mistake in digital marketing analytics is attempting to measure campaign effectiveness by controlling for whether a customer opened an email.

### The Marketing Pitch vs Causal Reality

- **Intuitive Pitch**: *"If a customer didn't open the email ($O=0$), the campaign couldn't possibly have influenced their purchase behavior ($Y$). Therefore, we should control for email opening or restrict our analysis to openers."*
- **Causal DAG Structure**:

$$A \text{ (Email Send)} \longrightarrow O \text{ (Email Open)} \longleftarrow U \text{ (Brand Affinity)} \longrightarrow Y \text{ (Purchase)}$$

```
    A (Email Send Campaign)
    │
    ▼
    O (Email Open Status) ◄─── U (Unobserved Brand Affinity / Engagement)
    │                          │
    └──────────────────────────┴──► Y (Customer Purchase)
```

### The Double Causal Violation

Adjusting for email open status $O$ in $Y \sim A + O$ damages causal identification in two distinct ways:

1. **Mediation Blocking**: Email opening $O$ is a primary mechanism through which campaign send $A$ drives purchase $Y$. Controlling for $O$ zeroes out the mediated causal effect.
2. **Collider Bias**: Email open $O$ is a collider between campaign assignment $A$ and unobserved brand affinity $U$. Conditioning on $O$ opens the noncausal path $A \rightarrow O \leftarrow U \rightarrow Y$.

> [!CAUTION]
> **Marketing Trap**: Comparing openers to non-openers makes the campaign look artificially effective because openers are self-selected high-affinity customers ($U$). To estimate the true Intent-to-Treat (ITT) effect of the campaign, analyze **all target recipients ($A=1$ vs $A=0$) without controlling for email opens**.

---

## 33. CRM application: eligibility as a collider

Enterprise CRM systems frequently enforce business eligibility rules $E \in \{0, 1\}$ before launching promotional offers.

### Structural Mechanism of Selection Bias

Suppose a customer qualifies for a campaign ($E=1$) if they satisfy *either* high historical spend ($V$) OR high digital engagement ($R$):

$$V \text{ (Historical Spend)} \longrightarrow E \text{ (Eligibility Rule)} \longleftarrow R \text{ (Digital Engagement)}$$

```
    V (Historical Spend) ───┐
                            ▼
                            E (Campaign Eligibility)
                            ▲
    R (Digital Engagement) ──┘
```

Eligibility $E$ is a structural **collider** between spend $V$ and engagement $R$.

### Within-Eligible Collider Bias

If an analyst restricts their evaluation dataset exclusively to eligible customers ($E=1$):
- They condition on collider $E$.
- Within the subset $E=1$, knowing a customer has low digital engagement ($R$) guarantees they must have high historical spend ($V$) to have qualified.
- Conditioning on $E$ induces a **spurious negative correlation** between $V$ and $R$ among eligible customers.

> [!WARNING]
> **Operational Rule**: Evaluative models should be fit across the full customer population, or eligibility mechanisms must be explicitly modeled as selection nodes.

---

## 34. CRM application: channel preference as precision parasite

In multi-channel marketing, companies record customer channel preferences $Z$ (e.g., SMS vs Email vs Push preference).

### Structural Mechanism

Channel preference $Z$ determines which delivery channel $A$ is selected for a campaign, but channel preference $Z$ has no direct causal impact on final purchase $Y$ independent of receiving the campaign:

$$Z \text{ (Channel Preference)} \longrightarrow A \text{ (Campaign Delivery)} \longrightarrow Y \text{ (Purchase)}$$

### Precision Parasite Effect

If a data scientist includes channel preference $Z$ as a control variable in $Y \sim A + Z$:
- **Bias**: Neutral (no backdoor path exists through $Z$).
- **Precision Loss**: $Z$ explains a large fraction of variance in treatment $A$, increasing the Variance Inflation Factor ($\mathrm{VIF} = \frac{1}{1 - R^2_{A \cdot Z}}$).
- **Consequence**: The standard error $\mathrm{SE}(\widehat{\beta}_A)$ expands, making it harder to establish statistical significance for a genuinely effective campaign.

---

## 35. CRM application: instrument-like assignment rule as bias amplifier

Automated CRM platforms often use algorithmic targeting scores $Z$ (e.g., propensity scores or randomized regional rollout waves) to assign campaigns.

### Structural DAG

$$Z \text{ (Algorithmic Assignment Score)} \longrightarrow A \text{ (Campaign Treatment)} \longleftarrow U \text{ (Unobserved Customer Need)} \longrightarrow Y \text{ (Conversion)}$$

```
    Z (Assignment Score / Shock)
    │
    ▼
    A (Campaign Treatment) ◄─── U (Unobserved Need / Intent)
    │                           │
    └───────────────────────────┴──► Y (Conversion / Revenue)
```

Targeting score $Z$ acts as an **instrument**: it influences treatment $A$, but shares no direct relationship with conversion $Y$ except through $A$.

### The Bias Amplifier Trap in OLS

If a data scientist includes assignment score $Z$ as an ordinary control covariate in OLS ($Y \sim A + Z$):
1. Controlling for $Z$ strips away the exogenous, clean variation in treatment $A$ driven by algorithm rule $Z$.
2. The remaining variation in treatment $A_{\perp Z}$ is now heavily dominated by unobserved customer need $U$.
3. The confounding bias $\frac{\mathrm{Cov}(U, Y)}{\mathrm{Var}(A \mid Z)}$ is **amplified**, producing a severely distorted estimate of campaign ROI.

### Correct Action: Two-Stage Least Squares (2SLS)

Rather than including $Z$ as an ordinary control in OLS, $Z$ must be used as an **Instrumental Variable** in a 2SLS pipeline:

$$\widehat{\beta}_{IV} = \frac{\mathrm{Cov}(Y, Z)}{\mathrm{Cov}(A, Z)}$$

Using $Z$ as an instrument purges the unobserved confounding $U$, yielding an unbiased estimate of the true campaign effect.

---

## 36. Implications for CausalOps

CausalOps must record structural roles, compare candidate adjustment sets via simulation, and maintain separate causal and predictive scoreboards.

---

## 37. Common errors exposed by A07

1. Adding every pre-treatment variable.
2. Selecting controls by p-value or feature importance.
3. Controlling for mediators or outcome descendants.
4. Using cross-validation to pick causal controls.
5. Assuming regularization repairs a bad DAG.

---

## 38. Lecture takeaways

1. Controls must be chosen relative to a DAG and estimand.
2. Controls affect both bias and precision.
3. Overfitting and causal bias are distinct problems requiring distinct scoreboards.

---

## 39. Review questions

Below is the list of 25 review questions covered in this section:

1. [**Question 1:** Why can a control be unbiased but harmful?](#q1)
2. [**Question 2:** What makes a common cause a good control?](#q2)
3. [**Question 3:** How can a cause of the outcome only improve precision?](#q3)
4. [**Question 4:** In the post-treatment simulation, why is the unadjusted estimate valid when $\beta_{ZY}=0$?](#q4)
5. [**Question 5:** Which collider path is opened by adjusting for $Z$?](#q5)
6. [**Question 6:** Why does a mediator control fail to estimate the total effect?](#q6)
7. [**Question 7:** Why is an adjusted mediator coefficient not automatically a direct effect?](#q7)
8. [**Question 8:** What is outcome-descendant bias?](#q8)
9. [**Question 9:** How does outcome-dependent sampling resemble conditioning on $Z$ in $X \rightarrow Y \rightarrow Z$?](#q9)
10. [**Question 10:** Why does a precision parasite increase coefficient variance?](#q10)
11. [**Question 11:** How does residualizing $X$ on an exposure predictor reduce identifying variation?](#q11)
12. [**Question 12:** What is a bias amplifier?](#q12)
13. [**Question 13:** Why can controlling for an instrument-like variable strengthen hidden confounding?](#q13)
14. [**Question 14:** Why is pre-treatment timing insufficient to identify good controls?](#q14)
15. [**Question 15:** How can randomization be undermined by post-treatment adjustment?](#q15)
16. [**Question 16:** What is the Table 2 fallacy?](#q16)
17. [**Question 17:** Which metrics should be compared in a simulation of candidate control sets?](#q17)
18. [**Question 18:** Why can a collider improve prediction while damaging causal estimation?](#q18)
19. [**Question 19:** What does leave-one-out cross-validation estimate?](#q19)
20. [**Question 20:** Why is overfitting different from causal bias?](#q20)
21. [**Question 21:** How does regularization affect polynomial curves?](#q21)
22. [**Question 22:** Why can WAIC or PSIS not select a causal DAG?](#q22)
23. [**Question 23:** In CRM, why is email opening usually a bad control for the total send effect?](#q23)
24. [**Question 24:** How can eligibility create collider selection?](#q24)
25. [**Question 25:** Why should CausalOps maintain separate causal and predictive scoreboards?](#q25)

---

### Question 1: Why can a control be unbiased but harmful? {#q1}
**Answer:** A control variable can leave the point estimate unbiased for the true causal effect while simultaneously increasing the estimator's variance (e.g., a precision parasite $Z \rightarrow X \rightarrow Y$). By absorbing exposure variation without reducing outcome residual variance, it degrades precision.

### Question 2: What makes a common cause a good control? {#q2}
**Answer:** A common cause $Z$ ($X \leftarrow Z \rightarrow Y$) generates a noncausal backdoor path between exposure $X$ and outcome $Y$. Conditioning on $Z$ blocks this backdoor path, identifying the true causal effect.

### Question 3: How can a cause of the outcome only improve precision? {#q3}
**Answer:** A variable $Z$ that causes $Y$ ($Z \rightarrow Y$) independently of $X$ ($Z \perp\!\!\!\perp X$) does not lie on any backdoor path between $X$ and $Y$. Including $Z$ in the model reduces outcome residual variance $\sigma^2$, thereby decreasing $\mathrm{Var}(\widehat{\beta}_X)$ and improving precision.

### Question 4: In the post-treatment simulation, why is the unadjusted estimate valid when $\beta_{ZY}=0$? {#q4}
**Answer:** When $\beta_{ZY}=0$, treatment $X$ affects $Z$, but $Z$ has no effect on $Y$. Unadjusted $X$ is independent of unobserved $U$ ($X \perp\!\!\!\perp U$), so $Y \sim X$ yields an unbiased estimate of zero.

### Question 5: Which collider path is opened by adjusting for $Z$? {#q5}
**Answer:** In $X \rightarrow Z \leftarrow U \rightarrow Y$, $Z$ is a collider between exposure $X$ and unobserved cause $U$. Conditioning on $Z$ opens the path $X \rightarrow Z \leftarrow U \rightarrow Y$, creating a spurious association between $X$ and $Y$.

### Question 6: Why does a mediator control fail to estimate the total effect? {#q6}
**Answer:** In $X \rightarrow Z \rightarrow Y$, $Z$ mediates $X$'s effect on $Y$. Conditioning on $Z$ blocks the mediated causal path, estimating only the direct effect (if unconfounded) rather than the total effect.

### Question 7: Why is an adjusted mediator coefficient not automatically a direct effect? {#q7}
**Answer:** If there is unobserved mediator-outcome confounding ($Z \leftarrow U \rightarrow Y$), conditioning on mediator $Z$ opens a collider path $X \rightarrow Z \leftarrow U \rightarrow Y$, biasing the direct effect estimate.

### Question 8: What is outcome-descendant bias? {#q8}
**Answer:** Conditioning on a variable $Z$ caused by the outcome ($X \rightarrow Y \rightarrow Z$) conditions on realized outcome variation, distorting the estimated effect of $X$ on $Y$.

### Question 9: How does outcome-dependent sampling resemble conditioning on $Z$ in $X \rightarrow Y \rightarrow Z$? {#q9}
**Answer:** Selecting rows into a sample based on outcome status $Y$ or its proxy $Z$ holds $Z$ constant or restricts its range, which acts as implicit conditioning on $Z$.

### Question 10: Why does a precision parasite increase coefficient variance? {#q10}
**Answer:** A precision parasite $Z$ ($Z \rightarrow X \rightarrow Y$) explains variation in $X$ without explaining independent variation in $Y$. Residualizing $X$ on $Z$ shrinks $\sum (X_i - \widehat{X}_i)^2$, increasing $\mathrm{Var}(\widehat{\beta}_X)$.

### Question 11: How does residualizing $X$ on an exposure predictor reduce identifying variation? {#q11}
**Answer:** Conditioning on an exposure cause removes the portion of $X$'s variation explained by that cause, leaving only the remaining residual variation in $X$ to estimate $\beta_X$.

### Question 12: What is a bias amplifier? {#q12}
**Answer:** A bias amplifier is an exposure predictor $Z$ ($Z \rightarrow X$) added to a model where $X$ and $Y$ share an unobserved confounder $U$ ($X \leftarrow U \rightarrow Y$). Adjusting for $Z$ removes exogenous variation in $X$, leaving $X$ more heavily dominated by $U$, which amplifies the confounding bias.

### Question 13: Why can controlling for an instrument-like variable strengthen hidden confounding? {#q13}
**Answer:** An instrument-like variable $Z$ affects $X$ but not $Y$ directly. If $X$ and $Y$ share hidden confounding $U$, adjusting for $Z$ acts as a bias amplifier.

### Question 14: Why is pre-treatment timing insufficient to identify good controls? {#q14}
**Answer:** Pre-treatment variables can be colliders of earlier latent variables, instruments, or precision parasites. Temporal ordering rules out post-treatment mediation but does not guarantee backdoor closure or lack of bias.

### Question 15: How can randomization be undermined by post-treatment adjustment? {#q15}
**Answer:** Randomization ensures $X \perp\!\!\!\perp U$ baseline unconfoundedness. However, adjusting for a post-treatment collider $Z$ ($X \rightarrow Z \leftarrow U \rightarrow Y$) opens $X \rightarrow Z \leftarrow U \rightarrow Y$, re-introducing confounding in the randomized experiment.

### Question 16: What is the Table 2 fallacy? {#q16}
**Answer:** The Table 2 fallacy is the mistaken assumption that all coefficients in a multivariable regression table can be interpreted as causal effects. Each coefficient requires a distinct adjustment set.

### Question 17: Which metrics should be compared in a simulation of candidate control sets? {#q17}
**Answer:** Candidate control sets should be compared using Monte Carlo bias, variance/SD, Root Mean Squared Error (RMSE), 95% interval coverage, and sign error rate.

### Question 18: Why can a collider improve prediction while damaging causal estimation? {#q18}
**Answer:** A collider $Z$ contains information about latent causes $U$ of outcome $Y$, helping predict $Y$. However, conditioning on $Z$ opens a noncausal path, biasing the causal estimate of $X$.

### Question 19: What does leave-one-out cross-validation estimate? {#q19}
**Answer:** LOO cross-validation estimates out-of-sample expected log predictive density ($\mathrm{elpd}_{\mathrm{LOO}}$), measuring predictive generalization to unseen data from the same data-generating distribution.

### Question 20: Why is overfitting different from causal bias? {#q20}
**Answer:** Overfitting is a sample variance problem where a model fits random noise. Causal bias is a structural identification problem where a model estimates the wrong causal estimand regardless of sample size.

### Question 21: How does regularization affect polynomial curves? {#q21}
**Answer:** Tight regularizing priors shrink higher-order polynomial coefficients toward zero, preventing extreme curve oscillations and reducing out-of-sample prediction error.

### Question 22: Why can WAIC or PSIS not select a causal DAG? {#q22}
**Answer:** WAIC and PSIS evaluate out-of-sample predictive accuracy, not causal identification. A model with collider bias or post-treatment controls may achieve better WAIC/PSIS scores than the causally correct model.

### Question 23: In CRM, why is email opening usually a bad control for the total send effect? {#q23}
**Answer:** Email opening $O$ is a post-treatment mediator of campaign $A$, and opening is often influenced by unobserved digital attention $U$ ($A \rightarrow O \leftarrow U \rightarrow Y$). Adjusting for $O$ both blocks campaign mediation and opens a collider backdoor path.

### Question 24: How can eligibility create collider selection? {#q24}
**Answer:** If campaign eligibility $E$ depends on both customer value $V$ and responsiveness $R$ ($V \rightarrow E \leftarrow R$), restricting analysis to eligible customers conditions on collider $E$, inducing a negative correlation between value and responsiveness.

### Question 25: Why should CausalOps maintain separate causal and predictive scoreboards? {#q25}
**Answer:** Causal scoreboards evaluate identification, backdoor closure, and causal simulation bias. Predictive scoreboards evaluate out-of-sample forecasting accuracy (LOO/PSIS). Conflating the two can lead to selecting predictively strong but causally biased models.

---

## 40. Book reading guide

### Primary reading
- Chapter 6, *The Haunted DAG & The Causal Terror* (Sections 6.2–6.4)

### Secondary transition
- Chapter 7, *Ulysses' Compass* (Sections 7.1–7.5)

---

## 41. Official code guide

### `06_simulations_bad_controls.r`
Simulates post-treatment confounding, outcome-descendant bias, precision parasites, and bias amplification across Monte Carlo trials.

### `07_copernican_model.R` & `07_overfitting_animations.r`
Animates Copernican predictive decentering, polynomial overfitting, LOO cross-validation, and regularizing priors.

---

## 42. Appendix — Homework A07 Question & Solutions

### Homework A07 Statement

This homework is based on the data in `data(foxes)`: 116 foxes from 30 different urban groups in England.

Variables:
- `area` ($A$): size of territory
- `avgfood` ($F$): average food in territory
- `groupsize` ($G$): number of foxes in territory
- `weight` ($W$): weight of each fox

**Assumed DAG:**

$$
\begin{array}{ccc}
 & A & \\
 & \downarrow & \\
F & \longrightarrow & G \\
\searrow & & \swarrow \\
 & W & 
\end{array}
$$

Specifically, directed edges:
- $A \rightarrow F$
- $F \rightarrow G$
- $F \rightarrow W$
- $G \rightarrow W$

**Questions:**
1. Use the backdoor criterion to estimate the **total causal influence** of avgfood ($F$) on weight ($W$). What is the minimal adjustment set?
2. What effect would increasing the food have on the weight of foxes inside it?
3. Can you explain the result?

---

### Theoretical & Graphical Path Analysis

Let's list all paths connecting exposure $F$ (avgfood) and outcome $W$ (weight):

1. **Path 1:** $F \rightarrow W$ (Direct causal path — **Open**)
2. **Path 2:** $F \rightarrow G \rightarrow W$ (Indirect causal path through mediator $G$ — **Open**)
3. **Path 3:** $F \leftarrow A$ (Backdoor path? No, $A$ is an upstream cause of $F$. $A$ has no other outgoing edge to $W$ or $G$. It is a dead end.)

#### Backdoor Criterion Analysis:
- There are **no open backdoor paths** between $F$ and $W$.
- Therefore, the **minimal sufficient adjustment set** for the total causal effect of $F$ on $W$ is the **empty set** $\emptyset$.
- Model for total effect: $W \sim F$.

#### What happens if we adjust for $G$ (groupsize)?
- $G$ is a mediator on the path $F \rightarrow G \rightarrow W$.
- Conditioning on $G$ blocks the indirect causal path $F \rightarrow G \rightarrow W$.
- Model $W \sim F + G$ estimates the **controlled direct effect** of $F$ on $W$, holding group size fixed.

#### Theoretical Explanation of the Result:
- **Total Effect ($W \sim F$):** Virtually **zero** ($\beta_F \approx 0$). Increasing food ($F$) increases territory group size ($G$) ($F \rightarrow G$), which spreads the food among more foxes, reducing weight per fox ($G \rightarrow W$ negative effect). The direct positive effect of food ($F \rightarrow W$) and the indirect negative effect ($F \rightarrow G \rightarrow W$) almost perfectly cancel out!
- **Direct Effect ($W \sim F + G$):** Strongly **positive** ($\beta_F > 0$). When group size $G$ is held fixed, more food directly leads to heavier foxes!

---

### Multi-Language Homework Implementation

```{r}
#| label: fig-homework-a07
#| fig-cap: "Homework A07: Foxes Total Effect vs Direct Effect Models"
library(rethinking)
library(dagitty)

data(foxes)
d <- foxes
d$F <- standardize(d$avgfood)
d$W <- standardize(d$weight)
d$G <- standardize(d$groupsize)
d$A <- standardize(d$area)

# Model 1: Total Effect of F on W (Minimal adjustment set = empty set)
m_total <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)

# Model 2: Direct Effect of F on W (Adjusting for mediator G)
m_direct <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F + bG * G,
    a ~ dnorm(0, 0.2),
    c(bF, bG) ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)

cat("--- Total Causal Effect of Food on Weight (Minimal set = empty) ---\n")
precis(m_total)

cat("\n--- Direct Causal Effect of Food on Weight (Adjusting for Group Size G) ---\n")
precis(m_direct)
```

::: {.panel-tabset}
### R (rethinking & dagitty)
```r
library(rethinking)
library(dagitty)

# Define Fox DAG
fox_dag <- dagitty("dag {
  A -> F
  F -> G
  F -> W
  G -> W
}")

print(adjustmentSets(fox_dag, exposure = "F", outcome = "W"))

data(foxes)
d <- foxes
d$F <- standardize(d$avgfood)
d$W <- standardize(d$weight)
d$G <- standardize(d$groupsize)

# Model 1: Total Effect (Minimal Set = empty)
m_total <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)
precis(m_total)

# Model 2: Direct Effect (Conditioning on Mediator G)
m_direct <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F + bG * G,
    a ~ dnorm(0, 0.2),
    c(bF, bG) ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)
precis(m_direct)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import pandas as pd
import arviz as az

# Standardized fox data simulation matching data(foxes) structure
np.random.seed(42)
N = 116
A_sim = np.random.normal(0, 1, size=N)
F_sim = np.random.normal(0.7 * A_sim, 0.7)
G_sim = np.random.normal(0.6 * F_sim, 0.7)
# Direct effect of F on W = +0.5, effect of G on W = -0.6 (Total effect = 0.5 + 0.6*(-0.6) ~ 0.14)
W_sim = np.random.normal(0.5 * F_sim - 0.6 * G_sim, 0.6)

# Model 1: Total Effect (W ~ F)
with pm.Model() as model_fox_total:
    a = pm.Normal("a", mu=0, sigma=0.2)
    bF = pm.Normal("bF", mu=0, sigma=0.5)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bF * F_sim
    w_obs = pm.Normal("w_obs", mu=mu, sigma=sigma, observed=W_sim)
    
    idata_total = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

# Model 2: Direct Effect (W ~ F + G)
with pm.Model() as model_fox_direct:
    a = pm.Normal("a", mu=0, sigma=0.2)
    bF = pm.Normal("bF", mu=0, sigma=0.5)
    bG = pm.Normal("bG", mu=0, sigma=0.5)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bF * F_sim + bG * G_sim
    w_obs = pm.Normal("w_obs", mu=mu, sigma=sigma, observed=W_sim)
    
    idata_direct = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print("--- Total Effect Model ---")
print(az.summary(idata_total, var_names=["bF"]))
print("\n--- Direct Effect Model ---")
print(az.summary(idata_direct, var_names=["bF", "bG"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] F;
  vector[N] G;
  vector[N] W;
}
parameters {
  real a;
  real bF;
  real bG;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 0.2);
  bF ~ normal(0, 0.5);
  bG ~ normal(0, 0.5);
  sigma ~ exponential(1);
  W ~ normal(a + bF * F + bG * G, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(42)
N = 116
F_t = torch.randn(N, 1)
G_t = torch.randn(N, 1) + 0.6 * F_t
W_t = torch.randn(N, 1) + 0.5 * F_t - 0.6 * G_t

# Direct effect model (W ~ F + G)
class FoxDirectModel(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = FoxDirectModel()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([F_t, G_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, W_t)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Weights [bF_direct, bG]:", model.linear.weight.data.numpy())
```
:::

---

## 43. Annex — Instrumental Variables (IV) and 2SLS Structural Logic

To complete the structural logic of the premise established in Section 13 (where controlling for an instrument $Z$ in OLS amplifies bias):

> **Core Premise:** Use an Instrument ($Z$) to isolate exogenous variation in an endogenous Treatment ($X$) via the Wald ratio or 2SLS, which eliminates confounding bias from Unobserved Confounders ($U$).

In observational data, when unmeasured confounding ($U$) affects both treatment ($X$) and outcome ($Y$), standard regression fails because $X$ is correlated with the error term. Instrumental Variables (IV) and Two-Stage Least Squares (2SLS) bypass this by extracting only the variation in $X$ that is driven by an external shock or assignment mechanism ($Z$), independent of $U$.

Here is the mathematical and structural breakdown of how this is correctly applied.

### 43.1 The Structural Causal Model (DAG)

For an instrument $Z$ to be valid, it must satisfy three core assumptions within the causal graph $Z \rightarrow X \rightarrow Y$ (with $U \rightarrow X$ and $U \rightarrow Y$):

1. **Relevance:** $Z$ has a strong causal effect on $X$ ($\mathrm{Cov}(Z, X) \neq 0$).
2. **Exclusion Restriction:** $Z$ affects $Y$ *only* through its effect on $X$. There is no direct path $Z \rightarrow Y$.
3. **Independence (Unconfoundedness):** $Z$ is independent of the unobserved confounders $U$ ($Z \perp\!\!\!\perp U$).

### 43.2 The Wald Estimator (The Ratio)

When $Z$ is a binary instrument (e.g., random assignment to a nudge), the causal effect of $X$ on $Y$ can be isolated using the **Wald Estimator**.

The logic is that any effect $Z$ has on $Y$ must pass entirely through $X$. Therefore, the total effect of $Z$ on $Y$ (the "intent-to-treat" effect), scaled by how much $Z$ actually changes $X$, yields the true causal effect of $X$ on $Y$:

$$\beta_{IV} = \frac{\mathrm{Cov}(Y, Z)}{\mathrm{Cov}(X, Z)}$$

For binary variables, this takes the form of a ratio of expectations:

$$\beta_{IV} = \frac{\mathbb{E}[Y \mid Z=1] - \mathbb{E}[Y \mid Z=0]}{\mathbb{E}[X \mid Z=1] - \mathbb{E}[X \mid Z=0]}$$

The numerator represents the variation in $Y$ caused by $Z$. The denominator represents the variation in $X$ caused by $Z$. By dividing the two, the confounding bias from $U$ is mathematically purged.

### 43.3 Two-Stage Least Squares (2SLS)

When dealing with continuous instruments, multiple instruments, or additional exogenous covariates, the Wald ratio is generalized using the 2SLS estimator. This is typically implemented via matrix algebra in automated decision pipelines.

Let $X$ be the endogenous treatment matrix, $Z$ be the matrix of instruments (including any exogenous control variables), and $Y$ be the outcome vector.

**Stage 1: Isolate Exogenous Variation**  
Regress the endogenous treatment $X$ on the instruments $Z$ to predict $\widehat{X}$. This purges $X$ of its correlation with the unobserved error term $U$.

$$\widehat{X} = Z(Z^T Z)^{-1} Z^T X$$

**Stage 2: Estimate the Causal Effect**  
Regress the outcome $Y$ on the fitted values $\widehat{X}$ to obtain the unbiased causal parameter $\widehat{\beta}_{2SLS}$.

$$\widehat{\beta}_{2SLS} = (\widehat{X}^T \widehat{X})^{-1} \widehat{X}^T Y$$

Combining these steps yields the direct closed-form 2SLS estimator:

$$\widehat{\beta}_{2SLS} = (X^T Z (Z^T Z)^{-1} Z^T X)^{-1} X^T Z (Z^T Z)^{-1} Z^T Y$$

### 43.4 Critical Failure Modes in Implementation

- **Weak Instruments:** If the correlation between $Z$ and $X$ is weak, the denominator in the Wald ratio approaches zero. This amplifies any tiny violations of the exclusion restriction, leading to massive asymptotic bias and inflated standard errors.
- **Monotonicity (LATE):** In heterogeneous populations with binary compliance, IV estimates the Local Average Treatment Effect (LATE)—the effect solely on the "compliers" (those whose treatment status was actually shifted by the instrument). It assumes no "defiers" exist.

