---
title: "A07 — Good and Bad Controls"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: true
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# A07 — Good and Bad Controls

**Course:** Statistical Rethinking 2026  
**Section:** A — Beginner  
**Lecture:** A07 — Good and Bad Controls  
**Duration:** approximately 1 hour 1 minute  
**Official reading:** Chapter 6 of *Statistical Rethinking*, second edition  
**Secondary transition:** Chapter 7 material on overfitting, regularization, and out-of-sample evaluation  
**Central question:** Which variables should be included in a statistical model for a specific causal estimand, and how can an apparently useful control create bias or destroy precision?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official scripts `06_simulations_bad_controls.r`, `07_copernican_model.R`, and `07_overfitting_animations.r`.
>
> The official A07 title and public description emphasize beneficial and harmful controls, including confounders, mediators, and colliders. The bad-control taxonomy below follows the simulations explicitly present in `06_simulations_bad_controls.r`: post-treatment confounding, case-control or outcome-descendant bias, precision parasites, and bias amplification.
>
> The repository also contains chapter-7 Copernican and overfitting scripts immediately after these materials. They are included as a clearly marked transition because they illuminate a separate question: whether additional variables or flexibility improve out-of-sample prediction. They should not be confused with causal control selection, and this guide does not claim that every chapter-7 animation appeared in the A07 recording.

---

## Table of Contents

- [1. Main thesis](#1-main-thesis)
- [2. Relationship to the book](#2-relationship-to-the-book)
  - [2.1 Two questions that must remain separate](#21-two-questions-that-must-remain-separate)
- [3. What statistical control means](#3-what-statistical-control-means)
- [4. Control quality has two dimensions](#4-control-quality-has-two-dimensions)
  - [4.1 Bias](#41-bias)
  - [4.2 Precision](#42-precision)
- [5. A genuinely good control: an observed common cause](#5-a-genuinely-good-control-an-observed-common-cause)
  - [5.1 Good does not mean perfect](#51-good-does-not-mean-perfect)
- [6. A useful precision variable](#6-a-useful-precision-variable)
- [7. Bad control 1: post-treatment confounding](#7-bad-control-1-post-treatment-confounding)
  - [7.1 Unadjusted model](#71-unadjusted-model)
  - [7.2 Adjusted model](#72-adjusted-model)
  - [7.3 Why “post-treatment” is not the complete explanation](#73-why-post-treatment-is-not-the-complete-explanation)
- [8. Blocking mediation and creating bias simultaneously](#8-blocking-mediation-and-creating-bias-simultaneously)
- [9. Bad control 2: case-control or outcome-descendant bias](#9-bad-control-2-case-control-or-outcome-descendant-bias)
  - [9.1 Unadjusted model](#91-unadjusted-model)
  - [9.2 Adjusted model](#92-adjusted-model)
  - [9.3 Connection to case-control sampling](#93-connection-to-case-control-sampling)
- [10. Outcome descendants and leakage](#10-outcome-descendants-and-leakage)
- [11. Bad or neutral control 3: the precision parasite](#11-bad-or-neutral-control-3-the-precision-parasite)
  - [11.1 Bias](#111-bias)
  - [11.2 Precision](#112-precision)
- [12. Why an exposure predictor can harm precision](#12-why-an-exposure-predictor-can-harm-precision)
- [13. Bad control 4: the bias amplifier](#13-bad-control-4-the-bias-amplifier)
  - [13.1 Unadjusted association](#131-unadjusted-association)
  - [13.2 Conditioning on $Z$](#132-conditioning-on-z)
- [14. Residualization view of bias amplification](#14-residualization-view-of-bias-amplification)
- [15. Pre-treatment is not sufficient for safety](#15-pre-treatment-is-not-sufficient-for-safety)
- [16. Post-treatment is a warning, not a full taxonomy](#16-post-treatment-is-a-warning-not-a-full-taxonomy)
- [17. Randomization does not make every control safe](#17-randomization-does-not-make-every-control-safe)
- [18. The Table 2 fallacy revisited](#18-the-table-2-fallacy-revisited)
- [19. A practical control-classification table](#19-a-practical-control-classification-table)
- [20. Simulation is the decisive teaching tool](#20-simulation-is-the-decisive-teaching-tool)
- [21. General simulation template for controls](#21-general-simulation-template-for-controls)
- [22. Control selection is not feature selection](#22-control-selection-is-not-feature-selection)
- [23. Transition: the Copernican principle for model evaluation](#23-transition-the-copernican-principle-for-model-evaluation)
  - [23.1 Causal and predictive humility](#231-causal-and-predictive-humility)
- [24. Overfitting is different from bad-control bias](#24-overfitting-is-different-from-bad-control-bias)
- [25. The official polynomial overfitting animation](#25-the-official-polynomial-overfitting-animation)
- [26. Leave-one-out cross-validation](#26-leave-one-out-cross-validation)
  - [26.1 What LOO can establish](#261-what-loo-can-establish)
- [27. In-sample and out-of-sample error](#27-in-sample-and-out-of-sample-error)
- [28. Regularization](#28-regularization)
  - [28.1 Regularization is not causal repair](#281-regularization-is-not-causal-repair)
- [29. WAIC and PSIS](#29-waic-and-psis)
- [30. Why predictive model comparison cannot choose the DAG](#30-why-predictive-model-comparison-cannot-choose-the-dag)
- [31. CRM application: risk-score adjustment](#31-crm-application-risk-score-adjustment)
- [32. CRM application: email-open control](#32-crm-application-email-open-control)
- [33. CRM application: eligibility as a collider](#33-crm-application-eligibility-as-a-collider)
- [34. CRM application: channel preference as precision parasite](#34-crm-application-channel-preference-as-precision-parasite)
- [35. CRM application: instrument-like assignment rule as bias amplifier](#35-crm-application-instrument-like-assignment-rule-as-bias-amplifier)
- [36. Implications for CausalOps](#36-implications-for-causalops)
  - [36.1 Proposed control-role schema](#361-proposed-control-role-schema)
  - [36.2 Candidate-set comparison](#362-candidate-set-comparison)
  - [36.3 Separate scoreboards](#363-separate-scoreboards)
- [37. Common errors exposed by A07](#37-common-errors-exposed-by-a07)
- [38. Lecture takeaways](#38-lecture-takeaways)
- [39. Review questions](#39-review-questions)
- [40. Book reading guide](#40-book-reading-guide)
- [41. Official code guide](#41-official-code-guide)
- [42. Appendix — Homework A07 Question & Solutions](#42-appendix--homework-a07-question--solutions)

---

## 1. Main thesis

The central argument is:

> **A control variable is good or bad only relative to a causal graph, an exposure, an outcome, and an estimand.**

There is no universal rule that more controls are better.

A control can:

- remove confounding;
- improve precision;
- leave the estimand unchanged but reduce precision;
- block part of the causal effect;
- open a collider path;
- amplify bias from an unmeasured common cause;
- or redefine the target population through selection.

The workflow is:

$$ \text{causal question} \rightarrow \text{estimand} \rightarrow \text{causal DAG} \rightarrow \text{candidate control} \rightarrow \text{path analysis} \rightarrow \text{bias and precision analysis} \rightarrow \text{simulation validation} \rightarrow \text{estimation} $$

The practical lesson is:

$$ \text{predictive association with }Y \not\Rightarrow \text{valid causal control} $$

and:

$$ \text{measured before treatment} \not\Rightarrow \text{safe to control} $$

---

## 2. Relationship to the book

A07 continues Chapter 6 and transitions toward Chapter 7.

| A07 topic | Second-edition counterpart | Role |
|---|---|---|
| Post-treatment controls | Section 6.2 | Controls can block causal paths or open colliders |
| Collider and selection bias | Section 6.3 | Conditioning can create association |
| Haunted DAG | Section 6.3.2 | Unobserved causes alter observed-variable roles |
| Backdoor adjustment | Section 6.4 | Identifies controls needed for a target effect |
| Variable selection by causal role | Chapter 6 overall | Replaces automatic adjustment rules |
| Overfitting | Sections 7.1–7.2 | Additional flexibility lowers training error but may worsen prediction |
| Regularization | Sections 7.2–7.3 | Priors constrain unstable complexity |
| Cross-validation, WAIC, PSIS | Sections 7.2–7.5 | Approximate out-of-sample predictive performance |

### 2.1 Two questions that must remain separate

A control-selection question asks:

> Does including $Z$ help identify the causal effect of $X$ on $Y$?

A predictive-model question asks:

> Does including $Z$ improve predictions for new observations?

The answers can differ. A collider may improve prediction while damaging causal identification. A valid confounder may be essential for causal identification even if it barely improves predictive accuracy. A variable can be causally neutral but increase the variance of the treatment-effect estimate.

---

## 3. What statistical control means

Consider:

$$ Y_i \sim \mathrm{Normal}(\mu_i,\sigma), \quad \mu_i = \alpha + \beta_X X_i + \beta_Z Z_i $$

Including $Z$ means that $\beta_X$ is identified from variation in $X$ among units with the same or similar values of $Z$, under the model.

This is statistical control. It is not equivalent to experimental control. An experiment manipulates or holds constant aspects of the data-generating process. A regression conditions on recorded values after the data have been generated.

---

## 4. Control quality has two dimensions

A candidate control should be evaluated along at least two dimensions:

### 4.1 Bias

Does conditioning on $Z$ move the estimate toward or away from the target causal effect?

$$ \mathrm{Bias} \left( \widehat{\tau} \right) = \mathbb{E} \left[ \widehat{\tau} \right] - \tau $$

### 4.2 Precision

How variable is the estimate across repeated samples? Mean squared error combines both:

$$ \mathrm{MSE} \left( \widehat{\tau} \right) = \mathrm{Bias} \left( \widehat{\tau} \right)^2 + \mathrm{Var} \left( \widehat{\tau} \right) $$

Causal identification places bias first. Precision cannot rescue a systematically wrong estimand.

---

## 5. A genuinely good control: an observed common cause

Suppose $Z \rightarrow X$, $Z \rightarrow Y$, and $X \rightarrow Y$. The backdoor path $X \leftarrow Z \rightarrow Y$ is open without adjustment. Conditioning on $Z$ closes it.

### 5.1 Good does not mean perfect

Adjustment can still fail when $Z$ is measured with error, important common causes remain unobserved, positivity is weak, or the conditional outcome model is misspecified.

---

## 6. A useful precision variable

Suppose $Z$ predicts $Y$ but is independent of $X$ ($Z \rightarrow Y, X \rightarrow Y$). Conditioning on $Z$ is not required to remove confounding, but it explains outcome variation unrelated to $X$, reducing residual variance and increasing precision for $\beta_X$.

---

## 7. Bad control 1: post-treatment confounding

The first official bad-control simulation is:

$$ X \rightarrow Z, \quad U \rightarrow Z, \quad U \rightarrow Y $$

with a possible path $Z \rightarrow Y$.

```{r}
#| label: fig-bad-control-1
#| fig-cap: "Bad Control 1: Post-treatment confounding simulation"
set.seed(123)
f_post <- function(n=100, bXZ=1, bZY=0) {
  X <- rnorm(n)
  u <- rnorm(n)
  Z <- rnorm(n, bXZ * X + u)
  Y <- rnorm(n, bZY * Z + u)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim1 <- replicate(1000, f_post(n=100, bZY=0))
cat("Unadjusted mean bX (True=0):", round(mean(sim1[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Biased):", round(mean(sim1[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
X <- rnorm(n)
U <- rnorm(n)
Z <- rnorm(n, X + U)
Y <- rnorm(n, 0 * Z + U) # True total effect of X on Y is 0

# Unadjusted model
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (opens collider X -> Z <- U -> Y)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.normal(0, 1, size=n)
U_data = np.random.normal(0, 1, size=n)
Z_data = np.random.normal(X_data + U_data, 1)
Y_data = np.random.normal(U_data, 1)

with pm.Model() as model_post:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_post = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_post, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.randn(n, 1)
U_t = torch.randn(n, 1)
Z_t = torch.randn(n, 1) + X_t + U_t
Y_t = torch.randn(n, 1) + U_t

class PostTreatmentRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = PostTreatmentRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Post-Treatment Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

### 7.1 Unadjusted model

The regression $Y \sim X$ recovers an effect near zero because $X$ is independent of $U$.

### 7.2 Adjusted model

The regression $Y \sim X + Z$ conditions on the collider $X \rightarrow Z \leftarrow U \rightarrow Y$, opening a noncausal path.

### 7.3 Why “post-treatment” is not the complete explanation

$Z$ is post-treatment, but it is harmful here because conditioning (1) may block mediation when $\beta_{ZY}\neq 0$, and (2) opens the collider path through $U$.

---

## 8. Blocking mediation and creating bias simultaneously

When $\beta_{ZY}\neq 0$, $Z$ is also a mediator ($X \rightarrow Z \rightarrow Y$). For the total effect of $X$ on $Y$, conditioning on $Z$ blocks a causal path and opens a noncausal collider path ($X \rightarrow Z \leftarrow U \rightarrow Y$).

---

## 9. Bad control 2: case-control or outcome-descendant bias

The second official simulation uses $X \rightarrow Y \rightarrow Z$:

```{r}
#| label: fig-bad-control-2
#| fig-cap: "Bad Control 2: Case-control / outcome-descendant bias simulation"
set.seed(123)
f_case <- function(n=100, bXY=1, bYZ=1) {
  X <- rnorm(n)
  Y <- rnorm(n, bXY * X)
  Z <- rnorm(n, bYZ * Y)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim2 <- replicate(1000, f_case())
cat("Unadjusted mean bX (True=1):", round(mean(sim2[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Biased):", round(mean(sim2[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
X <- rnorm(n)
Y <- rnorm(n, 1 * X)
Z <- rnorm(n, 1 * Y) # Z is a descendant of outcome Y

# Unadjusted model
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (conditions on outcome descendant Z)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
X_data = np.random.normal(0, 1, size=n)
Y_data = np.random.normal(1.0 * X_data, 1)
Z_data = np.random.normal(1.0 * Y_data, 1)

with pm.Model() as model_case:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_case = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_case, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
X_t = torch.randn(n, 1)
Y_t = torch.randn(n, 1) + 1.0 * X_t
Z_t = torch.randn(n, 1) + 1.0 * Y_t

class OutcomeDescendantRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = OutcomeDescendantRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Outcome-Descendant Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 10. Outcome descendants and leakage

In predictive modeling, an outcome descendant may dramatically improve apparent accuracy. For causal inference, it distorts the exposure coefficient. For prospective prediction, it is unavailable at scoring time.

---

## 11. Bad or neutral control 3: the precision parasite

The third official simulation has $Z \rightarrow X \rightarrow Y$:

```{r}
#| label: fig-bad-control-3
#| fig-cap: "Bad Control 3: Precision parasite simulation"
set.seed(123)
f_parasite <- function(n=50, bZX=1, bXY=1) {
  Z <- rnorm(n)
  X <- rnorm(n, bZX * Z)
  Y <- rnorm(n, bXY * X)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim3 <- replicate(1000, f_parasite())
cat("Unadjusted mean bX:", round(mean(sim3[1,]), 3), "SD:", round(sd(sim3[1,]), 3), "\n")
cat("Adjusted for Z mean bX:", round(mean(sim3[2,]), 3), "SD (Higher variance!):", round(sd(sim3[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 50
Z <- rnorm(n)
X <- rnorm(n, 1 * Z)
Y <- rnorm(n, 1 * X)

# Unadjusted model (more precise)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model (less precise, precision parasite Z)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 50
Z_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(1.0 * Z_data, 1)
Y_data = np.random.normal(1.0 * X_data, 1)

with pm.Model() as model_parasite:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_parasite = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_parasite, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 50
Z_t = torch.randn(n, 1)
X_t = torch.randn(n, 1) + 1.0 * Z_t
Y_t = torch.randn(n, 1) + 1.0 * X_t

class PrecisionParasiteRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = PrecisionParasiteRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Precision Parasite Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 12. Why an exposure predictor can harm precision

For a linear regression:

$$ \mathrm{Var} \left( \widehat{\beta}_X \right) \propto \frac{\sigma^2} {\sum_i \left( X_i-\widehat{X}_i(Z) \right)^2} $$

A variable $Z$ that strongly predicts $X$ without predicting $Y$ directly shrinks the denominator without shrinking outcome residual variance, increasing uncertainty.

---

## 13. Bad control 4: the bias amplifier

The fourth official simulation adds an unobserved common cause $U$:

$$ Z \rightarrow X \leftarrow U \rightarrow Y, \quad X \rightarrow Y $$

```{r}
#| label: fig-bad-control-4
#| fig-cap: "Bad Control 4: Bias amplifier simulation"
set.seed(123)
f_amp <- function(n=100, bZX=1, bXY=0) {
  Z <- rnorm(n)
  u <- rnorm(n)
  X <- rnorm(n, bZX * Z + u)
  Y <- rnorm(n, bXY * X + u)
  bX <- coef(lm(Y ~ X))['X']
  bXZ <- coef(lm(Y ~ X + Z))['X']
  return(c(bX, bXZ))
}

sim4 <- replicate(1000, f_amp(bXY=0))
cat("Unadjusted mean bX (Confounded by U):", round(mean(sim4[1,]), 3), "\n")
cat("Adjusted for Z mean bX (Bias amplified!):", round(mean(sim4[2,]), 3), "\n")
```

::: {.panel-tabset}
### R (rethinking)
```r
library(rethinking)
set.seed(123)
n <- 1000
Z <- rnorm(n)
U <- rnorm(n)
X <- rnorm(n, 1 * Z + U)
Y <- rnorm(n, 0 * X + U) # True total effect is 0

# Unadjusted model (shows baseline bias from U)
m_unadj <- lm(Y ~ X)
summary(m_unadj)

# Adjusted model for Z (amplifies bias from U!)
m_adj <- lm(Y ~ X + Z)
summary(m_adj)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import arviz as az

np.random.seed(123)
n = 1000
Z_data = np.random.normal(0, 1, size=n)
U_data = np.random.normal(0, 1, size=n)
X_data = np.random.normal(1.0 * Z_data + U_data, 1)
Y_data = np.random.normal(U_data, 1)

with pm.Model() as model_amp:
    a = pm.Normal("a", mu=0, sigma=1)
    bX = pm.Normal("bX", mu=0, sigma=1)
    bZ = pm.Normal("bZ", mu=0, sigma=1)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bX * X_data + bZ * Z_data
    y_obs = pm.Normal("y_obs", mu=mu, sigma=sigma, observed=Y_data)
    
    idata_amp = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print(az.summary(idata_amp, var_names=["bX", "bZ"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] X;
  vector[N] Y;
  vector[N] Z;
}
parameters {
  real a;
  real bX;
  real bZ;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 1);
  bX ~ normal(0, 1);
  bZ ~ normal(0, 1);
  sigma ~ exponential(1);
  Y ~ normal(a + bX * X + bZ * Z, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(123)
n = 1000
Z_t = torch.randn(n, 1)
U_t = torch.randn(n, 1)
X_t = torch.randn(n, 1) + 1.0 * Z_t + U_t
Y_t = torch.randn(n, 1) + U_t

class BiasAmplifierRegression(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = BiasAmplifierRegression()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([X_t, Z_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, Y_t)
    loss.backward()
    optimizer.step()

print("PyTorch Bias Amplifier Adjusted Weights [bX, bZ]:", model.linear.weight.data.numpy())
```
:::

---

## 14. Residualization view of bias amplification

Adjusting for $Z$ leaves $X_{\perp Z} = U + \varepsilon_X$. Removing $Z$'s variation increases the relative proportion of $X$'s variation explained by unobserved confounder $U$, amplifying bias.

---

## 15. Pre-treatment is not sufficient for safety

Pre-treatment variables can be colliders of prior events, instruments, or precision parasites. Pre-treatment timing does not guarantee a safe control.

---

## 16. Post-treatment is a warning, not a full taxonomy

Post-treatment variables require longitudinal causal models and estimand-specific methods (such as g-methods). Ordinary regression adjustment is not automatically valid.

---

## 17. Randomization does not make every control safe

Randomization protects the unadjusted treatment comparison. Conditioning on a post-treatment collider ($X \rightarrow Z \leftarrow U \rightarrow Y$) re-introduces correlation between treatment $X$ and $U$ given $Z$.

---

## 18. The Table 2 fallacy revisited

Each coefficient in a regression table requires its own DAG and identification argument. A single model cannot be interpreted as identifying causal effects for all listed predictors.

---

## 19. A practical control-classification table

| Candidate variable role | Bias consequence for total effect of $X$ on $Y$ | Typical precision consequence | Default action |
|---|---|---|---|
| Common cause of $X$ and $Y$ | Removes backdoor bias | May increase or decrease variance | Adjust |
| Cause of $Y$ only | Neutral for bias | Often improves precision | Often adjust |
| Cause of $X$ only | Neutral under no hidden confounding | Can reduce precision | Usually unnecessary |
| Cause of $X$ only with hidden $X$–$Y$ confounding | Can amplify bias | Often reduces precision | Avoid ordinary adjustment |
| Mediator | Blocks part of total effect | Varies | Do not adjust for total effect |
| Collider | Opens noncausal path | Can appear precise | Avoid |
| Collider descendant | Can partially open path | Varies | Usually avoid |
| Outcome descendant | Changes target comparison or selection | May look highly predictive | Avoid for ordinary causal effect |

---

## 20. Simulation is the decisive teaching tool

Monte Carlo simulation makes the true structural effect explicit, enabling empirical comparison of bias and precision across control sets.

---

## 21. General simulation template for controls

Define SCM $\rightarrow$ Set true parameter $\tau$ $\rightarrow$ Generate $R$ datasets $\rightarrow$ Fit candidate models $\rightarrow$ Evaluate $\mathrm{Bias}$, $\mathrm{RMSE}$, and coverage.

---

## 22. Control selection is not feature selection

Predictive feature selection optimizes outcome prediction. Causal control selection optimizes identification of an intervention contrast.

---

## 23. Transition: the Copernican principle for model evaluation

The observed sample is one realization, not the privileged center of the universe.

---

## 24. Overfitting is different from bad-control bias

Overfitting is fitting sample noise. Bad-control bias is estimating the wrong structural contrast.

---

## 25. The official polynomial overfitting animation

Polynomial models of increasing degree fit training noise better, lowering in-sample error while degrading out-of-sample prediction.

```{r}
#| label: fig-overfitting-demo
#| fig-cap: "Polynomial overfitting demo: High-degree polynomials overfit in-sample"
set.seed(123)
N <- 15
x <- seq(-2, 2, length.out=N)
y <- rnorm(N, 0.5 * x, 0.5)

m_poly1 <- lm(y ~ x)
m_poly3 <- lm(y ~ poly(x, 3))
cat("Linear In-Sample MSE:", round(mean(residuals(m_poly1)^2), 3), "\n")
cat("Polynomial Deg 3 In-Sample MSE:", round(mean(residuals(m_poly3)^2), 3), "\n")
```

---

## 26. Leave-one-out cross-validation

LOO estimates out-of-sample predictive generalization: $\mathrm{elpd}_{\mathrm{LOO}} = \sum_{i=1}^N \log p(y_i \mid D_{-i})$.

---

## 27. In-sample and out-of-sample error

Out-of-sample error follows a U-shape relative to model flexibility.

---

## 28. Regularization

Regularizing priors ($\beta_k \sim \mathrm{Normal}(0, 0.5)$) shrink coefficients, reducing overfitting. Regularization does not repair a wrong DAG.

---

## 29. WAIC and PSIS

WAIC and PSIS approximate out-of-sample predictive performance.

---

## 30. Why predictive model comparison cannot choose the DAG

Predictive criteria prefer models that predict $Y$ best, even if extra variables are colliders that introduce causal bias.

---

## 31. CRM application: risk-score adjustment

Pre-campaign risk score $R$ is a common cause. Post-campaign risk scores contain mediators or colliders.

---

## 32. CRM application: email-open control

Controlling for email opening $O$ ($A \rightarrow O \leftarrow U \rightarrow Y$) blocks the mediated effect and opens a collider path.

---

## 33. CRM application: eligibility as a collider

Eligibility rules ($R \rightarrow E \leftarrow V$) select on a collider, creating within-eligible selection.

---

## 34. CRM application: channel preference as precision parasite

Channel preference ($Z \rightarrow A \rightarrow Y$) reduces treatment variance without reducing bias.

---

## 35. CRM application: instrument-like assignment rule as bias amplifier

Adjusting for an assignment rule $Z$ ($Z \rightarrow A \leftarrow U \rightarrow Y$) amplifies affinity confounding.

---

## 36. Implications for CausalOps

CausalOps must record structural roles, compare candidate adjustment sets via simulation, and maintain separate causal and predictive scoreboards.

---

## 37. Common errors exposed by A07

1. Adding every pre-treatment variable.
2. Selecting controls by p-value or feature importance.
3. Controlling for mediators or outcome descendants.
4. Using cross-validation to pick causal controls.
5. Assuming regularization repairs a bad DAG.

---

## 38. Lecture takeaways

1. Controls must be chosen relative to a DAG and estimand.
2. Controls affect both bias and precision.
3. Overfitting and causal bias are distinct problems requiring distinct scoreboards.

---

## 39. Review questions

Below is the list of 25 review questions covered in this section:

1. [**Question 1:** Why can a control be unbiased but harmful?](#q1)
2. [**Question 2:** What makes a common cause a good control?](#q2)
3. [**Question 3:** How can a cause of the outcome only improve precision?](#q3)
4. [**Question 4:** In the post-treatment simulation, why is the unadjusted estimate valid when $\beta_{ZY}=0$?](#q4)
5. [**Question 5:** Which collider path is opened by adjusting for $Z$?](#q5)
6. [**Question 6:** Why does a mediator control fail to estimate the total effect?](#q6)
7. [**Question 7:** Why is an adjusted mediator coefficient not automatically a direct effect?](#q7)
8. [**Question 8:** What is outcome-descendant bias?](#q8)
9. [**Question 9:** How does outcome-dependent sampling resemble conditioning on $Z$ in $X \rightarrow Y \rightarrow Z$?](#q9)
10. [**Question 10:** Why does a precision parasite increase coefficient variance?](#q10)
11. [**Question 11:** How does residualizing $X$ on an exposure predictor reduce identifying variation?](#q11)
12. [**Question 12:** What is a bias amplifier?](#q12)
13. [**Question 13:** Why can controlling for an instrument-like variable strengthen hidden confounding?](#q13)
14. [**Question 14:** Why is pre-treatment timing insufficient to identify good controls?](#q14)
15. [**Question 15:** How can randomization be undermined by post-treatment adjustment?](#q15)
16. [**Question 16:** What is the Table 2 fallacy?](#q16)
17. [**Question 17:** Which metrics should be compared in a simulation of candidate control sets?](#q17)
18. [**Question 18:** Why can a collider improve prediction while damaging causal estimation?](#q18)
19. [**Question 19:** What does leave-one-out cross-validation estimate?](#q19)
20. [**Question 20:** Why is overfitting different from causal bias?](#q20)
21. [**Question 21:** How does regularization affect polynomial curves?](#q21)
22. [**Question 22:** Why can WAIC or PSIS not select a causal DAG?](#q22)
23. [**Question 23:** In CRM, why is email opening usually a bad control for the total send effect?](#q23)
24. [**Question 24:** How can eligibility create collider selection?](#q24)
25. [**Question 25:** Why should CausalOps maintain separate causal and predictive scoreboards?](#q25)

---

### Question 1: Why can a control be unbiased but harmful? {#q1}
**Answer:** A control variable can leave the point estimate unbiased for the true causal effect while simultaneously increasing the estimator's variance (e.g., a precision parasite $Z \rightarrow X \rightarrow Y$). By absorbing exposure variation without reducing outcome residual variance, it degrades precision.

### Question 2: What makes a common cause a good control? {#q2}
**Answer:** A common cause $Z$ ($X \leftarrow Z \rightarrow Y$) generates a noncausal backdoor path between exposure $X$ and outcome $Y$. Conditioning on $Z$ blocks this backdoor path, identifying the true causal effect.

### Question 3: How can a cause of the outcome only improve precision? {#q3}
**Answer:** A variable $Z$ that causes $Y$ ($Z \rightarrow Y$) independently of $X$ ($Z \perp\!\!\!\perp X$) does not lie on any backdoor path between $X$ and $Y$. Including $Z$ in the model reduces outcome residual variance $\sigma^2$, thereby decreasing $\mathrm{Var}(\widehat{\beta}_X)$ and improving precision.

### Question 4: In the post-treatment simulation, why is the unadjusted estimate valid when $\beta_{ZY}=0$? {#q4}
**Answer:** When $\beta_{ZY}=0$, treatment $X$ affects $Z$, but $Z$ has no effect on $Y$. Unadjusted $X$ is independent of unobserved $U$ ($X \perp\!\!\!\perp U$), so $Y \sim X$ yields an unbiased estimate of zero.

### Question 5: Which collider path is opened by adjusting for $Z$? {#q5}
**Answer:** In $X \rightarrow Z \leftarrow U \rightarrow Y$, $Z$ is a collider between exposure $X$ and unobserved cause $U$. Conditioning on $Z$ opens the path $X \rightarrow Z \leftarrow U \rightarrow Y$, creating a spurious association between $X$ and $Y$.

### Question 6: Why does a mediator control fail to estimate the total effect? {#q6}
**Answer:** In $X \rightarrow Z \rightarrow Y$, $Z$ mediates $X$'s effect on $Y$. Conditioning on $Z$ blocks the mediated causal path, estimating only the direct effect (if unconfounded) rather than the total effect.

### Question 7: Why is an adjusted mediator coefficient not automatically a direct effect? {#q7}
**Answer:** If there is unobserved mediator-outcome confounding ($Z \leftarrow U \rightarrow Y$), conditioning on mediator $Z$ opens a collider path $X \rightarrow Z \leftarrow U \rightarrow Y$, biasing the direct effect estimate.

### Question 8: What is outcome-descendant bias? {#q8}
**Answer:** Conditioning on a variable $Z$ caused by the outcome ($X \rightarrow Y \rightarrow Z$) conditions on realized outcome variation, distorting the estimated effect of $X$ on $Y$.

### Question 9: How does outcome-dependent sampling resemble conditioning on $Z$ in $X \rightarrow Y \rightarrow Z$? {#q9}
**Answer:** Selecting rows into a sample based on outcome status $Y$ or its proxy $Z$ holds $Z$ constant or restricts its range, which acts as implicit conditioning on $Z$.

### Question 10: Why does a precision parasite increase coefficient variance? {#q10}
**Answer:** A precision parasite $Z$ ($Z \rightarrow X \rightarrow Y$) explains variation in $X$ without explaining independent variation in $Y$. Residualizing $X$ on $Z$ shrinks $\sum (X_i - \widehat{X}_i)^2$, increasing $\mathrm{Var}(\widehat{\beta}_X)$.

### Question 11: How does residualizing $X$ on an exposure predictor reduce identifying variation? {#q11}
**Answer:** Conditioning on an exposure cause removes the portion of $X$'s variation explained by that cause, leaving only the remaining residual variation in $X$ to estimate $\beta_X$.

### Question 12: What is a bias amplifier? {#q12}
**Answer:** A bias amplifier is an exposure predictor $Z$ ($Z \rightarrow X$) added to a model where $X$ and $Y$ share an unobserved confounder $U$ ($X \leftarrow U \rightarrow Y$). Adjusting for $Z$ removes exogenous variation in $X$, leaving $X$ more heavily dominated by $U$, which amplifies the confounding bias.

### Question 13: Why can controlling for an instrument-like variable strengthen hidden confounding? {#q13}
**Answer:** An instrument-like variable $Z$ affects $X$ but not $Y$ directly. If $X$ and $Y$ share hidden confounding $U$, adjusting for $Z$ acts as a bias amplifier.

### Question 14: Why is pre-treatment timing insufficient to identify good controls? {#q14}
**Answer:** Pre-treatment variables can be colliders of earlier latent variables, instruments, or precision parasites. Temporal ordering rules out post-treatment mediation but does not guarantee backdoor closure or lack of bias.

### Question 15: How can randomization be undermined by post-treatment adjustment? {#q15}
**Answer:** Randomization ensures $X \perp\!\!\!\perp U$ baseline unconfoundedness. However, adjusting for a post-treatment collider $Z$ ($X \rightarrow Z \leftarrow U \rightarrow Y$) opens $X \rightarrow Z \leftarrow U \rightarrow Y$, re-introducing confounding in the randomized experiment.

### Question 16: What is the Table 2 fallacy? {#q16}
**Answer:** The Table 2 fallacy is the mistaken assumption that all coefficients in a multivariable regression table can be interpreted as causal effects. Each coefficient requires a distinct adjustment set.

### Question 17: Which metrics should be compared in a simulation of candidate control sets? {#q17}
**Answer:** Candidate control sets should be compared using Monte Carlo bias, variance/SD, Root Mean Squared Error (RMSE), 95% interval coverage, and sign error rate.

### Question 18: Why can a collider improve prediction while damaging causal estimation? {#q18}
**Answer:** A collider $Z$ contains information about latent causes $U$ of outcome $Y$, helping predict $Y$. However, conditioning on $Z$ opens a noncausal path, biasing the causal estimate of $X$.

### Question 19: What does leave-one-out cross-validation estimate? {#q19}
**Answer:** LOO cross-validation estimates out-of-sample expected log predictive density ($\mathrm{elpd}_{\mathrm{LOO}}$), measuring predictive generalization to unseen data from the same data-generating distribution.

### Question 20: Why is overfitting different from causal bias? {#q20}
**Answer:** Overfitting is a sample variance problem where a model fits random noise. Causal bias is a structural identification problem where a model estimates the wrong causal estimand regardless of sample size.

### Question 21: How does regularization affect polynomial curves? {#q21}
**Answer:** Tight regularizing priors shrink higher-order polynomial coefficients toward zero, preventing extreme curve oscillations and reducing out-of-sample prediction error.

### Question 22: Why can WAIC or PSIS not select a causal DAG? {#q22}
**Answer:** WAIC and PSIS evaluate out-of-sample predictive accuracy, not causal identification. A model with collider bias or post-treatment controls may achieve better WAIC/PSIS scores than the causally correct model.

### Question 23: In CRM, why is email opening usually a bad control for the total send effect? {#q23}
**Answer:** Email opening $O$ is a post-treatment mediator of campaign $A$, and opening is often influenced by unobserved digital attention $U$ ($A \rightarrow O \leftarrow U \rightarrow Y$). Adjusting for $O$ both blocks campaign mediation and opens a collider backdoor path.

### Question 24: How can eligibility create collider selection? {#q24}
**Answer:** If campaign eligibility $E$ depends on both customer value $V$ and responsiveness $R$ ($V \rightarrow E \leftarrow R$), restricting analysis to eligible customers conditions on collider $E$, inducing a negative correlation between value and responsiveness.

### Question 25: Why should CausalOps maintain separate causal and predictive scoreboards? {#q25}
**Answer:** Causal scoreboards evaluate identification, backdoor closure, and causal simulation bias. Predictive scoreboards evaluate out-of-sample forecasting accuracy (LOO/PSIS). Conflating the two can lead to selecting predictively strong but causally biased models.

---

## 40. Book reading guide

### Primary reading
- Chapter 6, *The Haunted DAG & The Causal Terror* (Sections 6.2–6.4)

### Secondary transition
- Chapter 7, *Ulysses' Compass* (Sections 7.1–7.5)

---

## 41. Official code guide

### `06_simulations_bad_controls.r`
Simulates post-treatment confounding, outcome-descendant bias, precision parasites, and bias amplification across Monte Carlo trials.

### `07_copernican_model.R` & `07_overfitting_animations.r`
Animates Copernican predictive decentering, polynomial overfitting, LOO cross-validation, and regularizing priors.

---

## 42. Appendix — Homework A07 Question & Solutions

### Homework A07 Statement

This homework is based on the data in `data(foxes)`: 116 foxes from 30 different urban groups in England.

Variables:
- `area` ($A$): size of territory
- `avgfood` ($F$): average food in territory
- `groupsize` ($G$): number of foxes in territory
- `weight` ($W$): weight of each fox

**Assumed DAG:**

$$
\begin{array}{ccc}
 & A & \\
 & \downarrow & \\
F & \longrightarrow & G \\
\searrow & & \swarrow \\
 & W & 
\end{array}
$$

Specifically, directed edges:
- $A \rightarrow F$
- $F \rightarrow G$
- $F \rightarrow W$
- $G \rightarrow W$

**Questions:**
1. Use the backdoor criterion to estimate the **total causal influence** of avgfood ($F$) on weight ($W$). What is the minimal adjustment set?
2. What effect would increasing the food have on the weight of foxes inside it?
3. Can you explain the result?

---

### Theoretical & Graphical Path Analysis

Let's list all paths connecting exposure $F$ (avgfood) and outcome $W$ (weight):

1. **Path 1:** $F \rightarrow W$ (Direct causal path — **Open**)
2. **Path 2:** $F \rightarrow G \rightarrow W$ (Indirect causal path through mediator $G$ — **Open**)
3. **Path 3:** $F \leftarrow A$ (Backdoor path? No, $A$ is an upstream cause of $F$. $A$ has no other outgoing edge to $W$ or $G$. It is a dead end.)

#### Backdoor Criterion Analysis:
- There are **no open backdoor paths** between $F$ and $W$.
- Therefore, the **minimal sufficient adjustment set** for the total causal effect of $F$ on $W$ is the **empty set** $\emptyset$.
- Model for total effect: $W \sim F$.

#### What happens if we adjust for $G$ (groupsize)?
- $G$ is a mediator on the path $F \rightarrow G \rightarrow W$.
- Conditioning on $G$ blocks the indirect causal path $F \rightarrow G \rightarrow W$.
- Model $W \sim F + G$ estimates the **controlled direct effect** of $F$ on $W$, holding group size fixed.

#### Theoretical Explanation of the Result:
- **Total Effect ($W \sim F$):** Virtually **zero** ($\beta_F \approx 0$). Increasing food ($F$) increases territory group size ($G$) ($F \rightarrow G$), which spreads the food among more foxes, reducing weight per fox ($G \rightarrow W$ negative effect). The direct positive effect of food ($F \rightarrow W$) and the indirect negative effect ($F \rightarrow G \rightarrow W$) almost perfectly cancel out!
- **Direct Effect ($W \sim F + G$):** Strongly **positive** ($\beta_F > 0$). When group size $G$ is held fixed, more food directly leads to heavier foxes!

---

### Multi-Language Homework Implementation

```{r}
#| label: fig-homework-a07
#| fig-cap: "Homework A07: Foxes Total Effect vs Direct Effect Models"
library(rethinking)
library(dagitty)

data(foxes)
d <- foxes
d$F <- standardize(d$avgfood)
d$W <- standardize(d$weight)
d$G <- standardize(d$groupsize)
d$A <- standardize(d$area)

# Model 1: Total Effect of F on W (Minimal adjustment set = empty set)
m_total <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)

# Model 2: Direct Effect of F on W (Adjusting for mediator G)
m_direct <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F + bG * G,
    a ~ dnorm(0, 0.2),
    c(bF, bG) ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)

cat("--- Total Causal Effect of Food on Weight (Minimal set = empty) ---\n")
precis(m_total)

cat("\n--- Direct Causal Effect of Food on Weight (Adjusting for Group Size G) ---\n")
precis(m_direct)
```

::: {.panel-tabset}
### R (rethinking & dagitty)
```r
library(rethinking)
library(dagitty)

# Define Fox DAG
fox_dag <- dagitty("dag {
  A -> F
  F -> G
  F -> W
  G -> W
}")

print(adjustmentSets(fox_dag, exposure = "F", outcome = "W"))

data(foxes)
d <- foxes
d$F <- standardize(d$avgfood)
d$W <- standardize(d$weight)
d$G <- standardize(d$groupsize)

# Model 1: Total Effect (Minimal Set = empty)
m_total <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F,
    a ~ dnorm(0, 0.2),
    bF ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)
precis(m_total)

# Model 2: Direct Effect (Conditioning on Mediator G)
m_direct <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a + bF * F + bG * G,
    a ~ dnorm(0, 0.2),
    c(bF, bG) ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d
)
precis(m_direct)
```

### Python (PyMC)
```python
import pymc as pm
import numpy as np
import pandas as pd
import arviz as az

# Standardized fox data simulation matching data(foxes) structure
np.random.seed(42)
N = 116
A_sim = np.random.normal(0, 1, size=N)
F_sim = np.random.normal(0.7 * A_sim, 0.7)
G_sim = np.random.normal(0.6 * F_sim, 0.7)
# Direct effect of F on W = +0.5, effect of G on W = -0.6 (Total effect = 0.5 + 0.6*(-0.6) ~ 0.14)
W_sim = np.random.normal(0.5 * F_sim - 0.6 * G_sim, 0.6)

# Model 1: Total Effect (W ~ F)
with pm.Model() as model_fox_total:
    a = pm.Normal("a", mu=0, sigma=0.2)
    bF = pm.Normal("bF", mu=0, sigma=0.5)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bF * F_sim
    w_obs = pm.Normal("w_obs", mu=mu, sigma=sigma, observed=W_sim)
    
    idata_total = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

# Model 2: Direct Effect (W ~ F + G)
with pm.Model() as model_fox_direct:
    a = pm.Normal("a", mu=0, sigma=0.2)
    bF = pm.Normal("bF", mu=0, sigma=0.5)
    bG = pm.Normal("bG", mu=0, sigma=0.5)
    sigma = pm.Exponential("sigma", lam=1)
    
    mu = a + bF * F_sim + bG * G_sim
    w_obs = pm.Normal("w_obs", mu=mu, sigma=sigma, observed=W_sim)
    
    idata_direct = pm.sample(draws=1000, tune=1000, return_inferencedata=True)

print("--- Total Effect Model ---")
print(az.summary(idata_total, var_names=["bF"]))
print("\n--- Direct Effect Model ---")
print(az.summary(idata_direct, var_names=["bF", "bG"]))
```

### Stan
```stan
data {
  int<lower=0> N;
  vector[N] F;
  vector[N] G;
  vector[N] W;
}
parameters {
  real a;
  real bF;
  real bG;
  real<lower=0> sigma;
}
model {
  a ~ normal(0, 0.2);
  bF ~ normal(0, 0.5);
  bG ~ normal(0, 0.5);
  sigma ~ exponential(1);
  W ~ normal(a + bF * F + bG * G, sigma);
}
```

### PyTorch
```python
import torch
import torch.nn as nn

torch.manual_seed(42)
N = 116
F_t = torch.randn(N, 1)
G_t = torch.randn(N, 1) + 0.6 * F_t
W_t = torch.randn(N, 1) + 0.5 * F_t - 0.6 * G_t

# Direct effect model (W ~ F + G)
class FoxDirectModel(nn.Module):
    def __init__(self):
        super().__init__()
        self.linear = nn.Linear(2, 1)
    def forward(self, inputs):
        return self.linear(inputs)

model = FoxDirectModel()
criterion = nn.MSELoss()
optimizer = torch.optim.Adam(model.parameters(), lr=0.05)
inputs = torch.cat([F_t, G_t], dim=1)

for epoch in range(500):
    optimizer.zero_grad()
    preds = model(inputs)
    loss = criterion(preds, W_t)
    loss.backward()
    optimizer.step()

print("PyTorch Estimated Weights [bF_direct, bG]:", model.linear.weight.data.numpy())
```
:::
