---
title: "B01 — Multilevel Models"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: false
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# B01 — Multilevel Models

**Course:** Statistical Rethinking 2026  
**Section:** B — Advanced  
**Lecture:** B01 — Multilevel Models  
**Official reading listed by the course:** Chapter 12  
**Closest second-edition correspondence:** Chapter 13, *Models With Memory*  
**Central question:** How can repeated observations and clustered data be modeled so that each group receives its own estimate while information is shared adaptively across groups?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official script `12_intro_multilevel_models.r`.
>
> The official schedule labels B01 with Chapter 12. In the uploaded second edition, however, the material in the script—varying intercepts, partial pooling, adaptive priors, tadpole tanks, multiple clusters, and non-centered parameterization—corresponds directly to Chapter 13, *Models With Memory*. Second-edition Chapter 12 concerns mixtures, overdispersion, zero inflation, and ordered categories.
>
> This guide therefore preserves the official chapter label while explicitly mapping the lecture to Chapter 13 of the uploaded edition. It does not assume the contents of an unavailable draft third edition.

---

## 0. Notation and reading conventions

This chapter uses the following notation. A symbol is restated locally whenever its meaning changes.

### Indices and observed units

- $i$ indexes individual observations (e.g., individual tadpoles, visits, or survey responses);
- $j$ indexes a cluster or group (e.g., tadpole tank, café, participant, or story);
- $J$ denotes the total number of observed groups/clusters;
- $N$ denotes total observations across all groups, while $N_j$ or $D_j$ denotes total observations within group $j$;
- $s$ indexes a retained MCMC posterior draw ($s = 1, \dots, S$).

### Model quantities

- $Y_{ij}$: outcome for observation $i$ in group $j$;
- $S_j$: count of successes (e.g., survivors) in group $j$ out of $D_j$ trials;
- $p_j$: event probability for group $j$;
- $\lambda_j$: rate or intensity parameter for group $j$ in a Poisson model;
- $\alpha_j$: group-specific intercept for group $j$;
- $\bar{\alpha}$: population mean intercept (hyperparameter);
- $\sigma_\alpha$: population standard deviation of group intercepts (hyperparameter);
- $z_j$: standardized group offset in a non-centered parameterization ($z_j \sim \mathrm{Normal}(0, 1)$);
- $\beta$: slope coefficient for a predictor variable;
- $X_{ij}$ or $P_j$: individual-level or group-level covariate.

### Acronyms and shorthand

- **GLMM:** generalized linear multilevel model;
- **HMC:** Hamiltonian Monte Carlo;
- **ESS:** effective sample size;
- **PSIS-LOO / WAIC:** Pareto-smoothed importance sampling leave-one-out cross-validation / Widely Applicable Information Criterion;
- **MSE:** mean squared error;
- **DAG:** directed acyclic graph.

---

## 1. Main thesis

The central argument is:

> **A multilevel model estimates group-specific effects and the population distribution of those effects simultaneously.**

This produces **partial pooling**.

Each group is informed by:

1. its own observations;
2. the observations from other groups;
3. the inferred amount of variation among groups.

The workflow is:

$$ \text{clustered generative process} \longrightarrow \text{group-specific parameters} \longrightarrow \text{population distribution of parameters} \longrightarrow \text{adaptive regularization} \longrightarrow \text{partial pooling} \longrightarrow \text{predictions for observed and new groups} $$

A multilevel model does not force every group to be identical.

It also does not treat every group as unrelated.

Instead:

$$ \text{complete pooling} \longleftrightarrow \text{partial pooling} \longleftrightarrow \text{no pooling} $$

The data learn where the model should lie between the two extremes.

---

## 2. Relationship to the book and the lecture

The B01 script aligns closely with Chapter 13 of *Statistical Rethinking* (2nd Edition).

| B01 topic | Closest second-edition section | Role |
|---|---|---|
| Clustered observations | Chapter 13 introduction | Motivates models with memory |
| Tadpole tanks | Section 13.1 | Introductory varying-intercept model |
| Partial pooling | Sections 13.1–13.2 | Adaptive regularization across groups |
| Underfitting–overfitting trade-off | Section 13.2 | Compares pooling strategies |
| More than one cluster type | Section 13.3 | Actor and block effects |
| Non-centered parameterization | Section 13.4 | Improves MCMC geometry |
| Same- and new-cluster prediction | Section 13.5 | Distinguishes conditional and population predictions |
| Trolley stories and participants | Script extension | Cross-classified varying intercepts |
| Café population model | 2026 script-specific teaching example | Sequentially illustrates shared learning |

### 2.1 Why the mismatch matters

The official label “Chapter 12” should not be used to locate this material in the uploaded second edition.

The practical mapping is:

$$ \text{B01 official Chapter 12} \approx \text{second-edition Chapter 13} $$

This is a course reorganization or edition difference, not a statistical contradiction.

---

## 3. Why clustered data require memory

Many datasets contain repeated observations within units.

Examples include:

- tadpoles within tanks;
- responses within participants;
- judgments within stories;
- transactions within customers;
- customers within stores;
- products within categories;
- repeated measurements within patients;
- students within schools.

Observations from the same cluster often share unmeasured causes.

If observations $i$ and $k$ belong to group $j$, then generally:

$$ Y_{ij} \not\!\perp\!\!\!\perp Y_{kj} $$

after conditioning only on global predictors.

A group-specific effect can represent the shared environment:

$$ Y_{ij} \sim p \left( Y_{ij} \mid \alpha_j, X_{ij}, \theta \right) $$

The model has “memory” because learning about group $j$ is retained and can also inform the population of groups.

---

## 4. Three pooling strategies

Suppose each group $j$ has its own baseline parameter $\alpha_j$.

### 4.1 Complete pooling

One parameter is shared by all groups:

$$ \alpha_j = \alpha $$

for every $j$.

The model assumes no residual group heterogeneity.

Advantages:

- low variance;
- simple;
- stable for sparse groups.

Disadvantages:

- underfits genuine between-group variation;
- can produce overconfident individual-group predictions;
- may mask important cluster structure.

### 4.2 No pooling

Every group receives an unrelated parameter:

$$ \alpha_j \sim p_0(\alpha_j) $$

independently under a fixed broad prior.

Advantages:

- allows arbitrary group differences;
- can reproduce group-level sample proportions closely.

Disadvantages:

- overfits sparse groups;
- does not share information;
- estimates extreme values from small samples;
- cannot learn a coherent population distribution adaptively.

### 4.3 Partial pooling

Group parameters share a learned population distribution:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

The hyperparameters are also estimated:

$$ \bar{\alpha} \sim p \left( \bar{\alpha} \right) $$

$$ \sigma_\alpha \sim p \left( \sigma_\alpha \right) $$

This model learns how similar the groups are.

---

## 5. Varying effects

McElreath generally prefers **varying effects** over “random effects.”

A varying intercept model gives every group its own intercept:

$$ \alpha_j $$

while assuming those intercepts arise from a common population.

The word “varying” emphasizes what the model does:

$$ \text{intercept varies by group} $$

It avoids the misleading impression that:

- some parameters are random while others are not;
- varying effects are merely nuisance terms;
- population-level parameters are fixed constants known without uncertainty.

In Bayesian modeling, every unknown parameter has a posterior distribution.

---

## 6. The multilevel generative model

A generic two-level Gaussian model is:

### Observation level

$$ Y_{ij} \sim \mathrm{Normal} \left( \alpha_j, \sigma_Y \right) $$

### Group level

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

### Hyperpriors

$$ \bar{\alpha} \sim \mathrm{Normal} \left( m_\alpha, s_\alpha \right) $$

$$ \sigma_\alpha \sim \mathrm{Exponential} \left( \lambda_\alpha \right) $$

The joint posterior is:

$$ p \left( \alpha_1,\ldots,\alpha_J, \bar{\alpha}, \sigma_\alpha, \sigma_Y \mid D \right) $$

All levels are estimated simultaneously.

---

## 7. Adaptive priors

For a no-pooling model, the prior on each group effect is fixed before fitting:

$$ \alpha_j \sim \mathrm{Normal}(0, 1.5) $$

In a multilevel model:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

The prior itself depends on unknown parameters learned from all groups.

It is therefore an **adaptive prior**.

The posterior for group $j$ is shaped by:

$$ p(\alpha_j \mid D) \propto p(D_j \mid \alpha_j) p \left( \alpha_j \mid \bar{\alpha}, \sigma_\alpha \right) $$

But $\bar{\alpha}$ and $\sigma_\alpha$ are themselves learned using all groups:

$$ D_1,\ldots,D_J $$

This is the mechanism by which the model remembers across groups.

---

## 8. Partial pooling and shrinkage

Partial pooling commonly moves group estimates toward the population mean.

This is called **shrinkage**.

For an approximately Gaussian group estimate $\widehat{\alpha}_j$ with sampling variance $V_j$ and a population model $\alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha^2 \right)$, the posterior mean has the schematic form:

$$ \mathbb{E} \left[ \alpha_j \mid D \right] \approx w_j \widehat{\alpha}_j + \left( 1-w_j \right) \bar{\alpha} $$

where:

$$ w_j = \frac{ \sigma_\alpha^2 }{ \sigma_\alpha^2 + V_j } $$

This formula is exact in linear Gaussian conjugate settings, but captures the general logic for all GLMMs.

### More data in group $j$

When $V_j$ is small (large $N_j$):

$$ w_j \rightarrow 1 $$

The group estimate remains near its own data mean.

### Less data in group $j$

When $V_j$ is large (small $N_j$):

$$ w_j \rightarrow 0 $$

The estimate moves more strongly toward the population mean $\bar{\alpha}$.

### More genuine group heterogeneity

When $\sigma_\alpha$ is large, pooling weakens ($w_j \rightarrow 1$).

### More homogeneous groups

When $\sigma_\alpha$ is small, pooling strengthens ($w_j \rightarrow 0$).

Shrinkage is therefore adaptive to both group sample size and estimated population variation.

---

## 9. Shrinkage is not bias correction

Partial pooling deliberately introduces bias toward the population mean for individual groups.

This reduces total expected error across all groups.

The bias–variance trade-off is:

$$ \mathrm{MSE} = \mathrm{Bias}^2 + \mathrm{Variance} $$

No-pooling estimates have little pooling bias but can have high variance.

Complete-pooling estimates have low variance but can have large bias for unusual groups.

Partial pooling aims to improve average out-of-sample accuracy by accepting modest shrinkage bias in exchange for a larger reduction in variance.

This is regularization, not an attempt to make every group look average.

---

## 10. The café golem

The official B01 script introduces a population of cafés.

Each café has a latent waiting-time intensity.

Visits produce observed waiting-time counts.

A simplified model is:

$$ Y_i \sim \mathrm{Poisson} \left( \lambda_i \right) $$

$$ \log(\lambda_i) = \alpha_{C_i} $$

where $C_i$ identifies the café.

The café effects arise from a population:

$$ \alpha_j = \bar{\alpha} + \sigma_\alpha z_j $$

$$ z_j \sim \mathrm{Normal}(0,1) $$

with:

$$ \bar{\alpha} \sim \mathrm{Normal}(2,0.5) $$

$$ \sigma_\alpha \sim \mathrm{Exponential}(10) $$

The script uses a non-centered representation directly.

---

## 11. Sequential learning in the café example

The animation in `12_intro_multilevel_models.r` begins with no café observations.

At that point, every café prediction is drawn from the same prior population.

As visits occur:

- the visited café's posterior changes directly;
- the population mean and variance may change;
- unvisited cafés can change indirectly because their prior population has changed;
- cafés with little information remain close to the population distribution;
- cafés with much information become more individually determined.

This is the key multilevel phenomenon:

> **An observation from one group can change predictions for another group through the learned population distribution.**

In a no-pooling model, a café observation would affect only that café.

In a multilevel model, the information propagates.

---

## 12. Population of cafés versus observed cafés

The model estimates two kinds of quantities.

### Café-specific effects

$$ \alpha_1,\ldots,\alpha_J $$

These describe cafés observed in the sample.

### Population distribution

$$ \alpha_{\mathrm{new}} \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

This describes a new café drawn from the modeled population.

The distinction becomes vital for prediction.

For an observed café $j$:

$$ p \left( Y_{\mathrm{new}} \mid C=j,D \right) $$

uses the posterior of $\alpha_j$.

For a new café:

$$ p \left( Y_{\mathrm{new}} \mid C=\mathrm{new},D \right) $$

must integrate over a newly drawn group effect $\alpha_{\mathrm{new}}$.

---

## 13. The reed-frog data

The official script uses 48 tadpole tanks from the `reedfrogs` dataset.

For tank $j$:

- $S_j$ is the number of survivors;
- $D_j$ is the initial number of tadpoles (density: 10, 20, or 35);
- $T_j$ identifies the tank.

The observation model is:

$$ S_j \sim \mathrm{Binomial} \left( D_j, p_j \right) $$

The varying-intercept model is:

$$ \mathrm{logit}(p_j) = \alpha_{T_j} $$

The group distribution is:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

with priors:

$$ \bar{\alpha} \sim \mathrm{Normal}(0,1.5) $$

$$ \sigma_\alpha \sim \mathrm{Exponential}(1) $$

---

## 14. No-pooling tadpole model

A no-pooling version gives every tank a separate parameter under a fixed prior:

$$ \alpha_j \sim \mathrm{Normal}(0, 1.5) $$

This model does not learn how much tank intercepts vary.

Each tank is estimated independently.

Observed survival proportions are:

$$ \widehat{p}_j = \frac{S_j}{D_j} $$

Small tanks can produce extreme proportions (0 or 1) from only a few outcomes.

No pooling preserves these extremes too strongly.

---

## 15. Partial-pooling tadpole model

The multilevel model estimates $\bar{\alpha}$ and $\sigma_\alpha$ from the 48 tanks.

Posterior tank probabilities are:

$$ p_j^{(s)} = \mathrm{logit}^{-1} \left( \alpha_j^{(s)} \right) $$

The posterior mean estimate is:

$$ \widehat{p}_j = \frac{1}{S} \sum_{s=1}^{S} p_j^{(s)} $$

The estimates move toward the population center, especially for tanks with fewer tadpoles.

The model does not shrink equally on the probability scale because the hierarchy is Gaussian on the log-odds scale.

---

## 16. Pooling and tank size

The tanks have different initial counts ($D_j \in \{10, 20, 35\}$).

A tank with $D_j=10$ contains less information about its survival probability than a tank with $D_j=35$.

Therefore, smaller tanks are pooled more strongly toward the population average.

This is visible in the difference between raw survival proportion and partially pooled posterior probability.

The model automatically accounts for uncertainty from the binomial likelihood.

No manual sample-size weighting is required.

---

## 17. Extreme groups shrink most visibly

Groups with extreme raw outcomes often move substantially toward the population mean.

This is not because the model forbids extreme true groups.

It is because extreme sample proportions from small groups are often sampling noise.

For example, $S_j=0$ out of $D_j=10$ does not imply $p_j=0$.

Likewise, $S_j=D_j$ does not imply $p_j=1$.

A multilevel posterior combines the group likelihood with evidence about the population of tanks.

---

## 18. Adaptive regularization

The script compares a model that estimates $\sigma_\alpha$ with one that fixes it.

Estimated hierarchy:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

Fixed hierarchy:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, 1 \right) $$

The estimated model learns the appropriate pooling strength.

This is adaptive regularization.

The hyperparameter $\sigma_\alpha$ controls flexibility:

- **Small $\sigma_\alpha$**: $\alpha_j \approx \bar{\alpha}$ (strong pooling, low effective complexity).
- **Large $\sigma_\alpha$**: group parameters differ widely (weak pooling, greater effective complexity).

---

## 19. Effective number of parameters

A varying-intercept model may contain many nominal parameters ($\alpha_1,\ldots,\alpha_J$).

But strong shrinkage means they do not behave like $J$ fully independent parameters.

Predictive criteria estimate an effective complexity such as $p_{\mathrm{WAIC}}$ or $p_{\mathrm{LOO}}$.

If $\sigma_\alpha$ is near zero, many group effects contribute little effective flexibility.

Thus:

$$ \text{nominal parameter count} \neq \text{effective parameter count} $$

Multilevel models can include many group effects without incurring the full overfitting cost of no pooling.

---

## 20. Partial pooling as an overfitting solution

The official script varies a fixed hierarchy scale across a sequence of $\sigma_\alpha$ values.

As $\sigma_\alpha$ increases:

- group estimates move closer to raw proportions;
- in-sample fit improves;
- model flexibility increases;
- out-of-sample predictive performance can eventually worsen.

As $\sigma_\alpha$ decreases:

- estimates collapse toward the common mean;
- variance decreases;
- underfitting can increase.

The learned posterior for $\sigma_\alpha$ allows the data and prior to negotiate this trade-off automatically.

---

## 21. Simulation evidence for partial pooling

Chapter 13 simulates ponds with known survival probabilities $p_j^{\mathrm{true}}$ and compares no-pooling versus partial-pooling estimates.

For pond $j$, absolute error is:

$$ E_j = \left| \widehat{p}_j - p_j^{\mathrm{true}} \right| $$

Across simulated groups, partial pooling reduces average error:

$$ \frac{1}{J} \sum_{j=1}^{J} E_j^{\mathrm{partial}} < \frac{1}{J} \sum_{j=1}^{J} E_j^{\mathrm{no\ pool}} $$

especially for small groups.

Partial pooling is not guaranteed to improve every individual group estimate, but it guarantees lower average error over the population of groups.

---

## 22. Adding a group-level predictor

The official script adds predator presence:

$$ P_j = \begin{cases} 1 & \text{predator present} \\ 0 & \text{predator absent} \end{cases} $$

The model is:

$$ S_j \sim \mathrm{Binomial} \left( D_j, p_j \right) $$

$$ \mathrm{logit}(p_j) = \alpha_j + \beta_P P_j $$

with:

$$ \beta_P \sim \mathrm{Normal}(0, 0.5) $$

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

The global predictor $\beta_P$ explains systematic variation due to predators.

The remaining $\alpha_j$ values represent residual tank heterogeneity after accounting for predators.

---

## 23. Explained and residual group variation

The posterior of $\sigma_\alpha$ shrinks after adding predictors.

Without predators:

$$ \sigma_\alpha $$

captures all between-tank variation.

With predators:

$$ \sigma_\alpha $$

captures variation not explained by $P_j$.

A reduction in $\sigma_\alpha$ indicates that predator status explains part of the between-tank variation.

---

## 24. Exchangeability

Partial pooling assumes groups are exchangeable conditional on modeled predictors.

Informally, before observing the group outcomes, group labels carry no additional information beyond their predictors.

Mathematically, the prior is symmetric:

$$ p \left( \alpha_1,\ldots,\alpha_J \mid \bar{\alpha}, \sigma_\alpha \right) = \prod_{j=1}^{J} p \left( \alpha_j \mid \bar{\alpha}, \sigma_\alpha \right) $$

The labels can be permuted without changing the joint prior.

### 24.1 When exchangeability is implausible

A single shared population may be inappropriate when groups differ structurally by country, product class, treatment regime, or measurement system.

These differences should be modeled with predictors, separate populations, or structured covariances.

---

## 25. Hierarchy is a scientific assumption

The statement:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

asserts that groups are generated by a population distribution.

It is a substantive modeling assumption.

The normal population model may be wrong when group effects are multimodal, skewed, heavy-tailed, or spatially/temporally correlated.

---

## 26. Prior predictive checks for a hierarchy

The official script examines priors on $\bar{\alpha}$ and $\sigma_\alpha$, then draws group effects $\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$.

For a logistic model $p_j = \mathrm{logit}^{-1}(\alpha_j)$, prior predictive checks should inspect:

- population-average survival;
- spread of survival probabilities across tanks;
- probability of near-zero or near-one groups;
- replicated survival counts by tank size.

---

## 27. Hyperpriors and near-zero variation

A hierarchy-scale prior might be:

$$ \sigma_\alpha \sim \mathrm{Exponential}(1) $$

This places density near zero while allowing larger values.

Near-zero group variation ($\sigma_\alpha \approx 0$) means $\alpha_j \approx \bar{\alpha}$ (complete pooling).

For logistic models, $\sigma_\alpha$ is in log-odds units. A standard deviation of $\sigma_\alpha = 1.5$ in log-odds implies very wide dispersion on the probability scale.

---

## 28. Centered parameterization

The centered hierarchy is:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

When data per group are weak and $\sigma_\alpha$ can approach zero, the posterior forms Neal's funnel:

- group effects must contract tightly around $\bar{\alpha}$ when $\sigma_\alpha \rightarrow 0$;
- they can spread widely when $\sigma_\alpha$ is large.

This creates sharp curvature that produces MCMC divergences, high autocorrelation, and low ESS.

---

## 29. Non-centered parameterization

The hierarchy is re-parameterized as:

$$ z_j \sim \mathrm{Normal}(0, 1) $$

$$ \alpha_j = \bar{\alpha} + \sigma_\alpha z_j $$

The scientific model is unchanged.

The sampled geometry changes: the standardized offsets $z_j$ are independent of $\sigma_\alpha$ in the prior.

This eliminates the funnel geometry and resolves MCMC divergences for sparse group data.

---

## 30. Centered versus non-centered is computational

Both parameterizations imply identical prior probability distributions for $\alpha_j$.

Therefore, when MCMC samples correctly from both, the target posterior distributions are identical.

The choice is strictly computational: non-centered parameterization excels when data per group are sparse, while centered parameterization can be more efficient when data per group are dense.

---

## 31. Multiple clustering dimensions

A dataset can contain more than one grouping structure.

In the trolley data:

- each response belongs to a participant $U_i$;
- each response also belongs to a story $S_i$.

The cluster dimensions are cross-classified ($U_i \times S_i$).

The model includes two varying intercepts in the same linear predictor:

$$ \phi_i = \text{predictors} + \alpha^{\mathrm{story}}_{S_i} + \alpha^{\mathrm{person}}_{U_i} $$

---

## 32. Trolley model with varying intercepts

The official script fits an ordered-logistic model:

$$ R_i \sim \mathrm{OrderedLogistic}(\phi_i, \mathbf{\alpha}_{\text{cut}}) $$

$$ \phi_i = \mathbf{X}_i^{\mathsf T} \mathbf{\beta} + \alpha^{\mathrm{story}}_{S_i} + \alpha^{\mathrm{person}}_{U_i} $$

$$ \alpha^{\mathrm{story}}_s \sim \mathrm{Normal}(0, \sigma_S), \quad \sigma_S \sim \mathrm{Exponential}(1) $$

$$ \alpha^{\mathrm{person}}_u \sim \mathrm{Normal}(0, \sigma_U), \quad \sigma_U \sim \mathrm{Exponential}(1) $$

---

## 33. What the two trolley hierarchies mean

- $\sigma_S$ measures residual variation among stories after accounting for experimental factors.
- $\sigma_U$ measures persistent variation among participants after accounting for participant predictors.

If $\sigma_U > \sigma_S$, there is greater unexplained heterogeneity among individuals than among scenario stories.

---

## 34. Repeated measures and pseudoreplication

Ignoring participant clustering treats repeated responses from the same person as independent observations.

This creates:

- underestimated uncertainty;
- overconfident treatment effect estimates;
- incorrect weighting of highly observed participants.

Including a participant varying intercept absorbs stable individual propensities and corrects pseudoreplication.

---

## 35. Varying intercepts versus fixed indicators

A fixed-effects indicator model estimates unrestricted parameters per group under broad fixed priors.

A multilevel model learns the population hyperparameters $\bar{\alpha}$ and $\sigma_\alpha$.

The core difference is not parameter count (both have $J$ intercepts), but adaptive shrinkage and shared information across groups.

---

## 36. Full Bayes versus empirical Bayes

Empirical Bayes plugs in point estimates $(\widehat{\bar{\alpha}}, \widehat{\sigma}_\alpha)$ and treats them as known.

Full Bayes retains hyperparameter uncertainty $p(\bar{\alpha}, \sigma_\alpha \mid D)$ and propagates it into posterior predictions.

Plugging in point estimates underestimates uncertainty when $J$ is small.

---

## 37. Prediction for observed groups

For an existing group $j$, posterior predictions draw from the group's specific posterior $\alpha_j^{(s)}$:

$$ S_{\mathrm{new}, j}^{(s)} \sim \mathrm{Binomial} \left( D_{\mathrm{new}, j}, \mathrm{logit}^{-1}\left(\alpha_j^{(s)}\right) \right) $$

This conditions on the group's observed data and partial pooling.

---

## 38. Prediction for new groups

For a new, unobserved group:

1. Draw hyperparameter samples: $\bar{\alpha}^{(s)}, \sigma_\alpha^{(s)}$;
2. Draw a new group intercept: $\alpha_{\mathrm{new}}^{(s)} \sim \mathrm{Normal}\left(\bar{\alpha}^{(s)}, \sigma_\alpha^{(s)}\right)$;
3. Simulate outcomes: $S_{\mathrm{new}}^{(s)} \sim \mathrm{Binomial}\left(D_{\mathrm{new}}, \mathrm{logit}^{-1}\left(\alpha_{\mathrm{new}}^{(s)}\right)\right)$.

New-group uncertainty is substantially wider because it incorporates population heterogeneity $\sigma_\alpha$.

---

## 39. Average-group prediction is not new-group prediction

Setting $\alpha_j = \bar{\alpha}$ describes a hypothetical average group on the linear scale.

It is NOT a prediction for a randomly selected new group.

For non-linear link functions (e.g., logit, log):

$$ \mathrm{logit}^{-1}(\bar{\alpha}) \neq \mathbb{E}_{\alpha_{\mathrm{new}}} \left[ \mathrm{logit}^{-1}(\alpha_{\mathrm{new}}) \right] $$

New-group prediction must integrate over the draw $\alpha_{\mathrm{new}} \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$.

---

## 40. Population-average versus cluster-specific effects

In non-linear GLMMs:

- **Cluster-specific effect:** $\Delta_j = \mathrm{logit}^{-1}(\alpha_j + \beta) - \mathrm{logit}^{-1}(\alpha_j)$
- **Population-average effect:** $\Delta_{\mathrm{pop}} = \mathbb{E}_{\alpha} \left[ \mathrm{logit}^{-1}(\alpha + \beta) - \mathrm{logit}^{-1}(\alpha) \right]$

Because of Jensen's inequality and link-function non-linearity, these two quantities differ.

---

## 41. Multilevel models and causality

Varying intercepts absorb group heterogeneity, but do not resolve group-level treatment confounding if treatment exposure $X_{ij}$ correlates with unobserved group traits.

Causal identification requires explicitly modeling between-group and within-group covariate relationships (e.g., Mundlak centering).

---

## 42. Varying effects are not fixed effects

Econometric fixed-effects models eliminate between-group variation to avoid omitted variable bias at the group level.

Standard multilevel models assume group intercepts are independent of treatment.

When group intercepts correlate with treatment, standard varying-intercept estimates of $\beta$ can be biased unless Mundlak terms ($\bar{X}_j$) are added.

---

## 43. Exchangeability and causal transport

Generalizing predictions to new clusters assumes the target cluster population matches the estimated hyperdistribution $p(\bar{\alpha}, \sigma_\alpha)$.

Transporting estimates to new domains requires including key domain-level predictors in the hyperdistribution.

---

## 44. CRM application: customers with repeated events

In customer relationship management (CRM):

- $i$ indexes transactions; $j$ indexes customers.
- $Y_{ij} \sim \mathrm{Bernoulli}(p_{ij})$, $\mathrm{logit}(p_{ij}) = \alpha_j + \beta_A A_{ij}$.
- $\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$ pools baseline customer conversion.

Customers with few interactions are shrunken toward $\bar{\alpha}$, avoiding extreme zero-or-one propensity estimates.

---

## 45. CRM application: partially pooled campaign effects

When testing $K$ marketing campaigns:

$$ \beta_k \sim \mathrm{Normal}(\bar{\beta}, \sigma_\beta) $$

Sparse campaigns with few impressions are shrunken toward the population campaign effect $\bar{\beta}$, preventing false positives from noisy small campaigns.

---

## 46. CRM application: stores and customers

Cross-classified hierarchy for retail transactions:

$$ \log(\lambda_i) = \mu + \alpha^{\mathrm{customer}}_{C_i} + \alpha^{\mathrm{store}}_{S_i} + X_i^{\mathsf T}\beta $$

Separates customer purchasing power from store-level performance.

---

## 47. CRM application: churn across products

Multi-way cross-classification for customer churn:

$$ \mathrm{logit}(p_{ipt}) = \alpha^{\mathrm{customer}}_i + \alpha^{\mathrm{product}}_p + \alpha^{\mathrm{month}}_t $$

Shares information across rare product subscriptions while tracking macro trends and customer propensities.

---

## 48. CausalOps implications

| B01 concept | CausalOps implication |
|---|---|
| Cluster variable | `ClusterArtifact` |
| Varying intercept | Group-specific latent parameter |
| Population distribution | `HierarchyArtifact` |
| Partial pooling | Adaptive regularization metadata |
| Hyperprior | Explicit population-scale prior |
| Exchangeability | Declared group-population assumption |
| Observed-group prediction | Conditional prediction mode |
| New-group prediction | Population predictive mode |
| Cross-classification | Multiple cluster memberships |
| Non-centered parameterization | Computational specification |

---

## 49. Validating partial pooling

Simulation validation workflow:

1. Draw true group parameters: $\alpha_j^{\mathrm{true}} \sim \mathrm{Normal}(\bar{\alpha}_{\text{true}}, \sigma_{\text{true}})$;
2. Simulate group sample sizes $N_j$ (including sparse groups);
3. Simulate outcomes $Y_{ij} \sim p(Y \mid \alpha_j^{\mathrm{true}})$;
4. Fit complete-pooling, no-pooling, and partial-pooling models;
5. Compute RMSE: $\sqrt{\frac{1}{J}\sum (\widehat{\alpha}_j - \alpha_j^{\mathrm{true}})^2}$;
6. Verify that partial pooling achieves minimal total RMSE.

---

## 50. Common errors exposed by B01

1. **Ignoring clustering:** Treating repeated measures as independent observations.
2. **Complete pooling assumption:** Assuming no differences between groups.
3. **No pooling overfitting:** Estimating separate parameters for sparse groups without shrinkage.
4. **Misunderstanding shrinkage:** Viewing shrinkage as an artifact rather than variance reduction.
5. **Assuming uniform shrinkage:** Expecting all groups to shrink by equal amounts regardless of $N_j$.
6. **Misinterpreting scale of $\sigma_\alpha$:** Interpreting log-odds standard deviations as probability standard deviations.
7. **Ignoring hyperparameter uncertainty:** Using plug-in empirical Bayes point estimates.
8. **Assuming exchangeability without checking:** Grouping heterogeneous sub-populations into a single hierarchy.
9. **Zeroing out new-group effects:** Setting $\alpha_{\text{new}} = \bar{\alpha}$ instead of drawing from the hyperdistribution.
10. **Expecting sample means from sparse groups:** Expecting shrunken estimates to equal sample averages.
11. **Parameter count confusion:** Equating number of nominal parameters with degrees of freedom.
12. **Ignoring MCMC funnel divergences:** Sampling centered hierarchies with sparse data without non-centering.
13. **Confounding varying intercepts with causal identification:** Expecting varying intercepts to fix treatment selection bias.
14. **Conflating observed and new group prediction:** Using conditional predictions for unobserved groups.

---

## 51. Code Demonstrations & Multi-Framework Implementation

### 51.1 Café Golem (R rethinking/ulam, Stan, PyMC)

#### R rethinking / ulam

```{r cafe-golem-ulam}
library(rethinking)

# Generate synthetic café data
set.seed(7)
n_cafes <- 5
w_true <- c(3, 15, 3, 1, 3)
n_visits <- 20
cafe_id <- c(rep(1, 1), rep(2, 1), 1, rep(c(1, 3, 4, 5), times = 4), 2)
y_obs <- rpois(n_visits, w_true[cafe_id]) + 1

dat_cafe <- list(
  N = n_visits,
  n_cafes = n_cafes,
  y = as.integer(y_obs),
  cafe = as.integer(cafe_id)
)

# Non-centered Poisson hierarchy in ulam
m_cafe_ulam <- ulam(
  alist(
    y ~ dpois(lambda),
    log(lambda) <- a_bar + z[cafe] * sigma,
    z[cafe] ~ dnorm(0, 1),
    a_bar ~ dnorm(2, 0.5),
    sigma ~ dexp(10)
  ),
  data = dat_cafe,
  chains = 4,
  cores = 4,
  iter = 2000
)

precis(m_cafe_ulam, depth = 2)
```

#### Stan (CmdStanR)

```stan
// cafe_golem.stan
data {
  int<lower=1> N;
  int<lower=1> n_cafes;
  array[N] int<lower=1, upper=n_cafes> cafe;
  array[N] int<lower=0> y;
}
parameters {
  real a_bar;
  real<lower=0> sigma;
  vector[n_cafes] z;
}
transformed parameters {
  vector[n_cafes] a = a_bar + z * sigma;
}
model {
  a_bar ~ normal(2, 0.5);
  sigma ~ exponential(10);
  z ~ normal(0, 1);
  
  for (i in 1:N) {
    y[i] ~ poisson_log(a[cafe[i]]);
  }
}
generated quantities {
  vector[N] log_lik;
  for (i in 1:N) {
    log_lik[i] = poisson_log_lpmf(y[i] | a[cafe[i]]);
  }
}
```

#### PyMC (Python)

```python
import pymc as pm
import numpy as np
import arviz as az

# Data setup
n_cafes = 5
n_visits = 20
cafe_id = np.array([0, 1, 0, 0, 2, 3, 4, 0, 2, 3, 4, 0, 2, 3, 4, 0, 2, 3, 4, 1])
y_obs = np.array([4, 18, 5, 3, 2, 4, 4, 3, 2, 3, 3, 4, 2, 4, 3, 4, 3, 4, 4, 15])

with pm.Model() as cafe_model:
    # Hyperpriors
    a_bar = pm.Normal("a_bar", mu=2.0, sigma=0.5)
    sigma = pm.Exponential("sigma", lam=10.0)
    
    # Non-centered parameterization
    z = pm.Normal("z", mu=0, sigma=1, shape=n_cafes)
    a = pm.Deterministic("a", a_bar + z * sigma)
    
    # Likelihood
    lambda_ = pm.math.exp(a[cafe_id])
    y = pm.Poisson("y", mu=lambda_, observed=y_obs)
    
    idata_cafe = pm.sample(draws=1000, tune=1000, chains=4, random_seed=7)

print(az.summary(idata_cafe, var_names=["a_bar", "sigma", "a"]))
```

---

### 51.2 Reed-Frog Multilevel Model (R ulam, Stan, PyMC)

#### R rethinking / ulam

```{r reedfrog-multilevel}
library(rethinking)
data(reedfrogs)
d <- reedfrogs
d$tank <- 1:nrow(d)

dat_frog <- list(
  S = d$surv,
  D = d$density,
  T = d$tank,
  P = ifelse(d$pred == "pred", 1L, 0L)
)

# Multilevel varying intercept model
m_frog_ulam <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T],
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

# Multilevel model with predator predictor
m_frog_pred_ulam <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T] + bP * P,
    bP ~ dnorm(0, 0.5),
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

compare(m_frog_ulam, m_frog_pred_ulam, func = PSIS)
```

#### Stan (CmdStanR)

```stan
// reedfrogs_multilevel.stan
data {
  int<lower=1> N_tanks;
  array[N_tanks] int<lower=0> S;
  array[N_tanks] int<lower=1> D;
  array[N_tanks] int<lower=0, upper=1> P;
}
parameters {
  real a_bar;
  real<lower=0> sigma;
  real bP;
  vector[N_tanks] z;
}
transformed parameters {
  vector[N_tanks] a = a_bar + z * sigma;
}
model {
  a_bar ~ normal(0, 1.5);
  sigma ~ exponential(1);
  bP ~ normal(0, 0.5);
  z ~ normal(0, 1);
  
  for (i in 1:N_tanks) {
    S[i] ~ binomial_logit(D[i], a[i] + bP * P[i]);
  }
}
generated quantities {
  vector[N_tanks] p;
  for (i in 1:N_tanks) {
    p[i] = inv_logit(a[i] + bP * P[i]);
  }
}
```

#### PyMC (Python)

```python
import pymc as pm
import pandas as pd
import numpy as np

# Load reedfrogs data equivalent
df_frogs = pd.DataFrame({
    'density': [10]*16 + [20]*16 + [35]*16,
    'surv': [9, 10, 7, 10, 9, 9, 10, 9, 4, 9, 7, 6, 7, 5, 9, 9,
             19, 14, 18, 19, 17, 18, 19, 16, 11, 17, 14, 13, 8, 8, 14, 13,
             32, 33, 31, 33, 32, 34, 33, 32, 13, 23, 27, 24, 20, 20, 22, 23],
    'pred': [0]*8 + [1]*8 + [0]*8 + [1]*8 + [0]*8 + [1]*8
})

n_tanks = len(df_frogs)
tank_id = np.arange(n_tanks)

with pm.Model() as frog_model:
    a_bar = pm.Normal("a_bar", mu=0, sigma=1.5)
    sigma = pm.Exponential("sigma", lam=1.0)
    bP = pm.Normal("bP", mu=0, sigma=0.5)
    
    z = pm.Normal("z", mu=0, sigma=1, shape=n_tanks)
    a = pm.Deterministic("a", a_bar + z * sigma)
    
    p = pm.Deterministic("p", pm.math.invlogit(a + bP * df_frogs['pred'].values))
    
    S = pm.Binomial("S", n=df_frogs['density'].values, p=p, observed=df_frogs['surv'].values)
    
    idata_frog = pm.sample(draws=1000, tune=1000, chains=4, random_seed=42)
```

---

### 51.3 Centered vs Non-Centered Funnel Demonstration

```{r funnel-demo}
library(rethinking)

# Centered hierarchy simulation
set.seed(42)
J <- 50
dat_funnel <- list(
  J = J,
  y = rnorm(J, 0, 0.1),
  N = rep(2, J)
)

# Centered parameterization (prone to divergences when data are sparse)
m_centered <- ulam(
  alist(
    y ~ dnorm(a, 1),
    a[J] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dat_funnel,
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.8)
)

# Non-centered parameterization (stable sampler geometry)
m_noncentered <- ulam(
  alist(
    y ~ dnorm(a, 1),
    a[J] <- a_bar + z[J] * sigma,
    z[J] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dat_funnel,
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.8)
)

# Inspect divergence comparison
cat("Centered divergences:", sum(get_divergent_iterations(m_centered@stanfit)), "\n")
cat("Non-centered divergences:", sum(get_divergent_iterations(m_noncentered@stanfit)), "\n")
```

---

### 51.4 Trolley Cross-Classified Varying Intercepts Model

```{r trolley-cross-classified}
library(rethinking)
data(Trolley)
d <- Trolley

dat_trolley <- list(
  R = d$response,
  A = d$action,
  I = d$intention,
  C = d$contact,
  S = as.integer(d$story),
  U = as.integer(d$id)
)

# Non-centered cross-classified model
m_trolley_nc <- ulam(
  alist(
    R ~ ordered_logistic(phi, cutpoints),
    phi <- bA * A + bI * I + bC * C + zs[S] * sigma_S + zu[U] * sigma_U,
    cutpoints ~ dnorm(0, 1.5),
    c(bA, bI, bC) ~ dnorm(0, 0.5),
    zs[S] ~ dnorm(0, 1),
    zu[U] ~ dnorm(0, 1),
    sigma_S ~ dexp(1),
    sigma_U ~ dexp(1)
  ),
  data = dat_trolley,
  chains = 4,
  cores = 4,
  sample = FALSE
)
```

---

### 51.5 Posterior Predictive Simulation: Same-Cluster vs New-Cluster

```{r posterior-predictive-simulation}
library(rethinking)

# Extract samples from fitted frog model
post <- extract.samples(m_frog_ulam)
n_samples <- length(post$a_bar)

# 1. Prediction for existing tank j=1 (density = 10)
p_existing_tank1 <- inv_logit(post$a[, 1])
surv_sim_existing <- rbinom(n_samples, size = 10, prob = p_existing_tank1)

# 2. Prediction for a NEW tank (density = 10)
a_new_tank <- rnorm(n_samples, mean = post$a_bar, sd = post$sigma)
p_new_tank <- inv_logit(a_new_tank)
surv_sim_new <- rbinom(n_samples, size = 10, prob = p_new_tank)

# Contrast predictive interval widths
cat("Existing tank 1 89% PI survival count:", HPDI(surv_sim_existing, prob = 0.89), "\n")
cat("New tank 89% PI survival count:", HPDI(surv_sim_new, prob = 0.89), "\n")
```

---

## 52. Homework Exercises and Complete Solutions

### Exercise 13M1

**Question:** Formulate a varying-intercept model for the reed frogs data including both density ($D$) and predation ($P$) as predictors. Write down the complete generative probability model including likelihood, link function, group distribution, and hyperpriors.

**Solution:**

Let $S_j$ be the number of surviving tadpoles in tank $j$ ($j = 1, \dots, 48$) out of initial density $D_j$. Let $P_j \in \{0, 1\}$ denote predator presence.

The mathematical model is:

$$ S_j \sim \mathrm{Binomial}(D_j, p_j) $$

$$ \mathrm{logit}(p_j) = \alpha_j + \beta_P P_j + \beta_D D_j^* $$

where $D_j^*$ is standardized density ($D_j^* = (D_j - \bar{D}) / s_D$).

The group distribution for residual tank intercepts is:

$$ \alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha) $$

The hyperpriors and prior specifications are:

$$ \bar{\alpha} \sim \mathrm{Normal}(0, 1.5) $$

$$ \sigma_\alpha \sim \mathrm{Exponential}(1) $$

$$ \beta_P \sim \mathrm{Normal}(0, 0.5) $$

$$ \beta_D \sim \mathrm{Normal}(0, 0.5) $$

---

### Exercise 13M2

**Question:** Write down the `ulam` formula list for the model in 13M1. Fit the model and compare its PSIS/WAIC score and estimated $\sigma_\alpha$ with the model containing only predation (`mSTP`).

**Solution:**

```{r solution-13m2}
library(rethinking)
data(reedfrogs)
d <- reedfrogs
d$tank <- 1:nrow(d)
d$D_std <- standardize(d$density)
d$P <- ifelse(d$pred == "pred", 1L, 0L)

dat_13m2 <- list(
  S = d$surv,
  D = d$density,
  D_std = d$D_std,
  P = d$P,
  T = d$tank
)

m_13m2 <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T] + bP * P + bD * D_std,
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1),
    bP ~ dnorm(0, 0.5),
    bD ~ dnorm(0, 0.5)
  ),
  data = dat_13m2,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

# Fit mSTP (predator only)
m_STP <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T] + bP * P,
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1),
    bP ~ dnorm(0, 0.5)
  ),
  data = dat_13m2,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

# Compare models
compare(m_STP, m_13m2, func = PSIS)

# Compare residual standard deviations
precis(m_STP, pars = "sigma")
precis(m_13m2, pars = "sigma")
```

**Interpretation:** Density is already accounted for in the binomial likelihood through trial count $D_j$. Adding density as a linear predictor in log-odds tests whether per-tadpole survival probability varies systematically with tank density. If $\sigma_\alpha$ changes little, initial density has minor additional log-odds effects beyond specifying the denominator.

---

### Exercise 13M3

**Question:** Re-estimate the varying-intercept reed frog model `mST` using a Half-Cauchy hyperprior on $\sigma_\alpha$ instead of Exponential: $\sigma_\alpha \sim \mathrm{HalfCauchy}(0, 1)$. Compare the posterior distribution of $\sigma_\alpha$ under both hyperpriors.

**Solution:**

```{r solution-13m3}
library(rethinking)
data(reedfrogs)
d <- reedfrogs
d$tank <- 1:nrow(d)

dat_frog <- list(
  S = d$surv,
  D = d$density,
  T = d$tank
)

# Model with Exponential(1)
m_exp <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T],
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4
)

# Model with Half-Cauchy(0, 1)
m_cauchy <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T],
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dcauchy(0, 1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4
)

plot(coeftab(m_exp, m_cauchy), pars = "sigma")
```

**Analytical explanation:** With 48 tanks, there is abundant data to estimate $\sigma_\alpha$. Consequently, the posterior of $\sigma_\alpha$ is dominated by the likelihood and changes very little between Exponential(1) and Half-Cauchy(0,1) priors. Half-Cauchy has a heavier tail, allowing very large $\sigma_\alpha$ when data are sparse, whereas Exponential concentrates more mass near zero.

---

### Exercise 13M4

**Question:** Fit a non-centered version of `mST` and compare sampling efficiency (number of effective samples `n_eff` and $\widehat{R}$) between the centered and non-centered parameterizations.

**Solution:**

```{r solution-13m4}
library(rethinking)
data(reedfrogs)
d <- reedfrogs
d$tank <- 1:nrow(d)

dat_frog <- list(
  S = d$surv,
  D = d$density,
  T = d$tank
)

# Centered model
m_centered <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T],
    a[T] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4
)

# Non-centered model
m_noncentered <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a_bar + z[T] * sigma,
    z[T] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_frog,
  chains = 4,
  cores = 4
)

# Compare n_eff for hyperparameters
precis(m_centered, pars = c("a_bar", "sigma"))
precis(m_noncentered, pars = c("a_bar", "sigma"))
```

---

### Exercise 13H1

**Question:** Using the `bangladesh` dataset in `rethinking`, fit a varying-intercept logistic regression model for baseline contraceptive use (`use`) across districts (`district`). Fit both a no-pooling model and a multilevel partial-pooling model. Plot district-level posterior probability predictions and describe the observed shrinkage.

**Solution:**

```{r solution-13h1}
library(rethinking)
data(bangladesh)
d <- bangladesh
d$district_id <- as.integer(as.factor(d$district))

dat_bangla <- list(
  use = d$use,
  district = d$district_id,
  N_districts = max(d$district_id)
)

# 1. No-pooling model
m_nopool <- ulam(
  alist(
    use ~ dbinom(1, p),
    logit(p) <- a[district],
    a[district] ~ dnorm(0, 1.5)
  ),
  data = dat_bangla,
  chains = 4,
  cores = 4
)

# 2. Multilevel partial-pooling model
m_partpool <- ulam(
  alist(
    use ~ dbinom(1, p),
    logit(p) <- a_bar + z[district] * sigma,
    z[district] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat_bangla,
  chains = 4,
  cores = 4
)

# Extract estimates and calculate shrinkage
post_np <- extract.samples(m_nopool)
post_pp <- extract.samples(m_partpool)

p_np <- apply(inv_logit(post_np$a), 2, mean)
p_pp <- apply(inv_logit(post_pp$a_bar + post_pp$z * post_pp$sigma), 2, mean)

# Plot raw vs shrunken
plot(p_np, p_pp, xlab = "No-pooling District Mean", ylab = "Partial-pooling District Mean",
     pch = 16, col = 4, xlim = c(0, 1), ylim = c(0, 1))
abline(a = 0, b = 1, lty = 2)
```

#### PyMC Code for 13H1

```python
import pymc as pm
import pandas as pd
import numpy as np

# PyMC Multilevel Bangladesh Model
# assuming df_bangladesh loaded with columns 'use' and 'district_id'
n_districts = df_bangladesh['district_id'].nunique()
district_idx = df_bangladesh['district_id'].values

with pm.Model() as bangladesh_pymc:
    a_bar = pm.Normal("a_bar", mu=0, sigma=1.5)
    sigma = pm.Exponential("sigma", lam=1.0)
    
    z = pm.Normal("z", mu=0, sigma=1, shape=n_districts)
    a = pm.Deterministic("a", a_bar + z * sigma)
    
    p = pm.Deterministic("p", pm.math.invlogit(a[district_idx]))
    
    use = pm.Bernoulli("use", p=p, observed=df_bangladesh['use'].values)
    
    idata_bangla = pm.sample(draws=1000, tune=1000, chains=4)
```

---

### Exercise 13H2

**Question:** Extend the Bangladesh model to include urban indicator $U_i$ (`urban`) and centered age ($A_i$). Compare fixed-slope versus varying-slope hypotheses across districts.

**Solution:**

```{r solution-13h2}
library(rethinking)
data(bangladesh)
d <- bangladesh
d$district_id <- as.integer(as.factor(d$district))
d$age_std <- standardize(d$age.center)

dat_13h2 <- list(
  use = d$use,
  district = d$district_id,
  urban = d$urban,
  age = d$age_std
)

# Varying intercept model with urban and age predictors
m_13h2 <- ulam(
  alist(
    use ~ dbinom(1, p),
    logit(p) <- a_bar + z[district] * sigma + bU * urban + bA * age,
    z[district] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1),
    bU ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5)
  ),
  data = dat_13h2,
  chains = 4,
  cores = 4
)

precis(m_13h2)
```

---

## 53. Lecture takeaways

1. Multilevel models estimate group effects and their population distribution jointly.
2. Complete pooling assumes no residual group heterogeneity.
3. No pooling treats groups as unrelated.
4. Partial pooling shares information adaptively.
5. Varying effects are group-specific parameters with learned population priors.
6. Shrinkage is stronger for weakly informed groups.
7. Shrinkage is weaker when the population is genuinely heterogeneous.
8. Adaptive priors learn the regularization strength from all groups.
9. Partial pooling usually improves average group prediction, especially for small groups.
10. It need not improve every individual group estimate.
11. The café example shows information propagating between groups through hyperparameters.
12. The tadpole model uses binomial observations and varying log-odds intercepts.
13. Adding predictors can explain part of the group-level variance.
14. Nominal parameter count differs from effective complexity.
15. Exchangeability is a scientific population assumption.
16. A centered hierarchy can create funnel geometry.
17. Non-centered parameterization often improves MCMC efficiency.
18. Multiple cluster types can be modeled simultaneously.
19. Participant and story effects in the trolley model are cross-classified.
20. Predictions for observed groups differ from predictions for new groups.
21. An average-group prediction is not a new-group prediction.
22. Nonlinear links make population-average and cluster-specific effects differ.
23. Multilevel structure improves modeling but does not itself identify causal effects.
24. The official B01 material maps to Chapter 13 of the uploaded second edition, despite the schedule's Chapter 12 label.

---

## 54. Review questions

1. What distinguishes complete, no, and partial pooling?
2. Why is a varying intercept different from an unrelated group indicator?
3. What makes the prior on $\alpha_j$ adaptive?
4. How do group sample size and $\sigma_\alpha$ affect shrinkage?
5. Why can biased shrinkage estimates have lower mean squared error?
6. How can one café's observation affect another café's posterior?
7. What is the difference between a café-specific parameter and the café population distribution?
8. What is the reed-frog likelihood?
9. Why do small tanks shrink more strongly?
10. Why do raw proportions of zero and one not imply true probabilities of zero and one?
11. What does $\sigma_\alpha$ control in the tadpole model?
12. Why can a model with 48 varying intercepts have far fewer than 48 effective parameters?
13. How does adding predator status change the meaning of tank variation?
14. What does exchangeability mean?
15. Which kinds of group differences can violate exchangeability?
16. What is the centered hierarchy?
17. How does the non-centered form represent the same model?
18. When is non-centering often advantageous?
19. What two clustering dimensions occur in the trolley model?
20. Why does ignoring participant clustering create pseudoreplication?
21. How does full Bayes differ from empirical Bayes?
22. How is prediction for an observed group constructed?
23. How is prediction for a new group constructed?
24. Why is setting a new group's effect to zero incorrect?
25. Why can cluster-specific and population-average probability effects differ?
26. Why does a varying intercept not automatically solve treatment confounding?
27. How does a random-intercept model differ causally from a unit fixed-effects design?
28. Which metadata should CausalOps store about a hierarchy?
29. How would you validate partial pooling by simulation?
30. Why is the official Chapter 12 label a mismatch for the uploaded second edition?

---

## 55. Book reading guide

### Official assigned reading

- Chapter 12, according to the 2026 schedule.

### Closest second-edition reading

- Chapter 13, *Models With Memory*
  - Section 13.1, *Example: Multilevel tadpoles*
  - Section 13.2, *Varying effects and the underfitting/overfitting trade-off*
  - Section 13.3, *More than one type of cluster*
  - Section 13.4, *Divergent transitions and non-centered priors*
  - Section 13.5, *Multilevel posterior predictions*

### Read before B01

Focus on:

- A04 categorical index models;
- A07 regularization and overfitting;
- A08 HMC diagnostics;
- A09 binomial and ordered-logistic likelihoods;
- the distinction between parameter uncertainty and predictive uncertainty.

---

## 56. Official code guide

### `12_intro_multilevel_models.r`

The script contains four principal blocks.

#### Trolley heterogeneity preview

Plots response distributions across stories and participants to motivate multiple cluster dimensions.

#### Café golem

Simulates café waiting-time counts, defines a non-centered Poisson hierarchy, updates sequentially, and demonstrates cross-group information propagation.

#### Reed-frog tanks

Displays raw survival proportions, fits a multilevel binomial model, compares learned and fixed pooling strength, plots partially shrunken estimates, and adds predator presence.

#### Full trolley model

Fits an ordered-logistic model with experimental predictors, varying story intercepts, and varying participant intercepts in centered and non-centered parameterizations.

---

## 57. Sources

- Richard McElreath, *Statistical Rethinking: A Bayesian Course with Examples in R and Stan*, second edition, Chapter 13.
- Official `stat_rethinking_2026` repository and B01 lecture listing.
- Official script `12_intro_multilevel_models.r`.

---

## 60. Annex: Complete Solutions to Review Questions

#### 1. What distinguishes complete, no, and partial pooling?

- **Complete pooling** assumes all groups share a single common parameter ($\alpha_j = \alpha$). It ignores all group variation, yielding low variance but severe underfitting if groups genuinely differ.
- **No pooling** estimates an independent parameter $\alpha_j$ for each group under fixed broad priors without sharing information across groups. It matches sample means closely but overfits small/sparse groups and yields high variance.
- **Partial pooling** treats group parameters as draws from a common population distribution $\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$ whose hyperparameters are estimated from the data. It adaptively balances group-specific data against population evidence.

#### 2. Why is a varying intercept different from an unrelated group indicator?

An unrelated group indicator assigns each group an independent parameter under a fixed, non-adaptive prior (no pooling). A varying intercept models group parameters as coming from a learned population distribution $\mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$. The adaptive population distribution shrinks estimates toward $\bar{\alpha}$ proportionally to group uncertainty, sharing information across groups and enabling valid predictions for unobserved groups.

#### 3. What makes the prior on $\alpha_j$ adaptive?

The prior $\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$ is adaptive because its location $\bar{\alpha}$ and scale $\sigma_\alpha$ are unknown hyperparameters estimated jointly from all groups. As data accumulate from any group, estimates of $\bar{\alpha}$ and $\sigma_\alpha$ update, which in turn updates the prior distribution governing every individual group $\alpha_j$.

#### 4. How do group sample size and $\sigma_\alpha$ affect shrinkage?

The shrinkage weight $w_j = \frac{\sigma_\alpha^2}{\sigma_\alpha^2 + V_j}$ determines how much weight is given to group $j$'s raw sample estimate $\widehat{\alpha}_j$ versus the population mean $\bar{\alpha}$:
- **Group sample size ($N_j$):** Larger sample size reduces sampling variance $V_j$, so $w_j \to 1$ (less shrinkage). Smaller $N_j$ increases $V_j$, so $w_j \to 0$ (stronger shrinkage toward $\bar{\alpha}$).
- **Population variance ($\sigma_\alpha^2$):** Larger $\sigma_\alpha^2$ means groups are genuinely heterogeneous, so $w_j \to 1$ (less shrinkage). Smaller $\sigma_\alpha^2$ means groups are homogeneous, so $w_j \to 0$ (stronger shrinkage).

#### 5. Why can biased shrinkage estimates have lower mean squared error?

Mean squared error decomposes into $\mathrm{MSE} = \mathrm{Bias}^2 + \mathrm{Variance}$. No-pooling estimates are unbiased for individual group means but have high variance for small samples. Shrinkage introduces a small amount of bias toward $\bar{\alpha}$, but drastically reduces estimation variance. When the variance reduction exceeds the squared bias, total MSE decreases substantially.

#### 6. How can one café's observation affect another café's posterior?

Observations from Café A update the posterior distributions of the population hyperparameters $\bar{\alpha}$ and $\sigma_\alpha$. Because Café B's prior is $\alpha_B \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$, any shift in $\bar{\alpha}$ or $\sigma_\alpha$ directly updates Café B's prior and posterior, even if Café B has no direct observations.

#### 7. What is the difference between a café-specific parameter and the café population distribution?

- A **café-specific parameter** $\alpha_j$ represents the latent log-wait time of a specific observed café $j$.
- The **café population distribution** $\mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$ describes the distribution of log-wait times across the entire infinite population of potential cafés from which the sampled cafés were drawn.

#### 8. What is the reed-frog likelihood?

The reed-frog observation model uses a Binomial likelihood:

$$ S_j \sim \mathrm{Binomial}(D_j, p_j) $$

where $S_j$ is the count of surviving tadpoles in tank $j$, $D_j$ is the initial density (number of trials), and $p_j = \mathrm{logit}^{-1}(\alpha_j)$ is the tank-specific survival probability.

#### 9. Why do small tanks shrink more strongly?

Small tanks ($D_j = 10$) have larger sampling variance $V_j \approx \frac{1}{D_j p_j (1-p_j)}$ on the logit scale compared to large tanks ($D_j = 35$). In the partial pooling weighting formula $w_j = \frac{\sigma_\alpha^2}{\sigma_\alpha^2 + V_j}$, larger $V_j$ yields a smaller $w_j$, causing estimates for small tanks to shrink more strongly toward the population mean $\bar{\alpha}$.

#### 10. Why do raw proportions of zero and one not imply true probabilities of zero and one?

In small samples ($D_j = 10$), observing $S_j = 0$ or $S_j = 10$ is easily produced by sampling variation even if the true probability $p_j$ is 0.15 or 0.85. On the log-odds scale, $p_j = 0$ corresponds to $\alpha_j = -\infty$ and $p_j = 1$ to $\alpha_j = +\infty$. Binomial likelihood uncertainty combined with the population prior prevents estimates from collapsing to infinity.

#### 11. What does $\sigma_\alpha$ control in the tadpole model?

$\sigma_\alpha$ measures the standard deviation of tank baseline survival propensities on the log-odds scale. It controls the degree of adaptive regularization: when $\sigma_\alpha \to 0$, all tanks are forced to share the same survival rate (complete pooling); when $\sigma_\alpha \to \infty$, tanks are estimated independently (no pooling).

#### 12. Why can a model with 48 varying intercepts have far fewer than 48 effective parameters?

Shrinkage restricts the freedom of the 48 intercepts by constraining them to lie within the population distribution $\mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$. When $\sigma_\alpha$ is small, the 48 parameters are pulled tightly together and vary mostly in unison, functioning effectively as a single parameter plus a scale term. Thus, $p_{\mathrm{WAIC}}$ or $p_{\mathrm{LOO}}$ is much smaller than 48.

#### 13. How does adding predator status change the meaning of tank variation?

In the intercept-only model $\mathrm{logit}(p_j) = \alpha_j$, $\sigma_\alpha$ measures *total* between-tank variation. When adding predator presence $\mathrm{logit}(p_j) = \alpha_j + \beta_P P_j$, $\alpha_j$ represents residual tank intercept after controlling for predators, and $\sigma_\alpha$ measures *unexplained residual* between-tank variation. The decrease in $\sigma_\alpha$ quantifies how much tank heterogeneity is explained by predation.

#### 14. What does exchangeability mean?

Exchangeability means that the joint prior distribution of group parameters $p(\alpha_1, \dots, \alpha_J)$ is invariant to permutations of the group indices $1, \dots, J$. Informally, it assumes that before observing outcomes, we have no background information distinguishing one group from another.

#### 15. Which kinds of group differences can violate exchangeability?

Exchangeability is violated when groups possess known structural ordering or grouping that is omitted from the model—such as spatial location, time of measurement, hierarchical nesting (tanks within ponds), or known group-level covariates (e.g., predator presence, country developmental status).

#### 16. What is the centered hierarchy?

The centered hierarchy parameters specifies group parameters directly from the population distribution:

$$ \alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha) $$

The sampler updates $\alpha_j$ conditional on $\bar{\alpha}$ and $\sigma_\alpha$.

#### 17. How does the non-centered form represent the same model?

The non-centered parameterization defines standardized parameters $z_j \sim \mathrm{Normal}(0, 1)$ and computes group intercepts deterministically:

$$ \alpha_j = \bar{\alpha} + \sigma_\alpha z_j $$

Because a linear transformation of a standard normal yields $\mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$, the joint prior density $p(\alpha, \bar{\alpha}, \sigma_\alpha)$ is mathematically identical.

#### 18. When is non-centering often advantageous?

Non-centering is advantageous when sample size per group ($N_j$) is small or data are sparse. Under centered parameterization, as $\sigma_\alpha \to 0$, the posterior forms a narrow geometry (Neal's funnel) that creates severe MCMC divergences. Non-centering decouples $z_j$ from $\sigma_\alpha$ in the proposal geometry, eliminating the funnel and resolving divergences.

#### 19. What two clustering dimensions occur in the trolley model?

The trolley dataset features cross-classified clustering across:
1. **Participants ($U_i$):** Repeated scenario evaluations by the same individual.
2. **Stories ($S_i$):** Repeated scenario evaluations using the same narrative script.

#### 20. Why does ignoring participant clustering create pseudoreplication?

Ignoring participant clustering treats multiple answers from one person as independent samples of human moral judgment. This inflates the nominal sample size, artificially shrinks standard errors, produces overconfident $p$-values/credible intervals, and misattributes individual personality bias to experimental factors.

#### 21. How does full Bayes differ from empirical Bayes?

- **Empirical Bayes** estimates hyperparameters $(\widehat{\bar{\alpha}}, \widehat{\sigma}_\alpha)$ via point optimization (e.g., marginal maximum likelihood) and plugs them in as fixed constants when estimating $\alpha_j$.
- **Full Bayes** assigns hyperpriors to $\bar{\alpha}$ and $\sigma_\alpha$ and integrates over their full posterior distribution $p(\bar{\alpha}, \sigma_\alpha \mid D)$, fully propagating hyperparameter uncertainty into group-level and predictive distributions.

#### 22. How is prediction for an observed group constructed?

For an observed group $j$, draw posterior samples of that specific group's intercept $\alpha_j^{(s)}$ from the MCMC chain, and evaluate the observation model:

$$ Y_{\mathrm{new}, j}^{(s)} \sim p\left( Y \mid \alpha_j^{(s)}, X \right) $$

This incorporates both hyperparameter uncertainty and specific knowledge about group $j$.

#### 23. How is prediction for a new group constructed?

For an unobserved new group:
1. Draw hyperparameter samples $(\bar{\alpha}^{(s)}, \sigma_\alpha^{(s)})$ from posterior draw $s$.
2. Simulate a new group intercept: $\alpha_{\mathrm{new}}^{(s)} \sim \mathrm{Normal}\left( \bar{\alpha}^{(s)}, \sigma_\alpha^{(s)} \right)$.
3. Simulate outcomes: $Y_{\mathrm{new}}^{(s)} \sim p\left( Y \mid \alpha_{\mathrm{new}}^{(s)}, X \right)$.

This incorporates population-level variance $\sigma_\alpha$.

#### 24. Why is setting a new group's effect to zero incorrect?

Setting $\alpha_{\mathrm{new}} = \bar{\alpha}$ (or 0) assumes the new group is exactly average, ignoring between-group variance $\sigma_\alpha$. This severely underestimates prediction uncertainty for new groups and, for non-linear link functions, gives a biased expectation of outcomes due to non-linear transformation.

#### 25. Why can cluster-specific and population-average probability effects differ?

Because the logit transformation $\mathrm{logit}^{-1}(x) = \frac{1}{1 + e^{-x}}$ is non-linear (convex for $x < 0$, concave for $x > 0$), Jensen's inequality implies:

$$ \mathbb{E}\left[ \mathrm{logit}^{-1}(\bar{\alpha} + \beta + \epsilon) \right] \neq \mathrm{logit}^{-1}\left( \bar{\alpha} + \beta \right) $$

A cluster-specific effect evaluates probability change for a fixed $\alpha_j$, whereas a population-average effect averages probability changes across the entire distribution of group intercepts $\alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha)$.

#### 26. Why does a varying intercept not automatically solve treatment confounding?

If treatment assignment $X_{ij}$ depends on unobserved group characteristics correlated with $\alpha_j$ (e.g., higher-income districts receiving more health clinics), the model $\mathrm{logit}(p_{ij}) = \alpha_j + \beta X_{ij}$ assumes $\alpha_j \perp X_{ij}$ in the prior hierarchy. This causes the group intercept to absorb part of the treatment effect or vice versa, resulting in biased estimates of $\beta$.

#### 27. How does a random-intercept model differ causally from a unit fixed-effects design?

- A **unit fixed-effects design** includes unrestricted group indicators, effectively conditioning on group identity and estimating $\beta$ exclusively from within-group variation. It is immune to time-invariant between-group confounding.
- A **standard random-intercept model** pools between-group and within-group variation. If between-group variation is confounded, the random-intercept estimate of $\beta$ is biased unless group-level covariate means $\bar{X}_j$ are explicitly included (Mundlak parameterization).

#### 28. Which metadata should CausalOps store about a hierarchy?

CausalOps should store:
- `cluster_id` and indexing columns;
- `observation_unit` and repeated measures identifiers;
- `exchangeability_assumptions` and conditioning variables;
- `hierarchy_family` (e.g., Normal, Student-t);
- `parameterization` (centered vs non-centered);
- `hyperprior_specifications` for mean and scale terms;
- `prediction_mode` rules (conditional vs new-cluster).

#### 29. How would you validate partial pooling by simulation?

1. Define a true population distribution $\alpha_j^{\mathrm{true}} \sim \mathrm{Normal}(\bar{\alpha}_{\mathrm{true}}, \sigma_{\text{true}})$.
2. Generate $J$ groups with varying sample sizes $N_j$, including very small $N_j$.
3. Simulate outcomes $Y_{ij} \sim p(Y \mid \alpha_j^{\mathrm{true}})$.
4. Fit complete-pooling, no-pooling, and partial-pooling models to the synthetic data.
5. Compute estimation error $\mathrm{RMSE} = \sqrt{\frac{1}{J}\sum (\widehat{\alpha}_j - \alpha_j^{\mathrm{true}})^2}$ and 95% posterior interval coverage for all three models.
6. Verify that partial pooling achieves minimal total RMSE and correct coverage across sample sizes.

#### 30. Why is the official Chapter 12 label a mismatch for the uploaded second edition?

The 2026 course outline uses a draft 3rd edition numbering where Chapter 12 introduces Multilevel Models. In the uploaded 2nd edition of *Statistical Rethinking*, Chapter 12 covers Logistic/Poisson GLMs, Overdispersion, Zero Inflation, and Ordered Categories, while Multilevel Models (varying intercepts, tadpoles, non-centered priors) are covered in **Chapter 13 (*Models With Memory*)**.
