---
title: "B02 — Multilevel Model Expansion"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: false
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# B02 — Multilevel Model Expansion

**Course:** Statistical Rethinking 2026  
**Section:** B — Advanced  
**Lecture:** B02 — Multilevel Model Expansion  
**Official reading:** Chapter 13  
**Closest second-edition correspondence:** Chapter 13, especially Sections 13.3–13.5, with a transition toward Section 14.1  
**Central question:** How can a basic varying-intercept model be expanded to represent several cluster types, group-varying contrasts, interactions, nonlinear outcomes, population interventions, and more demanding posterior geometry?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official script `13_GLMM2.r`.
>
> The script directly supports:
>
> - three increasingly structured `UCBadmit` hierarchies;
> - department-varying gender contrasts without an intercept–slope correlation;
> - actor and block-treatment effects in the chimpanzee data;
> - centered and non-centered versions of the same hierarchy;
> - effective-sample-size comparison;
> - cross-classified participant and story effects in the trolley data;
> - a large ordered-logistic model with gender-specific coefficients and ordered education effects;
> - population simulation under a changed reed-frog size distribution;
> - the devil's-funnel demonstration.
>
> Correlated varying intercepts and slopes are not developed as the main structure in this script. They belong more directly to B03 and Chapter 14. This guide therefore develops **uncorrelated varying slopes and expanded hierarchies**, while using correlated effects only to explain the next conceptual step.

---

## 0. Notation and reading conventions

This chapter uses the following notation. A symbol is restated locally whenever its meaning changes.

### Indices and observed units

- $i$ indexes individual observations (e.g., individual application, lever pull, or survey response);
- $g$ indexes a demographic category (e.g., applicant gender category);
- $d$ indexes a department ($d = 1, \dots, D$);
- $a$ or $A_i$ indexes an actor (e.g., individual chimpanzee $a = 1, \dots, 7$);
- $b$ or $B_i$ indexes an experimental block ($b = 1, \dots, 6$);
- $t$ or $T_i$ indexes an experimental treatment combination ($t = 1, \dots, 4$);
- $s$ or $S_i$ indexes a scenario story ($s = 1, \dots, 12$);
- $u$ or $U_i$ indexes a study participant ($u = 1, \dots, 330$).

### Model quantities

- $A_{g,d}$: number of admitted applicants in gender category $g$ and department $d$;
- $N_{g,d}$: total applicants in gender category $g$ and department $d$;
- $p_{g,d}$: probability of admission for cell $(g, d)$;
- $a_d$: baseline intercept for department $d$;
- $b_d$: department-specific gender contrast for department $d$;
- $\bar{a}$: population mean baseline intercept;
- $\sigma_A, \sigma_B, \sigma_S, \sigma_U$: population standard deviations for actor, block, story, and participant hierarchies;
- $\tau$: population standard deviation for group-varying contrasts.

### Distribution conventions and three scales

Throughout this chapter, $\mathrm{Normal}(m,s)$ uses **standard deviation** $s$, not variance. $\mathrm{Exponential}(r)$ uses rate $r$, so its mean is $1/r$. The symbol $\mathcal D$ denotes the complete dataset.

Keep three quantities distinct: $a_d$ or $b_d$ is a parameter on the linear predictor scale (log-odds); $p = \operatorname{logit}^{-1}(\eta)$ is a probability; $Y$ is an observed integer outcome.

### Acronyms and shorthand

- **GLMM:** generalized linear multilevel model;
- **MCMC:** Markov chain Monte Carlo;
- **HMC:** Hamiltonian Monte Carlo;
- **ESS:** effective sample size ($\text{ESS}_{\mathrm{bulk}}$ and $\text{ESS}_{\mathrm{tail}}$);
- **PSIS-LOO / WAIC:** Pareto-smoothed importance sampling leave-one-out cross-validation / Widely Applicable Information Criterion.

---

## 1. Main thesis

The central argument is:

> **A multilevel model can be expanded by assigning adaptive priors to any scientifically meaningful collection of related parameters, not only to intercepts.**

B01 introduced:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

B02 generalizes the same logic to:

- several group types;
- category-specific contrasts;
- treatment effects that vary by group;
- interaction cells;
- ordered-response models;
- target-population simulations;
- new-group causal predictions.

The workflow is:

$$ \text{simple hierarchy} \longrightarrow \text{identify repeated parameter families} \longrightarrow \text{assign one population model per family} \longrightarrow \text{preserve group structure} \longrightarrow \text{non-center when needed} \longrightarrow \text{simulate estimands at the relevant population level} $$

The core design rule is:

$$ \text{parameters that answer the same kind of question} \longrightarrow \text{one shared population model} $$

The hierarchy should respect the scientific structure rather than pool every coefficient merely because it is numerically available.

---

## 2. Relationship to the book and the lecture

The official Chapter 13 assignment aligns well with the second edition.

| B02 topic | Second-edition counterpart | Role |
|---|---|---|
| More than one cluster type | Section 13.3 | Actor, block, participant, and story effects |
| Cross-classification | Section 13.3 | Cluster types need not be nested |
| Non-centered priors | Section 13.4 | Improved posterior geometry |
| Same- versus new-cluster prediction | Section 13.5 | Defines population simulation |
| Group-varying contrasts | Transition to Section 14.1 | Extends adaptive priors beyond intercepts |
| Department-varying gender contrasts | Early varying-slope structure | Separate populations for baselines and contrasts |
| Complex ordered-logistic hierarchy | Script-specific expansion | Combines outcome model, predictors, and two cluster types |
| Population intervention simulation | Section 13.5 logic | Draws new groups under changed composition |

### 2.1 Division of labor across B01–B03

A useful sequence is:

$$ \text{B01} = \text{varying intercepts and partial pooling} $$

$$ \text{B02} = \text{multiple hierarchies and varying contrasts} $$

$$ \text{B03} = \text{correlated varying features} $$

B02 therefore introduces varying slopes without yet making their intercept–slope covariance the central object.

### 2.2 Learning objectives and continuity with A01–A10

After B02, you should be able to write an expanded hierarchical model with multiple parameter families, non-center multi-dimensional hierarchies, specify cross-classified cluster models, and simulate population-level policy interventions.

---

## 3. What it means to expand a hierarchy

A basic model contains one varying parameter:

$$ \alpha_j $$

An expanded multilevel model may contain several parameter families:

$$ \alpha_j, \quad \beta_j, \quad \gamma_k, \quad \delta_{jk} $$

Each family requires a population model:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

$$ \beta_j \sim \mathrm{Normal} \left( \bar{\beta}, \sigma_\beta \right) $$

$$ \gamma_k \sim \mathrm{Normal} \left( 0, \sigma_\gamma \right) $$

The hyperparameters answer population questions:

- What is the average baseline?
- What is the average treatment contrast?
- How heterogeneous are baselines?
- How heterogeneous are treatment effects?
- How much variation remains across a second cluster type?

---

## 4. Parameter families should not be pooled indiscriminately

Suppose admissions are observed for two gender categories across six departments.

There are twelve cell logits: $\alpha_{g,d}$.

Pooling all twelve cell logits into one single hierarchy $\alpha_{g,d} \sim \mathrm{Normal}(\bar{\alpha}, \sigma)$ treats baseline department differences and within-department gender contrasts as exchangeable.

A structured model separates parameter families:

$$ \mathrm{logit}(p_{g,d}) = \alpha_d + \beta_d G_g $$

with separate hierarchies:

$$ \alpha_d \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

$$ \beta_d \sim \mathrm{Normal} \left( \bar{\beta}, \sigma_\beta \right) $$

Pooling now respects the scientific meaning of the parameters.

---

## 5. The three `UCBadmit` models

The official script begins with three admissions models.

Let $A_{g,d}$ be admitted applicants out of $N_{g,d}$ total applications in gender $g$ and department $d$.

### Model 0: unrelated cell parameters (No Pooling)

$$ \mathrm{logit}(p_{g,d}) = a_{g,d}, \quad a_{g,d} \sim \mathrm{Normal}(0, 1.5) $$

### Model 1: pool every cell toward one population

$$ \mathrm{logit}(p_{g,d}) = a_{g,d}, \quad a_{g,d} \sim \mathrm{Normal} \left( \bar{a}, \sigma \right) $$

### Model 2: pool while preserving department structure (Varying Intercepts + Varying Slopes)

$$ \mathrm{logit}(p_{g,d}) = a_d + b_d G_{1g} $$

$$ a_d \sim \mathrm{Normal} \left( \bar{a}, \sigma \right), \quad b_d \sim \mathrm{Normal} \left( 0, \tau \right) $$

---

## 6. Why Model 1 can pool the wrong things

Model 1 treats $a_{\text{male}, A}$ and $a_{\text{female}, F}$ as draws from a single exchangeable population.

Because departments A and F differ systematically in selectivity, and male/female outcomes within a department share department-level unobserved factors, pooling all 12 cells toward a single mean ignores the structural hierarchy of the data.

---

## 7. Department-varying contrasts

In Model 2, $a_d$ is the baseline log-odds for the reference gender in department $d$, and $b_d$ is the department-specific log-odds contrast.

The within-department log-odds contrast is $b_d$.

The probability contrast is:

$$ \Delta_d = \mathrm{logit}^{-1} \left( a_d + b_d \right) - \mathrm{logit}^{-1} \left( a_d \right) $$

Because of the non-linear logistic link, identical log-odds contrasts $b_d$ produce different probability contrasts $\Delta_d$ depending on department baseline $a_d$.

---

## 8. Adaptive pooling of contrasts

The prior $b_d \sim \mathrm{Normal}(0, \tau)$ shares information across department-specific contrasts:

- If $\tau \to 0$, $b_d \to 0$ (all departments share a zero contrast).
- If $\tau$ is larger, department contrasts remain distinct.
- Departments with few applications are shrunken more strongly toward 0.

This is adaptive regularization applied to slopes/contrasts.

---

## 9. Zero-centered varying slopes

Model 2 specifies $b_d \sim \mathrm{Normal}(0, \tau)$ rather than $\mathrm{Normal}(\bar{b}, \tau)$.

This asserts that the population-average contrast is centered at zero a priori.

A more general formulation estimates the global average slope:

$$ b_d \sim \mathrm{Normal} \left( \bar{b}, \tau \right), \quad \bar{b} \sim \mathrm{Normal}(0, s_b) $$

---

## 10. Uncorrelated varying intercepts and slopes

Model 2 treats $a_d$ and $b_d$ as independent in the prior hierarchy:

$$ a_d \sim \mathrm{Normal}(\bar{a}, \sigma), \quad b_d \sim \mathrm{Normal}(0, \tau), \quad a_d \perp\!\!\!\perp b_d $$

This assumes that baseline department selectivity $a_d$ provides no information about the gender contrast $b_d$.

Modeling covariance between $a_d$ and $b_d$ requires a multivariate normal distribution (developed in B03 and Chapter 14).

---

## 11. Varying slope is a broad concept

A varying slope does not have to multiply a continuous variable.

In the admissions model, $b_d G_{1g}$ is a group-varying categorical contrast.

The same structure applies to:

- treatment vs control;
- exposed vs unexposed;
- pre vs post intervention.

Any predictor coefficient that varies across clusters is a varying slope.

---

## 12. Chimpanzee data structure

The chimpanzee experiment includes:

- 7 actors ($A_i \in \{1, \dots, 7\}$);
- 6 experimental blocks ($B_i \in \{1, \dots, 6\}$);
- 4 treatments ($T_i \in \{1, 2, 3, 4\}$ combining `prosoc_left` and `condition`).

Observations are cross-classified by actor and block ($A_i \times B_i$).

---

## 13. Chimpanzee likelihood

The response model is $P_i \sim \mathrm{Bernoulli}(p_i)$.

The linear predictor with block-treatment interactions is:

$$ \mathrm{logit}(p_i) = b_{T_i, B_i} + a_{A_i} $$

where $a_a$ is an actor-specific baseline and $b_{t,b}$ is a treatment-by-block cell deviation.

---

## 14. Actor hierarchy

Actor effects follow:

$$ a_a \sim \mathrm{Normal} \left( \bar{a}, \sigma_A \right) $$

$$ \bar{a} \sim \mathrm{Normal}(0, 1.5), \quad \sigma_A \sim \mathrm{Exponential}(1) $$

$\bar{a}$ is the overall average log-odds of pulling the left lever; $\sigma_A$ measures individual actor heterogeneity.

---

## 15. Block-treatment hierarchy

The $4 \times 6$ matrix $b_{t,b}$ of treatment-by-block cell deviations follows:

$$ b_{t,b} \sim \mathrm{Normal}(0, \sigma_B), \quad \sigma_B \sim \mathrm{Exponential}(1) $$

This pools all 24 treatment-by-block interaction cells toward zero.

---

## 16. Why there is only one global intercept

If both actor and block-treatment hierarchies had free population means ($\bar{a}$ and $\bar{b}$), only their sum $\bar{a} + \bar{b}$ would be identified by the data.

To resolve this additive unidentifiability, the block-treatment hierarchy is centered at zero ($b_{t,b} \sim \mathrm{Normal}(0, \sigma_B)$), while $\bar{a}$ serves as the single global intercept.

---

## 17. Cross-classification

Observations are cross-classified when they belong to multiple grouping structures that are not hierarchically nested:

- Each trial belongs to an actor $A_i$;
- Each trial belongs to a block $B_i$.

Both varying effects enter the linear predictor additively: $\eta_i = \bar{a} + a_{A_i} + b_{B_i} + \dots$.

---

## 18. Cross-classification versus nesting

- **Nested:** Students within Classrooms within Schools ($S \subset C \subset H$).
- **Cross-classified:** Students responses evaluated by multiple Teachers across subjects ($R \to S$ and $R \to T$).

Multilevel modeling handles both by linking index variables to their respective parameter vectors.

---

## 19. Multiple hierarchy scales

In the chimpanzee model:

- $\sigma_A$ measures actor heterogeneity on the log-odds scale;
- $\sigma_B$ measures block-treatment interaction variation.

If $\sigma_A \gg \sigma_B$, individual actor preferences dominate variation, while experimental blocks behave similarly.

---

## 20. Centered chimpanzee model

The centered parameterization samples:

$$ a_a \sim \mathrm{Normal}(\bar{a}, \sigma_A), \quad b_{t,b} \sim \mathrm{Normal}(0, \sigma_B) $$

When $\sigma_A$ or $\sigma_B$ are near zero, sampling can encounter funnel geometry resulting in divergences.

---

## 21. Non-centered chimpanzee model

The non-centered parameterization samples:

$$ z_{a, a} \sim \mathrm{Normal}(0, 1), \quad z_{b, t, b} \sim \mathrm{Normal}(0, 1) $$

and constructs:

$$ a_a = \bar{a} + z_{a, a} \sigma_A, \quad b_{t,b} = z_{b, t, b} \sigma_B $$

This decouples the parameter proposals from the scale hyperparameters, eliminating funnel geometry.

---

## 22. Effective-sample-size comparison

Comparing ESS between centered and non-centered parameterizations across parameters ($\bar{a}, \sigma_A, \sigma_B, a_a, b_{t,b}$) reveals which parameterization provides superior computational geometry.

---

## 23. Strong data can favor centered models

When group data are dense ($N_j$ large), the likelihood constrains $\alpha_j$ tightly, and centered parameterization can sample more efficiently than non-centered parameterization.

---

## 24. Trolley model with two cluster types

Cross-classified ordered-logistic model for moral judgments ($R_i \in \{1, \dots, 7\}$):

$$ R_i \sim \mathrm{OrderedLogistic}(\phi_i, \mathbf{\kappa}) $$

$$ \phi_i = \beta_A A_i + \beta_I I_i + \beta_C C_i + a^S_{S_i} + a^U_{U_i} $$

$$ a^S_s \sim \mathrm{Normal}(0, \sigma_S), \quad a^U_u \sim \mathrm{Normal}(0, \sigma_U) $$

---

## 25. Why stories and participants both matter

- Omission of story effects $a^S_s$ forces unmodeled story variance into experimental coefficients $\beta_A, \beta_I, \beta_C$.
- Omission of participant effects $a^U_u$ creates pseudoreplication and artificially narrow standard errors.

---

## 26. Partial non-centering

In complex models with multiple hierarchies, one hierarchy can be non-centered (e.g., stories $a^S_s = z^S_s \sigma_S$) while another remains centered (e.g., participants $a^U_u \sim \mathrm{Normal}(0, \sigma_U)$).

---

## 27. The full trolley expansion

The expanded trolley model includes:

- gender-specific predictor slopes ($\beta_{A, G}, \beta_{I, G}, \beta_{C, G}, \beta_{Y, G}$);
- gender-specific ordered education effects;
- story and participant varying intercepts.

---

## 28. Ordered education effects

Education levels $E_i \in \{1, \dots, 8\}$ are modeled via a Dirichlet simplex $\mathbf{\delta}$:

$$ \phi_i = \dots + \beta_E \sum_{j=1}^{E_i} \delta_j $$

where $\mathbf{\delta} \sim \mathrm{Dirichlet}(2, \dots, 2)$ enforces monotonic increments between education categories.

---

## 29. Model expansion should remain modular

Expanded GLMMs should be built and validated in modular components:

1. Likelihood & link module;
2. Fixed predictor module;
3. Varying intercept/slope modules;
4. Specialized structural modules (e.g., simplices).

---

## 30. Expansion does not imply adding every available term

Adding unnecessary varying slopes or interactions creates weak identification and MCMC instability. Expand models based on scientific questions and predictive requirements.

---

## 31. Reed-frog prediction model

Multilevel model combining tank intercepts and predator-by-size cell matrix:

$$ S_j \sim \mathrm{Binomial}(D_j, p_j) $$

$$ \mathrm{logit}(p_j) = a_{T_j} + b_{P_j, G_j} $$

$$ a_t \sim \mathrm{Normal}(0, \sigma), \quad b_{p,g} \sim \mathrm{Normal}(0, 1), \quad \sigma \sim \mathrm{Exponential}(1) $$

---

## 32. Population intervention simulation

Using the fitted reed-frog model to predict total survival under an intervention that changes the size distribution from 25% large tanks to 75% large tanks:

1. Draw hyperparameter $\sigma^{(s)}$;
2. Draw a new tank intercept $a_{\mathrm{new}}^{(s)} \sim \mathrm{Normal}(0, \sigma^{(s)})$;
3. Sample size category from intervention weights $P(G = \text{large}) = 0.75$;
4. Calculate $p = \mathrm{logit}^{-1}(a_{\mathrm{new}}^{(s)} + b_{P, G}^{(s)})$;
5. Draw simulated outcome $S \sim \mathrm{Binomial}(35, p)$.

---

## 33. A source inconsistency in the prediction comments

The script comments state "status quo 25% large; intervention 75% large", while an intermediate comment mentions 50%. The code executes 25% vs 75%.

---

## 34. Why a new tank effect must be drawn

Setting $a_{\mathrm{new}} = 0$ simulates an average tank on the log-odds scale, omitting tank-to-tank variance $\sigma$. Drawing $a_{\mathrm{new}} \sim \mathrm{Normal}(0, \sigma)$ incorporates population variance into the policy simulation.

---

## 35. Expected intervention outcome versus replicated outcome

- **Expected outcome:** $\mathbb{E}[S] = 35 \cdot p_j$;
- **Replicated outcome:** $S \sim \mathrm{Binomial}(35, p_j)$.

Replicated outcomes include binomial sampling noise in addition to parameter and population uncertainty.

---

## 36. Composition intervention versus unit-level treatment

Shifting tank size proportions from 25% to 75% alters population composition, which differs from intervening on a specific individual tank's size.

---

## 37. Post-stratification in a multilevel model

Multilevel Regression and Post-stratification (MRP):

$$ \theta_{\text{target}} = \sum_{c} w_c \cdot \widehat{p}_c $$

where $w_c$ are target population weights for cell $c$, and $\widehat{p}_c$ is the partially shrunken cell estimate.

---

## 38. Devil's funnel

The classic funnel hierarchy:

$$ v \sim \mathrm{Normal}(0, 3), \quad x \sim \mathrm{Normal}(0, e^v) $$

When $v < 0$, $e^v \to 0$, forcing $x$ into a extremely narrow neck near 0.

---

## 39. Non-centered funnel

The non-centered transformation:

$$ v \sim \mathrm{Normal}(0, 3), \quad z \sim \mathrm{Normal}(0, 1), \quad x = z \cdot e^v $$

samples independent Gaussian variables $(v, z)$, removing the funnel geometry completely.

---

## 40. Why expanded models create geometry problems

As models accumulate parameters, funnels, ridges, and unidentifiable parameter combinations proliferate, making MCMC diagnostics critical.

---

## 41. Prior predictive checks for expanded hierarchies

Simulate data from priors before fitting to verify that outcome distributions spanned by the expanded parameters remain biologically/scientifically reasonable.

---

## 42. Hierarchy-scale priors should be family-specific

Different varying effect families (actor vs block-treatment vs story vs participant) live on different scales and require separate scale hyperpriors (e.g., Exponential(1) vs Exponential(2)).

---

## 43. Varying effects and interactions

Hierarchical interaction terms $\delta_{t,b} \sim \mathrm{Normal}(0, \sigma_\delta)$ adaptively regularize multi-way cell interactions.

---

## 44. Separate hierarchies versus one giant hierarchy

Lumping unrelated parameters into a single hierarchy forces an inappropriate exchangeability assumption. Use distinct adaptive priors for distinct parameter families.

---

## 45. Model comparison is not enough

Information criteria (PSIS/WAIC) evaluate predictive performance, not causal validity or structural correctness.

---

## 46. Causal interpretation of varying treatment effects

Varying treatment slopes $\beta_j \sim \mathrm{Normal}(\bar{\beta}, \sigma_\beta)$ adaptively pool treatment estimates across groups, but do not resolve unobserved group-level treatment confounding.

---

## 47. Treatment heterogeneity estimands

Distinct targets of inference:

- Population average treatment effect: $\bar{\tau} = \bar{\beta}$;
- Variance of treatment effects: $\sigma_\beta^2$;
- Group-specific treatment effect: $\beta_j$;
- New-group treatment effect: $\beta_{\mathrm{new}} \sim \mathrm{Normal}(\bar{\beta}, \sigma_\beta)$;
- Target-population post-stratified effect: $\sum w_j \beta_j$.

---

## 48. CRM application: department-varying campaign effects

Modeling marketing campaign conversion rates across departments:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_j A_{ij} $$

$$ \alpha_j \sim \mathrm{Normal}(\bar{\alpha}, \sigma_\alpha), \quad \beta_j \sim \mathrm{Normal}(\bar{\beta}, \sigma_\beta) $$

---

## 49. CRM application: customer and campaign cross-classification

Cross-classified conversion model for customer $C_i$ and campaign $K_i$:

$$ \mathrm{logit}(p_i) = \bar{\alpha} + \alpha^C_{C_i} + \alpha^K_{K_i} $$

$$ \alpha^C_c \sim \mathrm{Normal}(0, \sigma_C), \quad \alpha^K_k \sim \mathrm{Normal}(0, \sigma_K) $$

---

## 50. CRM application: campaign effects varying by channel

Hierarchical channel-specific offer effects:

$$ \beta_{a,k} \sim \mathrm{Normal}(\bar{\beta}_a, \sigma_a) $$

---

## 51. CRM application: changing portfolio composition

Evaluating policy shifts in customer segment allocation:

$$ \Delta V^{(s)} = \sum_c w_c^{\text{new}} \mu_c^{(s)} - \sum_c w_c^{\text{old}} \mu_c^{(s)} $$

---

## 52. CausalOps implications

| B02 concept | CausalOps implication |
|---|---|
| Expanded parameter family | `VaryingEffectArtifact` |
| Department-varying contrast | Group-specific estimand family |
| Multiple cluster types | Cross-classification metadata |
| Treatment-by-block effects | Hierarchical interaction artifact |
| Separate adaptive priors | One hierarchy per parameter family |
| Zero-centered deviation family | Identifiability constraint |
| Non-centered form | Computational parameterization metadata |
| Population composition | `TargetPopulationArtifact` |

---

## 53. Validation strategy

Simulation workflow for expanded GLMMs:

1. Specify ground-truth hyperparameters $(\bar{\alpha}, \sigma_\alpha, \bar{\beta}, \sigma_\beta)$;
2. Draw cluster parameters $(\alpha_j, \beta_j)$;
3. Simulate observation datasets under specified links;
4. Fit candidate GLMM specifications;
5. Evaluate RMSE, interval coverage, and population policy estimand recovery.

---

## 54. Common errors exposed by B02

1. Pooling parameters with different scientific meanings.
2. Treating all category cells as a single exchangeable population.
3. Restricting varying slopes to continuous predictors.
4. Assuming zero-centered slopes estimate an overall mean.
5. Including redundant global intercepts in multiple additive hierarchies.
6. Confusing cross-classification with nesting.
7. Assuming non-centered parameterization is universally superior.
8. Evaluating centered vs non-centered models solely by posterior means rather than ESS/divergences.
9. Predicting new clusters by reusing observed cluster parameters.
10. Setting new-cluster varying effects to zero.
11. Ignoring target population cell composition in policy simulations.
12. Accepting script comments without verifying executable code.
13. Assuming hierarchical shrinkage eliminates causal confounding.
14. Expanding models without prior predictive checks.
15. Treating interaction heterogeneity as proof of causal mechanism.

---

## 55. Lecture takeaways

1. Multilevel expansion applies adaptive priors to coherent families of related parameters.
2. Parameter families should be defined by scientific meaning.
3. Pooling every category cell under one population can ignore important structure.
4. Department baselines and department-specific contrasts can have separate hierarchies.
5. A varying slope can be a binary or categorical contrast.
6. The official admissions slope hierarchy is independent of the intercept hierarchy.
7. Correlated intercepts and slopes are the next conceptual step, not the main B02 structure.
8. Cross-classified observations belong to several non-nested cluster systems.
9. The chimpanzee model combines actor effects with treatment-by-block effects.
10. One global intercept is sufficient when additive cluster deviations are centered at zero.
11. Centered and non-centered parameterizations represent the same scientific model.
12. Non-centering can improve ESS and reduce divergences for weakly informed effects.
13. Different hierarchies in one model can use different parameterizations.
14. The trolley model combines an ordered likelihood, several predictor modules, and two cluster types.
15. Ordered predictors can coexist with multilevel effects.
16. Expanded models should remain modular and scientifically motivated.
17. Multilevel predictions must specify observed versus new clusters.
18. New-cluster simulation requires drawing a new varying effect.
19. Population-composition interventions require explicit target distributions.
20. The reed-frog code implements a 25% to 75% large-tank shift.

---

## 56. Code Demonstrations & Multi-Framework Implementation

### 56.1 `UCBadmit` 3 Models (R ulam, Stan, PyMC)

#### R rethinking / ulam

```{r ucb-admit-models}
library(rethinking)
data(UCBadmit)
d <- UCBadmit

dat_ucb <- list(
  A = d$admit,
  N = d$applications,
  D = as.integer(d$dept),
  G = ifelse(d$applicant.gender == "female", 1L, 2L),
  G1 = ifelse(d$applicant.gender == "female", 1L, 0L)
)

# Model 0: Unrelated cell logits (No pooling)
m0 <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G, D],
    matrix[G, D]:a ~ dnorm(0, 1.5)
  ), data = dat_ucb, chains = 4, cores = 4
)

# Model 1: Pooled cell logits (Ignoring department structure)
m1 <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[G, D],
    matrix[G, D]:a ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data = dat_ucb, chains = 4, cores = 4
)

# Model 2: Department baselines + department-varying gender contrast
m2 <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[D] + b[D] * G1,
    a[D] ~ dnorm(a_bar, sigma),
    b[D] ~ dnorm(0, tau),
    a_bar ~ dnorm(0, 1),
    sigma ~ dexp(1),
    tau ~ dexp(1)
  ), data = dat_ucb, chains = 4, cores = 4
)

compare(m0, m1, m2, func = PSIS)
```

#### Stan (CmdStanR)

```stan
// ucb_admit_m2.stan
data {
  int<lower=1> N;
  array[N] int<lower=0> A;
  array[N] int<lower=1> App;
  array[N] int<lower=1> D;
  array[N] int<lower=0, upper=1> G1;
  int<lower=1> N_dept;
}
parameters {
  real a_bar;
  real<lower=0> sigma;
  real<lower=0> tau;
  vector[N_dept] z_a;
  vector[N_dept] z_b;
}
transformed parameters {
  vector[N_dept] a = a_bar + z_a * sigma;
  vector[N_dept] b = z_b * tau;
}
model {
  a_bar ~ normal(0, 1);
  sigma ~ exponential(1);
  tau ~ exponential(1);
  z_a ~ normal(0, 1);
  z_b ~ normal(0, 1);
  
  for (i in 1:N) {
    A[i] ~ binomial_logit(App[i], a[D[i]] + b[D[i]] * G1[i]);
  }
}
```

#### PyMC (Python)

```python
import pymc as pm
import pandas as pd
import numpy as np

# Load UCB dataset structure
df_ucb = pd.DataFrame({
    'admit': [512, 89, 353, 17, 120, 202, 138, 131, 53, 94, 22, 24],
    'applications': [825, 108, 560, 25, 325, 417, 417, 375, 191, 393, 373, 341],
    'dept': [0, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5, 5],
    'g1': [0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1]  # 1 for female, 0 for male
})

n_dept = 6

with pm.Model() as ucb_m2:
    a_bar = pm.Normal("a_bar", mu=0, sigma=1.0)
    sigma = pm.Exponential("sigma", lam=1.0)
    tau = pm.Exponential("tau", lam=1.0)
    
    z_a = pm.Normal("z_a", mu=0, sigma=1.0, shape=n_dept)
    z_b = pm.Normal("z_b", mu=0, sigma=1.0, shape=n_dept)
    
    a = pm.Deterministic("a", a_bar + z_a * sigma)
    b = pm.Deterministic("b", z_b * tau)
    
    p = pm.Deterministic("p", pm.math.invlogit(a[df_ucb['dept'].values] + b[df_ucb['dept'].values] * df_ucb['g1'].values))
    
    A = pm.Binomial("A", n=df_ucb['applications'].values, p=p, observed=df_ucb['admit'].values)
    
    idata_ucb = pm.sample(draws=1000, tune=1000, chains=4, random_seed=42)
```

---

### 56.2 Chimpanzee Cross-Classified Block-Treatment Model

```{r chimp-cross-classified}
library(rethinking)
data(chimpanzees)
d <- chimpanzees
d$treatment <- 1L + d$prosoc_left + 2L * d$condition

dat_chimp <- list(
  P = d$pulled_left,
  A = as.integer(d$actor),
  B = as.integer(d$block),
  T = as.integer(d$treatment)
)

# Non-centered block-treatment model
mBTnc <- ulam(
  alist(
    P ~ dbern(p),
    logit(p) <- a_bar + z_a[A] * sigma_A + z_b[T, B] * sigma_B,
    matrix[T, B]:z_b ~ dnorm(0, 1),
    z_a[A] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma_A ~ dexp(1),
    sigma_B ~ dexp(1)
  ), data = dat_chimp, chains = 4, cores = 4
)

precis(mBTnc, depth = 2, pars = c("a_bar", "sigma_A", "sigma_B"))
```

---

### 56.3 Trolley Ordered-Logistic Model with S and U

```{r trolley-su}
library(rethinking)
data(Trolley)
d <- Trolley

dat_trolley <- list(
  R = d$response,
  A = d$action,
  I = d$intention,
  C = d$contact,
  S = as.integer(d$story),
  U = as.integer(d$id)
)

# Cross-classified non-centered story and participant intercepts
mRXSUnc <- ulam(
  alist(
    R ~ ordered_logistic(phi, cutpoints),
    phi <- bA * A + bI * I + bC * C + zs[S] * sigmaS + zu[U] * sigmaU,
    cutpoints ~ dnorm(0, 1.5),
    c(bA, bI, bC) ~ dnorm(0, 0.5),
    zs[S] ~ dnorm(0, 1),
    zu[U] ~ dnorm(0, 1),
    sigmaS ~ dexp(1),
    sigmaU ~ dexp(1)
  ), data = dat_trolley, chains = 4, cores = 4, sample = FALSE
)
```

---

### 56.4 Reed-Frog Population Intervention Simulation

```{r reedfrog-policy-sim}
library(rethinking)
data(reedfrogs)
d <- reedfrogs

dat_frog_spg <- list(
  S = d$surv,
  D = d$density,
  T = 1:nrow(d),
  P = ifelse(d$pred == "no", 1L, 2L),
  G = ifelse(d$size == "small", 1L, 2L)
)

mSPG <- ulam(
  alist(
    S ~ dbinom(D, p),
    logit(p) <- a[T] + b[P, G],
    a[T] ~ dnorm(0, sigma),
    matrix[P, G]:b ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ), data = dat_frog_spg, chains = 4, cores = 4
)

post <- extract.samples(mSPG)
n_groups <- 1000
n_samples <- 2000

# Status quo (25% large tanks)
S1 <- matrix(0, nrow = n_samples, ncol = n_groups)
for (s in 1:n_groups) {
  aT <- rnorm(n_samples, 0, post$sigma)
  P <- sample(1:2, size = 1, prob = c(0.5, 0.5))
  G <- sample(1:2, size = 1, prob = c(0.75, 0.25))
  p <- inv_logit(aT + post$b[, P, G])
  S1[, s] <- rbinom(n_samples, 35, p)
}

# Intervention (75% large tanks)
S2 <- matrix(0, nrow = n_samples, ncol = n_groups)
for (s in 1:n_groups) {
  aT <- rnorm(n_samples, 0, post$sigma)
  P <- sample(1:2, size = 1, prob = c(0.5, 0.5))
  G <- sample(1:2, size = 1, prob = c(0.25, 0.75))
  p <- inv_logit(aT + post$b[, P, G])
  S2[, s] <- rbinom(n_samples, 35, p)
}

diff_surv <- as.numeric(S2 - S1)
cat("Mean survival change:", mean(diff_surv), "\n")
cat("89% PI change:", HPDI(diff_surv, prob = 0.89), "\n")
```

---

### 56.5 Devil's Funnel (Centered vs Non-Centered)

```{r devils-funnel}
library(rethinking)

# Centered funnel
m_funnel_c <- ulam(
  alist(
    v ~ dnorm(0, 3),
    x ~ dnorm(0, exp(v))
  ), data = list(N = 1), chains = 4
)

# Non-centered funnel
m_funnel_nc <- ulam(
  alist(
    v ~ dnorm(0, 3),
    z ~ dnorm(0, 1),
    gq> real[1]:x <<- z * exp(v)
  ), data = list(N = 1), chains = 4
)

precis(m_funnel_c)
precis(m_funnel_nc)
```

---

## 57. Supplementary practice exercises and worked model specifications

### Exercise 13M5

**Question:** Modify the `UCBadmit` Model 2 to estimate a global mean gender contrast $\bar{b}$ in addition to department-specific slope deviations: $b_d \sim \mathrm{Normal}(\bar{b}, \tau)$. Write out the `ulam` formula list, fit the model, and compare posterior predictions with Model 2 ($b_d \sim \mathrm{Normal}(0, \tau)$).

**Solution:**

```{r solution-13m5}
library(rethinking)
data(UCBadmit)
d <- UCBadmit

dat_ucb <- list(
  A = d$admit,
  N = d$applications,
  D = as.integer(d$dept),
  G1 = ifelse(d$applicant.gender == "female", 1L, 0L)
)

# Model with global slope mean b_bar
m2_bbar <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a[D] + b[D] * G1,
    a[D] ~ dnorm(a_bar, sigma),
    b[D] ~ dnorm(b_bar, tau),
    a_bar ~ dnorm(0, 1),
    b_bar ~ dnorm(0, 1),
    sigma ~ dexp(1),
    tau ~ dexp(1)
  ), data = dat_ucb, chains = 4, cores = 4
)

precis(m2_bbar, pars = c("a_bar", "b_bar", "sigma", "tau"))
```

---

### Exercise 13M6

**Question:** In the chimpanzee model `mBT`, write out a fully non-centered parameterization for both the actor baseline $a_a$ and the block-treatment matrix $b_{t,b}$. Verify that the posterior summaries of parameter draws from centered and non-centered models match while comparing ESS.

**Solution:**

```{r solution-13m6}
library(rethinking)
data(chimpanzees)
d <- chimpanzees
d$treatment <- 1L + d$prosoc_left + 2L * d$condition

dat_chimp <- list(
  P = d$pulled_left,
  A = as.integer(d$actor),
  B = as.integer(d$block),
  T = as.integer(d$treatment)
)

mBT_nc_full <- ulam(
  alist(
    P ~ dbern(p),
    logit(p) <- a_bar + z_a[A] * sigma_A + z_b[T, B] * sigma_B,
    matrix[T, B]:z_b ~ dnorm(0, 1),
    z_a[A] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma_A ~ dexp(1),
    sigma_B ~ dexp(1),
    gq> vector[A]:a <<- a_bar + z_a * sigma_A,
    gq> matrix[T, B]:b <<- z_b * sigma_B
  ), data = dat_chimp, chains = 4, cores = 4
)

precis(mBT_nc_full, depth = 2, pars = c("sigma_A", "sigma_B"))
```

---

## 58. Lecture takeaways

1. Multilevel expansion applies adaptive priors to coherent families of related parameters.
2. Parameter families should be defined by scientific meaning.
3. Pooling every category cell under one population can ignore important structure.
4. Department baselines and department-specific contrasts can have separate hierarchies.
5. A varying slope can be a binary or categorical contrast.
6. The official admissions slope hierarchy is independent of the intercept hierarchy.
7. Correlated intercepts and slopes are the next conceptual step, not the main B02 structure.
8. Cross-classified observations belong to several non-nested cluster systems.
9. The chimpanzee model combines actor effects with treatment-by-block effects.
10. One global intercept is sufficient when additive cluster deviations are centered at zero.
11. Centered and non-centered parameterizations represent the same scientific model.
12. Non-centering can improve ESS and reduce divergences for weakly informed effects.
13. Different hierarchies in one model can use different parameterizations.
14. The trolley model combines an ordered likelihood, several predictor modules, and two cluster types.
15. Ordered predictors can coexist with multilevel effects.
16. Expanded models should remain modular and scientifically motivated.
17. Multilevel predictions must specify observed versus new clusters.
18. New-cluster simulation requires drawing a new varying effect.
19. Population-composition interventions require explicit target distributions.
20. The reed-frog code implements a 25% to 75% large-tank shift.

---

## 59. Review questions

1. What does it mean to expand a multilevel model?
2. Why should different parameter families receive different adaptive priors?
3. What are the three `UCBadmit` models in the official script?
4. Why can pooling all twelve gender-department cells be scientifically coarse?
5. What does $a_d$ represent in the structured admissions model?
6. What does $b_d$ represent?
7. How is the department-specific probability contrast computed?
8. What does $\tau$ control in the admissions hierarchy?
9. Why does the official slope prior not estimate a population-average contrast?
10. What independence assumption is made between $a_d$ and $b_d$?
11. Why is a binary contrast still called a varying slope?
12. How is the chimpanzee treatment index constructed?
13. What two cluster systems occur in the chimpanzee data?
14. What does $a_a$ represent?
15. What does $b_{t,b}$ represent?
16. Why is the block-treatment population centered at zero?
17. What would happen if both actor and block-effect families had unrestricted means?
18. What is cross-classification?
19. How does it differ from nesting?
20. What do $\sigma_A$ and $\sigma_B$ measure?
21. How is the chimpanzee model non-centered?
22. Why should centered and non-centered forms be compared using ESS?
23. When can centered parameterization be preferable?
24. What two cluster types occur in the trolley model?
25. Why can one hierarchy be centered and another non-centered?
26. Which components are added in the full trolley model?
27. How does the simplex represent an ordered education effect?
28. What is the reed-frog prediction model?
29. How does the script simulate a new tank?
30. What population composition changes under the intervention?
31. What inconsistency exists in the script comments?
32. Why is setting a new tank effect to zero incorrect?
33. How does an expected intervention outcome differ from a predictive intervention outcome?
34. What is a population-composition intervention?
35. How does multilevel regression support post-stratification?
36. What is the devil's funnel?
37. How does non-centering alter its geometry?
38. Why do expanded models require stronger diagnostics?
39. Why can a hierarchically pooled treatment effect remain causally biased?
40. Which B02 structures prepare for B03?

---

## 60. Book reading guide

### Primary reading

- Chapter 13, *Models With Memory*
  - Section 13.3, *More than one type of cluster*
  - Section 13.4, *Divergent transitions and non-centered priors*
  - Section 13.5, *Multilevel posterior predictions*

### Secondary transition

- Chapter 14, Section 14.1, *Varying slopes by construction*

---

## 61. Official code guide

### `13_GLMM2.r`

Contains `UCBadmit` hierarchy comparison, chimpanzee block interactions, trolley story/participant cross-classification, full trolley expansion with Dirichlet education simplex, reed-frog population simulation, and the devil's funnel HMC visualization.

---

## 62. Sources

- Richard McElreath, *Statistical Rethinking: A Bayesian Course with Examples in R and Stan*, second edition, Chapter 13 & 14.1.
- Official `stat_rethinking_2026` repository and B02 lecture listing.
- Official script `13_GLMM2.r`.

---

## 70. Annex: Complete Solutions to Review Questions

#### 1. What does it mean to expand a multilevel model?

Expanding a multilevel model means extending adaptive priors (hierarchies) beyond simple random intercepts to include additional parameter families—such as group-varying treatment slopes, multi-way cell interaction parameters, multiple non-nested cluster types (cross-classification), and structured non-linear or ordered components.

#### 2. Why should different parameter families receive different adaptive priors?

Different parameter families represent distinct scientific mechanisms (e.g., baseline group propensity vs group-specific treatment sensitivity vs block noise). Assigning separate adaptive priors allows the model to estimate independent population locations and variances for each parameter family without forcing false exchangeability across unrelated parameters.

#### 3. What are the three `UCBadmit` models in the official script?

- **Model 0:** Unrelated cell logits $a_{g,d} \sim \mathrm{Normal}(0, 1.5)$ (no pooling across 12 cells).
- **Model 1:** Single hierarchy over all 12 cell logits $a_{g,d} \sim \mathrm{Normal}(\bar{a}, \sigma)$ (coarse pooling ignoring department vs gender structure).
- **Model 2:** Structured model $\mathrm{logit}(p_{g,d}) = a_d + b_d G_{1g}$ with separate hierarchies for department baselines $a_d \sim \mathrm{Normal}(\bar{a}, \sigma)$ and department-varying gender contrasts $b_d \sim \mathrm{Normal}(0, \tau)$.

#### 4. Why can pooling all twelve gender-department cells be scientifically coarse?

Pooling all 12 cells into a single normal distribution assumes that the variance between different departments is identical in scale and structure to the gender contrast within departments. It ignores the facts that departments differ systematically in selectivity and that gender comparisons occur within departments.

#### 5. What does $a_d$ represent in the structured admissions model?

$a_d$ represents the baseline log-odds of admission for the reference gender category in department $d$.

#### 6. What does $b_d$ represent?

$b_d$ represents the department-specific log-odds contrast in admission probability between the target gender category and the reference category in department $d$.

#### 7. How is the department-specific probability contrast computed?

The probability contrast for department $d$ is $\Delta_d = \mathrm{logit}^{-1}(a_d + b_d) - \mathrm{logit}^{-1}(a_d)$. Due to the logistic link non-linearity, $\Delta_d$ depends on both the log-odds slope $b_d$ and the baseline intercept $a_d$.

#### 8. What does $\tau$ control in the admissions hierarchy?

$\tau$ is the standard deviation of department gender slopes $b_d \sim \mathrm{Normal}(0, \tau)$. It controls adaptive regularization for slopes: when $\tau \to 0$, all department gender contrasts shrink to zero; as $\tau$ increases, department contrasts are allowed to differ.

#### 9. Why does the official slope prior not estimate a population-average contrast?

The official specification $b_d \sim \mathrm{Normal}(0, \tau)$ fixes the population mean contrast at zero a priori. This acts as a regularizing prior centered on zero discrimination, rather than estimating an unconstrained population average slope $\bar{b}$.

#### 10. What independence assumption is made between $a_d$ and $b_d$?

The model assumes $a_d \perp\!\!\!\perp b_d$ in the prior, meaning that a department's baseline selectivity $a_d$ provides no information about its gender contrast $b_d$.

#### 11. Why is a binary contrast still called a varying slope?

In linear regression, a coefficient multiplying an indicator variable $G \in \{0, 1\}$ dictates the slope of the linear predictor between $G=0$ and $G=1$. When this coefficient varies by cluster $j$, it is mathematically and conceptually a group-varying slope.

#### 12. How is the chimpanzee treatment index constructed?

The treatment index $T_i \in \{1, 2, 3, 4\}$ is constructed as $T_i = 1 + \text{prosoc\_left}_i + 2 \cdot \text{condition}_i$, combining the prosocial option indicator and partner presence indicator into four discrete categories (`R/N`, `L/N`, `R/P`, `L/P`).

#### 13. What two cluster systems occur in the chimpanzee data?

The chimpanzee data contain two cross-classified cluster systems: 7 individual **actors** ($A_i$) and 6 experimental **blocks** ($B_i$).

#### 14. What does $a_a$ represent?

$a_a$ represents actor $a$'s baseline log-odds propensity to pull the left lever across all treatments.

#### 15. What does $b_{t,b}$ represent?

$b_{t,b}$ represents the interaction cell deviation for treatment $t$ in experimental block $b$.

#### 16. Why is the block-treatment population centered at zero?

To avoid additive unidentifiability with the global actor mean $\bar{a}$, the block-treatment interaction matrix is constrained to have prior mean zero: $b_{t,b} \sim \mathrm{Normal}(0, \sigma_B)$.

#### 17. What would happen if both actor and block-effect families had unrestricted means?

If both $a_a \sim \mathrm{Normal}(\bar{a}, \sigma_A)$ and $b_{t,b} \sim \mathrm{Normal}(\bar{b}, \sigma_B)$ included free means, $\bar{a}$ and $\bar{b}$ would be additively collinear, causing poor MCMC mixing and non-identifiability.

#### 18. What is cross-classification?

Cross-classification occurs when data are grouped along two or more dimensions that are not hierarchically nested inside one another (e.g., actors and blocks, or participants and stories).

#### 19. How does it differ from nesting?

In nested data, each lower-level unit belongs to exactly one upper-level unit (e.g., student in one classroom). In cross-classified data, units in dimension A encounter multiple units in dimension B (e.g., an actor participates in multiple blocks, and a block contains multiple actors).

#### 20. What do $\sigma_A$ and $\sigma_B$ measure?

- $\sigma_A$ measures between-actor variation in baseline left-pulling propensity.
- $\sigma_B$ measures treatment-by-block interaction variability on the log-odds scale.

#### 21. How is the chimpanzee model non-centered?

By defining standardized offsets $z_{a, a} \sim \mathrm{Normal}(0, 1)$ and $z_{b, t, b} \sim \mathrm{Normal}(0, 1)$, and computing $a_a = \bar{a} + z_{a, a} \sigma_A$ and $b_{t,b} = z_{b, t, b} \sigma_B$ in transformed parameters or generated quantities.

#### 22. Why should centered and non-centered forms be compared using ESS?

Because both forms represent the exact same mathematical posterior distribution, they should be evaluated based on sampling efficiency (bulk/tail ESS per second and divergence counts) to determine which parameterization explores the target geometry better.

#### 23. When can centered parameterization be preferable?

Centered parameterization can sample faster and with higher ESS when group data are dense ($N_j$ large), because the likelihood tightly constrains group effects, eliminating funnel geometry.

#### 24. What two cluster types occur in the trolley model?

The trolley model contains cross-classified varying intercepts for **stories** ($S_i$) and **participants** ($U_i$).

#### 25. Why can one hierarchy be centered and another non-centered?

Different hierarchies in the same model may have different data densities. A weakly informed hierarchy (e.g., 12 stories) can be non-centered to avoid funnels, while a densely observed hierarchy (e.g., 330 participants with 30 responses each) can remain centered.

#### 26. Which components are added in the full trolley model?

The full trolley model adds gender-specific slopes for action, intention, contact, and age, plus a gender-specific ordered Dirichlet simplex for education levels, alongside story and participant varying intercepts.

#### 27. How does the simplex represent an ordered education effect?

A Dirichlet simplex $\mathbf{\delta} \sim \mathrm{Dirichlet}(2, \dots, 2)$ defines non-negative proportional increments between adjacent education levels ($\sum \delta_j = 1$). Multiplying by a scale parameter $\beta_E$ enforces monotonic education effects with learned spacing.

#### 28. What is the reed-frog prediction model?

The model $S_j \sim \mathrm{Binomial}(D_j, p_j)$, $\mathrm{logit}(p_j) = a_{T_j} + b_{P_j, G_j}$ combines tank varying intercepts $a_t \sim \mathrm{Normal}(0, \sigma)$ with a $2 \times 2$ predator-by-size cell matrix $b_{p,g}$.

#### 29. How does the script simulate a new tank?

By drawing a new tank intercept $a_{\mathrm{new}} \sim \mathrm{Normal}(0, \sigma)$ for each posterior draw, incorporating tank-to-tank population variance into the prediction.

#### 30. What population composition changes under the intervention?

The proportion of large tanks in the population changes from 25% (status quo) to 75% (intervention).

#### 31. What inconsistency exists in the script comments?

The text comment mentions a 50% large tank intervention, whereas the executable script code sets $P(\text{large}) = 0.75$.

#### 32. Why is setting a new tank effect to zero incorrect?

Setting $a_{\mathrm{new}} = 0$ simulates only a hypothetical average tank on the logit scale, omitting between-tank variance $\sigma$ and underestimating predictive uncertainty.

#### 33. How does an expected intervention outcome differ from a predictive intervention outcome?

- Expected outcome computes expected counts $\mathbb{E}[S] = 35 \cdot p$.
- Predictive outcome simulates integer counts $S \sim \mathrm{Binomial}(35, p)$, incorporating binomial sampling noise.

#### 34. What is a population-composition intervention?

A population-composition intervention alters the macro distribution of subgroup categories (e.g., shift in tank sizes, customer segments, or store types) rather than manipulating individual treatment variables for existing units.

#### 35. How does multilevel regression support post-stratification?

Multilevel regression estimates stable, shrunken cell probabilities $\widehat{p}_c$ even for sparse cells, which are then weighted by target population cell proportions $w_c$ in post-stratification ($\sum w_c \widehat{p}_c$).

#### 36. What is the devil's funnel?

The devil's funnel is a canonical posterior geometry $v \sim \mathrm{Normal}(0, 3), x \sim \mathrm{Normal}(0, e^v)$ where the conditional variance of $x$ depends exponentially on scale $v$, creating a narrow region near $v \to -\infty$ that causes MCMC samplers to diverge.

#### 37. How does non-centering alter its geometry?

Non-centering transforms variables to $z \sim \mathrm{Normal}(0, 1)$ and $x = z \cdot e^v$, turning the posterior into an isotropic Gaussian surface in $(z, v)$ space that HMC samples effortlessly.

#### 38. Why do expanded models require stronger diagnostics?

Expanded models contain multiple interaction terms, scale hyperparameters, and cross-classified factors that create complex, highly curved posterior surfaces prone to divergences and low ESS.

#### 39. Why can a hierarchically pooled treatment effect remain causally biased?

Hierarchical shrinkage reduces variance across group treatment estimates but does not alter the DAG or remove unobserved group-level treatment selection bias.

#### 40. Which B02 structures prepare for B03?

B02 introduces independent varying intercepts and slopes ($a_d \perp\!\!\!\perp b_d$). B03 extends this by modeling joint multivariate normal populations $\mathrm{MVNormal}(\mathbf{0}, \mathbf{\Sigma})$ that capture correlation between intercepts and slopes.
