---
title: "B03 — Correlated Features"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: false
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# B03 — Correlated Features

**Course:** Statistical Rethinking 2026  
**Section:** B — Advanced  
**Lecture:** B03 — Correlated Features  
**Official reading listed by the course:** Chapter 13  
**Closest second-edition correspondence:** Chapter 14, especially Sections 14.1–14.2  
**Central question:** How can group-specific intercepts, slopes, treatment responses, and other features be modeled jointly so that their covariance becomes a source of partial pooling and prediction?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the uploaded second edition of *Statistical Rethinking*, and the official script `14_GLMM_slopes_.r`.
>
> The principal script directly supports:
>
> - a correlated intercept-and-slope hierarchy for `UCBadmit`;
> - a multivariate normal population of department features;
> - separate population means, standard deviations, and a correlation matrix;
> - an LKJ prior on correlations;
> - sequential learning of the population covariance;
> - two-dimensional shrinkage compared with no pooling;
> - correlated four-feature actor and block hierarchies in the chimpanzee data;
> - centered and non-centered covariance models;
> - Cholesky-factor construction.
>
> The repository script `14_varying_slopes_bangladesh.R` is used as a supplementary worked example. It directly compares uncorrelated, centered-correlated, and non-centered-correlated district models for rural and urban contraceptive use.
>
> The course schedule assigns Chapter 13. In the uploaded second edition, the exact mathematical treatment of correlated varying intercepts and slopes is Chapter 14, *Adventures in Covariance*. Chapter 13 supplies the multilevel foundation, while Chapter 14 supplies the multivariate population prior, LKJ correlation prior, and varying-slope machinery.

---

## 0. Notation and reading conventions

This chapter uses the following notation. A symbol is restated locally whenever its meaning changes.

### Indices and observed units

- $i$ indexes individual observations (e.g., individual application, lever pull, or survey response);
- $j$ or $d$ indexes a cluster or group (e.g., department, district, actor, or block);
- $J$ denotes the total number of observed groups/clusters;
- $k$ indexes a feature or dimension within a group vector ($k = 1, \dots, K$);
- $s$ indexes a retained MCMC posterior draw ($s = 1, \dots, S$).

### Model quantities

- $\boldsymbol{\theta}_j = \begin{bmatrix} \alpha_j \\ \beta_j \end{bmatrix}$: vector of group-specific parameters for cluster $j$;
- $\boldsymbol{\mu} = \begin{bmatrix} \bar{\alpha} \\ \bar{\beta} \end{bmatrix}$: vector of population means;
- $\mathbf{S}$: $K \times K$ population covariance matrix;
- $\mathbf{D}$: diagonal matrix of marginal feature standard deviations $\operatorname{diag}(\sigma_1, \dots, \sigma_K)$;
- $\mathbf{R}$: $K \times K$ population correlation matrix;
- $\mathbf{L}_R$: lower-triangular Cholesky factor of correlation matrix $\mathbf{R}$ ($\mathbf{R} = \mathbf{L}_R \mathbf{L}_R^{\mathsf T}$);
- $\rho$: population correlation coefficient between two group features ($\rho_{12} \in [-1, 1]$);
- $\eta$: shape parameter of the LKJ correlation prior ($\mathrm{LKJCorr}(\eta)$).

### Distribution conventions and three scales

Throughout this chapter, $\mathrm{Normal}(m,s)$ uses standard deviation $s$, while $\mathrm{MVNormal}(\boldsymbol{\mu}, \mathbf{S})$ uses covariance matrix $\mathbf{S}$. $\mathrm{Exponential}(r)$ uses rate $r$. The symbol $\mathcal D$ denotes the complete dataset.

Keep three quantities distinct: $\alpha_j, \beta_j$ are linear predictor parameters (log-odds); $p_j = \mathrm{logit}^{-1}(\eta_j)$ is a probability; $Y$ is an observed integer outcome.

### Acronyms and shorthand

- **GLMM:** generalized linear multilevel model;
- **MVN / MVNormal:** multivariate normal distribution;
- **LKJ:** Lewandowski-Kurowicka-Joe correlation matrix distribution;
- **HMC:** Hamiltonian Monte Carlo;
- **ESS:** effective sample size;
- **PSIS-LOO / WAIC:** Pareto-smoothed importance sampling leave-one-out cross-validation / Widely Applicable Information Criterion.

---

## 1. Main thesis

The central argument is:

> **Group features often vary together, and modeling their covariance allows information to flow not only across groups but also across parameter types.**

B01 pooled group intercepts:

$$ \alpha_j \sim \mathrm{Normal} \left( \bar{\alpha}, \sigma_\alpha \right) $$

B02 added separate populations of varying contrasts:

$$ \beta_j \sim \mathrm{Normal} \left( \bar{\beta}, \sigma_\beta \right) $$

B03 models the pair jointly:

$$ \begin{bmatrix} \alpha_j \\ \beta_j \end{bmatrix} \sim \mathrm{MVNormal} \left( \begin{bmatrix} \bar{\alpha} \\ \bar{\beta} \end{bmatrix}, \mathbf{S} \right) $$

The covariance matrix permits:

$$ \mathrm{Cov} \left( \alpha_j, \beta_j \right) \neq 0 $$

Learning a group intercept can therefore improve inference about its slope, and vice versa.

The workflow is:

$$ \text{group-specific feature vector} \longrightarrow \text{multivariate population model} \longrightarrow \text{scale-correlation decomposition} \longrightarrow \text{joint partial pooling} \longrightarrow \text{new-group feature prediction} $$

---

## 2. Relationship to the book and the lecture

The lecture lies at the boundary between Chapters 13 and 14 of *Statistical Rethinking* (2nd Edition).

| B03 topic | Closest second-edition counterpart | Role |
|---|---|---|
| Partial pooling and group populations | Chapter 13 | Foundation |
| Varying slopes | Section 14.1 | Adds group-specific responses |
| Joint intercept-slope population | Section 14.1 | Models correlated group features |
| Covariance decomposition | Section 14.1 | Separates scales and correlations |
| LKJ prior | Section 14.1 | Regularizes correlation matrices |
| Advanced varying slopes | Section 14.2 | Larger feature vectors and computation |
| Cholesky non-centering | Sections 14.1–14.2 | Efficient MCMC implementation |
| Bangladesh district model | Chapter 14 practice | Interprets urban-rural covariance |
| Chimpanzee four-feature hierarchies | 2026 script extension | Correlated treatment patterns by actor and block |

The practical mapping is:

$$ \text{official B03 Chapter 13} = \text{Chapter 13 multilevel foundation} + \text{Chapter 14 correlated varying features} $$

---

## 3. From varying intercepts to varying lines

A varying-intercept model is:

$$ \mu_{ij} = \alpha_j + \beta x_{ij} $$

Every group has its own baseline, but all groups share one slope.

A varying-intercept-and-slope model is:

$$ \mu_{ij} = \alpha_j + \beta_j x_{ij} $$

Now every group has its own line.

The pair $\left( \alpha_j, \beta_j \right)$ is a vector of group features.

The population model must describe:

- average intercept ($\bar{\alpha}$);
- average slope ($\bar{\beta}$);
- intercept heterogeneity ($\sigma_\alpha$);
- slope heterogeneity ($\sigma_\beta$);
- intercept-slope correlation ($\rho$).

---

## 4. Why intercepts and slopes may covary

Consider cafés with morning and afternoon waiting times.

Let $\alpha_j$ be the morning wait at café $j$, and $\beta_j$ be the afternoon-minus-morning difference.

A popular café may have a long morning wait and a large afternoon improvement:

$$ \alpha_j \text{ high}, \qquad \beta_j \text{ strongly negative} $$

An unpopular café may have a short morning wait and little afternoon change:

$$ \alpha_j \text{ low}, \qquad \beta_j \text{ near zero} $$

Across cafés, $\mathrm{Cor} \left( \alpha_j, \beta_j \right) < 0$.

This covariance is population information: a morning observation from a new café updates an afternoon prediction before any afternoon observation is collected.

---

## 5. Varying slopes as interaction machines

The model $\mu_{ij} = \alpha_j + \beta_j x_{ij}$ is a regularized interaction between group identity and predictor $x$.

Every group receives a different predictor response.

The hierarchy estimates $\bar{\beta}$ and $\sigma_\beta$ while shrinking noisy group slopes toward the population pattern.

When $\sigma_\beta$ is large, the distribution of slopes may be more informative than the average slope alone.

---

## 6. The multivariate population model

For group $j$, define $\boldsymbol{\theta}_j = \begin{bmatrix} \alpha_j \\ \beta_j \end{bmatrix}$.

The population model is:

$$ \boldsymbol{\theta}_j \sim \mathrm{MVNormal} \left( \boldsymbol{\mu}, \mathbf{S} \right) $$

where:

$$ \boldsymbol{\mu} = \begin{bmatrix} \bar{\alpha} \\ \bar{\beta} \end{bmatrix}, \qquad \mathbf{S} = \begin{bmatrix} \sigma_\alpha^2 & \rho\sigma_\alpha\sigma_\beta \\ \rho\sigma_\alpha\sigma_\beta & \sigma_\beta^2 \end{bmatrix} $$

The diagonal entries are marginal feature variances; the off-diagonal entries are feature covariances.

---

## 7. Correlation and covariance

Covariance is:

$$ \mathrm{Cov} \left( \alpha, \beta \right) = \rho \sigma_\alpha \sigma_\beta $$

Correlation is:

$$ \rho = \frac{ \mathrm{Cov} \left( \alpha, \beta \right) }{ \sigma_\alpha\sigma_\beta }, \qquad -1 \leq \rho \leq 1 $$

The model separates:

- how much intercepts vary ($\sigma_\alpha$);
- how much slopes vary ($\sigma_\beta$);
- how their variations align ($\rho$).

---

## 8. Scale-correlation decomposition

Define:

$$ \mathbf{D} = \begin{bmatrix} \sigma_\alpha & 0 \\ 0 & \sigma_\beta \end{bmatrix}, \qquad \mathbf{R} = \begin{bmatrix} 1 & \rho \\ \rho & 1 \end{bmatrix} $$

Then:

$$ \mathbf{S} = \mathbf{D} \mathbf{R} \mathbf{D} $$

This separates marginal variation from dependence and guarantees a valid covariance matrix when the scales are positive and $\mathbf{R}$ is a valid correlation matrix.

In Stan, this is efficiently computed via `quad_form_diag(Rho, Sigma)`.

---

## 9. Why a multivariate Gaussian?

A multivariate Gaussian is the maximum-entropy choice when only the mean vector and covariance matrix are specified.

However, it remains an empirical assumption. It can be inappropriate when group features are multimodal, heavy-tailed, strongly skewed, or divided into discrete latent classes.

---

## 10. The `UCBadmit` correlated-feature model

The model uses admitted count $A_i$, total applications $N_i$, department $D_i$, and female indicator $F_i$:

$$ A_i \sim \mathrm{Binomial} \left( N_i, p_i \right) $$

$$ \mathrm{logit}(p_i) = \bar{a} + a_{D_i} + \left( \bar{b} + b_{D_i} \right) F_i $$

Here:

- $\bar{a}$ is the overall baseline log-odds;
- $a_d$ is department $d$'s baseline deviation;
- $\bar{b}$ is the overall female-minus-male contrast;
- $b_d$ is department $d$'s contrast deviation.

---

## 11. Department feature vectors

For department $d$, define $\boldsymbol{u}_d = \begin{bmatrix} a_d \\ b_d \end{bmatrix}$.

The hierarchy is:

$$ \boldsymbol{u}_d \sim \mathrm{MVNormal} \left( \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \mathbf{S} \right) $$

Total department parameters are $\alpha_d = \bar{a} + a_d$ and $\beta_d = \bar{b} + b_d$.

---

## 12. Population interpretation in admissions

- $\bar{a}$: average department baseline log-odds;
- $\bar{b}$: average department category contrast;
- $\sigma_a$: heterogeneity in department selectivity;
- $\sigma_b$: heterogeneity in department category contrasts;
- $\rho_{ab}$: correlation asking whether selective departments (high $\bar{a} + a_d$) tend to have larger or smaller female contrasts $b_d$.

---

## 13. Probability-scale department contrasts

Reference probability: $p_{M,d} = \mathrm{logit}^{-1}(\bar{a} + a_d)$.

Female probability: $p_{F,d} = \mathrm{logit}^{-1}(\bar{a} + a_d + \bar{b} + b_d)$.

Probability difference: $\Delta_d = p_{F,d} - p_{M,d}$.

Logit-scale correlation $\rho_{ab}$ does not equal probability-scale correlation because of the non-linear inverse-logit transformation.

---

## 14. Joint population ellipses

The bivariate normal population is visualized as a set of concentric confidence ellipses in the $(a_d, b_d)$ plane:

- Positive correlation ($\rho > 0$): ellipse tilted upward;
- Negative correlation ($\rho < 0$): ellipse tilted downward;
- Zero correlation ($\rho = 0$): axis-aligned ellipse.

---

## 15. Sequential learning of covariance

As data accumulate across groups:

1. Group posterior points contract toward their true values;
2. Population standard deviations $\sigma_a, \sigma_b$ adjust;
3. Correlation $\rho$ narrows;
4. The population ellipse rotates;
5. Sparse groups borrow directional strength along the major axis of the ellipse.

---

## 16. Number of groups versus observations per group

- **Within-group observations ($N_j$):** identify individual group parameters $(a_j, b_j)$;
- **Number of groups ($J$):** identifies population hyperparameters $(\bar{a}, \bar{b}, \sigma_a, \sigma_b, \rho)$.

Estimating a correlation $\rho$ accurately requires many groups ($J \ge 15-20$), not merely many observations in 2 or 3 groups.

---

## 17. Two-dimensional shrinkage

Under independent hierarchies, intercepts and slopes shrink orthogonally toward $(\bar{a}, \bar{b})$.

Under a correlated hierarchy, shrinkage occurs along the principal axes of the population covariance ellipse. If $\rho < 0$, a high intercept pulls the associated slope estimate downward.

---

## 18. Partial pooling versus no pooling

Correlated partial pooling balances data likelihood against the joint bivariate normal population. In low-sample regimes, joint partial pooling produces substantially lower estimation error than independent no-pooling or independent partial pooling models.

---

## 19. Cross-feature borrowing

If group $j$ has dense reference-category observations but sparse target-category observations, its intercept $a_j$ is well-identified while its slope $b_j$ is noisy.

Through population covariance $\rho$, $a_j$ informs the estimate of $b_j$.

---

## 20. The LKJ correlation prior

The LKJ distribution puts a prior over valid correlation matrices $\mathbf{R} \sim \mathrm{LKJCorr}(\eta)$:

- $\eta = 1$: uniform density over all valid positive-definite correlation matrices;
- $\eta > 1$: concentrates density around the identity matrix $\mathbf{I}$ ($\rho \to 0$);
- $\eta = 4$: standard regularizing prior that penalizes extreme correlations ($\rho \approx \pm 1$).

---

## 21. LKJ is a matrix prior

For $K$ features, there are $K(K-1)/2$ correlation parameters.

Because the correlation matrix $\mathbf{R}$ must remain symmetric positive-definite ($\boldsymbol{x}^{\mathsf T} \mathbf{R} \boldsymbol{x} > 0$), correlations cannot be specified independently. LKJ places a joint probability measure over the entire positive-definite manifold.

---

## 22. Correlation priors require scale priors

The full covariance is $\mathbf{S} = \mathbf{D} \mathbf{R} \mathbf{D}$.

A high correlation $\rho = 0.9$ creates negligible covariance if scale $\sigma_\beta \to 0$. Calibrating priors requires evaluating both marginal scales $\mathbf{D}$ and correlation matrix $\mathbf{R}$ jointly.

---

## 23. Correlation is weakly meaningful when a scale is near zero

If $\sigma_\beta \approx 0$, all group slopes are virtually identical ($\beta_j \approx \bar{\beta}$). In this regime, the numerical value of $\rho$ is poorly identified by data and carries minor predictive impact.

---

## 24. Cholesky decomposition

Every symmetric positive-definite correlation matrix $\mathbf{R}$ can be decomposed into a lower-triangular matrix $\mathbf{L}_R$ such that:

$$ \mathbf{R} = \mathbf{L}_R \mathbf{L}_R^{\mathsf T} $$

If $\boldsymbol{z}_j \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{I})$ is a vector of independent standard normal variables, then:

$$ \boldsymbol{u}_j = \mathbf{D} \mathbf{L}_R \boldsymbol{z}_j \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{D} \mathbf{R} \mathbf{D}) $$

---

## 25. Two-dimensional Cholesky intuition

For two features with scales $\sigma_1, \sigma_2$ and correlation $\rho$:

$$ \mathbf{L}_R = \begin{bmatrix} 1 & 0 \\ \rho & \sqrt{1-\rho^2} \end{bmatrix} $$

$$ \begin{bmatrix} u_{1j} \\ u_{2j} \end{bmatrix} = \begin{bmatrix} \sigma_1 z_{1j} \\ \sigma_2 \left( \rho z_{1j} + \sqrt{1-\rho^2} z_{2j} \right) \end{bmatrix} $$

$z_{1j}$ generates shared covariance, while $z_{2j}$ supplies independent variation.

---

## 26. Centered correlated hierarchy

The centered parameterization samples group vectors directly:

$$ \boldsymbol{u}_j \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}) $$

When scales $\sigma$ are small or correlations $\rho$ are extreme, posterior sampling suffers from multivariate funnel geometry, producing divergences and low ESS.

---

## 27. Non-centered correlated hierarchy

The non-centered Cholesky parameterization samples independent standardized matrices $\mathbf{Z} \sim \mathrm{Normal}(0, 1)$ and constructs correlated effects deterministically:

$$ \mathbf{V} = \left( \mathbf{D} \mathbf{L}_R \mathbf{Z} \right)^{\mathsf T} $$

This decouples the structural parameters from scale and correlation hyperparameters, eliminating multivariate funnels.

---

## 28. Why Stan samples the Cholesky factor

Sampling $\mathbf{L}_R \sim \mathrm{LKJCorrCholesky}(\eta)$ in Stan:

- guarantees positive-definiteness automatically;
- avoids expensive matrix inversions;
- accelerates gradient calculations in HMC;
- integrates seamlessly into non-centered representations.

---

## 29. Coding changes the intercept-slope correlation

Consider $\eta_{ij} = \alpha_j + \beta_j x_{ij}$. If $x$ is shifted by constant $c$ ($x^* = x - c$), the new intercept becomes $\alpha_j^* = \alpha_j + c \beta_j$.

The new covariance is:

$$ \mathrm{Cov}(\alpha^*, \beta) = \mathrm{Cov}(\alpha, \beta) + c \mathrm{Var}(\beta) $$

Intercept-slope correlation depends directly on the location of $x = 0$.

---

## 30. Meaningful zero points

Because the intercept is evaluated at $x = 0$, $x$ should be centered at a scientifically meaningful value (e.g., sample mean $\bar{x}$, baseline age, or control group). Otherwise, $\rho$ measures correlation at an unobserved or arbitrary point.

---

## 31. Binary-predictor coding

For a binary indicator $U \in \{0, 1\}$ (e.g., rural=0, urban=1):

$$ \mathrm{logit}(p_{ij}) = a_j + b_j U_{ij} $$

- $a_j$: log-odds in the reference category ($U = 0$);
- $b_j$: log-odds contrast between $U = 1$ and $U = 0$.

Switching reference categories ($U^* = 1 - U$) flips the sign of $\mathrm{Cov}(a^*, b^*)$.

---

## 32. Bangladesh data structure

The Bangladesh dataset (`bangladesh`) contains contraceptive use $C_i \in \{0, 1\}$ across 61 districts $D_i$, with urban indicator $U_i \in \{0, 1\}$, centered age $A_i$, and living children $K_i$.

---

## 33. Uncorrelated Bangladesh model

Model with independent district intercepts and urban slopes:

$$ \mathrm{logit}(p_i) = a_{D_i} + b_{D_i} U_i $$

$$ a_d \sim \mathrm{Normal}(\bar{a}, \sigma_a), \quad b_d \sim \mathrm{Normal}(\bar{b}, \sigma_b), \quad a_d \perp\!\!\!\perp b_d $$

---

## 34. Correlated Bangladesh model

Joint district hierarchy:

$$ \begin{bmatrix} a_d \\ b_d \end{bmatrix} \sim \mathrm{MVNormal} \left( \begin{bmatrix} \bar{a} \\ \bar{b} \end{bmatrix}, \mathbf{S} \right) $$

$$ \mathbf{S} = \operatorname{diag}(\sigma_a, \sigma_b) \, \mathbf{R} \, \operatorname{diag}(\sigma_a, \sigma_b), \quad \mathbf{R} \sim \mathrm{LKJCorr}(4) $$

---

## 35. Rural and urban probabilities

For district $d$:

$$ p_{\text{rural}, d} = \mathrm{logit}^{-1}(a_d), \qquad p_{\text{urban}, d} = \mathrm{logit}^{-1}(a_d + b_d) $$

Plotting $(p_{\text{rural}, d}, p_{\text{urban}, d})$ displays district predictions in the probability plane relative to the identity line $p_{\text{urban}} = p_{\text{rural}}$.

---

## 36. Logit-scale versus probability-scale patterns

Logit-scale correlation $\rho_{ab}$ measures association between baseline log-odds and log-odds contrasts. Because the logit function is non-linear, a linear correlation on logit scale maps into curved, bounded probability patterns.

---

## 37. Correlated shrinkage in sparse districts

Districts with few urban observations borrow strength from their own rural observations through the learned population covariance $\mathbf{S}$.

---

## 38. Correlation is a population feature

$\rho_{ab}$ measures variation **across group parameters in the population**, not correlation among individual observations within a group or posterior parameter correlation within MCMC chains.

---

## 39. Correlation is not causation

A non-zero correlation $\rho_{ab} \ne 0$ indicates that group baselines and slopes co-occur in the population. It does not establish that baseline propensity causes slope sensitivity.

---

## 40. Omitted variables can create feature covariance

An omitted group-level variable $Z_j$ that influences both baseline $\alpha_j$ and slope $\beta_j$ creates non-zero covariance $\mathrm{Cov}(\alpha_j, \beta_j) \ne 0$ even without direct causal interaction between $\alpha$ and $\beta$.

---

## 41. Ceiling and floor effects

Groups with extreme baseline probabilities ($p \to 0$ or $p \to 1$) face mathematical bounds on potential probability changes, inducing natural logit-scale and probability-scale covariances.

---

## 42. Imbalance and missing cells

If a district has zero urban observations, an uncorrelated model cannot estimate its urban slope $b_d$ from data. A correlated model infers $b_d$ deterministically from $a_d$ via the population covariance matrix.

---

## 43. New-group prediction

To predict for a new, unobserved group:

1. Draw hyperparameter samples $(\boldsymbol{\mu}^{(s)}, \mathbf{D}^{(s)}, \mathbf{L}_R^{(s)})$;
2. Draw independent standard normal scores $\boldsymbol{z}_{\text{new}}^{(s)} \sim \mathrm{Normal}(\mathbf{0}, \mathbf{I})$;
3. Construct joint group parameters $\boldsymbol{\theta}_{\text{new}}^{(s)} = \boldsymbol{\mu}^{(s)} + \mathbf{D}^{(s)} \mathbf{L}_R^{(s)} \boldsymbol{z}_{\text{new}}^{(s)}$.

Drawing $\alpha_{\text{new}}$ and $\beta_{\text{new}}$ independently would destroy learned population correlation.

---

## 44. Conditional prediction of one feature from another

Given an observed group intercept $\alpha_j$, the conditional distribution of its slope is:

$$ \mathbb{E}[\beta_j \mid \alpha_j] = \bar{\beta} + \rho \frac{\sigma_\beta}{\sigma_\alpha} (\alpha_j - \bar{\alpha}) $$

$$ \operatorname{Var}(\beta_j \mid \alpha_j) = \sigma_\beta^2 (1 - \rho^2) $$

---

## 45. The chimpanzee four-feature hierarchy

In the chimpanzee model with 4 treatment conditions:

$$ \mathrm{logit}(p_{ij}) = \bar{a}_{A_i} + a_{A_i, T_i} + \bar{b}_{B_i} + b_{B_i, T_i} $$

Each actor $a$ has a 4-dimensional treatment vector $\boldsymbol{a}_a = [a_{a,1}, a_{a,2}, a_{a,3}, a_{a,4}]^{\mathsf T}$.

---

## 46. Actor covariance matrix

The 4-dimensional actor treatment vectors follow:

$$ \boldsymbol{a}_a \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_A) $$

where $\mathbf{S}_A = \mathbf{D}_A \mathbf{R}_A \mathbf{D}_A$ is a $4 \times 4$ covariance matrix containing 4 scale parameters and 6 unique correlation parameters.

---

## 47. Block covariance matrix

Similarly, experimental blocks have a $4 \times 4$ covariance matrix $\mathbf{S}_B = \mathbf{D}_B \mathbf{R}_B \mathbf{D}_B$ describing block-treatment interaction covariance.

---

## 48. Actor and block baselines

Separate scalar priors govern actor baseline $\bar{a}_A \sim \mathrm{Normal}(0, \tau_A)$ and block baseline $\bar{b}_B \sim \mathrm{Normal}(0, \tau_B)$, while the $4 \times 4$ matrices treat treatment deviations as zero-centered.

---

## 49. Correlation matrices grow quickly

A $K$-dimensional feature vector contains $K(K-1)/2$ unique correlation parameters:

- $K=2$: 1 correlation;
- $K=4$: 6 correlations;
- $K=10$: 45 correlations.

High-dimensional correlation matrices require strong regularization ($\eta \ge 4$) when group count $J$ is small.

---

## 50. Non-centered chimpanzee covariance model

The non-centered formulation uses Cholesky factors $\mathbf{L}_{R_A} \sim \mathrm{LKJCorrCholesky}(4)$ and $\mathbf{L}_{R_B} \sim \mathrm{LKJCorrCholesky}(4)$ with standard normal parameter matrices $\mathbf{Z}_A$ and $\mathbf{Z}_B$.

---

## 51. ESS comparison

Comparing bulk and tail ESS across centered and non-centered MVN implementations highlights parameters where Cholesky non-centering resolves sampling bottlenecks.

---

## 52. Component plots versus full predictions

Plotting individual actor components $\mathrm{logit}^{-1}(\bar{a}_a + a_{a,t})$ displays relative actor preferences, but full outcome probabilities require combining actor, block, and baseline terms: $\mathrm{logit}^{-1}(\bar{a}_a + a_{a,t} + \bar{b}_b + b_{b,t})$.

---

## 53. Prior predictive checks

Simulate covariance matrices $\mathbf{R} \sim \mathrm{LKJCorr}(\eta)$ and feature vectors $\boldsymbol{u}_j \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S})$ from priors to verify that implied outcome distributions across group lines remain scientifically reasonable.

---

## 54. Posterior predictive checks

Verify that replicated data generated from the fitted correlated model reproduce observed group-level means, contrasts, and cross-feature covariances.

---

## 55. Prior-posterior correlation comparison

Overlaying the posterior density of correlation $\rho$ against its LKJ prior density quantifies how much information the group data provided about feature alignment.

---

## 56. Code Demonstrations & Multi-Framework Implementation

### 56.1 `UCBadmit` Correlated Intercept-Slope Model (R ulam, Stan, PyMC)

#### R rethinking / ulam

```{r ucb-admit-correlated}
library(rethinking)
data(UCBadmit)
d <- UCBadmit

dat_ucb <- list(
  A = d$admit,
  N = d$applications,
  D = as.integer(d$dept),
  F = ifelse(d$applicant.gender == "female", 1L, 0L)
)

# Non-centered Cholesky correlated intercept-slope model
m_ucb_cov <- ulam(
  alist(
    A ~ dbinom(N, p),
    logit(p) <- a_bar + a[D] + (b_bar + b[D]) * F,
    
    # Non-centered Cholesky representation
    transpars> vector[6]:a <<- v[, 1],
    transpars> vector[6]:b <<- v[, 2],
    transpars> matrix[6, 2]:v <- compose_noncentered(sigma, L_Rho, Z),
    
    # Priors
    matrix[2, 6]:Z ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1),
    b_bar ~ dnorm(0, 1),
    vector[2]:sigma ~ dexp(1),
    cholesky_factor_corr[2]:L_Rho ~ lkj_corr_cholesky(4),
    
    # Generated correlation matrix
    gq> matrix[2, 2]:Rho <<- Chol_to_Corr(L_Rho)
  ),
  data = dat_ucb,
  chains = 4,
  cores = 4
)

precis(m_ucb_cov, depth = 3, pars = c("a_bar", "b_bar", "sigma", "Rho"))
```

#### Stan (CmdStanR)

```stan
// ucb_admit_correlated.stan
data {
  int<lower=1> N;
  array[N] int<lower=0> A;
  array[N] int<lower=1> App;
  array[N] int<lower=1> D;
  array[N] int<lower=0, upper=1> F;
  int<lower=1> N_dept;
}
parameters {
  real a_bar;
  real b_bar;
  vector<lower=0>[2] sigma;
  cholesky_factor_corr[2] L_Rho;
  matrix[2, N_dept] Z;
}
transformed parameters {
  matrix[N_dept, 2] v = (diag_pre_multiply(sigma, L_Rho) * Z)';
  vector[N_dept] a = v[, 1];
  vector[N_dept] b = v[, 2];
}
model {
  a_bar ~ normal(0, 1);
  b_bar ~ normal(0, 1);
  sigma ~ exponential(1);
  L_Rho ~ lkj_corr_cholesky(4);
  to_vector(Z) ~ normal(0, 1);
  
  for (i in 1:N) {
    A[i] ~ binomial_logit(App[i], a_bar + a[D[i]] + (b_bar + b[D[i]]) * F[i]);
  }
}
generated quantities {
  matrix[2, 2] Rho = multiply_lower_tri_self_transpose(L_Rho);
}
```

#### PyMC (Python)

```python
import pymc as pm
import pandas as pd
import numpy as np

# Data setup
df_ucb = pd.DataFrame({
    'admit': [512, 89, 353, 17, 120, 202, 138, 131, 53, 94, 22, 24],
    'applications': [825, 108, 560, 25, 325, 417, 417, 375, 191, 393, 373, 341],
    'dept': [0, 0, 1, 1, 2, 2, 3, 3, 4, 4, 5, 5],
    'female': [0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1]
})

n_dept = 6

with pm.Model() as ucb_cov_pymc:
    a_bar = pm.Normal("a_bar", mu=0, sigma=1.0)
    b_bar = pm.Normal("b_bar", mu=0, sigma=1.0)
    
    sigma = pm.Exponential("sigma", lam=1.0, shape=2)
    
    # LKJ Cholesky covariance factor
    L_Rho, Rho, _ = pm.LKJCholeskyCov("L_Rho", n=2, eta=4.0, sd_dist=pm.Exponential.dist(lam=1.0))
    
    # Non-centered group offsets
    Z = pm.Normal("Z", mu=0, sigma=1.0, shape=(2, n_dept))
    v = pm.math.dot(L_Rho, Z).T
    
    a = v[:, 0]
    b = v[:, 1]
    
    dept_idx = df_ucb['dept'].values
    female_idx = df_ucb['female'].values
    
    logit_p = a_bar + a[dept_idx] + (b_bar + b[dept_idx]) * female_idx
    p = pm.Deterministic("p", pm.math.invlogit(logit_p))
    
    A = pm.Binomial("A", n=df_ucb['applications'].values, p=p, observed=df_ucb['admit'].values)
    
    idata_ucb_cov = pm.sample(draws=1000, tune=1000, chains=4, random_seed=42)
```

---

### 56.2 Bangladesh Contraception District Model (R ulam)

```{r bangladesh-correlated}
library(rethinking)
data(bangladesh)
d <- bangladesh

dat_bangla <- list(
  C = d$use.contraception,
  D = as.integer(as.factor(d$district)),
  U = d$urban
)
n_districts <- max(dat_bangla$D)

m_bangla_cov <- ulam(
  alist(
    C ~ dbern(p),
    logit(p) <- abar[1] + v[D, 1] + (abar[2] + v[D, 2]) * U,
    
    transpars> matrix[n_districts, 2]:v <- compose_noncentered(sigma, L_Rho, Z),
    
    matrix[2, n_districts]:Z ~ dnorm(0, 1),
    vector[2]:abar ~ dnorm(0, 1),
    vector[2]:sigma ~ dexp(1),
    cholesky_factor_corr[2]:L_Rho ~ lkj_corr_cholesky(4),
    
    gq> matrix[2, 2]:Rho <<- Chol_to_Corr(L_Rho)
  ),
  data = dat_bangla,
  chains = 4,
  cores = 4
)

precis(m_bangla_cov, depth = 3, pars = c("abar", "sigma", "Rho"))
```

---

### 56.3 Chimpanzee 4-Feature Actor & Block Covariance Matrix Model

```{r chimp-4feature-cov}
library(rethinking)
data(chimpanzees)
d <- chimpanzees
d$treatment <- 1L + d$prosoc_left + 2L * d$condition

dat_chimp4 <- list(
  P = d$pulled_left,
  A = as.integer(d$actor),
  B = as.integer(d$block),
  T = as.integer(d$treatment)
)

m_chimp_4cov <- ulam(
  alist(
    P ~ dbern(p),
    logit(p) <- a_bar[A] + a[A, T] + b_bar[B] + b[B, T],
    
    # Non-centered 4x4 actor covariance
    transpars> matrix[7, 4]:a <- compose_noncentered(sigma_A, L_Rho_A, Z_A),
    # Non-centered 4x4 block covariance
    transpars> matrix[6, 4]:b <- compose_noncentered(sigma_B, L_Rho_B, Z_B),
    
    matrix[4, 7]:Z_A ~ dnorm(0, 1),
    matrix[4, 6]:Z_B ~ dnorm(0, 1),
    
    vector[7]:a_bar ~ dnorm(0, 1),
    vector[6]:b_bar ~ dnorm(0, 1),
    
    vector[4]:sigma_A ~ dexp(1),
    vector[4]:sigma_B ~ dexp(1),
    
    cholesky_factor_corr[4]:L_Rho_A ~ lkj_corr_cholesky(4),
    cholesky_factor_corr[4]:L_Rho_B ~ lkj_corr_cholesky(4),
    
    gq> matrix[4, 4]:Rho_A <<- Chol_to_Corr(L_Rho_A),
    gq> matrix[4, 4]:Rho_B <<- Chol_to_Corr(L_Rho_B)
  ),
  data = dat_chimp4,
  chains = 4,
  cores = 4,
  sample = FALSE
)
```

---

### 56.4 Cholesky Decomposition & Non-Centered Gaussian Transformation Demonstration

```{r cholesky-demo}
library(rethinking)

# Defined 2x2 correlation matrix
R_true <- matrix(c(1, -0.6, -0.6, 1), nrow = 2)
sigma_true <- c(1.5, 0.8)
D_true <- diag(sigma_true)
S_true <- D_true %*% R_true %*% D_true

# Lower-triangular Cholesky factor
L_R <- t(chol(R_true))  # base R chol() returns upper-triangular, so transpose

# Draw independent standard normal scores Z
set.seed(42)
N_sim <- 1000
Z_scores <- matrix(rnorm(2 * N_sim), nrow = 2, ncol = N_sim)

# Construct correlated feature vectors U = D * L_R * Z
U_correlated <- t(D_true %*% L_R %*% Z_scores)

# Empirical covariance check
cat("Empirical Covariance:\n")
print(round(cov(U_correlated), 3))
cat("Target Covariance:\n")
print(round(S_true, 3))
```

---

### 56.5 Joint Posterior Predictive Simulation for New Clusters

```{r new-cluster-mvn-sim}
library(rethinking)

# Extract samples from fitted UCB correlated model
post <- extract.samples(m_ucb_cov)
n_samples <- length(post$a_bar)
n_new_dept <- 1000

# Draw new department intercepts and slopes jointly
a_new <- numeric(n_samples)
b_new <- numeric(n_samples)

set.seed(2026)
for (s in 1:n_samples) {
  # Reconstruct covariance matrix S = D * R * D for draw s
  D_s <- diag(post$sigma[s, ])
  R_s <- post$Rho[s, , ]
  S_s <- D_s %*% R_s %*% D_s
  
  # Draw joint feature vector (a_dev, b_dev)
  dev_draw <- rmvnorm(1, mean = c(0, 0), sigma = S_s)
  
  a_new[s] <- post$a_bar[s] + dev_draw[1]
  b_new[s] <- post$b_bar[s] + dev_draw[2]
}

# Compute joint probability predictions for female applicants in new departments
p_female_new <- inv_logit(a_new + b_new)
cat("New department female admission probability 89% PI:", HPDI(p_female_new, prob = 0.89), "\n")
```

---

## 57. Supplementary practice exercises and worked model specifications

### Exercise 14M1

**Question:** Write down the complete generative specification for a linear regression model with a continuous predictor $x$ and correlated varying intercepts and slopes across groups $j$. Specify the likelihood, link, multivariate normal prior, and LKJ hyperpriors.

**Solution:**

Let $Y_{ij}$ be observation $i$ in group $j$ ($j = 1, \dots, J$). Let $x_{ij}^* = x_{ij} - \bar{x}$ be the centered continuous predictor.

$$ Y_{ij} \sim \mathrm{Normal}(\mu_{ij}, \sigma_Y) $$

$$ \mu_{ij} = \alpha_j + \beta_j x_{ij}^* $$

The group feature vector $\boldsymbol{\theta}_j = \begin{bmatrix} \alpha_j \\ \beta_j \end{bmatrix}$ follows a bivariate normal distribution:

$$ \begin{bmatrix} \alpha_j \\ \beta_j \end{bmatrix} \sim \mathrm{MVNormal} \left( \begin{bmatrix} \bar{\alpha} \\ \bar{\beta} \end{bmatrix}, \mathbf{S} \right) $$

where $\mathbf{S} = \mathbf{D} \mathbf{R} \mathbf{D}$, $\mathbf{D} = \operatorname{diag}(\sigma_\alpha, \sigma_\beta)$, and $\mathbf{R} = \begin{bmatrix} 1 & \rho \\ \rho & 1 \end{bmatrix}$.

Hyperpriors:

$$ \bar{\alpha} \sim \mathrm{Normal}(0, 10), \qquad \bar{\beta} \sim \mathrm{Normal}(0, 2) $$

$$ \sigma_Y \sim \mathrm{Exponential}(1), \qquad \sigma_\alpha \sim \mathrm{Exponential}(1), \qquad \sigma_\beta \sim \mathrm{Exponential}(1) $$

$$ \mathbf{R} \sim \mathrm{LKJCorr}(2) $$

---

### Exercise 14M2

**Question:** Write down the `ulam` formula list for the model in 14M1 using Cholesky non-centered parameterization.

**Solution:**

```{r solution-14m2}
library(rethinking)

# Formula specification for Cholesky non-centered correlated line model
f_14m2 <- alist(
  Y ~ dnorm(mu, sigma_Y),
  mu <- abar + v[j, 1] + (bbar + v[j, 2]) * x_std,
  
  transpars> matrix[J, 2]:v <- compose_noncentered(sigma_group, L_Rho, Z),
  
  matrix[2, J]:Z ~ dnorm(0, 1),
  abar ~ dnorm(0, 10),
  bbar ~ dnorm(0, 2),
  sigma_Y ~ dexp(1),
  vector[2]:sigma_group ~ dexp(1),
  cholesky_factor_corr[2]:L_Rho ~ lkj_corr_cholesky(2),
  
  gq> matrix[2, 2]:Rho <<- Chol_to_Corr(L_Rho)
)
```

---

### Exercise 14H1

**Question:** Using the `bangladesh` dataset, fit a correlated varying-intercept-and-slope model for contraception use (`use.contraception`) across districts (`district`) with urban indicator (`urban`) and centered age (`age.centered`). Compare the estimated correlation $\rho_{ab}$ between baseline district use and urban contrast with an uncorrelated model.

**Solution:**

```{r solution-14h1}
library(rethinking)
data(bangladesh)
d <- bangladesh

dat_14h1 <- list(
  C = d$use.contraception,
  D = as.integer(as.factor(d$district)),
  U = d$urban,
  A = standardize(d$age.centered)
)
n_dist <- max(dat_14h1$D)

# Correlated model with age covariate
m_14h1_cov <- ulam(
  alist(
    C ~ dbern(p),
    logit(p) <- abar[1] + v[D, 1] + (abar[2] + v[D, 2]) * U + bA * A,
    
    transpars> matrix[n_dist, 2]:v <- compose_noncentered(sigma, L_Rho, Z),
    
    matrix[2, n_dist]:Z ~ dnorm(0, 1),
    vector[2]:abar ~ dnorm(0, 1),
    bA ~ dnorm(0, 0.5),
    vector[2]:sigma ~ dexp(1),
    cholesky_factor_corr[2]:L_Rho ~ lkj_corr_cholesky(4),
    
    gq> matrix[2, 2]:Rho <<- Chol_to_Corr(L_Rho)
  ),
  data = dat_14h1,
  chains = 4,
  cores = 4
)

precis(m_14h1_cov, depth = 3, pars = c("abar", "bA", "sigma", "Rho"))
```

---

## 58. Lecture takeaways

1. Group-specific features can be modeled jointly rather than independently.
2. A varying-slope model assigns each group its own predictor response.
3. Varying slopes are regularized interactions.
4. Intercepts and slopes may covary through shared group mechanisms.
5. A multivariate Gaussian population models means, scales, and covariance.
6. Covariance can be decomposed into marginal scales and a correlation matrix.
7. Joint pooling allows information to flow across parameter types.
8. Shrinkage occurs toward a population ellipse.
9. Learning correlation requires several groups.
10. Within-group observations and number of groups identify different components.
11. The admissions model estimates correlated department baselines and category contrasts.
12. Probability contrasts require inverse-link transformation.
13. LKJ priors regularize valid correlation matrices.
14. Larger $\eta$ values favor weaker correlations.
15. Correlation is weakly meaningful when one marginal scale is near zero.
16. Cholesky factors provide an efficient matrix representation.
17. Non-centering constructs correlated effects from independent standardized scores.
18. Predictor centering changes intercept-slope covariance.
19. Binary reference coding changes the interpretation of correlation.
20. The Bangladesh model correlates rural baseline use with the urban-rural contrast.
21. Logit-scale correlation differs from probability-scale patterns.
22. Correlated pooling is valuable in sparse or missing cells.
23. The chimpanzee model estimates four-dimensional actor and block feature populations.
24. High-dimensional covariance requires strong regularization when groups are few.
25. New-group features must be simulated jointly.
26. Feature correlation can suggest omitted variables but does not prove a mechanism.
27. Causal varying slopes still require treatment identification.
28. CausalOps should label correlation level, scale, reference point, and epistemic status.
29. The official reading says Chapter 13, while the exact second-edition treatment is Chapter 14.

---

## 59. Review questions

1. What does a correlated varying-slope model add beyond independent hierarchies?
2. Why can learning an intercept improve inference about a slope?
3. What is the group feature vector?
4. How is the covariance matrix built from scales and correlations?
5. What does the off-diagonal covariance equal?
6. Why is the multivariate Gaussian convenient?
7. What are the components of the `UCBadmit` predictor?
8. What do $a_d$ and $b_d$ represent?
9. How is the probability-scale department contrast computed?
10. What does the population ellipse represent?
11. Why are several groups required to estimate correlation?
12. How does joint shrinkage differ from separate shrinkage?
13. What does the LKJ prior regularize?
14. What does $\eta=1$ imply?
15. Why can a high correlation be unimportant when one scale is near zero?
16. What is a Cholesky factor?
17. How does multivariate non-centering work?
18. Why does Stan use Cholesky-factor correlation parameters?
19. How does centering $x$ change the intercept?
20. How does it change covariance with the slope?
21. What does the intercept mean in the Bangladesh model?
22. What does the slope mean?
23. Why is the modeled correlation not a rural-urban probability correlation?
24. How does correlated pooling help sparse districts?
25. What are the chimpanzee actor and block feature vectors?
26. How many correlations exist in a four-by-four matrix?
27. Why are these difficult to identify with few actors and blocks?
28. Why are component probabilities not full response probabilities?
29. What should a prior predictive simulation examine?
30. What does strong posterior-prior overlap for $\rho$ imply?
31. How should new group intercepts and slopes be simulated?
32. Why is feature correlation not a causal conclusion?
33. How can omitted group variables generate covariance?
34. Why can targeted assignment distort slope covariance?
35. What does a within-between decomposition separate?
36. How would you model correlated campaign response across channels?
37. Which metadata should CausalOps store for a correlation matrix?

---

## 60. Book reading guide

### Primary reading

- Chapter 13, according to the 2026 course schedule.

### Closest second-edition reading

- Chapter 14, *Adventures in Covariance*
  - Section 14.1, *Varying slopes by construction*
  - Section 14.2, *Advanced varying slopes*

---

## 61. Official code guide

### `14_GLMM_slopes_.r` & `14_varying_slopes_bangladesh.R`

Contains `UCBadmit` correlated hierarchy, sequential covariance animation, 2D shrinkage animation, chimpanzee 4-feature actor/block covariance models, Cholesky matrix decomposition, and Bangladesh rural-urban district covariance models.

---

## 62. Sources

- Richard McElreath, *Statistical Rethinking: A Bayesian Course with Examples in R and Stan*, second edition, Chapters 13–14.
- Official `stat_rethinking_2026` repository and B03 lecture listing.
- Official scripts `14_GLMM_slopes_.r` and `14_varying_slopes_bangladesh.R`.

---

## 70. Annex: Complete Solutions to Review Questions

#### 1. What does a correlated varying-slope model add beyond independent hierarchies?

A correlated varying-slope model estimates off-diagonal covariance terms ($\mathrm{Cov}(\alpha_j, \beta_j) = \rho \sigma_\alpha \sigma_b$) in a joint multivariate normal population prior. This allows information to share cross-dimensionally between parameters (e.g., intercept informing slope) and enables directional 2D shrinkage along joint confidence ellipses.

#### 2. Why can learning an intercept improve inference about a slope?

If population correlation $\rho \ne 0$, observing a group's baseline intercept $\alpha_j$ constrains the conditional distribution of its slope $p(\beta_j \mid \alpha_j) = \mathrm{Normal}\left(\bar{\beta} + \rho \frac{\sigma_\beta}{\sigma_\alpha}(\alpha_j - \bar{\alpha}), \sigma_\beta \sqrt{1-\rho^2}\right)$, reducing uncertainty about $\beta_j$ even before dense slope data arrive.

#### 3. What is the group feature vector?

The group feature vector $\boldsymbol{\theta}_j = [\alpha_j, \beta_{1j}, \dots, \beta_{Kj}]^{\mathsf T}$ collects all parameters assigned to cluster $j$ that vary together across groups in a joint multivariate population distribution.

#### 4. How is the covariance matrix built from scales and correlations?

The covariance matrix $\mathbf{S}$ is constructed via scale-correlation decomposition $\mathbf{S} = \mathbf{D} \mathbf{R} \mathbf{D}$, where $\mathbf{D} = \operatorname{diag}(\sigma_1, \dots, \sigma_K)$ is the diagonal matrix of marginal standard deviations and $\mathbf{R}$ is the symmetric positive-definite correlation matrix.

#### 5. What does the off-diagonal covariance equal?

The off-diagonal element at row $m$ and column $n$ equals $S_{mn} = \rho_{mn} \sigma_m \sigma_n$.

#### 6. Why is the multivariate Gaussian convenient?

The multivariate Gaussian is maximum entropy for continuous unbounded parameters specified by mean vector $\boldsymbol{\mu}$ and covariance matrix $\mathbf{S}$. It provides analytical conditional distributions and linear Cholesky parameterizations.

#### 7. What are the components of the `UCBadmit` predictor?

In $\mathrm{logit}(p_i) = \bar{a} + a_{D_i} + (\bar{b} + b_{D_i}) F_i$: $\bar{a}$ is global baseline log-odds, $a_d$ is department baseline deviation, $\bar{b}$ is global female contrast, and $b_d$ is department contrast deviation.

#### 8. What do $a_d$ and $b_d$ represent?

- $a_d$: department $d$'s baseline log-odds deviation from global mean $\bar{a}$;
- $b_d$: department $d$'s female-minus-male contrast deviation from global contrast $\bar{b}$.

#### 9. How is the probability-scale department contrast computed?

$\Delta_d = \mathrm{logit}^{-1}(\bar{a} + a_d + \bar{b} + b_d) - \mathrm{logit}^{-1}(\bar{a} + a_d)$.

#### 10. What does the population ellipse represent?

The population ellipse represents contours of constant probability density under the bivariate normal distribution $\mathrm{MVNormal}(\mathbf{0}, \mathbf{S})$ in feature space $(a_d, b_d)$.

#### 11. Why are several groups required to estimate correlation?

Because $\rho$ describes variation *across groups*, observing 2 or 3 groups provides minimal degrees of freedom to identify population orientation. Accurately estimating $\rho$ requires $J \ge 15-20$ groups.

#### 12. How does joint shrinkage differ from separate shrinkage?

- **Separate shrinkage:** Intercepts and slopes shrink independently toward $(\bar{a}, \bar{b})$ along coordinate axes.
- **Joint shrinkage:** Features shrink directionally along the major axis of the population covariance ellipse.

#### 13. What does the LKJ prior regularize?

The LKJ prior $\mathrm{LKJCorr}(\eta)$ regularizes valid correlation matrices $\mathbf{R}$, penalizing extreme correlations ($\rho \to \pm 1$) when $\eta > 1$.

#### 14. What does $\eta=1$ imply?

$\eta = 1$ assigns uniform probability density over the entire space of valid symmetric positive-definite correlation matrices.

#### 15. Why can a high correlation be unimportant when one scale is near zero?

If slope standard deviation $\sigma_\beta \to 0$, the covariance $\mathrm{Cov}(\alpha, \beta) = \rho \sigma_\alpha \sigma_\beta \to 0$, making all group slopes essentially identical regardless of $\rho$.

#### 16. What is a Cholesky factor?

A Cholesky factor $\mathbf{L}_R$ is a lower-triangular matrix satisfying $\mathbf{R} = \mathbf{L}_R \mathbf{L}_R^{\mathsf T}$.

#### 17. How does multivariate non-centering work?

By sampling independent standard normal matrix $\mathbf{Z} \sim \mathrm{Normal}(0, 1)$ and constructing correlated group vectors deterministically: $\mathbf{V} = (\mathbf{D} \mathbf{L}_R \mathbf{Z})^{\mathsf T}$.

#### 18. Why does Stan use Cholesky-factor correlation parameters?

Stan samples $\mathbf{L}_R \sim \mathrm{LKJCorrCholesky}(\eta)$ because lower-triangular factors maintain positive-definiteness automatically and simplify matrix multiplication in non-centered transforms.

#### 19. How does centering $x$ change the intercept?

Shifting $x^* = x - c$ changes the intercept to $\alpha_j^* = \alpha_j + c \beta_j$, moving the zero-point of the linear predictor.

#### 20. How does it change covariance with the slope?

New covariance is $\mathrm{Cov}(\alpha^*, \beta) = \mathrm{Cov}(\alpha, \beta) + c \operatorname{Var}(\beta)$, altering correlation magnitude and sign.

#### 21. What does the intercept mean in the Bangladesh model?

In $\mathrm{logit}(p_i) = a_d + b_d U_i$ with $U \in \{0, 1\}$, $a_d$ is the rural baseline log-odds of contraceptive use in district $d$.

#### 22. What does the slope mean?

$b_d$ is the urban-minus-rural log-odds contrast in contraceptive use for district $d$.

#### 23. Why is the modeled correlation not a rural-urban probability correlation?

Because $\rho_{ab}$ measures correlation between log-odds intercept $a_d$ and log-odds contrast $b_d$, whereas probability correlation involves non-linear transformations $(p_{\text{rural}}, p_{\text{urban}})$.

#### 24. How does correlated pooling help sparse districts?

Districts with zero urban observations can infer their urban slope $b_d$ from their rural baseline $a_d$ through population covariance $\mathbf{S}$.

#### 25. What are the chimpanzee actor and block feature vectors?

Each actor $a$ has a 4-vector $\boldsymbol{a}_a = [a_{a,1}, a_{a,2}, a_{a,3}, a_{a,4}]^{\mathsf T}$ of treatment responses, and each block $b$ has a matching 4-vector $\boldsymbol{b}_b$.

#### 26. How many correlations exist in a four-by-four matrix?

A $4 \times 4$ correlation matrix contains $4(4-1)/2 = 6$ unique off-diagonal correlation parameters.

#### 27. Why are these difficult to identify with few actors and blocks?

With only 7 actors and 6 blocks, estimating 6 correlations per matrix suffers from small sample size at the group level, making inferences heavily reliant on the LKJ prior.

#### 28. Why are component probabilities not full response probabilities?

Component displays $\mathrm{logit}^{-1}(\bar{a}_a + a_{a,t})$ isolate actor contributions, whereas full response probabilities require adding block effects: $\mathrm{logit}^{-1}(\bar{a}_a + a_{a,t} + \bar{b}_b + b_{b,t})$.

#### 29. What should a prior predictive simulation examine?

Prior predictive simulation should sample correlation matrices $\mathbf{R}$ and scale vectors $\mathbf{D}$ to inspect generated population ellipses, group line distributions, and outcome probability spreads.

#### 30. What does strong posterior-prior overlap for $\rho$ imply?

High overlap indicates that group data provided minimal information to update correlation $\rho$, leaving posterior inference dominated by the LKJ prior.

#### 31. How should new group intercepts and slopes be simulated?

Draw joint feature vectors $\boldsymbol{\theta}_{\text{new}} \sim \mathrm{MVNormal}(\boldsymbol{\mu}, \mathbf{S})$ using the joint covariance matrix rather than drawing intercepts and slopes independently.

#### 32. Why is feature correlation not a causal conclusion?

Correlation $\rho_{ab} \ne 0$ describes co-occurrence of baseline and slope parameters in a population. It does not prove that baseline propensity causes slope sensitivity.

#### 33. How can omitted group variables generate covariance?

An unmeasured group factor $Z_j$ influencing both baseline $\alpha_j$ and slope $\beta_j$ creates non-zero feature covariance $\mathrm{Cov}(\alpha, \beta) \ne 0$ without direct causation between $\alpha$ and $\beta$.

#### 34. Why can targeted assignment distort slope covariance?

If policy targets treatment to high-baseline groups, baseline and slope parameters correlate due to selection rather than intrinsic response heterogeneity.

#### 35. What does a within-between decomposition separate?

A Mundlak decomposition $\eta_{ij} = \alpha_j + \beta_W (X_{ij} - \bar{X}_j) + \beta_B \bar{X}_j$ separates individual within-group treatment effects $\beta_W$ from group-level contextual effects $\beta_B$.

#### 36. How would you model correlated campaign response across channels?

Assign a joint $K$-dimensional MVNormal prior $\boldsymbol{\beta}_j \sim \mathrm{MVNormal}(\boldsymbol{\bar{\beta}}, \mathbf{S}_\beta)$ over channel treatment slopes for customer segment $j$, using Cholesky LKJ non-centering.

#### 37. Which metadata should CausalOps store for a correlation matrix?

CausalOps should store feature names, predictor coding/centering reference points, parameter scale (log-odds), LKJ shape parameter $\eta$, Cholesky decomposition status, ESS/Rhat diagnostics, and prior-posterior overlap metrics.
