---
title: "B04 — Group-level Confounds / Social Networks I"
subtitle: "Statistical Rethinking 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  eval: false
  warning: false
  message: false
knitr:
  opts_chunk:
    python.reticulate: false
---

# B04 — Group-level Confounds / Social Networks I

**Course:** Statistical Rethinking 2026  
**Section:** B — Advanced  
**Lecture:** B04 — Group-level Confounds / Social Networks I  
**Meeting date:** 30 January 2026  
**Official reading:** Chapter 14  
**Closest second-edition correspondence:** Chapter 14, especially Section 14.4, with Chapters 6, 13, and Sections 14.1–14.2 as foundations  
**Central question:** How can a multilevel model distinguish within-group from between-group information when group membership is confounded, and how can correlated varying effects represent directed social relationships?

> **Source note**
>
> This guide is grounded in the official 2026 course outline, the supplied second edition of *Statistical Rethinking*, the official scripts `B04_bonus_mundlak.r` and `B04_social_networks.r`, and `homework_B04.pdf`.
>
> The Mundlak script directly supports:
>
> - simulation of an unobserved group-level cause of both an individual exposure and an outcome;
> - comparison of naive, fixed-effects, ordinary varying-effects, Mundlak, and latent-Mundlak models;
> - intervention-based evaluation rather than coefficient recovery alone;
> - partial pooling of group intercepts;
> - comparison of group-specific probability-scale intervention effects.
>
> The social-network script directly supports:
>
> - paired directed outcomes within unordered dyads;
> - correlated dyad effects as a model of reciprocity or relationship intensity;
> - generalized giving and generalized receiving effects for households;
> - non-centered multivariate varying effects with Cholesky factors;
> - posterior network reconstruction;
> - household wealth and dyad production predictors;
> - separation of actor, receiver, and dyad variation.
>
> The book's Section 14.4 develops the social-relations model. The group-confounding and Mundlak material is a 2026 lecture extension of the multilevel and causal foundations in Chapters 6, 13, and 14. The term **Mundlak** is therefore script- and lecture-specific rather than a section heading in the supplied second edition.

---

## 1. Main thesis

The lecture joins two problems that initially look different:

1. observations from the same group share causes;
2. observations from the same dyad share a relationship.

Both problems require the model to represent dependence at the level where it arises.

The central causal warning is:

> **Partial pooling is not automatically deconfounding. If group effects are associated with an exposure, an ordinary varying-intercept model can mix within-group and between-group information.**

The central network insight is:

> **A directed interaction is generated jointly by a sender, a receiver, and their dyad. Correlated varying effects let the model learn each source of dependence.**

The workflow is:

$$ \text{causal structure} \rightarrow \text{level of variation} \rightarrow \text{within-between decomposition} \rightarrow \text{correlated group or dyad effects} \rightarrow \text{posterior intervention or network summary} $$

---

## 2. Relationship to the book and earlier lectures

| B04 topic | Closest source | Role |
|---|---|---|
| Confounding and DAG logic | Chapter 6 | Defines open non-causal paths |
| Varying intercepts | Chapter 13 | Models group heterogeneity and pooling |
| Same- versus new-group prediction | Section 13.5 | Clarifies the target population |
| Correlated varying features | Sections 14.1–14.2 and B03 | Supplies covariance machinery |
| Group-level confounding | 2026 Mundlak script | Separates within- and between-group information |
| Directed dyads | Section 14.4 | Models both directions of exchange jointly |
| Generalized giving and receiving | Section 14.4 | Adds household sender/receiver features |
| Reciprocity | Section 14.4 | Learned as covariance within dyads |
| Wealth and production predictors | B04 social-network script | Explains part of household and dyad variation |

The conceptual sequence is:

$$ \text{B01: partial pooling} \rightarrow \text{B02: multiple hierarchies} \rightarrow \text{B03: correlated features} \rightarrow \text{B04: causal group structure and relational data} $$

B03 asks whether group features covary.

B04 asks two harder questions:

- Is that covariance associated with the exposure, so that ordinary pooling can bias a causal estimate?
- Can covariance itself encode a social relation, such as reciprocal exchange?

---

## 3. Why ordinary regression fails with grouped data

Let individual $i$ belong to group $j$.

Suppose:

- $X_{ij}$ is an individual exposure;
- $Y_{ij}$ is an outcome;
- $U_j$ is an unobserved group cause.

A simple causal structure is:

$$ U_j \rightarrow X_{ij} $$

$$ U_j \rightarrow Y_{ij} $$

and possibly:

$$ X_{ij} \rightarrow Y_{ij} $$

The backdoor path is:

$$ X_{ij} \leftarrow U_j \rightarrow Y_{ij} $$

The naive model:

$$ \mathrm{logit}(p_{ij}) = \alpha+\beta_X X_{ij} $$

cannot distinguish:

- differences among people inside the same group;
- differences between groups with different average exposure.

It therefore combines two associations that need not have the same causal meaning.

---

## 4. The official simulation

The script generates group membership:

```r
g <- sample(1:N_groups, size=N_id, replace=TRUE)
```

and an unobserved group feature:

```r
Ug <- rnorm(N_groups, 1.5)
```

The individual exposure depends on that group feature:

```r
X <- rnorm(N_id, Ug[g])
```

The binary outcome also depends on it:

```r
Y <- rbern(
  N_id,
  p = inv_logit(a0 + bXY*X + Ug[g] + bZY*Z[g])
)
```

In mathematical form:

$$ X_{ij} \sim \mathrm{Normal}(U_j,1) $$

$$ Y_{ij} \sim \mathrm{Bernoulli}(p_{ij}) $$

$$ \mathrm{logit}(p_{ij}) = \alpha_0 + \beta_{XY}X_{ij} + U_j + \beta_{ZY}Z_j $$

The lecture sets the true individual effect to zero in the main demonstration:

$$ \beta_{XY}=0 $$

Yet $X$ and $Y$ remain associated because both inherit variation from $U_j$.

This is a calibration exercise: because the data-generating truth is known, competing estimators can be judged against the correct intervention.

---

## 5. Three levels of variables

The simulation contains variables at different levels.

| Variable | Level | Varies within group? | Varies between groups? |
|---|---:|---:|---:|
| $Y_{ij}$ | Individual | Yes | Yes |
| $X_{ij}$ | Individual | Yes | Yes |
| $Z_j$ | Group | No | Yes |
| $U_j$ | Group, unobserved | No | Yes |
| $g_i$ | Membership index | No | Identifies group |

This classification determines what the data can identify.

A group-level predictor such as $Z_j$ cannot be distinguished from a complete set of unrestricted group intercepts using only cross-sectional variation.

An individual predictor such as $X_{ij}$ can be compared within a group, provided there is sufficient within-group variation.

---

## 6. The within-between decomposition

For each group, define the exposure mean:

$$ \bar X_j = \frac{1}{n_j} \sum_{i:g_i=j}X_{ij} $$

Then decompose every exposure value:

$$ X_{ij} = \bar X_j + \left(X_{ij}-\bar X_j\right) $$

### Within-group component

$$ X_{ij}-\bar X_j $$

compares people to others in the same group.

### Between-group component

$$ \bar X_j $$

compares groups with different average exposure.

If $U_j$ causes both $\bar X_j$ and $Y_{ij}$, the between-group association is confounded.

The within-group contrast can still be useful because $U_j$ is constant inside group $j$.

---

## 7. The naive model

The official naive model is:

$$ Y_i \sim \mathrm{Bernoulli}(p_i) $$

$$ \mathrm{logit}(p_i) = \alpha + \beta_X X_i + \beta_Z Z_{g_i} $$

```{r naive-model}
m0 <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a + bxy*X + bzy*Z_g,
    a ~ dnorm(0, 10),
    c(bxy, bzy) ~ dnorm(0, 1)
  ), data=dat, chains=4, cores=4
)
```

The coefficient $\beta_X$ is informed by both $\mathrm{Cov} \left( X_{ij}-\bar X_j, Y_{ij} \right)$ and $\mathrm{Cov} \left( \bar X_j, \bar Y_j \right)$. The latter contains the open path through $U_j$.

---

## 8. Fixed group effects

A fixed-effects model assigns each group its own intercept:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_W X_{ij} $$

where the $\alpha_j$ are estimated without a shared population distribution.

```{r fixed-effects-model}
mf <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a[g] + bxy*X + bzy*Z_g,
    a[g] ~ dnorm(0, 10),
    c(bxy, bzy) ~ dnorm(0, 1)
  ), data=dat, chains=4, cores=4
)
```

### What fixed effects accomplish
- They block all time-constant group-level backdoor paths for the within-group exposure contrast.
- They do not require the group intercepts to be independent of $X$.

### What fixed effects cost
- There is no partial pooling of group intercepts.
- Sparse groups can have unstable intercept estimates.
- New-group prediction is not naturally defined.
- A purely group-level predictor is collinear with the group indicators.

---

## 9. Why a group-level coefficient becomes unidentified

Suppose $Z_j$ is constant for everyone in group $j$. The linear predictor is $\alpha_j+\beta_ZZ_j$.

Define a new intercept $\alpha_j^* = \alpha_j+\beta_ZZ_j$. Every value of $\beta_Z$ can be offset by a corresponding change in $\alpha_j$. The likelihood cannot distinguish them.

Fixed effects can identify a within-group exposure effect while making a group-level effect unidentified.

---

## 10. Demeaning as a fixed-effects transformation

The script also fits:

$$ \mathrm{logit}(p_{ij}) = \alpha + \beta_W \left(X_{ij}-\bar X_j\right) + \beta_Z Z_j $$

```{r demeaning-model}
mf2 <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a + bxy*(X-Xbar_g) + bzy*Z_g,
    a ~ dnorm(0, 10),
    c(bxy, bzy) ~ dnorm(0, 1)
  ), data=dat, chains=4, cores=4
)
```

The centered exposure has zero mean in each group: $\sum_{i:g_i=j} \left(X_{ij}-\bar X_j\right) =0$. Therefore its coefficient is learned strictly from within-group deviations.

---

## 11. The ordinary varying-intercept model

The multilevel model is:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_X X_{ij} + \beta_Z Z_j $$

$$ \alpha_j \sim \mathrm{Normal} \left( \bar\alpha, \tau \right) $$

```{r ordinary-varying-effects}
mr <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a[g] + bxy*X + bzy*Z_g,
    transpars> vector[Ng]:a <<- abar + z*tau,
    z[g] ~ dnorm(0, 1),
    c(bxy, bzy) ~ dnorm(0, 1),
    abar ~ dnorm(0, 1),
    tau ~ dexp(1)
  ), data=dat, chains=4, cores=4
)
```

This model pools noisy group intercepts, but its population prior assumes $\alpha_j \perp X_{ij} \mid \bar\alpha,\tau,\text{included predictors}$. When $U_j$ shifts $X$, partial pooling draws intercept variation and exposure variation together, biasing $\beta_X$.

---

## 12. The false dichotomy of fixed versus random effects

| Model | Group intercepts pooled? | Allows intercept–exposure association? | New-group prediction? |
|---|---:|---:|---:|
| Fixed group effects | No | Yes, implicitly | Poorly defined |
| Ordinary varying effects | Yes | No, conditionally | Yes |
| Mundlak varying effects | Yes | Yes, through group exposure mean | Yes |

The choice is which population assumptions are scientifically defensible for the estimand.

---

## 13. The Mundlak model

The Mundlak strategy adds the group exposure mean to the varying-intercept model:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_W X_{ij} + \beta_M\bar X_j + \beta_ZZ_j $$

$$ \alpha_j \sim \mathrm{Normal} \left( \bar\alpha, \tau \right) $$

```{r mundlak-model}
mrx <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a[g] + bxy*X + buy*Xbar_g + bzy*Z_g,
    transpars> vector[Ng]:a <<- abar + z*tau,
    z[g] ~ dnorm(0, 1),
    c(bxy, buy, bzy) ~ dnorm(0, 1),
    abar ~ dnorm(0, 1),
    tau ~ dexp(1)
  ), data=dat, chains=4, cores=4
)
```

The group mean acts as an observed proxy for group context. Conditional on $\bar X_j$, the residual varying intercept is less associated with individual $X_{ij}$.

---

## 14. Interpreting Mundlak coefficients

Substitute $X_{ij} = \left(X_{ij}-\bar X_j\right) + \bar X_j$ into $\eta_{ij} = \alpha_j + \beta_WX_{ij} + \beta_M\bar X_j$:

$$ \eta_{ij} = \alpha_j + \beta_W \left(X_{ij}-\bar X_j\right) + \left(\beta_W+\beta_M\right)\bar X_j $$

- $\beta_W$: within-group association;
- $\beta_W+\beta_M$: between-group association;
- $\beta_M$: difference between the between- and within-group slopes.

---

## 15. Equivalent centered Mundlak form

Written transparently:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_W \left(X_{ij}-\bar X_j\right) + \beta_B\bar X_j $$

where $\beta_B = \beta_W+\beta_M$. Here $\beta_W$ is the within-group slope and $\beta_B$ is the between-group slope.

---

## 16. Why Mundlak can keep pooling

Mundlak decomposes the group intercept: $\alpha_j^{\text{total}} = \bar\alpha + \beta_M\bar X_j + u_j$, with $u_j\sim\mathrm{Normal}(0,\tau)$. The residual $u_j$ can be pooled without demanding that total group intercept be unrelated to exposure composition.

---

## 17. What Mundlak does not guarantee

Mundlak does not solve: unmeasured individual confounding, time-varying group confounding, measurement error, positivity failures, or reverse causation. Its causal value depends on unobserved group causes being adequately represented by $\bar X_j$.

---

## 18. Positivity and within-group information

If $X_{ij}-\bar X_j = 0$ for all $i$ in group $j$, that group provides zero information about within-group slope. Partial pooling stabilizes estimation, but cannot manufacture missing within-group contrasts.

---

## 19. Nonlinear links and effect scales

In a logistic model, $\beta_W$ is a log-odds contrast. The group-specific probability contrast $\Delta_{ij} = \mathrm{logit}^{-1}(\eta_1) - \mathrm{logit}^{-1}(\eta_0)$ varies across groups because the inverse-logit transformation is non-linear.

---

## 20. The real estimand is an intervention

Evaluating a 1-unit intervention on $X$:

```{r intervention-eval}
# Copy data and intervene
dat1 <- dat
dat1$X <- dat1$X + 1

# Average Treatment Effect (ATE) across posterior draws
# ATE^(s) = (1/N) * sum_i ( p_i^(s)(X+1) - p_i^(s)(X) )
```

---

## 21. Conditional versus compositional interventions

- **Conditional intervention**: $X_{ij} \mapsto X_{ij} + 1$, holding $\bar X_j$ fixed.
- **Compositional intervention**: $X_{ij} \mapsto X_{ij} + 1$ and $\bar X_j \mapsto \bar X_j + 1$.

Both answer distinct scientific questions.

---

## 22. The latent Mundlak model

Joint generative model of exposure and outcome via latent factor $u_j$:

$$ \mathrm{logit}(p_{ij}) = \alpha_j + \beta_X X_{ij} + \beta_ZZ_j + \beta_{UY}u_j $$

$$ X_{ij} \sim \mathrm{Normal}(\alpha_X + \beta_{UX}u_j, \sigma_X), \qquad u_j \sim \mathrm{Normal}(0, 1) $$

```{r latent-mundlak}
mru <- ulam(
  alist(
    Y ~ bernoulli(p),
    logit(p) <- a[g] + bxy*X + bzy*Z_g + buy*u[g],
    transpars> vector[Ng]:a <<- abar + z*tau,
    X ~ normal(mu, sigma),
    mu <- aX + bux*u[g],
    vector[Ng]:u ~ normal(0, 1),
    z[g] ~ dnorm(0, 1),
    c(aX, bxy, buy, bzy) ~ dnorm(0, 1),
    bux ~ dexp(1),
    abar ~ dnorm(0, 1),
    tau ~ dexp(1),
    sigma ~ dexp(1)
  ), data=dat, chains=4, cores=4
)
```

---

## 23. Ordinary Mundlak versus latent Mundlak

| Feature | Observed-mean Mundlak | Latent Mundlak |
|---|---|---|
| Group context | $\bar X_j$ | Latent $u_j$ |
| Exposure model | Not required | Explicitly modeled |
| Distributional burden | Lower | Higher |
| Identification | Within-between variation | Joint exposure-outcome assumptions |

---

## 24. Intercept accuracy and slope accuracy

Partial pooling improves intercept accuracy, but a model can estimate intercepts well while estimating the causal slope poorly. Model performance is strictly estimand-specific.

---

## 25. Group-specific treatment effects on the probability scale

For group $j$: $\Delta_j = \mathrm{logit}^{-1}(\alpha_j+\beta_X) - \mathrm{logit}^{-1}(\alpha_j)$. A constant link-scale slope produces varying probability-scale treatment effects across groups.

---

## 26. A simulation comparison protocol

To compare models under simulation: compute Bias, RMSE, and Interval Coverage for the target probability-scale intervention across simulated datasets.

---

## 27. Homework B04: the Bangladesh estimand

The assignment evaluates the causal influence of living children ($K_i$) on contraceptive use ($C_i \in \{0, 1\}$) across districts ($D_i$).

---

## 28. Homework model 1: district fixed effects

$$ \mathrm{logit}(p_i) = \alpha_{D_i} + \beta_W K_i $$

Learned strictly from within-district contrasts.

---

## 29. Homework model 2: ordinary varying effects

$$ \mathrm{logit}(p_i) = \alpha_{D_i} + \beta_X K_i, \qquad \alpha_d \sim \mathrm{Normal}(\bar\alpha, \sigma_D) $$

Mixes between-district variation if district child averages are correlated with unobserved district factors.

---

## 30. Homework model 3: the Mundlak machine

$$ \mathrm{logit}(p_i) = \alpha_{D_i} + \beta_W K_i + \beta_M \bar K_{D_i}, \qquad \alpha_d \sim \mathrm{Normal}(\bar\alpha, \sigma_D) $$

Isolates within-district slope $\beta_W$ while retaining partial pooling of residual district intercepts.

---

## 31. How to compare the homework models

Compare posteriors of $\beta_W$, differences between within- and between-district slopes ($\beta_M$), district shrinkage, and probability-scale intervention contrasts.

---

## 32. Transition from groups to networks

Dyadic data assign each observation to two actors ($i \to j$). Standard regression ignoring actor/dyad cross-classification fails to represent network dependence.

---

## 33. The Koster–Leckie gift data

The dataset `kl_dyads` and `kl_households` records directed gift exchange counts $G_{AB}$ and $G_{BA}$ between households.

---

## 34. Ordered and unordered dyads

Among $N$ actors, there are $\binom{N}{2} = \frac{N(N-1)}{2}$ unordered dyads $\{A, B\}$, each containing two ordered directions $A \to B$ and $B \to A$.

---

## 35. A synthetic latent friendship process

Simulating shared dyad state $F_d \sim \mathrm{Bernoulli}(0.1)$ shows how mutual ties induce positive directed association without forcing exact directional equality.

---

## 36. The dyad-only social-relations model

$$ G_{AB,d} \sim \mathrm{Poisson}(\lambda_{AB,d}), \qquad G_{BA,d} \sim \mathrm{Poisson}(\lambda_{BA,d}) $$

$$ \begin{bmatrix} \log\lambda_{AB,d} \\ \log\lambda_{BA,d} \end{bmatrix} = \begin{bmatrix} \alpha + T_{d,AB} \\ \alpha + T_{d,BA} \end{bmatrix}, \qquad \begin{bmatrix} T_{d,AB} \\ T_{d,BA} \end{bmatrix} \sim \mathrm{MVNormal}\left(\mathbf{0}, \mathbf{S}_T\right) $$

---

## 37. Reciprocity as dyad correlation

$\rho_T = \mathrm{Cor}(T_{d,AB}, T_{d,BA})$ measures conditional directional balance within dyads after accounting for global baseline and covariates.

---

## 38. Dyad scale and dyad correlation

Covariance is $\mathrm{Cov}(T_{d,AB}, T_{d,BA}) = \rho_T \sigma_T^2$. High correlation with tiny $\sigma_T$ has minor empirical impact.

---

## 39. Non-centered dyad effects

```{r noncentered-dyad}
# Non-centered Cholesky representation for dyads
# T = compose_noncentered( [sigma_T, sigma_T], L_Rho_T, Z )
```

---

## 40. From dyads to generalized giving and receiving

Households have generalized giving ($g_h$) and receiving ($r_h$) tendencies:

$$ \begin{bmatrix} g_h \\ r_h \end{bmatrix} \sim \mathrm{MVNormal}\left(\mathbf{0}, \mathbf{S}_{gr}\right) $$

---

## 41. The full social-relations linear predictor

$$ \log\lambda_{AB,d} = \alpha + T_{d,AB} + g_A + r_B $$

$$ \log\lambda_{BA,d} = \alpha + T_{d,BA} + g_B + r_A $$

---

## 42. The complete varying-effects structure

- **Dyad population**: $\mathbf{T}_d \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_T)$ with correlation $\rho_T$ (reciprocity).
- **Household population**: $\mathbf{gr}_h \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_{gr})$ with correlation $\rho_{gr}$ (giving-receiving balance).

---

## 43. Interpreting generalized giving–receiving correlation

$\rho_{gr} = \mathrm{Cor}(g_h, r_h)$:
- Positive: High givers are high receivers (high network activity/centrality).
- Negative: High givers are low receivers (asymmetric/need-based transfers).

---

## 44. Cross-classification in a network

Each household appears repeatedly as sender and receiver across multiple dyads, creating a cross-classified structure.

---

## 45. Adding household wealth

$$ G_A = g_A + \beta_{WG} W_A, \qquad R_B = r_B + \beta_{WR} W_B $$

$$ \log\lambda_{AB,d} = \alpha + T_{d,AB} + G_A + R_B $$

---

## 46. Adding a dyad-level production feature

Symmetric dyad predictor $A_d$: $\log\lambda_{AB,d} = \alpha + T_{d,AB} + \beta_A A_d + G_A + R_B$.

---

## 47. Explained and residual network variation

Comparing posterior standard deviations $(\sigma_T, \sigma_g, \sigma_r)$ before and after adding covariates quantifies explained relational variation.

---

## 48. Posterior network reconstruction

Reconstructing network graphs from posterior draws of matrix $\mathbf{T}^{(s)}$ reveals estimation uncertainty in edge weights.

---

## 49. A crucial network-scale caution

$T_{ij}$ is a log-rate deviation ($\exp(T_{ij})$ multiplier), not a binary edge indicator or probability.

---

## 50. Posterior expected rates

$$ \lambda_{AB,d}^{(s)} = \exp\left(\alpha^{(s)} + T_{AB,d}^{(s)} + g_A^{(s)} + r_B^{(s)}\right) $$

Transformation must occur per posterior draw before averaging ($\mathbb{E}[\exp(X)] \ne \exp(\mathbb{E}[X])$).

---

## 51. Posterior ellipses for household roles

Plotting posterior draws of $(\alpha+g_h, \alpha+r_h)$ displays estimation uncertainty for household $h$, while across-household means reflect population correlation $\rho_{gr}$.

---

## 52. Prior predictive checks for the network model

Simulate whole networks to ensure expected gift counts, zero proportions, and max rates remain scientifically realistic under exponential and Gaussian priors.

---

## 53. Posterior predictive checks for relational data

Check total flow, household outflow/inflow, dyadic balance $|G_{ij}-G_{ji}|$, and zero proportion against observed network data.

---

## 54. Dependence not represented by the model

The basic social-relations model does not include triadic closure/transitivity, community block structure, or temporal dynamics.

---

## 55. Confounding inside networks

Network ties are non-random; observed tie-outcome associations may reflect homophily, shared environment, or selection rather than peer contagion.

---

## 56. Multi-Framework Code Implementations

### 56.1 Mundlak Machine Model (R ulam, Stan, PyMC)

#### R rethinking / ulam

```{r mundlak-full-r}
library(rethinking)

# Generate synthetic group-confounded data
set.seed(2026)
N_groups <- 20
N_id <- 400
g <- sample(1:N_groups, size=N_id, replace=TRUE)
Ug <- rnorm(N_groups, mean=1.5, sd=1.0)
Zg <- rnorm(N_groups, mean=0, sd=1.0)
X <- rnorm(N_id, mean=Ug[g], sd=1.0)
# True bXY = 0, but Ug confounds X and Y
Y <- rbern(N_id, p = inv_logit(0 + 0*X + Ug[g] + 0.5*Zg[g]))

Xbar_g <- tapply(X, g, mean)[g]

dat_mundlak <- list(
  Y = Y,
  X = X,
  Xbar_g = Xbar_g,
  Z_g = Zg[g],
  g = g,
  Ng = N_groups
)

m_mundlak_ulam <- ulam(
  alist(
    Y ~ dbern(p),
    logit(p) <- abar + z[g]*tau + bW*(X - Xbar_g) + bB*Xbar_g + bZ*Z_g,
    z[g] ~ dnorm(0, 1),
    abar ~ dnorm(0, 1),
    tau ~ dexp(1),
    c(bW, bB, bZ) ~ dnorm(0, 1)
  ),
  data = dat_mundlak,
  chains = 4,
  cores = 4
)
precis(m_mundlak_ulam, depth = 2, pars = c("abar", "bW", "bB", "bZ", "tau"))
```

#### Stan (CmdStanR)

```stan
// mundlak_model.stan
data {
  int<lower=1> N;
  int<lower=1> N_groups;
  array[N] int<lower=0, upper=1> Y;
  vector[N] X;
  vector[N] Xbar_g;
  vector[N] Z_g;
  array[N] int<lower=1, upper=N_groups> g;
}
parameters {
  real abar;
  real<lower=0> tau;
  vector[N_groups] z;
  real bW;
  real bB;
  real bZ;
}
transformed parameters {
  vector[N_groups] a = abar + z * tau;
}
model {
  abar ~ normal(0, 1);
  tau ~ exponential(1);
  z ~ normal(0, 1);
  bW ~ normal(0, 1);
  bB ~ normal(0, 1);
  bZ ~ normal(0, 1);
  
  for (i in 1:N) {
    real logit_p = a[g[i]] + bW * (X[i] - Xbar_g[i]) + bB * Xbar_g[i] + bZ * Z_g[i];
    Y[i] ~ binomial_logit(1, logit_p);
  }
}
```

#### PyMC (Python)

```python
import pymc as pm
import numpy as np

# Synthetic setup
np.random.seed(2026)
n_groups = 20
n_id = 400
g_idx = np.random.choice(n_groups, size=n_id)
Ug = np.random.normal(1.5, 1.0, size=n_groups)
Zg = np.random.normal(0, 1.0, size=n_groups)
X_val = np.random.normal(Ug[g_idx], 1.0)
prob_true = 1.0 / (1.0 + np.exp(-(0 + 0*X_val + Ug[g_idx] + 0.5*Zg[g_idx])))
Y_val = np.random.binomial(1, prob_true)

Xbar_per_group = np.bincount(g_idx, weights=X_val) / np.bincount(g_idx)
Xbar_val = Xbar_per_group[g_idx]
Z_val = Zg[g_idx]

with pm.Model() as mundlak_pymc:
    abar = pm.Normal("abar", mu=0, sigma=1.0)
    tau = pm.Exponential("tau", lam=1.0)
    z = pm.Normal("z", mu=0, sigma=1.0, shape=n_groups)
    a = pm.Deterministic("a", abar + z * tau)
    
    bW = pm.Normal("bW", mu=0, sigma=1.0)
    bB = pm.Normal("bB", mu=0, sigma=1.0)
    bZ = pm.Normal("bZ", mu=0, sigma=1.0)
    
    logit_p = a[g_idx] + bW * (X_val - Xbar_val) + bB * Xbar_val + bZ * Z_val
    Y_obs = pm.Bernoulli("Y_obs", logit_p=logit_p, observed=Y_val)
    
    idata_mundlak = pm.sample(draws=1000, tune=1000, chains=4, random_seed=42)
```

---

### 56.2 Full Social Relations Model (R ulam)

```{r srm-full-r}
library(rethinking)
data(KosterLeckie)

# Dyad structure setup
N_hh <- nrow(kl_households)
dyads <- t(combn(N_hh, 2))
N_dyads <- nrow(dyads)

# Build paired gifts array
y_G <- matrix(0, nrow=N_dyads, ncol=2)
for (d in 1:N_dyads) {
  hA <- dyads[d, 1]
  hB <- dyads[d, 2]
  # Gift A -> B
  rowAB <- which(kl_dyads$hidA == hA & kl_dyads$hidB == hB)
  if (length(rowAB) > 0) y_G[d, 1] <- kl_dyads$giftsAB[rowAB]
  # Gift B -> A
  rowBA <- which(kl_dyads$hidA == hB & kl_dyads$hidB == hA)
  if (length(rowBA) > 0) y_G[d, 2] <- kl_dyads$giftsAB[rowBA]
}

dat_srm <- list(
  N_hh = N_hh,
  N_dyads = N_dyads,
  G = y_G,
  hA = dyads[, 1],
  hB = dyads[, 2]
)

m_srm_ulam <- ulam(
  alist(
    # Likelihood for paired directions within dyad
    G[, 1] ~ dpois(lambda1),
    G[, 2] ~ dpois(lambda2),
    
    log(lambda1) <- a + T[d, 1] + gr[hA, 1] + gr[hB, 2],
    log(lambda2) <- a + T[d, 2] + gr[hB, 1] + gr[hA, 2],
    
    # Non-centered 2D dyad effects (reciprocity)
    transpars> matrix[N_dyads, 2]:T <- compose_noncentered(rep_vector(sigma_T, 2), L_Rho_T, Z_T),
    matrix[2, N_dyads]:Z_T ~ dnorm(0, 1),
    
    # Non-centered 2D household effects (giving/receiving)
    transpars> matrix[N_hh, 2]:gr <- compose_noncentered(sigma_gr, L_Rho_gr, Z_gr),
    matrix[2, N_hh]:Z_gr ~ dnorm(0, 1),
    
    # Hyperpriors
    a ~ dnorm(0, 1),
    sigma_T ~ dexp(1),
    vector[2]:sigma_gr ~ dexp(1),
    cholesky_factor_corr[2]:L_Rho_T ~ lkj_corr_cholesky(2),
    cholesky_factor_corr[2]:L_Rho_gr ~ lkj_corr_cholesky(2),
    
    gq> matrix[2, 2]:Rho_T <<- Chol_to_Corr(L_Rho_T),
    gq> matrix[2, 2]:Rho_gr <<- Chol_to_Corr(L_Rho_gr)
  ),
  data = dat_srm,
  chains = 4,
  cores = 4,
  sample = FALSE # Compilation setup
)
```

---

## 57. Supplementary practice exercises and worked model specifications

### Homework B04 Solution: Bangladesh Contraceptive Use with District Mundlak Machine

```{r homework-b04-solution}
library(rethinking)
data(bangladesh)
d <- bangladesh

dat_hw <- list(
  C = d$use.contraception,
  D = as.integer(as.factor(d$district)),
  K = d$living.children,
  U = d$urban
)
n_districts <- max(dat_hw$D)

# District mean living children
Kbar_dist <- tapply(dat_hw$K, dat_hw$D, mean)[dat_hw$D]
dat_hw$Kbar <- Kbar_dist

# Fit Mundlak model for surviving children effect
m_hw_mundlak <- ulam(
  alist(
    C ~ dbern(p),
    logit(p) <- abar + z[D]*sigma_D + bW*K + bM*Kbar + bU*U,
    z[D] ~ dnorm(0, 1),
    abar ~ dnorm(0, 1),
    sigma_D ~ dexp(1),
    c(bW, bM, bU) ~ dnorm(0, 1),
    gq> real:bB <<- bW + bM
  ),
  data = dat_hw,
  chains = 4,
  cores = 4
)

precis(m_hw_mundlak, depth = 2, pars = c("abar", "bW", "bM", "bB", "bU", "sigma_D"))
```

---

## 58. Lecture takeaways

1. Grouped observations contain both within-group and between-group information.
2. An unobserved group cause of exposure and outcome creates group-level confounding.
3. Fixed group effects isolate within-group comparisons by absorbing stable group causes.
4. Purely group-level predictors are unidentified alongside unrestricted group intercepts.
5. Ordinary varying-intercept models assume residual group effects are independent of exposure.
6. The Mundlak model adds group exposure means to separate within-group ($\beta_W$) and between-group ($\beta_B = \beta_W + \beta_M$) associations.
7. Intervention simulation on the probability scale is the correct way to evaluate causal estimands.
8. Relational data assign observations to senders, receivers, and dyads in a cross-classified hierarchy.
9. Dyad correlation $\rho_T$ measures reciprocity; household correlation $\rho_{gr}$ measures giving-receiving balance.
10. Non-centered Cholesky parameterizations allow efficient sampling of multi-level covariance structures.

---

## 59. Review questions

1. How can $U_j$ confound an individual exposure-outcome association?
2. What information does a naive slope combine?
3. Define $\bar X_j$ and $X_{ij}-\bar X_j$.
4. Which component identifies a within-group contrast?
5. Why do fixed group intercepts absorb stable group causes?
6. Why can they not identify a purely group-level coefficient?
7. What independence assumption is implicit in an ordinary varying-intercept model?
8. Why can partial pooling bias a causal slope under group confounding?
9. What variable does the Mundlak model add?
10. How does the Mundlak model retain partial pooling?
11. In the official parameterization, what is the within slope?
12. What is the between slope?
13. What does a nonzero contextual coefficient imply descriptively?
14. Which causal problems remain after Mundlak adjustment?
15. Why is within-group positivity necessary?
16. How does a link-scale coefficient differ from an outcome-scale effect?
17. What intervention is implemented by increasing only individual $X$?
18. When should $\bar X_j$ also change under intervention?
19. What additional assumptions enter a latent Mundlak model?
20. Why should simulation evaluate intervention error?
21. What are the three homework models?
22. How should categorical surviving-children values be decomposed?
23. Why is information-criterion ranking insufficient for the homework?
24. What is an unordered dyad?
25. How many unordered dyads exist among $N$ actors?
26. Why are $A\to B$ and $B\to A$ modeled jointly?
27. What does $\rho_T$ measure?
28. What does $\rho_{gr}$ measure?
29. Why are these correlations not interchangeable?
30. What are generalized giving and receiving effects?
31. How does the predictor change when direction reverses?
32. Why is this network hierarchy cross-classified?
33. How does wealth enter sender and receiver roles?
34. What does a reduction in $\sigma_T$ after adding predictors suggest?
35. Why is $T_{ij}$ not an edge probability?
36. How should a new dyad between known households be predicted?
37. How should two new households be predicted?
38. Which network summaries belong in posterior predictive checks?
39. What dependence remains unmodeled without a transitivity term?
40. Why does a social-relations model not identify contagion?

---

## 60. Book reading guide

### Official assigned reading
- Chapter 14, *Adventures in Covariance*.

### Essential second-edition sections
- Section 14.1, varying slopes by construction;
- Section 14.2, advanced varying slopes;
- Section 14.4, social relations as correlated varying effects.

---

## 61. Official code guide

### `B04_bonus_mundlak.r`
Simulates group confounding, compares naive, fixed-effects, varying-intercepts, Mundlak, and latent Mundlak models, and evaluates 1-unit probability-scale interventions.

### `B04_social_networks.r`
Loads Koster–Leckie dyad data, fits dyad reciprocity and household giving/receiving MVN models, reconstructs posterior weighted networks, and incorporates wealth and production covariates.

---

## 62. Sources

- Richard McElreath, *Statistical Rethinking*, second edition, Chapters 6, 13, 14.
- Official `stat_rethinking_2026` repository scripts `B04_bonus_mundlak.r` and `B04_social_networks.r`.

---

## 70. Annex: Complete Solutions to Review Questions

#### 1. How can $U_j$ confound an individual exposure-outcome association?
An unobserved group feature $U_j$ that influences both individual exposure $X_{ij}$ and outcome $Y_{ij}$ creates an open non-causal backdoor path $X_{ij} \leftarrow U_j \rightarrow Y_{ij}$. This causes $X$ and $Y$ to co-vary across groups even if the true individual-level causal effect $\beta_{XY} = 0$.

#### 2. What information does a naive slope combine?
A naive single-equation regression slope $\beta_X$ combines within-group covariance $\mathrm{Cov}(X_{ij}-\bar X_j, Y_{ij})$ and between-group covariance $\mathrm{Cov}(\bar X_j, \bar Y_j)$ into a single weighted average. When group context is confounded, the between-group component distorts the estimate.

#### 3. Define $\bar X_j$ and $X_{ij}-\bar X_j$.
- $\bar X_j = \frac{1}{n_j} \sum_{i:g_i=j} X_{ij}$ is the group mean exposure (between-group component).
- $X_{ij} - \bar X_j$ is the individual deviation from the group mean (within-group component).

#### 4. Which component identifies a within-group contrast?
The within-group deviation $X_{ij} - \bar X_j$ identifies the within-group contrast, because unobserved group-level factor $U_j$ is constant inside group $j$ and cancels out when comparing individuals within the same group.

#### 5. Why do fixed group intercepts absorb stable group causes?
Fixed group intercepts $\alpha_j$ assign a unique parameter to each group without a shared distribution. Any group-level variable $U_j$ or $Z_j$ that is constant across members of group $j$ is completely absorbed into $\alpha_j$, blocking all time-constant group-level backdoor paths.

#### 6. Why can they not identify a purely group-level coefficient?
Because a group-level predictor $Z_j$ takes only one value per group, a linear combination $\alpha_j + \beta_Z Z_j$ is strictly collinear with unrestricted group dummy variables $\alpha_j$. Any shift in $\beta_Z$ can be perfectly offset by shifts in $\alpha_j$, leaving $\beta_Z$ unidentifiable.

#### 7. What independence assumption is implicit in an ordinary varying-intercept model?
An ordinary varying-intercept model assumes $\alpha_j \perp\!\!\!\perp X_{ij} \mid \bar\alpha, \tau$, meaning the population distribution of group intercepts is conditionally independent of individual exposure levels.

#### 8. Why can partial pooling bias a causal slope under group confounding?
When $\alpha_j$ and $X_{ij}$ are correlated in the data due to $U_j$, forcing group intercepts toward a shared mean through partial pooling forces the exposure slope $\beta_X$ to absorb part of the between-group baseline differences, pulling $\beta_X$ away from the true within-group causal effect.

#### 9. What variable does the Mundlak model add?
The Mundlak model adds the group-level exposure mean $\bar X_j$ to the linear predictor of the varying-intercept model.

#### 10. How does the Mundlak model retain partial pooling?
By explicitly conditioning on $\bar X_j$, the model models group baseline as $\alpha_j^{\text{total}} = \bar\alpha + \beta_M \bar X_j + u_j$. Partial pooling is applied only to the residual group noise $u_j \sim \mathrm{Normal}(0, \tau)$, which is now conditionally uncorrelated with exposure.

#### 11. In the official parameterization, what is the within slope?
In the raw parameterization $\eta_{ij} = \alpha_j + \beta_{XY} X_{ij} + \beta_{UY} \bar X_j$, the within-group slope is $\beta_{XY}$.

#### 12. What is the between slope?
The between-group slope is $\beta_B = \beta_{XY} + \beta_{UY}$ (the sum of the raw exposure slope and the contextual adjustment slope).

#### 13. What does a nonzero contextual coefficient imply descriptively?
A nonzero contextual coefficient $\beta_M \ne 0$ indicates that the between-group association between exposure and outcome differs from the within-group association, revealing group-level confounding, contextual environmental effects, or model misspecification.

#### 14. Which causal problems remain after Mundlak adjustment?
Mundlak does not eliminate unobserved individual-level confounding within groups, time-varying group confounding, measurement error in $X_{ij}$, exposure-dependent selection into groups, positivity failures, or reverse causation.

#### 15. Why is within-group positivity necessary?
If all individuals in group $j$ have identical exposure ($X_{ij} - \bar X_j = 0$), group $j$ contains zero internal exposure variance and contributes no mathematical information toward estimating the within-group slope $\beta_W$.

#### 16. How does a link-scale coefficient differ from an outcome-scale effect?
A link-scale coefficient $\beta_W$ (e.g. log-odds in logistic regression) represents a constant linear additive change in $\mathrm{logit}(p)$. On the outcome probability scale, $\Delta p = \mathrm{logit}^{-1}(\eta + \beta_W) - \mathrm{logit}^{-1}(\eta)$ varies nonlinearly depending on the baseline probability $\mathrm{logit}^{-1}(\eta)$.

#### 17. What intervention is implemented by increasing only individual $X$?
Increasing individual $X_{ij} \mapsto X_{ij} + 1$ while holding $\bar X_j$ constant simulates a conditional micro-level intervention that alters an individual's exposure without changing their overall group composition or environment.

#### 18. When should $\bar X_j$ also change under intervention?
$\bar X_j$ should also change ($\bar X_j \mapsto \bar X_j + 1$) when evaluating a macro-level or policy-wide intervention that alters exposure for all members of group $j$ simultaneously, thereby modifying group composition.

#### 19. What additional assumptions enter a latent Mundlak model?
A latent Mundlak model assumes a joint parametric distribution for exposure $X_{ij} \sim \mathrm{Normal}(\alpha_X + \beta_{UX} u_j, \sigma_X)$ and outcome, requiring correct specification of the exposure model and assuming $u_j \sim \mathrm{Normal}(0, 1)$ captures all unobserved group confounding.

#### 20. Why should simulation evaluate intervention error?
Because on nonlinear outcome scales, coefficient recovery error does not map linearly to outcome prediction error. Evaluating intervention error ($\mathrm{ATE}$) directly measures performance on the actual decision-relevant causal quantity.

#### 21. What are the three homework models?
1. District fixed-effects model.
2. Ordinary district varying-intercept model.
3. Mundlak machine varying-intercept model with district mean children $\bar K_d$.

#### 22. How should categorical surviving-children values be decomposed?
For categorical surviving children, decomposition requires constructing district proportions for each child-count category (e.g., proportion of households in district $d$ with 1, 2, or $3+$ children) rather than treating arbitrary integer category codes as a continuous metric variable.

#### 23. Why is information-criterion ranking insufficient for the homework?
Information criteria (WAIC/LOO) measure out-of-sample predictive accuracy for observing existing data distributions under a specific likelihood. They do not evaluate whether a model has closed backdoor paths or correctly identified a causal estimand under confounding.

#### 24. What is an unordered dyad?
An unordered dyad $\{A, B\}$ is a pair of two actors $A$ and $B$ considered symmetrically without specifying sender or receiver roles.

#### 25. How many unordered dyads exist among $N$ actors?
Among $N$ actors, there are $\binom{N}{2} = \frac{N(N-1)}{2}$ unique unordered dyads.

#### 26. Why are $A\to B$ and $B\to A$ modeled jointly?
$A \to B$ and $B \to A$ are modeled jointly because directed interactions within the same dyad share mutual unobserved relational features (such as friendship, reciprocity, or kinship) and involve the same pair of actors.

#### 27. What does $\rho_T$ measure?
$\rho_T$ measures dyadic reciprocity: the correlation between directed deviations $T_{d, AB}$ and $T_{d, BA}$ within unordered dyads after accounting for global baselines and actor features.

#### 28. What does $\rho_{gr}$ measure?
$\rho_{gr}$ measures household role covariance: the correlation between a household's generalized giving propensity $g_h$ and its generalized receiving propensity $r_h$ across the population.

#### 29. Why are these correlations not interchangeable?
$\rho_T$ operates within pairs of directed interactions inside specific dyads (dyadic level), whereas $\rho_{gr}$ operates across actor traits in the node population (individual/household level).

#### 30. What are generalized giving and receiving effects?
- **Generalized giving ($g_h$)**: Household $h$'s baseline propensity to send gifts to all other households in the network.
- **Generalized receiving ($r_h$)**: Household $h$'s baseline propensity to receive gifts from all other households in the network.

#### 31. How does the predictor change when direction reverses?
When direction reverses from $A \to B$ to $B \to A$, household $A$ shifts from sender role ($g_A$) to receiver role ($r_A$), household $B$ shifts from receiver role ($r_B$) to sender role ($g_B$), and the dyad effect switches from $T_{d, AB}$ to $T_{d, BA}$.

#### 32. Why is this network hierarchy cross-classified?
Because every directed interaction $i \to j$ belongs simultaneously to sender $i$, receiver $j$, and dyad $\{i, j\}$. Since actors appear as senders and receivers across many different dyads, units cannot be grouped into a single clean nested tree.

#### 33. How does wealth enter sender and receiver roles?
Wealth enters via role-specific slopes: $G_A = g_A + \beta_{WG} W_A$ for giving and $R_B = r_B + \beta_{WR} W_B$ for receiving, allowing wealth to have distinct effects on giving versus receiving.

#### 34. What does a reduction in $\sigma_T$ after adding predictors suggest?
A reduction in residual dyad standard deviation $\sigma_T$ indicates that the added dyadic covariates (e.g. dyad production $A_d$) explain part of the unobserved variation in pairwise relationship intensity.

#### 35. Why is $T_{ij}$ not an edge probability?
In a Poisson social-relations model, $T_{ij}$ is a continuous additive log-rate deviation. Its exponent $\exp(T_{ij})$ acts as a multiplicative rate factor on expected counts, not a probability bounded in $[0, 1]$.

#### 36. How should a new dyad between known households be predicted?
Sample dyad effects from the learned dyad population $\mathbf{T}_{d,\text{new}} \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_T)$ while reusing the posterior samples of $g_A, r_A, g_B, r_B$ for the known households.

#### 37. How should two new households be predicted?
Draw new household vectors $\begin{bmatrix} g_A \\ r_A \end{bmatrix}, \begin{bmatrix} g_B \\ r_B \end{bmatrix} \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_{gr})$ and a new dyad vector $\mathbf{T}_{d,\text{new}} \sim \mathrm{MVNormal}(\mathbf{0}, \mathbf{S}_T)$ from the hyperparameter posteriors.

#### 38. Which network summaries belong in posterior predictive checks?
Relational summaries such as total network gift flow, household in-degree/out-degree distributions, dyadic balance $|G_{ij} - G_{ji}|$, zero-dyad proportions, and network reciprocity coefficients.

#### 39. What dependence remains unmodeled without a transitivity term?
Triadic closure or transitivity (the tendency for friends of friends to exchange gifts, $A \to B \text{ and } B \to C \implies A \to C$) is not modeled by separate sender, receiver, and dyad parameters.

#### 40. Why does a social-relations model not identify contagion?
Because observed tie-outcome associations in cross-sectional network data can be produced by homophily (similar actors forming ties), unobserved shared environmental causes, or partner selection, rather than causal social influence or peer contagion.
